### 2023-02-23 abstract.tex
abstract draft

\begin{abstract}
Type systems typically only define the conditions under which an expression is well-typed, leaving ill-typed expressions formally meaningless. This approach is insufficient for developing language servers for modern programming environments, which are expected to sensibly localize type errors and continue to provide a variety of semantic services in their presence. Lacking formal foundations, error localization has largely become a matter of folklore in the community. Recent work on the Hazelnut calculus takes a step toward assigning formal meaning to erroneous programs, but it stops short of a comprehensive account. In particular,  (1) Hazelnut is able to reason around marked type inconsistencies, but not several other kinds of static errors, and (2) Hazelnut provides an insufficient account of the marking procedure itself, relying on a limited structure editor, or on the user, to correctly insert marks. This paper addresses these deficiencies, contributing the first comprehensive formal account of type error localization, called the marked lambda calculus. This contribution is independently useful as a core calculus for language servers. In addition, we deploy its total marking procedure to fix several instances of undefined behavior in the Hazelnut structure editor calculus, namely by allowing edits that introduce non-local errors. We mechanize the calculus in Agda and implement an extension in Hazel, a full-scale live functional programming environment previously based on Hazelnut.

The marked lambda calculus supports local type inference, i.e. it is bidirectionally typed. Constraint-based type inference schemes (e.g. Hindley-Milner) are notorious for complicating error localization. We approach the problem of  constraint-based type inference in a novel manner: we deploy constraint solving as a layer atop a gradual bidirectionally typed core, where it infers type hole fillings. This system is able to elegantly handle situations where the system of constraints cannot be solved, because the unification algorithm we develop localizes errors arising from inconsistent constraints to individual type holes, rather than attempting to localize them to particular expressions, which is fundamentally \emph{ad hoc}. The user can resolve these errors by selecting from suggested partial solutions, or leave the errors unresolved, in which case they operate as the unknown type from gradual typing. We formally specify this approach and implement it in Hazel. 
\end{abstract}

### 2023-02-25 calculus.tex
Initial progress on marked lambda calculus

\section{Motivating Examples}
In this section, we motivate a formal calculus for bidirectional type error localization through a
number of examples.

% TODO Talk about idea that bidirectional typing makes error localization easy?

\subsection{Unbound Variables}
% A simple example of an error that may occur in programs is the use of an unbound variable.

\subsection{Inconsistent Types}
Often, the type an expression \emph{should} be is known. For example, in an expression such as
$\EPlus{\EMV_1}{\EMV_2}$, both $\EMV_1$ and $\EMV_2$ should have the type $\TNum$. In a
bidirectional system, $\EMV_1$ and $\EMV_2$ would be analysed against $\TNum$.

\subsection{Inconsistent Branches}

\section{Marking Calculus}
\label{sec:calculus}

The previous section introduced different error localization cases by example. Now, we formally
introduce a \emph{marking calculus} for bidirectional type error localization on a lambda calculus
extended with numbers, conditionals, and let bindings, given by a two-layer system:
%
\begin{itemize}
  \item \textbf{Unmarked expressions} ($\EMName$), which are the surface-level expression syntax.
    Types classify unmarked expressions according to a standard bidirectional gradual typing
    semantics. Note that a non-gradual system may be recovered straightforwardly by removing the
    unknown type and replacing type consistency with equality (discussed in
    \cref{sec:calculus-non-gradual}).

  \item \textbf{Marked expressions} ($\ECMName$), which mirror the structure of unmarked expressions
    but are extended with \textbf{error markings}. Every marked expression corresponds to an
    unmarked expression by \emph{mark erasure}.

  \item \textbf{Mark insertion}, which \textbf{marks} any unmarked expression into a marked
    expression, inserting error markings where appropriate.
\end{itemize}

In the interest of space, this overview focuses on the ``interesting'' cases; the whole system,
alongside the metatheorems discussed have been mechanized in the Agda proof assistant (discussed in
\cref{sec:calculus-agda}).

\subsection{Unmarked Expressions}
\label{sec:calculus-uexp}

\Cref{fig:calculus-syntax-uexp} gives the syntax of types, $\TMV$, and unmarked expressions, $\EMV$,
which extend the simply typed lambda calculus with $\TNum$ and $\TBool$ base types of numbers and
booleans. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$, and
there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$ correspond to
the boolean values $\textsf{true}$ and $\textsf{false}$, and the branching expression
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives an if-else form.

In addition, types are extended with an \emph{unknown type}, $\TUnknown$, which

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \mid \ELet{x}{\EMV}{\EMV} \\
             &       & \mid         & \ENumMV \mid \EPlus{\EMV}{\EMV}
                       \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
  \end{array}\]
  \caption{Syntax of types and unmarked expressions.}
  \label{fig:calculus-syntax-uexp}
\end{figure}

Unmarked expressions are governed by a standard bidirectional gradual type system, as given in
\Cref{fig:calculus-typing-uexp}.

% TODO Not sure how explicitly we want to spell out the type system, in lieu of background
% section on bidirectional typing?

% \input{figs/calculus-type-judgments.tex}
% \input{figs/calculus-typing-uexp.tex}

\subsection{Marked Expressions}
\label{sec:calculus-mexp}

Marked expressions, $\ECMV$, extend the syntax of unmarked expressions with explicit \emph{error
markings}. As given in the syntax definition of \cref{fig:calculus-syntax-mexp}, there are three
such forms, which mirror the examples above:
%
\begin{itemize}
  \item $\ECUnbound{x}$, denoting the usage of an unbound variable.

  \item $\ECInconType{\ECMV}$, which occurs when the synthesized type of $\ECMV$ is inconsistent
    with the required type.

  \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, which occurs when the types of the two branches
    $\ECMV_2$ and $\ECMV_3$ are inconsistent.
\end{itemize}
%
Additional forms may arise as required by the language.

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV} \\
             &       & \mid         & \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV} \\
             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
  \end{array}\]
  \caption{Syntax of marked expressions}
  \label{fig:calculus-syntax-mexp}
\end{figure}

Marked expressions are also typed by a bidirectional system, which, for most forms, closely mirrors
the rules for corresponding unmarked forms. \Cref{fig:calculus-typing-mexp} contains a selection of
the ``interesting'' cases, particularly those concerned with error markings.

\input{figs/calculus-typing-mexp.tex}

% TODO Not really sure how to explain the syntax and typing of marked expressions in an intuitive
% way in relation to unmarked expressions.

For unbound variables $\ECUnbound{x}$, nothing may be said about their types; hence, they synthesize
the unknown type. Inconsistent branch forms behave similarly, as inconsistent types have no join.

There are three rules concerned with the typing of lambda abstractions. MALam1, applies to
$\ECLam{x}{\TMV}{\ECMV}$ if the type $\TMV_3$ analysed against is a matched arrow type
$\TArrow{\TMV_1}{\TMV_2}$ and input type $\TMV_1$ is consistent with the binding ascription $\TMV$.
This corresponds to the lambda rule for unmarked expressions. However, we are additionally concerned
with the cases when $\TMV_3$ is not a matched arrow type or $\TMV_1$ is inconsistent with $\TMV$.
These are described by MALam2 and MALam3.

Alongside the standard subsumption rule, in which an expression that synthesizes a type may be
analysed against any consistent type, MAInconsistentTypes governs most occurrences of the
inconsistent types marking: $\ECInconType{\ECMV}$ analyzes against any type as long as $\ECMV$
synthesizes an inconsistent type. In both cases, we stipulate that subsumption may be applied to
$\ECMV$, written $\subsumable{\ECMV}$. In particular, lambda abstractions and if-else forms are
\emph{not} subsumable, formalizing these rules as ones of ``last resort'', preserving determinism of
typing.

Similarly to lambda abstractions, two rules govern application. In the unmarked syntax, the
application $\EAp{\EMV_1}{\EMV_2}$ was well-typed if the type $\TMV$ synthesized by $\EMV_1$ was a
matched arrow type. This corresponds to the MSAp1 rule for marked expressions. In the case that
$\TMV$ is not a matched arrow type, we require an inconsistent types marking around the

\subsection{Mark Insertion}

\subsection{Well-Formedness}
\label{sec:calculus-wellformedness}

\subsection{Alternative Formulations}

\subsection{Non-Gradual System}
\label{sec:calculus-non-gradual}

\subsection{Mechanization}
\label{sec:calculus-agda}

The metatheory of the calculus has been fully mechanized in the Agda proof assistant [citation]. The
mechanization's documentation contains more details regarding technical decisions made therein.

### 2023-02-25 abstract.tex
Updates from Overleaf

@@
 Recent work on the Hazelnut calculus takes a step toward assigning formal meaning to erroneous programs, but it stops short of a comprehensive account.
-In particular,  (1) Hazelnut is able to reason around marked type inconsistencies, but not several other kinds of static errors, and (2) Hazelnut provides an insufficient account of the marking procedure itself, relying on a limited structure editor, or on the user, to correctly insert marks.
-This paper addresses these deficiencies, contributing the first comprehensive formal account of type error localization, called the marked lambda calculus.
+In particular,  (1) Hazelnut is able to reason around marked type inconsistencies, but not several other kinds of static errors, and (2) Hazelnut provides an insufficient account of the marking procedure itself, relying on a limited structure editor, or on the user, to correctly insert marks.\todo{decenter Hazelnut?} This paper addresses these deficiencies, contributing the first comprehensive formal account of type error localization, called the marked lambda calculus.
 This contribution is independently useful as a core calculus for language servers.
@@
 We approach the problem of  constraint-based type inference in a novel manner: we deploy constraint solving as a layer atop a gradual bidirectionally typed core, where it infers type hole fillings.
-This system is able to elegantly handle situations where the system of constraints cannot be solved, because the unification algorithm we develop localizes errors arising from inconsistent constraints to individual type holes, rather than attempting to localize them to particular expressions, which is fundamentally \emph{ad hoc}.
+This system is able to elegantly handle situations where the system of constraints cannot be solved, because the unification algorithm we develop localizes errors arising from inconsistent constraints to individual type holes, rather than attempting to localize them to one particular expression, which is fundamentally \emph{ad hoc}.
 The user can resolve these errors by selecting from suggested partial solutions, or leave the errors unresolved, in which case they operate as the unknown type from gradual typing.

### 2023-02-25 calculus.tex
Updates from Overleaf

@@
-\section{Motivating Examples}
-In this section, we motivate a formal calculus for bidirectional type error localization through a
-number of examples.
-
-% TODO Talk about idea that bidirectional typing makes error localization easy?
-
-\subsection{Unbound Variables}
-% A simple example of an error that may occur in programs is the use of an unbound variable.
-
-\subsection{Inconsistent Types}
-Often, the type an expression \emph{should} be is known. For example, in an expression such as
-$\EPlus{\EMV_1}{\EMV_2}$, both $\EMV_1$ and $\EMV_2$ should have the type $\TNum$. In a
-bidirectional system, $\EMV_1$ and $\EMV_2$ would be analysed against $\TNum$.
-
-\subsection{Inconsistent Branches}
-
-\section{Marking Calculus}
+\section{Marked Lambda Calculus}
 \label{sec:calculus}
-
-The previous section introduced different error localization cases by example. Now, we formally
-introduce a \emph{marking calculus} for bidirectional type error localization on a lambda calculus
-extended with numbers, conditionals, and let bindings, given by a two-layer system:
-%
-\begin{itemize}
-  \item \textbf{Unmarked expressions} ($\EMName$), which are the surface-level expression syntax.
-    Types classify unmarked expressions according to a standard bidirectional gradual typing
-    semantics. Note that a non-gradual system may be recovered straightforwardly by removing the
-    unknown type and replacing type consistency with equality (discussed in
-    \cref{sec:calculus-non-gradual}).
-
-  \item \textbf{Marked expressions} ($\ECMName$), which mirror the structure of unmarked expressions
-    but are extended with \textbf{error markings}. Every marked expression corresponds to an
-    unmarked expression by \emph{mark erasure}.
-
-  \item \textbf{Mark insertion}, which \textbf{marks} any unmarked expression into a marked
-    expression, inserting error markings where appropriate.
-\end{itemize}
-
-In the interest of space, this overview focuses on the ``interesting'' cases; the whole system,
-alongside the metatheorems discussed have been mechanized in the Agda proof assistant (discussed in
-\cref{sec:calculus-agda}).
-
-\subsection{Unmarked Expressions}
-\label{sec:calculus-uexp}
-
-\Cref{fig:calculus-syntax-uexp} gives the syntax of types, $\TMV$, and unmarked expressions, $\EMV$,
-which extend the simply typed lambda calculus with $\TNum$ and $\TBool$ base types of numbers and
-booleans. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$, and
-there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$ correspond to
-the boolean values $\textsf{true}$ and $\textsf{false}$, and the branching expression
-$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives an if-else form.
-
-In addition, types are extended with an \emph{unknown type}, $\TUnknown$, which
-
-\begin{figure}[htbp]
-  \[\begin{array}{rrcl}
-    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
-    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \mid \ELet{x}{\EMV}{\EMV} \\
-             &       & \mid         & \ENumMV \mid \EPlus{\EMV}{\EMV}
-                       \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
-  \end{array}\]
-  \caption{Syntax of types and unmarked expressions.}
-  \label{fig:calculus-syntax-uexp}
-\end{figure}
-
-Unmarked expressions are governed by a standard bidirectional gradual type system, as given in
-\Cref{fig:calculus-typing-uexp}.
-
-% TODO Not sure how explicitly we want to spell out the type system, in lieu of background
-% section on bidirectional typing?
-
-% \input{figs/calculus-type-judgments.tex}
-% \input{figs/calculus-typing-uexp.tex}
-
-\subsection{Marked Expressions}
-\label{sec:calculus-mexp}
-
-Marked expressions, $\ECMV$, extend the syntax of unmarked expressions with explicit \emph{error
-markings}. As given in the syntax definition of \cref{fig:calculus-syntax-mexp}, there are three
-such forms, which mirror the examples above:
-%
-\begin{itemize}
-  \item $\ECUnbound{x}$, denoting the usage of an unbound variable.
-
-  \item $\ECInconType{\ECMV}$, which occurs when the synthesized type of $\ECMV$ is inconsistent
-    with the required type.
-
-  \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, which occurs when the types of the two branches
-    $\ECMV_2$ and $\ECMV_3$ are inconsistent.
-\end{itemize}
-%
-Additional forms may arise as required by the language.
-
-\begin{figure}[htbp]
-  \[\begin{array}{rrcl}
-    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV} \\
-             &       & \mid         & \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
-                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV} \\
-             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
-  \end{array}\]
-  \caption{Syntax of marked expressions}
-  \label{fig:calculus-syntax-mexp}
-\end{figure}
-
-Marked expressions are also typed by a bidirectional system, which, for most forms, closely mirrors
-the rules for corresponding unmarked forms. \Cref{fig:calculus-typing-mexp} contains a selection of
-the ``interesting'' cases, particularly those concerned with error markings.
-
-\input{figs/calculus-typing-mexp.tex}
-
-% TODO Not really sure how to explain the syntax and typing of marked expressions in an intuitive
-% way in relation to unmarked expressions.
-
-For unbound variables $\ECUnbound{x}$, nothing may be said about their types; hence, they synthesize
-the unknown type. Inconsistent branch forms behave similarly, as inconsistent types have no join.
-
-There are three rules concerned with the typing of lambda abstractions. MALam1, applies to
-$\ECLam{x}{\TMV}{\ECMV}$ if the type $\TMV_3$ analysed against is a matched arrow type
-$\TArrow{\TMV_1}{\TMV_2}$ and input type $\TMV_1$ is consistent with the binding ascription $\TMV$.
-This corresponds to the lambda rule for unmarked expressions. However, we are additionally concerned
-with the cases when $\TMV_3$ is not a matched arrow type or $\TMV_1$ is inconsistent with $\TMV$.
-These are described by MALam2 and MALam3.
-
-Alongside the standard subsumption rule, in which an expression that synthesizes a type may be
-analysed against any consistent type, MAInconsistentTypes governs most occurrences of the
-inconsistent types marking: $\ECInconType{\ECMV}$ analyzes against any type as long as $\ECMV$
-synthesizes an inconsistent type. In both cases, we stipulate that subsumption may be applied to
-$\ECMV$, written $\subsumable{\ECMV}$. In particular, lambda abstractions and if-else forms are
-\emph{not} subsumable, formalizing these rules as ones of ``last resort'', preserving determinism of
-typing.
-
-Similarly to lambda abstractions, two rules govern application. In the unmarked syntax, the
-application $\EAp{\EMV_1}{\EMV_2}$ was well-typed if the type $\TMV$ synthesized by $\EMV_1$ was a
-matched arrow type. This corresponds to the MSAp1 rule for marked expressions. In the case that
-$\TMV$ is not a matched arrow type, we require an inconsistent types marking around the
-
-\subsection{Mark Insertion}
-
-\subsection{Well-Formedness}
-\label{sec:calculus-wellformedness}
-
-\subsection{Alternative Formulations}
-
-\subsection{Non-Gradual System}
-\label{sec:calculus-non-gradual}
-
-\subsection{Mechanization}
-\label{sec:calculus-agda}
-
-The metatheory of the calculus has been fully mechanized in the Agda proof assistant [citation]. The
-mechanization's documentation contains more details regarding technical decisions made therein.

### 2023-02-25 related.tex
Type Hole Inference: Fixes screenshots; Adjusts headings; Adds todos

\section{Related Work}
\label{sec:related}

Jeremy Siek Gradual Typing with Inference
- Siek's type inference halts when it can't solve, ours doesn't

### 2023-02-25 calculus.tex
Add more figures and discussion of marking calculus rules

--- 2023-02-25 calculus.tex
@@
 \section{Motivating Examples}
-In this section, we motivate a formal calculus for bidirectional type error localization through a
-number of examples.
+\emph{Bidirectional typing} combines type checking, which determines if a program satisfies a given
+type, and type synthesis, which generates a type from the program. This gives a simple algorithmic
+system in which type information is recursively propagated through terms and allows for local type
+inference.
 
-% TODO Talk about idea that bidirectional typing makes error localization easy?
+% TODO talk about idea that bidirectional typing makes error localization easy? i.e. why choose
+% bidirectional?
+
+Often, programmers are working with program states that are syntactically well-formed but not
+well-typed. However, formal typing semantics usually only give meaning to well-typed programs,
+leaving ill-typed ones meaningless. This is insufficient for editor services and language servers,
+which give users feedback on where and why a program is ill-typed. Traditionally, this has been
+performed in an \emph{ad hoc} manner [citations?].
+
+In this section, we motivate a formal calculus for bidirectional type error localization by
+exploring different kinds of type errors that may occur.
+
+% TODO state goals: prefer fine-grained over coarser blaming/marking
 
 \subsection{Unbound Variables}
-% A simple example of an error that may occur in programs is the use of an unbound variable.
+A simple and common example of an error that may occur in programs is the use of an unbound
+variable. In this case, editor services often indicate visually that such usages are invalid.
+
+% TODO image of an editor for this example? not sure what else to say about this
 
 \subsection{Inconsistent Types}
-Often, the type an expression \emph{should} be is known. For example, in an expression such as
-$\EPlus{\EMV_1}{\EMV_2}$, both $\EMV_1$ and $\EMV_2$ should have the type $\TNum$. In a
-bidirectional system, $\EMV_1$ and $\EMV_2$ would be analysed against $\TNum$.
+Often, the type an expression \emph{should} be is known. Both operands of an addition expression
+$\EPlus{\ETrue}{\ENumMV}$, for example, should have the type $\TNum$. In a bidirectional system,
+$\ETrue$ and $\ENumMV$ would be analysed against $\TNum$, in which the former check would fail:
+$\ETrue$ synthesizes $\TBool$, which is inconsistent with $\TNum$.
+
+% An editor may indicate such an error in different ways. Two obvious ones are considered below: (1)
+% indicating an error on $\ETrue$, or (2) indicating an error on the entire expression
+% $\EPlus{\ETrue}{\ENumMV}$.
+% \[%
+  % \EPlus{\ETrue}{\ENumMV} \qquad \EPlus{\ETrue}{\ENumMV}
+% \]%
+% The former ``blames'' the error on the operand,
+
+% Since the ``fault'' of the error
+
+% TODO cite real editor/language?
+
+Since expressions typed in analytic mode are checked to satisfy a correct, known type, they are
+given ``blame'' when ill-typed.
 
@@
 
-\section{Marking Calculus}
+\subsection{Correctness}
+\begin{theorem}[name=Marking Totality]
+  Any syntactically well-formed program may be marked for errors.
+\end{theorem}
+
+In the next section, we will formally develop
+
+\section{Marked Lambda Calculus}
 \label{sec:calculus}
 
-The previous section introduced different error localization cases by example. Now, we formally
-introduce a \emph{marking calculus} for bidirectional type error localization on a lambda calculus
-extended with numbers, conditionals, and let bindings, given by a two-layer system:
+The previous section surveyed different type error cases. Now, we formally introduce a \emph{marking
+calculus} for bidirectional type error localization on the gradually typed lambda calculus above.
+This comes as a two-layer system:
 %
 \begin{itemize}
-  \item \textbf{Unmarked expressions} ($\EMName$), which are the surface-level expression syntax.
-    Types classify unmarked expressions according to a standard bidirectional gradual typing
-    semantics. Note that a non-gradual system may be recovered straightforwardly by removing the
-    unknown type and replacing type consistency with equality (discussed in
-    \cref{sec:calculus-non-gradual}).
+  \item \textbf{Unmarked expressions} ($\EMName$), which compose the surface-level syntax. Types
+    classify unmarked expressions according to a standard bidirectional gradual typing semantics.
 
@@
 \end{itemize}
+%
+That is, we extend the ordinary typing semantics of a language with a secondary ``marked'' language
+and define a mapping from ``unmarked'' programs to ``marked'' programs. While the semantics that
+follow concern a gradually typed language, a non-gradual system may be recovered straightforwardly
+by removing the unknown type and replacing type consistency with equality.
 
-In the interest of space, this overview focuses on the ``interesting'' cases; the whole system,
-alongside the metatheorems discussed have been mechanized in the Agda proof assistant (discussed in
-\cref{sec:calculus-agda}).
+In the interest of space, this overview focuses on the ``interesting'' cases. All rules and
+metatheorems may be found in the supplementary appendix, alongside a complete mechanization in the
+Agda proof assistant (discussed in \cref{sec:calculus-agda}).
 
@@
 
-\Cref{fig:calculus-syntax-uexp} gives the syntax of types, $\TMV$, and unmarked expressions, $\EMV$,
-which extend the simply typed lambda calculus with $\TNum$ and $\TBool$ base types of numbers and
-booleans. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$, and
-there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$ correspond to
-the boolean values $\textsf{true}$ and $\textsf{false}$, and the branching expression
-$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives an if-else form.
-
-In addition, types are extended with an \emph{unknown type}, $\TUnknown$, which
+We consider a gradually typed lambda calculus extended with numbers and conditionals, whose syntax
+and bidirectional typing are given in \cref{fig:calculus-syntax-example} and
+\cref{fig:calculus-typing-example}, respectively. The base types $\TNum$ and $\TBool$ classify
+number and boolean expressions. The number literal corresponding to the mathematical number $n$ is
+given by $\ENumMV$, and there is a single addition operation on numeric expressions. $\ETrue$ and
+$\EFalse$ correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, and
+$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives an if-else structure. In addition to these forms, $\TUnknown$
+indicates an unknown type.
 
@@
   \[\begin{array}{rrcl}
-    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
-    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \mid \ELet{x}{\EMV}{\EMV} \\
+    \textsf{Type} & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
+    \textsf{Exp}  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \\
          &       & \mid         & \ENumMV \mid \EPlus{\EMV}{\EMV}
@@
   \end{array}\]
-  \caption{Syntax of types and unmarked expressions.}
+  \caption{Syntax of the extended gradually typed lambda calculus}
   \label{fig:calculus-syntax-uexp}
@@
 
-Unmarked expressions are governed by a standard bidirectional gradual type system, as given in
-\Cref{fig:calculus-typing-uexp}.
+Bidirectional typing employs two mutually defined judgments. \emph{Type synthesis}, written
+$\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under the typing context $\ctx$, the expression
+$\EMV$ synthesizes or infers the type $\TMV$. \emph{Type analysis}, written
+$\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear where an expression of
+type $\TMV$ is expected. The \emph{subsumption} rule (ASubsume) states that if an expression
+synthesizes a type, it may be analysed against that type or any \emph{consistent} type.
 
-% TODO Not sure how explicitly we want to spell out the type system, in lieu of background
-% section on bidirectional typing?
+Note that the subsumption rule requires that $\EMV$ be ``subsumable'', a judgment defined on all
+forms except lambda abstractions and if-else expressions. This restricts the usage of subsumption to
+a rule of ``last resort'', admitting the unicity of typing.
 
-% \input{figs/calculus-type-judgments.tex}
-% \input{figs/calculus-typing-uexp.tex}
+% \begin{theorem}[name=Unmarked Expression Typing Unicity] \
+  % If $\ctxSynType{\ctx}{\EMV}{\TMV}$ and $\ctxSynType{\ctx}{\EMV}{\TMV'}$, then $\TMV = \TMV'$.
+% \end{theorem}
+
+\input{figs/calculus-typing-uexp}
+
+The \emph{consistency} judgment $\consistent{\TMV}{\TMV'}$ that appears in the subsumption rule
+defines a reflexive and symmetric (but not transitive) relation between types. The unknown type is
+additionally consistent with any other type, and arrow types are consistent if their components are.
+
+$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$ gives that $\TMV$ is a \emph{matched arrow type}, which means
+that it may be considered an arrow type. It is purely a technical mechanism to avoid duplication of
+rules related to arrow types.
+
+\input{figs/calculus-type-judgments}
 
@@
 
-Marked expressions, $\ECMV$, extend the syntax of unmarked expressions with explicit \emph{error
-markings}. As given in the syntax definition of \cref{fig:calculus-syntax-mexp}, there are three
-such forms, which mirror the examples above:
+Marked expressions, $\ECMV$, extend the syntax of unmarked expressions with \emph{explicit error
+marks}. As given in the syntax definition of \cref{fig:calculus-syntax-mexp}, there are three such
+forms, which mirror the examples discussed previously:
 %
-\begin{itemize}
-  \item $\ECUnbound{x}$, denoting the usage of an unbound variable.
+\begin{itemize} % TODO not the snappiest names
+  \item $\ECUnbound{x}$, the \emph{unbound variable mark}, denotes the usage of an unbound variable.
 
-  \item $\ECInconType{\ECMV}$, which occurs when the synthesized type of $\ECMV$ is inconsistent
-    with the required type.
+  \item $\ECInconType{\ECMV}$, the \emph{inconsistent types mark}, occurs when the synthesized type
+    of $\ECMV$ is inconsistent with the required type.
 
-  \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, which occurs when the types of the two branches
-    $\ECMV_2$ and $\ECMV_3$ are inconsistent.
+  \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, corresponds
+    to an if-else expression where the types of the two branches $\ECMV_2$ and $\ECMV_3$ are
+    inconsistent.
 \end{itemize}
@@
 Additional forms may arise as required by the language.
+
+% TODO examples of such?
 
@@
 
+The unmarked version of a marked expression $\ECMV$ may be recovered through \emph{mark erasure},
+given by $\erase{\ECMV}$. As in \Cref{fig:calculus-mark-erasure}, this total function may be defined
+inductively in a straightforward manner. Observe that erasing the inconsistent branches mark gives
+an if-else expression with sub-expression marks recursively erased.
+
+\input{figs/calculus-mark-erasure}
+
 Marked expressions are also typed by a bidirectional system, which, for most forms, closely mirrors
@@
 
-\input{figs/calculus-typing-mexp.tex}
+\input{figs/calculus-typing-mexp}
 
@@
 
-\subsection{Well-Formedness}
-\label{sec:calculus-wellformedness}
+\subsection{Correctness}
+\label{sec:calculus-correctness}
+
+% Totality
+
+% Well-formedness conditions
 
@@
 
+% TODO extension to non-typing semantics?
+
 \subsection{Mechanization}

### 2023-02-26 calculus.tex
Further describe uexp and mexp typing rules

@@
 $\ETrue$ and $\ENumMV$ would be analysed against $\TNum$, in which the former check would fail:
-$\ETrue$ synthesizes $\TBool$, which is inconsistent with $\TNum$.
+$\ETrue$ synthesises $\TBool$, which is inconsistent with $\TNum$.
 
@@
 \begin{itemize}
-  \item \textbf{Unmarked expressions} ($\EMName$), which compose the surface-level syntax. Types
-    classify unmarked expressions according to a standard bidirectional gradual typing semantics.
+  \item \textbf{Unmarked expressions} (\cref{sec:calculus-uexp}), which compose the surface-level
+    syntax. Types classify unmarked expressions according to a standard bidirectional gradual typing
+    semantics.
 
-  \item \textbf{Marked expressions} ($\ECMName$), which mirror the structure of unmarked expressions
-    but are extended with \textbf{error markings}. Every marked expression corresponds to an
-    unmarked expression by \emph{mark erasure}.
+  \item \textbf{Marked expressions} (\cref{sec:calculus-mexp}), which mirror the structure of
+    unmarked expressions but are extended with \textbf{error markings}. Every marked expression
+    corresponds to an unmarked expression by \emph{mark erasure}.
 
-  \item \textbf{Mark insertion}, which \textbf{marks} any unmarked expression into a marked
+  \item \textbf{Mark insertion} (\cref{sec:calculus-marking}), which \textbf{marks} any unmarked expression into a marked
     expression, inserting error markings where appropriate.
@@
 That is, we extend the ordinary typing semantics of a language with a secondary ``marked'' language
-and define a mapping from ``unmarked'' programs to ``marked'' programs. While the semantics that
-follow concern a gradually typed language, a non-gradual system may be recovered straightforwardly
-by removing the unknown type and replacing type consistency with equality.
+and define a mapping from ``unmarked'' programs to ``marked'' programs. Note that while the
+semantics that follow concern a gradually typed language, a non-gradual system may be recovered
+straightforwardly by removing the unknown type and replacing type consistency with equality.
 
-In the interest of space, this overview focuses on the ``interesting'' cases. All rules and
-metatheorems may be found in the supplementary appendix, alongside a complete mechanization in the
-Agda proof assistant (discussed in \cref{sec:calculus-agda}).
+In \cref{sec:calculus-correctness}, we discuss the metatheory that ensures the system---and any such
+error localization semantics---to be correct. We follow then by extending the calculus with support
+for product types. In the interest of space, this overview focuses on ``interesting'' cases. All
+rules and theorems may be found in the supplementary appendix, alongside a complete mechanization
+(including the extension with products) in the Agda proof assistant, which is discussed in
+\cref{sec:calculus-agda}.
+
+% TODO extension to non-typing semantics?
 
@@
 $\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under the typing context $\ctx$, the expression
-$\EMV$ synthesizes or infers the type $\TMV$. \emph{Type analysis}, written
+$\EMV$ synthesises or infers the type $\TMV$. \emph{Type analysis}, written
 $\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear where an expression of
-type $\TMV$ is expected. The \emph{subsumption} rule (ASubsume) states that if an expression
-synthesizes a type, it may be analysed against that type or any \emph{consistent} type.
+type $\TMV$ is expected.
 
-Note that the subsumption rule requires that $\EMV$ be ``subsumable'', a judgment defined on all
-forms except lambda abstractions and if-else expressions. This restricts the usage of subsumption to
-a rule of ``last resort'', admitting the unicity of typing.
+The \emph{subsumption} rule (ASubsume) states that if an expression synthesises a type, it may be
+analysed against that type or any \emph{consistent} type. To invoke subsumption, the expression
+$\EMV$ must be ``subsumable'', written $\subsumable{\EMV}$, a judgment defined on all forms except
+lambda abstractions and if-else expressions. This restricts the usage of subsumption to a rule of
+``last resort'', admitting the unicity of typing.
 
@@
 additionally consistent with any other type, and arrow types are consistent if their components are.
-
-$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$ gives that $\TMV$ is a \emph{matched arrow type}, which means
-that it may be considered an arrow type. It is purely a technical mechanism to avoid duplication of
-rules related to arrow types.
+The USAp and UALam rules for application and lambda abstractions are nearly standard, but employ the
+judgment $\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which gives that $\TMV$ is a \emph{matched arrow
+type}, i.e. it may be considered an arrow type. It is purely a technical mechanism to avoid
+duplication of rules related to arrow types.
 
 \input{figs/calculus-type-judgments}
+
+The branches of if-else expressions may synthesize different types $\TMV_1$ and $\TMV_2$ as long as
+they are consistent. The entire expression then synthesizes the join, or lowest upper bound,
+$\TJoin{\TMV_1}{\TMV_2}$.
 
@@
 
-  \item $\ECInconType{\ECMV}$, the \emph{inconsistent types mark}, occurs when the synthesized type
+  \item $\ECInconType{\ECMV}$, the \emph{inconsistent types mark}, occurs when the synthesised type
     of $\ECMV$ is inconsistent with the required type.
@@
 Marked expressions are also typed by a bidirectional system, which, for most forms, closely mirrors
-the rules for corresponding unmarked forms. \Cref{fig:calculus-typing-mexp} contains a selection of
-the ``interesting'' cases, particularly those concerned with error markings.
+the rules for corresponding unmarked forms. Additional rules handle the cases when certain premises
+are not met, requiring then that expressions be wrapped in appropriate error marks.
+\Cref{fig:calculus-typing-mexp} contains a selection of these ``interesting'' cases.
+
+In parallel with the rule governing bound variables, MSUnbound types unbound variables. Such
+variables must be marked unbound, and since nothing may be said about their types, they synthesise
+the unknown type. MSInconsistentBranches corresponds to the USIf, handling the case that the
+branches synthesise inconsistent types and hence have no join. Such forms are required to be marked
+as such.
 
@@
 
-For unbound variables $\ECUnbound{x}$, nothing may be said about their types; hence, they synthesize
-the unknown type. Inconsistent branch forms behave similarly, as inconsistent types have no join.
+In addition to a synthesis rule for lambda abstractions that mirrors USLam, observe that there are
+three, instead of one, analytic lambda rules. MALam1 applies to $\ECLam{x}{\TMV}{\ECMV}$ if the type
+$\TMV_3$ analysed against is a matched arrow type $\TArrow{\TMV_1}{\TMV_2}$ and the input type
+$\TMV_1$ is consistent with the binding ascription $\TMV$. This corresponds to UALam.
 
-There are three rules concerned with the typing of lambda abstractions. MALam1, applies to
-$\ECLam{x}{\TMV}{\ECMV}$ if the type $\TMV_3$ analysed against is a matched arrow type
-$\TArrow{\TMV_1}{\TMV_2}$ and input type $\TMV_1$ is consistent with the binding ascription $\TMV$.
-This corresponds to the lambda rule for unmarked expressions. However, we are additionally concerned
-with the cases when $\TMV_3$ is not a matched arrow type or $\TMV_1$ is inconsistent with $\TMV$.
-These are described by MALam2 and MALam3.
+However, we are additionally concerned with the cases when $\TMV_3$ is not a matched arrow type or
+$\TMV_1$ is inconsistent with $\TMV$. These are described by MALam2 and MALam3. The absence of a
+matched arrow type means that $\TMV_3$ must be inconsistent the type synthesised by the lambda, so
+the former appropriately requires that the lambda be inside an inconsistent types mark. In this
+case, as no output type is known, the body is analysed against the unknown type. Similarly,
+inconsistency between the matched input type and binding ascription also implies that $\TMV_3$ is
+inconsistent with the synthesised type; MALam3 also requires that the lambda be marked inconsistent,
+and the body is analysed against $\TMV_2$.
 
-Alongside the standard subsumption rule, in which an expression that synthesizes a type may be
-analysed against any consistent type, MAInconsistentTypes governs most occurrences of the
-inconsistent types marking: $\ECInconType{\ECMV}$ analyzes against any type as long as $\ECMV$
-synthesizes an inconsistent type. In both cases, we stipulate that subsumption may be applied to
-$\ECMV$, written $\subsumable{\ECMV}$. In particular, lambda abstractions and if-else forms are
-\emph{not} subsumable, formalizing these rules as ones of ``last resort'', preserving determinism of
-typing.
+% TODO mention that every expression synthesises?
 
-Similarly to lambda abstractions, two rules govern application. In the unmarked syntax, the
-application $\EAp{\EMV_1}{\EMV_2}$ was well-typed if the type $\TMV$ synthesized by $\EMV_1$ was a
-matched arrow type. This corresponds to the MSAp1 rule for marked expressions. In the case that
-$\TMV$ is not a matched arrow type, we require an inconsistent types marking around the
+Two synthesis rules govern application. In the unmarked language, the application
+$\EAp{\EMV_1}{\EMV_2}$ is well-typed if the type $\TMV$ synthesised by $\EMV_1$ is a matched arrow
+type $\TArrow{\TMV_1}{\TMV_2}$ and the argument $\EMV_2$ analyses against $\TMV_1$. This corresponds
+to the MSAp1 rule for marked expressions. In the case that $\TMV$ is not a matched arrow type, i.e.
+it may not be applied, we require $\ECMV_1$ to be inside an inconsistent types mark. In this way, the
+blame is placed upon $\ECMV_1$, and this approach is in accordance with the meaning of the
+inconsistent types mark.
+
+It is worth noting, however, that $\ECMV_1$ is not actually in analytic position but in synthetic
+position. Treating such ``constrained synthetic'' mode specially, we may instead designate a
+specialized mark, written $\ECInconMatchedArrow{\ECMV_1}$, for such applications. This yields the
+following rule:
+\[%
+  \judgment{
+    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
+    \notMatchedArrow{\TMV} \\
+    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
+  }{
+    \ctxSynType{\ctx}{\ECApNonMatchedAltAlt{\ECMV_1}{\ECMV_2}}{\TUnknown}
+  }{MSAp2'}
+\]%
+It then becomes natural to extend this to other elimination forms that require the handling of
+unmatched types, such as products. We demonstrate such an extension in \cref{sec:calculus-products}.
+
+Finally, alongside the standard subsumption rule, MAInconsistentTypes governs most occurrences of
+the inconsistent types mark $\ECInconType{\ECMV}$ analyses against any type as long as $\ECMV$
+synthesises an inconsistent type. In both cases, we again stipulate that subsumption may be applied
+to $\ECMV$, written $\subsumable{\ECMV}$, for similar reasons as before.
 
 \subsection{Mark Insertion}
+\label{sec:calculus-marking}
 
@@
 
-\subsection{Alternative Formulations}
+% Well-typed expressions have no resulting marks
 
-\subsection{Non-Gradual System}
-\label{sec:calculus-non-gradual}
-
-% TODO extension to non-typing semantics?
+\subsection{Adding Products}
+\label{sec:calculus-products}
 

### 2023-02-26 calculus.tex
Begin describing mark insertion

@@
 The branches of if-else expressions may synthesize different types $\TMV_1$ and $\TMV_2$ as long as
-they are consistent. The entire expression then synthesizes the join, or lowest upper bound,
+they are consistent. The entire expression then synthesizes the \emph{join}
 $\TJoin{\TMV_1}{\TMV_2}$.
@@
 
+The central component of the calculus is \emph{mark insertion}, a bidirectional system given by the
+following mutually defined judgments. The synthetic insertion judgment
+$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that under the context $\Gamma$, the
+unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
+$\TMV$. Analogously, the analytic insertion judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
+states that the unmarked expression $\EMV$ is marked into $\ECMV$, which analyses against type
+$\TMV$.
+
+% TODO need to better explain why it is bidirectional
+
+\input{figs/calculus-mark-insertion}
+
+The rules for mark insertion are given in \Cref{fig:calculus-mark-insertion}. Observe that they
+mirror the typing rules for marked expressions closely: all possible outcomes for fallible premises
+are handled and markings are inserted as needed. For example, ISVar and ISUnbound together handle
+variable usages. The former yields the same variable in the marked language when it is bound. The
+latter marks the variable as unbound if it is not found in the context.
+
+We again find two rules governing the marking of function application. ISAp1 corresponds to MSAp1 as
+the ``success case'', recursively marking $\EMV_1$ and $\EMV_2$. Accordingly, ISAp2 parallels MSAp2
+as the ``error case''. Similarly, ISIf and ISInconsistentBranches correspond to MSIf and
+MSInconsistentBranches, and IALam1, IALam2, and IALam3 correspond to MSLam1, MSLam2, and MSLam3.
+
 \subsection{Correctness}

### 2023-02-26 calculus.tex
Describe marking metatheorems, start product extension section

@@
+% TODO standardize naming: marking vs mark insertion, marks vs error marks vs markings
+% TODO uniform style for referencing rules
+
 \section{Motivating Examples}
@@
 performed in an \emph{ad hoc} manner [citations?].
+
+% TODO examples of this ad hoc stuff
 
@@
 
-\subsection{Correctness}
-\begin{theorem}[name=Marking Totality]
-  Any syntactically well-formed program may be marked for errors.
-\end{theorem}
-
-In the next section, we will formally develop
-
 \section{Marked Lambda Calculus}
@@
   \item \textbf{Marked expressions} (\cref{sec:calculus-mexp}), which mirror the structure of
-    unmarked expressions but are extended with \textbf{error markings}. Every marked expression
+    unmarked expressions but are extended with \textbf{error marks}. Every marked expression
     corresponds to an unmarked expression by \emph{mark erasure}.
+    % TODO more intuition for what marked expressions are?
 
-  \item \textbf{Mark insertion} (\cref{sec:calculus-marking}), which \textbf{marks} any unmarked expression into a marked
+  \item \textbf{Marking} (\cref{sec:calculus-marking}), which \textbf{marks} any unmarked expression into a marked
     expression, inserting error markings where appropriate.
@@
 
-In \cref{sec:calculus-correctness}, we discuss the metatheory that ensures the system---and any such
-error localization semantics---to be correct. We follow then by extending the calculus with support
-for product types. In the interest of space, this overview focuses on ``interesting'' cases. All
-rules and theorems may be found in the supplementary appendix, alongside a complete mechanization
-(including the extension with products) in the Agda proof assistant, which is discussed in
+Key to ensuring correctness of the system---and, indeed, any such error localization semantics---is
+a metatheorem, \emph{marking totality}, which guides the construction of rules:
+%
+\begin{theorem}[name=Marking Totality]
+  Any syntactically well-formed unmarked expression may be marked into a marked expression.
+\end{theorem}
+%
+We will give a more concrete statement of the metatheorem after formally introducing the components
+of the system.
+
+In \cref{sec:calculus-products}, we follow with an extension of the calculus to support binary
+product types. In the interest of space, this overview focuses on ``interesting'' cases. All rules
+and theorems may be found in the supplementary appendix, alongside a complete mechanization
+(including the extension with binary products) in the Agda proof assistant, which is discussed in
 \cref{sec:calculus-agda}.
@@
 
-\subsection{Mark Insertion}
+\subsection{Marking}
 \label{sec:calculus-marking}
 
-The central component of the calculus is \emph{mark insertion}, a bidirectional system given by the
+The central component of the calculus is \emph{marking}, a bidirectional system given by the
 following mutually defined judgments. The synthetic insertion judgment
-$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that under the context $\Gamma$, the
+$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
 unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
@@
 
-The rules for mark insertion are given in \Cref{fig:calculus-mark-insertion}. Observe that they
+The rules for mark insertion are given in \cref{fig:calculus-mark-insertion}. Observe that they
 mirror the typing rules for marked expressions closely: all possible outcomes for fallible premises
@@
 MSInconsistentBranches, and IALam1, IALam2, and IALam3 correspond to MSLam1, MSLam2, and MSLam3.
+The rules IAInconsistentTypes and IASubsume, of course, act in parallel to MAInconsistentTypes and
+MASubsume.
 
-\subsection{Correctness}
-\label{sec:calculus-correctness}
+We may now formally restate the critical metatheorem of marking totality:
+%
+\begin{theorem}[name=Marking Totality] \
+  \label{thm:calculus-marking-totality}
+  \begin{enumerate}
+    \item For all $\ctx$ and $\EMV$, there exist $\ECMV$ and $\TMV$ such that
+      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
+  \item For all $\ctx$, $\EMV$, and $\TMV$, there exists $\ECMV$ such that
+    $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
+  \end{enumerate}
+\end{theorem}
 
-% Totality
+Additionally, it is important that marking preserves the original syntactic structure modulo error
+marks; it would be trivial to satisfy totality by marking all expressions into numbers, for example.
+This property is ensured by the following well-formedness metatheorem:
+%
+\begin{theorem}[name=Marking Well-Formedness] \
+  \label{thm:calculus-marking-well-formedness}
+  \begin{enumerate}
+    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynType{\ctx}{\ECMV}{\TMV}$ and
+      $\erasesTo{\ECMV}{\EMV}$.
+    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ and
+      $\erasesTo{\ECMV}{\EMV}$.
+  \end{enumerate}
+  % TODO split this into two theorems?
+  % TODO not really sure how to intuitively describe the part about types matching up
+\end{theorem}
+%
+In other words, if an unmarked expression is marked into some marked expression, erasing error marks
+from the marked expression gives the original unmarked expression. To ensure that we place error
+marks in the correct spots, we furthermore require the following theorem:
+%
+\begin{theorem}[name=Marking of Well-Typed/Ill-Typed Expressions] \ % TODO better name for this
+  \label{thm:calculus-marking-well-ill-typed}
+  \begin{enumerate}
+    \item \begin{enumerate}
+        \item If $\ctxSynType{\ctx}{\EMV}{\TMV}$, then $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
+          such that $\imageIs{\EMV}{\ECMV}$.
+        \item If $\ctxAnaType{\ctx}{\EMV}{\TMV}$, then $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
+          such that $\imageIs{\EMV}{\ECMV}$.
+      \end{enumerate}
 
-% Well-formedness conditions
+    \item \begin{enumerate}
+        \item If there does not exist $\TMV$ such that $\ctxSynType{\ctx}{\EMV}{\TMV}$, then for all
+          $\ECMV$ and $\TMV'$ such that $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, we have
+          $\imageIsNot{\EMV}{\ECMV}$.
+        \item If there does not exist $\TMV$ such that $\ctxAnaType{\ctx}{\EMV}{\TMV}$, then for all
+          $\ECMV$ and $\TMV'$ such that $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, we have
+          $\imageIsNot{\EMV}{\ECMV}$.
+      \end{enumerate}
+  \end{enumerate}
+\end{theorem}
+%
+That is, since error marks indicate type errors in the program, a well-typed program should, of
+course, have no error marks. The converse---that marks are inserted into ill-typed programs---is
+also true. We have chosen to formulate the absence of error marks through an inclusion mapping of
+unmarked expressions into marked expressions; alternatively, a predicate over marked expressions
+could be used. % TODO latter is easier to do in Agda, but maybe it is weaker?
 
-% Well-typed expressions have no resulting marks
+Finally, no less importantly, marking is deterministic. This is given by
+\cref{thm:calculus-marking-unicity}.
+%
+\begin{theorem}[name=Marking Unicity] \
+  \label{thm:calculus-marking-unicity}
+  \begin{enumerate}
+    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV_1}$ and
+      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV_2}$, then $\ECMV_1 = \ECMV_2$ and $\TMV_1 =
+      \TMV_2$.
+    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV_1}$ and
+      $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV_2}$, then $\ECMV_1 = \ECMV_2$ and $\TMV_1 =
+      \TMV_2$.
+  \end{enumerate}
+\end{theorem}
+%
+Together with totality, marking may be considered and implemented as a total function.
 
-\subsection{Adding Products}
+\subsection{Adding Binary Products}
 \label{sec:calculus-products}
+
+We now demonstrate how one might extend or adapt the marking calculus for new typing features by
+adding binary product types. First, we extend in \cref{fig:calculus-products-judgments} the syntax
+of types and notions of consistency and join. As with arrow types, we additionally define an
+auxiliary notion of \emph{matched binary product types}: the judgment
+$\matchedPair{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be considered the binary product
+type $\TPair{\TMV_1}{\TMV_2}$.
+
+\input{figs/calculus-products-judgments}
+
+The unmarked language with a pair constructor and projection operators.
+\Cref{fig:calculus-products-uexp} gives their syntax and typing rules. Similarly to application,
+\textsc{USProjL} and \textsc{USProjR} employ the matched binary product type; a projection
+$\EProjL{\EMV}$ or $\EProjR{\EMV}$ is then well-typed if the $\EMV$ may be considered to be a pair.
+
+\input{figs/calculus-products-uexp}
+
+\input{figs/calculus-products-mexp}
+
+\input{figs/calculus-products-marking}
 
@@
 mechanization's documentation contains more details regarding technical decisions made therein.
+
+\subsection{Implementation in Hazel}
+
+\subsection{Reified Hazelnut Action Semantics}

### 2023-02-26 abstract.tex
Updates from Overleaf

--- 2023-02-23 abstract.tex
@@
 \begin{abstract}
-Type systems typically only define the conditions under which an expression is well-typed, leaving ill-typed expressions formally meaningless.
-This approach is insufficient for developing language servers for modern programming environments, which are expected to sensibly localize type errors and continue to provide a variety of semantic services in their presence.
-Lacking formal foundations, error localization has largely become a matter of folklore in the community.
-Recent work on the Hazelnut calculus takes a step toward assigning formal meaning to erroneous programs, but it stops short of a comprehensive account.
-In particular,  (1) Hazelnut is able to reason around marked type inconsistencies, but not several other kinds of static errors, and (2) Hazelnut provides an insufficient account of the marking procedure itself, relying on a limited structure editor, or on the user, to correctly insert marks.
-This paper addresses these deficiencies, contributing the first comprehensive formal account of type error localization, called the marked lambda calculus.
-This contribution is independently useful as a core calculus for language servers.
-In addition, we deploy its total marking procedure to fix several instances of undefined behavior in the Hazelnut structure editor calculus, namely by allowing edits that introduce non-local errors.
-We mechanize the calculus in Agda and implement an extension in Hazel, a full-scale live functional programming environment previously based on Hazelnut.
+Type systems typically only define the conditions under which an expression is well-typed, leaving ill-typed expressions entirely meaningless. 
+This approach is insufficient as the basis for language servers for modern programming environments, which are expected to sensibly localize type errors and recover, i.e.
+continue to provide a variety of semantic services in their presence.
+This paper addresses this problem, contributing the first comprehensive formal account of type error localization and recovery, called the marked lambda calculus.
+This calculus, which is intended as a core calculus for language servers, consists of a gradual type system for expressions with marked errors and a total marking procedure on unmarked expressions with holes.
+We mechanize the metatheory of the marked lambda calculus in Agda, deploy its total marking procedure to fix several instances of undefined behavior in the Hazelnut structure editor calculus, and implement the system as the new basis for Hazel, a full-scale live functional programming environment.
 
@@
 Hindley-Milner) are notorious for complicating error localization.
-We approach the problem of  constraint-based type inference in a novel manner: we deploy constraint solving as a layer atop a gradual bidirectionally typed core, where it infers type hole fillings.
-This system is able to elegantly handle situations where the system of constraints cannot be solved, because the unification algorithm we develop localizes errors arising from inconsistent constraints to individual type holes, rather than attempting to localize them to particular expressions, which is fundamentally \emph{ad hoc}.
-The user can resolve these errors by selecting from suggested partial solutions, or leave the errors unresolved, in which case they operate as the unknown type from gradual typing.
-We formally specify this approach and implement it in Hazel.
+We approach the problem of incorporating constraint-based type inference by deploying constraint solving as a type hole filling layer atop a gradual bidirectionally typed core.
+When the system of constraints cannot be fully solved, the unification algorithm we develop localizes errors arising from inconsistent constraints to individual type holes, rather than attempting to localize them in an \emph{ad hoc} manner to one particular expression and then failing to recover from there.
+The user can resolve these errors by selecting from suggested partial solutions, or leave the errors unresolved, in which case the type hole recovers gradually.
+We formally specify this type hole inference system and implement it in Hazel.
 \end{abstract}
+
+% Lacking formal foundations, error localization has largely become a matter of folklore in the community. 
+% Recent work on the Hazelnut calculus takes a step toward assigning formal meaning to erroneous programs, but it stops short of a comprehensive account. 
+% In particular,  (1) Hazelnut is able to reason around marked type inconsistencies, but not several other kinds of static errors, and (2) Hazelnut provides an insufficient account of the marking procedure itself, relying on a limited structure editor, or on the user, to correctly insert marks.\todo{decenter Hazelnut?}

### 2023-02-26 appendixA.tex
Updates from Overleaf

\newpage
\section{Appendix A: An implementation of union-findable Possible Type Sets}

\begin{figure}[h!]
\begin{lstlisting}[escapeinside={(*}{*)}]
open UnionFind

type t = possible_types UnionFind.elem
and possible_type_set = possible_type list
and possible_type =
  | Unknown(provenance)
  | Num
  | Bool
  | Arrow of t * t;

let rec merge_sets_UF (s1: t, s2: t): t =
    let merged_typs = merge_sets (UnionFind.get s1) (UnionFind.get s2) in
    let final_rep = UnionFind.union s1 s2 in
    UnionFind.set final_rep merged_typs;
    final_rep
and merge_sets (s1: possible_type_set, s2: possible_type_set): possible_types = 
    match (s2)
    | [] -> s1
    | s2_hd::s2_hd -> merge_sets (add s1 s2_hd) s2_hd
and add (s: possible_type_set, t: possible_type): possible_type_set =
    match (s)
    | [] -> [t]
    | s_hd::s_tl -> (
        match (s_hd, t)
        | (Arrow(s_hd_L, s_hd_R), Arrow(t_L, t_R)) ->
            let left = merge_sets_UF s_hd_L t_L in
            let right = merge_sets_UF s_hd_R t_R in
            Arrow(left, right)::s_tl
        | _ -> if (s_hd = t) then s else s_hd::(add s_tl t)
    )

let union => if (UnionFind.eq t1 t2) then () else merge_sets_UF t1 t2;
\end{lstlisting}
\vspace{-4px}
 \hrule
\caption{An Implementation of \textit{PossibleTypeSet} using UnionFind}
\label{fig:ufalgcode}
\end{figure}

### 2023-02-26 intro.tex
Updates from Overleaf

\section{Introduction}
\label{sec:introduction}

Type errors are both a boon and a bane to developers.
At their best, type errors help programmers quickly and accurately identify the location and cause of a genuine problem in their program that is blocking progress on their task. 
At their worst, type errors mislead the programmer / distract them away from the parts of the program that are of interest, and 
prevent evaluation and editor services from working, even when the error is benign or located at an irrelevant location in the program. 
(May be one reason some developers favor dynamic languages, especially when prototyping)

For these reasons, accurate type error localization and robust type error recovery have been a topic of considerable practical interest. 
Indeed, essentially all widely-used programming environments (often via an underlying language server) have some support for type error 
localization and many also attempt recovery in some situations. However, the design of these systems has largely remained a matter of folklore or empirical iteration. 
One reason for this is 
that type system definitions typically only define the conditions under which an expression is well-typed, e.g. in the style of the simply typed lambda calculus and its derivatives. When there is a type error anywhere in the program, it is everywhere meaningless according to definitions in this style, i.e. there is no direct specification for how to reason within an erroneous program. Different language servers or typecheckers for the same language can and do attempt to localize, report, and recover from these errors in different ways, with little in the way of unifying theory.

The primary contribution of this paper is to provide these missing theoretical foundations for type error localization and recovery. 
Start assuming a system with local type inference / bidirectional typing, marked lambda calculus(Give judgement forms right off the bat??) 
(Mention gradual typing here?)

We build on the foundation of the Hazelnut type system, which allows for local type inconsistencies to be placed within a non-empty hole, 
e.g. $1 + false$ is meaningless but $1 + \{false\}$ can be given type Num. In other words, the non-empty holes serves as a membrane around type inconsistencies, here between the expected type, Num, and the synthesized type, Bool. Problems:

1. does not handle every kind of error that can arise, e.g. inconsistent branch types, unbound variables, etc.

2. does not provide a general account for how to correctly place the non-empty holes -- in that paper, they assume that a structure editor will be used and show how the structure editor can determine whether a non-empty hole is needed at the cursor after an edit action, but edits that might cause non-local errors to appear are undefined. users can also manually mark these locations, similar to Agda's non-empty holes. (Examples?)

To solve both of these problems, we establish totality of marking -- we prove it to be total (and to insert only marks, without making any other changes), i.e. it must handle 
every possible syntactically well-formed program and thus every error. Mechanized in Agda.

Marking does not rely on a structure editor, so it is suitable as a foundational calculus for language servers supporting editors of any 
design, both text editors (with syntax error recovery, a well-studied topic that we consider outside of the scope of this paper) and structure editors. To support this claim, we have implemented the calculus in a version of Hazel built atop tylr, which features both a text editor with explicit syntax for holes (a la Agda, GHC, etc.) and a textual structure editor with syntax error recovery via automatic hole insertion. Maximal liveness guarantee!

Additional foundational contribution, we have also defined two versions of Hazelnut that eliminate the undefined behavior just mentioned. Composed with marking, this solves the problem. Also briefly consider incremental remarking in the style of Hazelnut.

As we go, we'll discuss various localization design decisions that the language designer should consider when specifying the marked type system for their language. Overall, though, our work validates the folklore that systems that rely on local type inference tend to have an easier time deciding how to localize type errors (Pierce paper). Many modern languages feature local type inference, including .... 
However, non-local constraint-based type inference is also a common language feature, and it is known to substantially complicate error localization and recovery. 
Heuristic approaches based on ML-based approaches, weighting, etc.
In this paper, we consider an alternative approach to incorporating non-local constraints. Rather than ...

### 2023-02-27 calculus.tex
Rewrite marked lambda caclulus section as discussed

@@
 
-\section{Motivating Examples}
+\section{Marked Lambda Calculus}
+\label{sec:calculus}
+
+Programmers often work with program states that are syntactically well-formed but not well-typed.
+\Cref{fig:calculus-examples} shows common type errors as they appear in the Hazel programming
+environment. For example, a simple error that may occur is the use of an unbound variable. As in
+\cref{fig:calculus-examples-unbound}, editor services indicate the usage of unbound $y$ is invalid
+by highlighting it in red.
+
+\input{figs/calculus-examples}
+
+Sometimes, the type an expression \emph{should} be is known. Consider
+\cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
+$+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
+indicates a mismatch between the expected and actual types of the operand. The application in
+\cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
+this case, however, the expectation is not any singular type, but any member of a family of function
+types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error of inconsistent
+branches: the branches of the if-else expression are of different types. Hazel opts to highlight the
+entire expression in such a situation, while other tools may choose one branch to be ``correct'' and
+indicate a type mismatch error on the other.
+
+% TODO elaborate on this part
+
+Unfortunately, formal typing semantics usually only give meaning to well-typed programs, leaving
+such ill-typed ones meaningless. This is insufficient for editor services and language servers,
+which must provide users feedback on where and why a program is ill-typed. Moreover, they must
+continue providing editor services even when an error is encountered. Indeed, similarly to
+error-recovering parsers, real-world type checkers must recover from type errors. Traditionally,
+this has been performed in an \emph{ad hoc} manner [citations?], with different languages and tools
+adopting varying approaches in the localization, reportage, and recovery from errors.
+
+% TODO examples of this ad hoc stuff
+
+The key contribution of this section is a formal calculus, called the \emph{marking calculus}, for
+type error localization in bidirectional systems, a judgmental framework provides a unified theory
+for the design of typing semantics that supports ill-typed programs. \Cref{sec:calculus-calculus}
+describes its form and metatheory by example on a minimal gradually typed lambda calculus, extending
+with binary product types in \cref{sec:calculus-products}. All rules and theorems may be found in
+the supplementary appendix, alongside a complete mechanization in the Agda proof assistant, which is
+discussed in \cref{sec:calculus-agda}. Finally, in \cref{sec:calculus-poly}, we consider how the
+system might be extended to more complex typing features, such as System F-style polymorphism.
+
+% TODO not really sure when to start talking about bidirectional typing
+
 \emph{Bidirectional typing} combines type checking, which determines if a program satisfies a given
@@
 
-Often, programmers are working with program states that are syntactically well-formed but not
-well-typed. However, formal typing semantics usually only give meaning to well-typed programs,
-leaving ill-typed ones meaningless. This is insufficient for editor services and language servers,
-which give users feedback on where and why a program is ill-typed. Traditionally, this has been
-performed in an \emph{ad hoc} manner [citations?].
+\subsection{The Calculus}
+\label{sec:calculus-calculus}
 
-% TODO examples of this ad hoc stuff
+% TODO outline the structure of this section at the beginning
+% TODO make it clear unmarked/marked languages share same types
 
-In this section, we motivate a formal calculus for bidirectional type error localization by
-exploring different kinds of type errors that may occur.
-
-% TODO state goals: prefer fine-grained over coarser blaming/marking
-
-\subsection{Unbound Variables}
-A simple and common example of an error that may occur in programs is the use of an unbound
-variable. In this case, editor services often indicate visually that such usages are invalid.
-
-% TODO image of an editor for this example? not sure what else to say about this
-
-\subsection{Inconsistent Types}
-Often, the type an expression \emph{should} be is known. Both operands of an addition expression
-$\EPlus{\ETrue}{\ENumMV}$, for example, should have the type $\TNum$. In a bidirectional system,
-$\ETrue$ and $\ENumMV$ would be analysed against $\TNum$, in which the former check would fail:
-$\ETrue$ synthesises $\TBool$, which is inconsistent with $\TNum$.
-
-% An editor may indicate such an error in different ways. Two obvious ones are considered below: (1)
-% indicating an error on $\ETrue$, or (2) indicating an error on the entire expression
-% $\EPlus{\ETrue}{\ENumMV}$.
-% \[%
-  % \EPlus{\ETrue}{\ENumMV} \qquad \EPlus{\ETrue}{\ENumMV}
-% \]%
-% The former ``blames'' the error on the operand,
-
-% Since the ``fault'' of the error
-
-% TODO cite real editor/language?
-
-Since expressions typed in analytic mode are checked to satisfy a correct, known type, they are
-given ``blame'' when ill-typed.
-
-\subsection{Inconsistent Branches}
-
-\section{Marked Lambda Calculus}
-\label{sec:calculus}
-
-The previous section surveyed different type error cases. Now, we formally introduce a \emph{marking
-calculus} for bidirectional type error localization on the gradually typed lambda calculus above.
-This comes as a two-layer system:
-%
-\begin{itemize}
-  \item \textbf{Unmarked expressions} (\cref{sec:calculus-uexp}), which compose the surface-level
-    syntax. Types classify unmarked expressions according to a standard bidirectional gradual typing
-    semantics.
-
-  \item \textbf{Marked expressions} (\cref{sec:calculus-mexp}), which mirror the structure of
-    unmarked expressions but are extended with \textbf{error marks}. Every marked expression
-    corresponds to an unmarked expression by \emph{mark erasure}.
-    % TODO more intuition for what marked expressions are?
-
-  \item \textbf{Marking} (\cref{sec:calculus-marking}), which \textbf{marks} any unmarked expression into a marked
-    expression, inserting error markings where appropriate.
-\end{itemize}
-%
-That is, we extend the ordinary typing semantics of a language with a secondary ``marked'' language
-and define a mapping from ``unmarked'' programs to ``marked'' programs. Note that while the
-semantics that follow concern a gradually typed language, a non-gradual system may be recovered
-straightforwardly by removing the unknown type and replacing type consistency with equality.
-
-Key to ensuring correctness of the system---and, indeed, any such error localization semantics---is
-a metatheorem, \emph{marking totality}, which guides the construction of rules:
-%
-\begin{theorem}[name=Marking Totality]
-  Any syntactically well-formed unmarked expression may be marked into a marked expression.
-\end{theorem}
-%
-We will give a more concrete statement of the metatheorem after formally introducing the components
-of the system.
-
-In \cref{sec:calculus-products}, we follow with an extension of the calculus to support binary
-product types. In the interest of space, this overview focuses on ``interesting'' cases. All rules
-and theorems may be found in the supplementary appendix, alongside a complete mechanization
-(including the extension with binary products) in the Agda proof assistant, which is discussed in
-\cref{sec:calculus-agda}.
-
-% TODO extension to non-typing semantics?
-
-\subsection{Unmarked Expressions}
-\label{sec:calculus-uexp}
-
-We consider a gradually typed lambda calculus extended with numbers and conditionals, whose syntax
-and bidirectional typing are given in \cref{fig:calculus-syntax-example} and
-\cref{fig:calculus-typing-example}, respectively. The base types $\TNum$ and $\TBool$ classify
-number and boolean expressions. The number literal corresponding to the mathematical number $n$ is
-given by $\ENumMV$, and there is a single addition operation on numeric expressions. $\ETrue$ and
-$\EFalse$ correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, and
+We now formally introduce the \emph{marking calculus} through its application to a gradually typed
+lambda calculus extended with numbers and booleans. Consider the syntax for such a language, as
+shown in \cref{fig:calculus-syntax-uexp}. The base types $\TNum$ and $\TBool$ classify number and
+boolean expressions. The number literal corresponding to the mathematical number $n$ is given by
+$\ENumMV$, and there is a single addition operation on numeric expression. $\ETrue$ and $\EFalse$
+correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
 $\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives an if-else structure. In addition to these forms, $\TUnknown$
-indicates an unknown type.
+is the unknown type from gradual type theory.
 
@@
 
-Bidirectional typing employs two mutually defined judgments. \emph{Type synthesis}, written
-$\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under the typing context $\ctx$, the expression
-$\EMV$ synthesises or infers the type $\TMV$. \emph{Type analysis}, written
+The language is typed bidirectionally, which employs two mutually defined judgments. \emph{Type
+synthesis}, written $\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under the typing context
+$\ctx$, the expression $e$ synthesises or infers the type $\TMV$. \emph{Type analysis}, written
 $\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear where an expression of
-type $\TMV$ is expected.
+type $\TMV$ is expected. We may give these typing semantics traditionally, without considering error
+localization and recovery.
 
-The \emph{subsumption} rule (ASubsume) states that if an expression synthesises a type, it may be
-analysed against that type or any \emph{consistent} type. To invoke subsumption, the expression
-$\EMV$ must be ``subsumable'', written $\subsumable{\EMV}$, a judgment defined on all forms except
-lambda abstractions and if-else expressions. This restricts the usage of subsumption to a rule of
-``last resort'', admitting the unicity of typing.
+The \emph{marking calculus} builds atop these typing semantics into a two-layer system:
+%
+\begin{itemize}
+  \item The \textbf{unmarked language}, which is the original language. Expressions of this language
+    are \emph{unmarked expressions}.
 
-% \begin{theorem}[name=Unmarked Expression Typing Unicity] \
-  % If $\ctxSynType{\ctx}{\EMV}{\TMV}$ and $\ctxSynType{\ctx}{\EMV}{\TMV'}$, then $\TMV = \TMV'$.
-% \end{theorem}
+  \item The \textbf{marked language}, which mirrors the structure of the unmarked language but is
+    extended with explicit \emph{error marks}. We call expressions of this language to be
+    \emph{marked expressions}, denoted $\ECMV$.
+    % TODO more intuition for what marked expressions are?
 
-\input{figs/calculus-typing-uexp}
-
-The \emph{consistency} judgment $\consistent{\TMV}{\TMV'}$ that appears in the subsumption rule
-defines a reflexive and symmetric (but not transitive) relation between types. The unknown type is
-additionally consistent with any other type, and arrow types are consistent if their components are.
-The USAp and UALam rules for application and lambda abstractions are nearly standard, but employ the
-judgment $\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which gives that $\TMV$ is a \emph{matched arrow
-type}, i.e. it may be considered an arrow type. It is purely a technical mechanism to avoid
-duplication of rules related to arrow types.
-
-\input{figs/calculus-type-judgments}
-
-The branches of if-else expressions may synthesize different types $\TMV_1$ and $\TMV_2$ as long as
-they are consistent. The entire expression then synthesizes the \emph{join}
-$\TJoin{\TMV_1}{\TMV_2}$.
-
-\subsection{Marked Expressions}
-\label{sec:calculus-mexp}
-
-Marked expressions, $\ECMV$, extend the syntax of unmarked expressions with \emph{explicit error
-marks}. As given in the syntax definition of \cref{fig:calculus-syntax-mexp}, there are three such
-forms, which mirror the examples discussed previously:
-%
-\begin{itemize} % TODO not the snappiest names
-  \item $\ECUnbound{x}$, the \emph{unbound variable mark}, denotes the usage of an unbound variable.
-
-  \item $\ECInconType{\ECMV}$, the \emph{inconsistent types mark}, occurs when the synthesised type
-    of $\ECMV$ is inconsistent with the required type.
-
-  \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, corresponds
-    to an if-else expression where the types of the two branches $\ECMV_2$ and $\ECMV_3$ are
-    inconsistent.
+  \item \textbf{Marking}, which \emph{marks} any unmarked expression into a marked expression,
+    inserting error markings where appropriate.
 \end{itemize}
 %
-Additional forms may arise as required by the language.
+In other words, we extend the ordinary typing semantics of a language with a secondary ``marked''
+language and ``mark'' unmarked programs into marked ones. This corresponds to a type checking
+process with error localization and recovery.
 
-% TODO examples of such?
+% TODO more intuition
 
+We give, for now, only part of the marked expression syntax in \cref{fig:calculus-syntax-mexp-part}.
+Throughout the remainder of this section, as we consider the formulation of marking, we will add the
+necessary error mark forms.
+
+% TODO include holes
 \begin{figure}[htbp]
@@
                        \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV} \\
-             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
+             &       & \mid         & \cdots
+             % &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
   \end{array}\]
-  \caption{Syntax of marked expressions}
-  \label{fig:calculus-syntax-mexp}
+  \caption{Partial syntax of marked expressions}
+  \label{fig:calculus-syntax-mexp-part}
 \end{figure}
 
-The unmarked version of a marked expression $\ECMV$ may be recovered through \emph{mark erasure},
-given by $\erase{\ECMV}$. As in \Cref{fig:calculus-mark-erasure}, this total function may be defined
-inductively in a straightforward manner. Observe that erasing the inconsistent branches mark gives
-an if-else expression with sub-expression marks recursively erased.
-
-\input{figs/calculus-mark-erasure}
-
-Marked expressions are also typed by a bidirectional system, which, for most forms, closely mirrors
-the rules for corresponding unmarked forms. Additional rules handle the cases when certain premises
-are not met, requiring then that expressions be wrapped in appropriate error marks.
-\Cref{fig:calculus-typing-mexp} contains a selection of these ``interesting'' cases.
-
-In parallel with the rule governing bound variables, MSUnbound types unbound variables. Such
-variables must be marked unbound, and since nothing may be said about their types, they synthesise
-the unknown type. MSInconsistentBranches corresponds to the USIf, handling the case that the
-branches synthesise inconsistent types and hence have no join. Such forms are required to be marked
-as such.
-
-\input{figs/calculus-typing-mexp}
-
-% TODO Not really sure how to explain the syntax and typing of marked expressions in an intuitive
-% way in relation to unmarked expressions.
-
-In addition to a synthesis rule for lambda abstractions that mirrors USLam, observe that there are
-three, instead of one, analytic lambda rules. MALam1 applies to $\ECLam{x}{\TMV}{\ECMV}$ if the type
-$\TMV_3$ analysed against is a matched arrow type $\TArrow{\TMV_1}{\TMV_2}$ and the input type
-$\TMV_1$ is consistent with the binding ascription $\TMV$. This corresponds to UALam.
-
-However, we are additionally concerned with the cases when $\TMV_3$ is not a matched arrow type or
-$\TMV_1$ is inconsistent with $\TMV$. These are described by MALam2 and MALam3. The absence of a
-matched arrow type means that $\TMV_3$ must be inconsistent the type synthesised by the lambda, so
-the former appropriately requires that the lambda be inside an inconsistent types mark. In this
-case, as no output type is known, the body is analysed against the unknown type. Similarly,
-inconsistency between the matched input type and binding ascription also implies that $\TMV_3$ is
-inconsistent with the synthesised type; MALam3 also requires that the lambda be marked inconsistent,
-and the body is analysed against $\TMV_2$.
-
-% TODO mention that every expression synthesises?
-
-Two synthesis rules govern application. In the unmarked language, the application
-$\EAp{\EMV_1}{\EMV_2}$ is well-typed if the type $\TMV$ synthesised by $\EMV_1$ is a matched arrow
-type $\TArrow{\TMV_1}{\TMV_2}$ and the argument $\EMV_2$ analyses against $\TMV_1$. This corresponds
-to the MSAp1 rule for marked expressions. In the case that $\TMV$ is not a matched arrow type, i.e.
-it may not be applied, we require $\ECMV_1$ to be inside an inconsistent types mark. In this way, the
-blame is placed upon $\ECMV_1$, and this approach is in accordance with the meaning of the
-inconsistent types mark.
-
-It is worth noting, however, that $\ECMV_1$ is not actually in analytic position but in synthetic
-position. Treating such ``constrained synthetic'' mode specially, we may instead designate a
-specialized mark, written $\ECInconMatchedArrow{\ECMV_1}$, for such applications. This yields the
-following rule:
-\[%
-  \judgment{
-    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
-    \notMatchedArrow{\TMV} \\
-    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
-  }{
-    \ctxSynType{\ctx}{\ECApNonMatchedAltAlt{\ECMV_1}{\ECMV_2}}{\TUnknown}
-  }{MSAp2'}
-\]%
-It then becomes natural to extend this to other elimination forms that require the handling of
-unmatched types, such as products. We demonstrate such an extension in \cref{sec:calculus-products}.
-
-Finally, alongside the standard subsumption rule, MAInconsistentTypes governs most occurrences of
-the inconsistent types mark $\ECInconType{\ECMV}$ analyses against any type as long as $\ECMV$
-synthesises an inconsistent type. In both cases, we again stipulate that subsumption may be applied
-to $\ECMV$, written $\subsumable{\ECMV}$, for similar reasons as before.
-
-\subsection{Marking}
-\label{sec:calculus-marking}
-
-The central component of the calculus is \emph{marking}, a bidirectional system given by the
-following mutually defined judgments. The synthetic insertion judgment
+The marked language possesses its own typing semantics, also formulated bidirectionally. We write
+$\ctxSynType{\ctx}{\ECMV}{\TMV}$ for synthesis and $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ for analysis.
+Finally, marking is bidirectional as well. The synthetic marking judgment
 $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
 unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
-$\TMV$. Analogously, the analytic insertion judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
-states that the unmarked expression $\EMV$ is marked into $\ECMV$, which analyses against type
-$\TMV$.
+$\TMV$. Analogously, the analytic marking judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
+states that $\EMV$ is marked into $\ECMV$, which analyses against $\TMV$.
 
-% TODO need to better explain why it is bidirectional
-
-\input{figs/calculus-mark-insertion}
-
-The rules for mark insertion are given in \cref{fig:calculus-mark-insertion}. Observe that they
-mirror the typing rules for marked expressions closely: all possible outcomes for fallible premises
-are handled and markings are inserted as needed. For example, ISVar and ISUnbound together handle
-variable usages. The former yields the same variable in the marked language when it is bound. The
-latter marks the variable as unbound if it is not found in the context.
-
-We again find two rules governing the marking of function application. ISAp1 corresponds to MSAp1 as
-the ``success case'', recursively marking $\EMV_1$ and $\EMV_2$. Accordingly, ISAp2 parallels MSAp2
-as the ``error case''. Similarly, ISIf and ISInconsistentBranches correspond to MSIf and
-MSInconsistentBranches, and IALam1, IALam2, and IALam3 correspond to MSLam1, MSLam2, and MSLam3.
-The rules IAInconsistentTypes and IASubsume, of course, act in parallel to MAInconsistentTypes and
-MASubsume.
-
-We may now formally restate the critical metatheorem of marking totality:
+How can we ensure that the marking procedure is correctly defined? There are two critical
+metatheorems to keep in mind as we continue. The first is a \emph{totality} of marking:
 %
@@
 \end{theorem}
-
-Additionally, it is important that marking preserves the original syntactic structure modulo error
-marks; it would be trivial to satisfy totality by marking all expressions into numbers, for example.
-This property is ensured by the following well-formedness metatheorem:
+%
+That is, we may mark, i.e. type check and localize errors for \emph{any} syntactically well-formed
+program. Furthermore, marking should preserve the syntactic structure of programs, modulo error
+marks. Thus, we must define a notion of \emph{mark erasure}, which converts marked expressions back
+into unmarked expression by removing error marks. \Cref{fig:calculus-mark-erasure-part} gives the
+definition of mark erasure, written $\erase{\ECMV}$, for our initial marked expression syntax. Then,
+we may define a \emph{well-formedness} criterion for marking:
 %
@@
 In other words, if an unmarked expression is marked into some marked expression, erasing error marks
-from the marked expression gives the original unmarked expression. To ensure that we place error
-marks in the correct spots, we furthermore require the following theorem:
+from the marked expression gives the original unmarked expression. We also ensure that marking gives
+marked expressions that are well-typed; indeed, their types should match the types synthesised or
+analysed against in marking. % TODO more intuition for this theorem
+
+\input{figs/calculus-mark-erasure-part}
+
+In the remainder of this subsection, we consider the different type errors that may arise in the
+unmarked language; in this way, the typing semantics of the unmarked language guide the construction
+of the marking process.
+
+\subsubsection{Variables}
+\label{sec:calculus-variables}
+
+% TODO subsumption is discussed later...
+The typing of unmarked variables in the gradually typed lambda calculus is standard. Because of
+subsumption, we need only define a synthesis rule:
+\[%
+  \judgment{
+    \inCtx{\ctx}{x}{\TMV}
+  }{
+    \ctxSynType{\ctx}{x}{\TMV}
+  }{USVar}
+\]%
+If $x$ is bound to a type in the typing context, it synthesizes that type. Straightforwardly,
+marking would convert the unmarked $x$ into the same marked $x$ via the following marking and typing
+rules:
 %
+\begin{mathpar}
+  \judgment{
+    \inCtx{\ctx}{x}{\TMV}
+  }{
+    \ctxSynFixedInto{\ctx}{x}{x}{\TMV}
+  }{ISVar}
+
+  \judgment{
+    \inCtx{\ctx}{x}{\TMV}
+  }{
+    \ctxSynType{\ctx}{x}{\TMV}
+  }{MSVar}
+\end{mathpar}
+% TODO more intuition on why this makes sense?
+
+But what happens a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
+bound? Then, \textsc{USVar} does not apply, and the typing semantics say nothing about such
+expressions. To satisfy totality, such a possibility motivates an \emph{unbound variable mark}:
+%
+\[\begin{array}{rrcl}
+  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECUnbound{x}
+\end{array}\]
+%
+accompanied by the following rules:
+%
+\begin{mathpar}
+  \judgment{
+    \notInCtx{\ctx}{x}
+  }{
+    \ctxSynFixedInto{\ctx}{x}{\ECUnbound{x}}{\TUnknown}
+  }{ISUnbound}
+
+  \judgment{
+    \notInCtx{\ctx}{x}
+  }{
+    \ctxSynType{\ctx}{\ECUnbound{x}}{\TUnknown}
+  }{MSUnbound}
+\end{mathpar}
+%
+\textsc{ISUnbound} states that an variable unbound in the context is marked into that variable
+inside an unbound variable mark. \textsc{MSUnbound} then gives the typing of unbound variable marks.
+Since nothing may be said about their types, we synthesize the unknown type, which allows type
+checking to continue---instead of simply failing.
+
+To prove \cref{thm:calculus-marking-well-formedness}, we extend the definition of mark erasure to
+handle the new mark:
+%
+\[\begin{array}{rcl}
+  & \cdots & \\
+  \erasesToRow{\ECUnbound{x}}{x}
+\end{array}\]
+
+\subsubsection{Lambda Abstractions}
+\label{sec:calculus-lambda-abstractions}
+
+There are explicit synthesis and analysis rules for lambda abstractions:
+%
+\begin{mathpar}
+  \judgment{
+    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
+  }{
+    \ctxSynType{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\TArrow{\TMV_1}{\TMV_2}}
+  }{USLam}
+
+  \judgment{
+    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
+    \consistent{\TMV}{\TMV_1} \\
+    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
+  }{
+    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
+  }{UALam}
+\end{mathpar}
+%
+The synthesis rule is standard. In the premises of \textsc{UALam}, we use the notion of
+\emph{type consistency} from gradual type theory, which defines a reflexive and symmetric (but not
+transitive) relation between types, writing $\consistent{\TMV_1}{\TMV_2}$ to mean that $\TMV_1$ is
+consistent with $\TMV_2$. This replaces the notion of \emph{type equality} to relate the unknown
+type to all other types. The analysis rule also employs the judgment
+$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which establishes that $\TMV$ is a \emph{matched arrow type},
+i.e. it may be considered an arrow type. This notion is purely a technical mechanism to avoid
+duplication of rules related to arrow types. \Cref{fig:calculus-consistency-matched-arrow} gives the
+rules for both judgments.
+
+% TODO mention explicitly that by inductive hypothesis, the marking of body cannot fail?
+% TODO more intuition on these two judgments
+% TODO more intuition on USLam and UALam
+
+\begin{figure}[htbp]
+  \raggedright
+  \judgbox{\ensuremath{\consistent{\TMV_1}{\TMV_2}}} $\TMV_1$ is consistent with $\TMV_2$
+  %
+  \begin{mathpar}
+    \judgment{ }{
+      \consistent{\TUnknown}{\TMV}
+    }{TCUnknown1}
+
+    \judgment{ }{
+      \consistent{\TMV}{\TUnknown}
+    }{TCUnknown2}
+
+    \judgment{ }{
+      \consistent{\TMV}{\TMV}
+    }{TCRefl}
+
+    \judgment{
+      \consistent{\TMV_1}{\TMV_1'} \\
+      \consistent{\TMV_2}{\TMV_2'} \\
+    }{
+      \consistent{\TArrow{\TMV_1}{\TMV_2}}{\TArrow{\TMV_1'}{\TMV_2'}}
+    }{TCArr} \\
+  \end{mathpar}
+
+  \judgbox{\ensuremath{\matchedArrow{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched arrow type $\TArrow{\TMV_1}{\TMV_2}$
+  %
+  \begin{mathpar}
+    \judgment{ }{
+      \matchedArrow{\TUnknown}{\TUnknown}{\TUnknown}
+    }{TMAHole}
+
+    \judgment{ }{
+      \matchedArrow{\TArrow{\TMV}{\TMV}}{\TMV}{\TMV}
+    }{TMAArr}
+  \end{mathpar}
+  %
+  \caption{Type consistency and matched arrow types.}
+  \label{fig:calculus-consistency-matched-arrow}
+\end{figure}
+
+The synthesis rule for marking intuitively follows \textsc{USLam} closely. We should recursively
+mark the body with an extended context binding $x$ to $\TMV$ and construct a new marked lambda
+abstraction:
+%
+\begin{mathpar}
+  \judgment{
+    \ctxSynFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
+  }{
+    \ctxSynFixedInto{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\ELam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
+  }{ISLam}
+
+  \judgment{
+    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
+  }{
+    \ctxSynType{\ctx}{\ECLam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
+  }{MSLam}
+\end{mathpar}
+%
+Similarly, we construct corresponding analytic rules:
+%
+\begin{mathpar}
+  \judgment{
+    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
+    \consistent{\TMV}{\TMV_1} \\\\
+    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
+  }{
+    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
+  }{IALam1}
+
+  \judgment{
+    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
+    \consistent{\TMV}{\TMV_1} \\\\
+    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
+  }{
+    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
+  }{MALam1}
+\end{mathpar}
+%
+However, recalling totality, we look again to the premises of \textsc{UALam} to see what type errors
+may arise. First, what if $\TMV_3$ is not a matched arrow type? This would be the case if it were
+$\TNum$, for example, and the lambda abstraction would be inconsistent with the expected type. Thus,
+to the syntax we add an \emph{inconsistent type mark}, which indicates that the expression does not
+synthesize a type consistent with the expected type:
+%
+\[\begin{array}{rrcl}
+  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECInconType{\ECMV}
+\end{array}\]
+%
+along with the following rules, where the lambda abstraction is placed inside the inconsistent type
+mark.
+%
+\begin{mathpar}
+  \judgment{
+    \notMatchedArrow{\TMV_3} \\
+    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TUnknown}
+  }{
+    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
+  }{IALam2}
+
+  \judgment{
+    \notMatchedArrow{\TMV_3} \\
+    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
+  }{
+    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
+  }{MALam2*}
+\end{mathpar}
+%
+Though no expected output type is known, we still need to check and mark the body; it is thus
+analysed against the unknown type in \textsc{IALam2}.
+
+A similar situation arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
+$\inconsistent{\TMV}{\TMV_1}$, i.e. the expected input type is inconsistent with the actual input
+type, giving the last pair of analytic rules for lambda abstractions:
+%
+\begin{mathpar}
+  \judgment{
+    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
+    \inconsistent{\TMV}{\TMV_1} \\\\
+    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
+  }{
+    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
+  }{IALam3}
+
+  \judgment{
+    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
+    \inconsistent{\TMV}{\TMV_1} \\\\
+    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
+  }{
+    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
+  }{MALam3}
+\end{mathpar}
+
+Finally, as with the unbound variable mark, we extend mark erasure:
+%
+\[\begin{array}{rcl}
+  & \cdots & \\
+  \erasesToRow{\ECInconType{\ECMV}}{\erase{\ECMV}}
+\end{array}\]
+
+\subsubsection{Applications}
+\label{sec:calculus-applications}
+
+The case of application is somewhat interesting. In the unmarked language, only a synthesis rule is
+necessary:
+\[%
+  \judgment{
+    \ctxSynType{\ctx}{\EMV_1}{\TMV} \\
+    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\
+    \ctxAnaType{\ctx}{\EMV_2}{\TMV_1}
+  }{
+    \ctxSynType{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\TMV_2}
+  }{USAp}
+\]%
+Following the same methodology, we can straightforwardly give the following marking and typing
+rules:
+%
+\begin{mathpar}
+  \judgment{
+    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
+    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\\\
+    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
+  }{
+    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
+  }{ISAp1}
+
+  \judgment{
+    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
+    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\\\
+    \ctxAnaType{\ctx}{\ECMV_2}{\TMV_1}
+  }{
+    \ctxSynType{\ctx}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
+  }{MSAp1}
+\end{mathpar}
+
+Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type. There
+is no expected type for the argument, so we perform analytic marking on $\EMV_2$ against the
+unknown type. The applied expression is placed inside the inconsistent type mark, which is in
+accordance with the meaning of that mark.
+%
+\begin{mathpar}
+  \judgment{
+    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
+    \notMatchedArrow{\TMV} \\\\
+    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
+  }{
+    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApNonMatched{\ECMV_1}{\ECMV_2}}{\TUnknown}
+  }{ISAp2}
+
+  \judgment{
+    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
+    \notMatchedArrow{\TMV} \\\\
+    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
+  }{
+    \ctxSynType{\ctx}{\ECApNonMatched{\ECMV_1}{\ECMV_2}}{\TUnknown}
+  }{MSAp2}
+\end{mathpar}
+
+It is worth noting, however, that $\EMV_1$ is not actually in analytic position, but in a synthetic
+position where any arrow type is expected. Treating such ``constrained synthetic'' cases specially,
+we may instead designate a specialized mark, written $\ECApNonMatchedAltAlt{\ECMV}$. This yields the
+following alternative rules:
+%
+\[\begin{array}{rrcl}
+  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECApNonMatchedAltAlt{\ECMV}
+\end{array}\]
+%
+\begin{mathpar}
+  \judgment{
+    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
+    \notMatchedArrow{\TMV} \\\\
+    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
+  }{
+    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECApNonMatchedAltAlt{\ECMV_1}}{\ECMV_2}}{\TUnknown}
+  }{ISAp2'}
+
+  \judgment{
+    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
+    \notMatchedArrow{\TMV} \\\\
+    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
+  }{
+    \ctxSynType{\ctx}{\ECApNonMatchedAltAlt{\ECMV_1}{\ECMV_2}}{\TUnknown}
+  }{MSAp2'}
+\end{mathpar}
+
+It is natural to extend this approach to other elimination forms that require the handling of
+unmatched types, such as binary products. Indeed, in \cref{sec:calculus-products}, we opt for this
+approach. Another approach is to mark an error around the entire application expression.
+
+\subsubsection{Numbers}
+\label{sec:calculus-numbers}
+
+The marking and typing of numbers is straightforward:
+%
+\begin{mathpar}
+  \judgment{ }{
+    \ctxSynType{\ctx}{\ENumMV}{\TNum}
+  }{USNum}
+
+  \judgment{
+  }{
+    \ctxSynFixedInto{\ctx}{\ENumMV}{\ECNumMV}{\TNum}
+  }{ISNum}
+
+  \judgment{ }{
+    \ctxSynType{\ctx}{\ECNumMV}{\TNum}
+  }{MSNum}
+\end{mathpar}
+
+For addition expressions, we simply recursively mark each operand, which must analyse against
+$\TNum$, and build the marked addition expression. In this way, blame for type errors is shift into
+the operands:
+%
+\begin{mathpar}
+  \judgment{
+    \ctxAnaType{\ctx}{\EMV_1}{\TNum} \\\\
+    \ctxAnaType{\ctx}{\EMV_2}{\TNum}
+  }{
+    \ctxSynType{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
+  }{USPlus}
+
+  \judgment{
+    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\\\
+    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
+  }{
+    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
+  }{ISPlus}
+
+  \judgment{
+    \ctxAnaType{\ctx}{\ECMV_1}{\TNum} \\\\
+    \ctxAnaType{\ctx}{\ECMV_2}{\TNum}
+  }{
+    \ctxSynType{\ctx}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
+  }{MSPlus}
+\end{mathpar}
+%
+Again, we need only synthetic rules; subsumption handles the analytic case.
+
+\subsubsection{Booleans}
+\label{sec:calculus-booleans}
+
+Boolean values are similarly straightforward:
+%
+\begin{mathpar}
+  \judgment{ }{
+    \ctxSynType{\ctx}{\ETrue}{\TBool}
+  }{USTrue}
+
+  \judgment{
+  }{
+    \ctxSynFixedInto{\ctx}{\ETrue}{\ECTrue}{\TBool}
+  }{ISTrue}
+
+  \judgment{ }{
+    \ctxSynType{\ctx}{\ECTrue}{\TBool}
+  }{MSTrue} \\
+
+  \judgment{ }{
+    \ctxSynType{\ctx}{\EFalse}{\TBool}
+  }{USFalse}
+
+  \judgment{
+  }{
+    \ctxSynFixedInto{\ctx}{\EFalse}{\ECFalse}{\TBool}
+  }{ISFalse}
+
+  \judgment{ }{
+    \ctxSynType{\ctx}{\ECFalse}{\TBool}
+  }{MSFalse}
+\end{mathpar}
+
+If-else expressions, however, offer another interesting case. In the unmarked language, we have
+both explicit synthetic and analytic rules:
+%
+\begin{mathpar}
+  \judgment{
+    \ctxAnaType{\ctx}{\EMV_1}{\TBool} \\\\
+    \ctxSynType{\ctx}{\EMV_2}{\TMV_1} \\
+    \ctxSynType{\ctx}{\EMV_3}{\TMV_2}
+  }{
+    \ctxSynType{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
+  }{USIf}
+
+  \judgment{
+    \ctxAnaType{\ctx}{\EMV_1}{\TBool} \\\\
+    \ctxAnaType{\ctx}{\EMV_1}{\TMV} \\
+    \ctxAnaType{\ctx}{\EMV_2}{\TMV}
+  }{
+    \ctxAnaType{\ctx}{\ECIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TMV}
+  }{UAIf}
+\end{mathpar}
+%
+In synthetic position, conditionals synthesize the \emph{join} of the branch types $\TMV_1$ and
+$\TMV_2$, which we define inductively in \cref{fig:calculus-type-join}. In analytic position, since
+there is an expected type for both branches, we shift the blame for any errors to them.
+
+% TODO intuition why we use join
+
+\begin{figure}[htbp]
+  \raggedright
+  \judgbox{\ensuremath{\TJoin{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
+  %
+  \newcommand{\joinsTo}[3]{\ensuremath{\TJoin{#1}{#2} & = & #3}}
+  \[\begin{array}{rcl}
+    \joinsTo{\TUnknown}{\TMV}{\TMV} \\
+    \joinsTo{\TMV}{\TUnknown}{\TMV} \\
+    \joinsTo{\TNum}{\TNum}{\TNum} \\
+    \joinsTo{\TBool}{\TBool}{\TBool} \\
+    \joinsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}} \\
+  \end{array}\]
+  %
+  \caption{Type join.}
+  \label{fig:calculus-type-join}
+\end{figure}
+% TODO intuition for type join?
+
+Following \textsc{USIf} and \textsc{UAIf}, we may derive the following marking and typing rules.
+Similar to previous cases, we simply recursively mark the guard and branch expressions and construct
+the marked if-else expression from the results.
+%
+\begin{mathpar}
+  \judgment{
+    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
+    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
+    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2}
+  }{
+    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
+  }{ISIf}
+
+  \judgment{
+    \ctxAnaType{\ctx}{\ECMV_1}{\TBool} \\\\
+    \ctxSynType{\ctx}{\ECMV_2}{\TMV_1} \\
+    \ctxSynType{\ctx}{\ECMV_3}{\TMV_2}
+  }{
+    \ctxSynType{\ctx}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
+  }{MSIf} \\
+
+  \judgment{
+    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
+    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV} \\
+    \ctxAnaFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV} \\
+  }{
+    \ctxAnaFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
+  }{IAIf}
+
+  \judgment{
+    \ctxAnaType{\ctx}{\ECMV_1}{\TBool} \\\\
+    \ctxAnaType{\ctx}{\ECMV_1}{\TMV} \\
+    \ctxAnaType{\ctx}{\ECMV_2}{\TMV}
+  }{
+    \ctxAnaType{\ctx}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
+  }{MAIf}
+\end{mathpar}
+
+However, in synthetic mode, it may be the case that the two branch types have no join. This occurs,
+in fact, when they are inconsistent. This motivates \emph{the inconsistent branches mark}:
+%
+\[\begin{array}{rrcl}
+  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
+\end{array}\]
+
+Then, adding the following rules ensures marking totality on conditionals:
+%
+\begin{mathpar}
+  \judgment{
+    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
+    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\\\
+    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2} \\
+    \inconsistent{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TUnknown}
+  }{ISInconsistentBranches}
+
+  \judgment{
+    \ctxAnaType{\ctx}{\ECMV_1}{\TBool} \\
+    \ctxSynType{\ctx}{\ECMV_2}{\TMV_1} \\\\
+    \ctxSynType{\ctx}{\ECMV_3}{\TMV_2} \\
+    \inconsistent{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynType{\ctx}{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TUnknown}
+  }{MSInconsistentBranches}
+\end{mathpar}
+
+The extension of mark erasure is again straightforward:
+%
+\[\begin{array}{rcl}
+  & \cdots & \\
+  \erasesToRow{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\EIf{(\erase{\ECMV_1})}{(\erase{\ECMV_2})}{(\erase{\ECMV_3})}}
+\end{array}\]
+
+\subsubsection{Subsumption}
+\label{sec:calculus-subsumption}
+
+Finally, we consider the subsumption rule in typing unmarked expressions, which states that if an
+expression synthesizes a type, it may also be analysed against that type or any consistent type.
+\[%
+  \judgment{
+    \ctxSynType{\ctx}{\EMV}{\TMV'} \\
+    \consistent{\TMV}{\TMV'} \\
+    \subsumable{\EMV}
+  }{
+    \ctxAnaType{\ctx}{\EMV}{\TMV}
+  }{UASubsume}
+\]%
+We also restrict the usage of subsumption to ``subsumable'' syntactic forms, written
+$\subsumable{\EMV}$. This judgment is defined on all syntactic forms except lambda abstractions and
+conditionals, the only ones with explicit analysis rules. In other words, we restrict
+subsumption to be the analytic rule of ``last resort''. This is critical in establishing that
+marking and typing are deterministic (discussed later in\cref{thm:calculus-marking-unicity}).
+
+Since we did not provide explicit analytic rules for marking expressions that are not lambda
+abstractions or conditionals (or typing their marked versions) either, we give corresponding
+subsumption rules:
+%
+\begin{mathpar}
+  \judgment{
+    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
+    \consistent{\TMV}{\TMV'} \\
+    \subsumable{\EMV}
+  }{
+    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}
+  }{IASubsume}
+
+  \judgment{
+    \ctxSynType{\ctx}{\ECMV}{\TMV'} \\
+    \consistent{\TMV}{\TMV'} \\
+    \subsumable{\ECMV}
+  }{
+    \ctxAnaType{\ctx}{\ECMV}{\TMV}
+  }{MASubsume}
+\end{mathpar}
+%
+Note that we define a analogous notion of ``subsumability'' for marked expressions, written
+$\subsumable{\ECMV}$.
+
+However, it may be the case that the expression is inconsistent with the expected type. Recalling
+the example of \cref{fig:calculus-examples-inconsistent-types}, subsumption would used to analyse
+$\ETrue$ against $\TNum$ in $\EPlus{\ETrue}{5}$, which fails since $\inconsistent{\TNum}{\TBool}$,
+and $\ETrue$ should be marked with an inconsistent type mark. Hence, the following final set of
+rules is required to handle such cases:
+%
+\begin{mathpar}
+  \judgment{
+    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
+    \inconsistent{\TMV}{\TMV'} \\\\
+    \subsumable{\EMV}
+  }{
+    \ctxAnaFixedInto{\ctx}{\EMV}{\ECInconType{\ECMV}}{\TMV}
+  }{IAInconsistentTypes}
+
+  \judgment{
+    \ctxSynType{\ctx}{\ECMV}{\TMV'} \\
+    \inconsistent{\TMV}{\TMV'} \\\\
+    \subsumable{\ECMV}
+  }{
+    \ctxAnaType{\ctx}{\ECInconType{\ECMV}}{\TMV}
+  }{MAInconsistentTypes}
+\end{mathpar}
+%
+These rules complete the development of the marking calculus for the gradually typed lambda
+calculus.
+
+% TODO mini conclusion on methodology
+
+To conclude this section, we present two more metatheorems that help ensure correctness of the
+system. Though \cref{thm:calculus-marking-well-formedness} guarantees that marking does not change
+the syntactic structure of a program, it makes no statement about the presence of error marks in the
+resulting marked program. Following the logic of constructing marking rules, we have
+\cref{thm:calculus-marking-well-ill-typed}.
+
 \begin{theorem}[name=Marking of Well-Typed/Ill-Typed Expressions] \ % TODO better name for this
@@
 %
-That is, since error marks indicate type errors in the program, a well-typed program should, of
-course, have no error marks. The converse---that marks are inserted into ill-typed programs---is
-also true. We have chosen to formulate the absence of error marks through an inclusion mapping of
-unmarked expressions into marked expressions; alternatively, a predicate over marked expressions
-could be used. % TODO latter is easier to do in Agda, but maybe it is weaker?
+That is, since error marks indicate type errors in the program, no marks should be inserted
+into a well-typed program. The converse---that at least one mark is inserted into ill-typed
+programs---is also true. We have chosen to formulate the absence of error marks through an inclusion
+mapping of unmarked expressions into marked expressions; alternatively, a predicate over marked
+expressions may be used. % TODO latter is easier to do in Agda, but maybe it is weaker?
 
@@
       \TMV_2$.
-    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV_1}$ and
-      $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV_2}$, then $\ECMV_1 = \ECMV_2$ and $\TMV_1 =
-      \TMV_2$.
+    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV}$ and
+      $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV}$, then $\ECMV_1 = \ECMV_2$.
   \end{enumerate}
@@
 Together with totality, marking may be considered and implemented as a total function.
+
+% TODO non-gradual semantics?
+
+% Note that while the semantics that follow concern a gradually typed language, a non-gradual system
+% may be recovered straightforwardly by removing the unknown type and replacing type consistency with
+% equality.
 
@@
 
+\subsection{Polymorphism and Other Typing Features}
+\label{sec:calculus-poly}
+
 \subsection{Implementation in Hazel}

### 2023-02-27 figs/calculus-examples.tex
Rewrite marked lambda caclulus section as discussed

\begin{figure}[htbp]
  \begin{tabular}[b]{cc}
    \begin{subfigure}[b]{0.3\columnwidth}
      \includegraphics[width=\columnwidth]{images/haz3l-unbound-variable.png}
      \caption{Unbound variable error.}
      \label{fig:calculus-examples-unbound}
    \end{subfigure}
    &
    \begin{subfigure}[b]{0.3\columnwidth}
      \includegraphics[width=\columnwidth]{images/haz3l-unbound-variable.png}
      \caption{Inconsistent types error.}
      \label{fig:calculus-examples-inconsistent-types}
    \end{subfigure} \\
    \begin{subfigure}[b]{0.3\columnwidth}
      \includegraphics[width=\columnwidth]{images/haz3l-unbound-variable.png}
      \caption{Application of a non-lambda.}
      \label{fig:calculus-examples-app-non-lambda}
    \end{subfigure}
    &
    \begin{subfigure}[b]{0.3\columnwidth}
      \includegraphics[width=\columnwidth]{images/haz3l-unbound-variable.png}
      \caption{Inconsistent branches error.}
      \label{fig:calculus-examples-inconsistent-branches}
    \end{subfigure}
  \end{tabular}
  %
  \caption{Examples of common type errors.}
  \label{fig:calculus-examples}
\end{figure}

### 2023-02-27 calculus.tex
Draft mechanization section

@@
 
-\section{Marked Lambda Calculus}
+\section{The Marking Calculus}
 \label{sec:calculus}
@@
 
-Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type. There
-is no expected type for the argument, so we perform analytic marking on $\EMV_2$ against the
-unknown type. The applied expression is placed inside the inconsistent type mark, which is in
-accordance with the meaning of that mark.
+Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type, such
+as in the example of \cref{fig:calculus-examples-app-non-lambda}. There is no expected type for the
+argument, so we perform analytic marking on $\EMV_2$ against the unknown type. The applied
+expression is placed inside the inconsistent type mark, which is in accordance with the meaning of
+that mark.
 %
@@
 
+\newcommand{\joinsTo}[3]{\ensuremath{\TJoin{#1}{#2} & = & #3}}
 \begin{figure}[htbp]
@@
   %
-  \newcommand{\joinsTo}[3]{\ensuremath{\TJoin{#1}{#2} & = & #3}}
   \[\begin{array}{rcl}
@@
 subsumption to be the analytic rule of ``last resort''. This is critical in establishing that
-marking and typing are deterministic (discussed later in\cref{thm:calculus-marking-unicity}).
+marking and typing are deterministic (discussed later in \cref{thm:calculus-marking-unicity}).
 
@@
     \item \begin{enumerate}
-        \item If $\ctxSynType{\ctx}{\EMV}{\TMV}$, then $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
-          such that $\imageIs{\EMV}{\ECMV}$.
-        \item If $\ctxAnaType{\ctx}{\EMV}{\TMV}$, then $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
-          such that $\imageIs{\EMV}{\ECMV}$.
+        \item If $\ctxSynType{\ctx}{\EMV}{\TMV}$ and $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
+          then $\imageIs{\EMV}{\ECMV}$.
+        \item If $\ctxAnaType{\ctx}{\EMV}{\TMV}$ and $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
+          then $\imageIs{\EMV}{\ECMV}$.
       \end{enumerate}
@@
 We now demonstrate how one might extend or adapt the marking calculus for new typing features by
-adding binary product types. First, we extend in \cref{fig:calculus-products-judgments} the syntax
-of types and notions of consistency and join. As with arrow types, we additionally define an
-auxiliary notion of \emph{matched binary product types}: the judgment
-$\matchedPair{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be considered the binary product
-type $\TPair{\TMV_1}{\TMV_2}$.
+adding binary product types. First, we extend the syntax of types and the accompanying notions of
+consistency and join, which closely mirror the definitions for arrow types.
+%
+\[\begin{array}{rrcl}
+  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TPair{\TMV}{\TMV}
+\end{array}\]
 
-\input{figs/calculus-products-judgments}
+We additionally define an auxiliary notion of \emph{matched binary product types} in
+\cref{fig:calculus-matched-product}: the judgment $\matchedPair{\TMV}{\TMV_1}{\TMV_2}$ establishes
+that $\TMV$ may be considered the binary product type $\TPair{\TMV_1}{\TMV_2}$.
 
-The unmarked language with a pair constructor and projection operators.
-\Cref{fig:calculus-products-uexp} gives their syntax and typing rules. Similarly to application,
-\textsc{USProjL} and \textsc{USProjR} employ the matched binary product type; a projection
-$\EProjL{\EMV}$ or $\EProjR{\EMV}$ is then well-typed if the $\EMV$ may be considered to be a pair.
+\begin{figure}[htbp]
+  \raggedright
+  \judgbox{\ensuremath{\matchedPair{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched binary product type $\TPair{\TMV_1}{\TMV_2}$
+  %
+  \begin{mathpar}
+    \judgment{ }{
+      \matchedPair{\TUnknown}{\TUnknown}{\TUnknown}
+    }{TMPHole}
 
-\input{figs/calculus-products-uexp}
+    \judgment{ }{
+      \matchedPair{\TPair{\TMV_1}{\TMV_2}}{\TMV_1}{\TMV_2}
+    }{TMPProd}
+  \end{mathpar}
+  %
+  \caption{Matched product type.}
+  \label{fig:calculus-matched-product}
+\end{figure}
+
+The unmarked language is extended with a pair constructor and projection operators.
+%
+\[\begin{array}{rrcl}
+  \EMName  & \EMV  & \Coloneqq & \cdots
+                               \mid \EPair{\EMV}{\EMV}
+                               \mid \EProjL{\EMV} \mid \EProjR{\EMV}
+\end{array}\]
+%
+\begin{mathpar}
+  \judgment{
+    \ctxSynType{\ctx}{\EMV_1}{\TMV_1} \\
+    \ctxSynType{\ctx}{\EMV_2}{\TMV_2}
+  }{
+    \ctxSynType{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TPair{\TMV_1}{\TMV_2}}
+  }{USPair}
+
+  \judgment{
+    \ctxSynType{\ctx}{\EMV}{\TMV} \\
+    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynType{\ctx}{\EProjL{\EMV}}{\TMV_1}
+  }{USProjL}
+
+  \judgment{
+    \ctxSynType{\ctx}{\EMV}{\TMV} \\
+    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynType{\ctx}{\EProjR{\EMV}}{\TMV_2}
+  }{USProjR} \\
+\end{mathpar}
+%
+Similarly to application, \textsc{USProjL} and \textsc{USProjR} employ the matched binary product
+type; a projection $\EProjL{\EMV}$ or $\EProjR{\EMV}$ is then well-typed if the $\EMV$ may be
+considered to be a pair.
 
@@
 
+\subsection{Polymorphism and Other Typing Features}
+\label{sec:calculus-poly}
+
+We now consider how the calculus might be extended to richer typing features such as parametric
+polymorphism. Specifically, we consider System F-style explicit polymorphism.
+
 \subsection{Mechanization}
@@
 
-The metatheory of the calculus has been fully mechanized in the Agda proof assistant [citation]. The
-mechanization's documentation contains more details regarding technical decisions made therein.
+The semantics and metatheory presented above, including the extension with binary products, have
+been fully mechanized in the Agda proof assistant [citation]. Though the mechanization's
+documentation contains more detailed discussion regarding technical decisions made therein, we
+highlight some important aspects here.
 
-\subsection{Polymorphism and Other Typing Features}
-\label{sec:calculus-poly}
+The standard approach of modeling judgments as inductive datatypes and rules as constructors for
+those datatypes is taken. For marked expressions, we opt for implicitly typed terms [citation],
+which gives us part of \cref{thm:calculus-marking-well-formedness} for free. For the convenience of
+readers interested in browsing the mechanization, all rule names match those presented here.
+
+\section{Marking Calculus in Structured Editing}
+\label{sec:calculus-structured-editing}
+
+The previous section presented a formal calculus for error localization and recovery. We now discuss
+its usage in a structured editing setting.
+
+In particular, the calculus has been integrated into the Hazel programming environment, a live
+programming environment with support for typed holes.
 

### 2023-02-27 abstract.tex
Updates from Overleaf

@@
 Type systems typically only define the conditions under which an expression is well-typed, leaving ill-typed expressions entirely meaningless. 
-This approach is insufficient as the basis for language servers for modern programming environments, which are expected to sensibly localize type errors and recover, i.e.
-continue to provide a variety of semantic services in their presence.
-This paper addresses this problem, contributing the first comprehensive formal account of type error localization and recovery, called the marked lambda calculus.
-This calculus, which is intended as a core calculus for language servers, consists of a gradual type system for expressions with marked errors and a total marking procedure on unmarked expressions with holes.
-We mechanize the metatheory of the marked lambda calculus in Agda, deploy its total marking procedure to fix several instances of undefined behavior in the Hazelnut structure editor calculus, and implement the system as the new basis for Hazel, a full-scale live functional programming environment.
+This approach is insufficient as the basis for language servers for modern programming environments, which are expected to sensibly localize any type errors and recover, i.e.
+continue to provide a variety of downstream semantic services.
+This paper addresses this problem, contributing the first comprehensive formal account of type error localization and recovery: the marked lambda calculus.
+In particular, we define a gradual type system for expressions with marked errors and holes, together with a total procedure for marking unmarked expressions.
+We mechanize the metatheory of the marked lambda calculus in Agda and implement it, scaled up, as the new basis for Hazel, a full-scale live functional programming environment.
+As a secondary contribution, we show how total marking resolves 
+the problem of undefined behavior in the Hazelnut structure editor calculus
+(though marking itself does not require a structure editor).
 
@@
 it is bidirectionally typed.
-Constraint-based type inference schemes (e.g.
-Hindley-Milner) are notorious for complicating error localization.
-We approach the problem of incorporating constraint-based type inference by deploying constraint solving as a type hole filling layer atop a gradual bidirectionally typed core.
-When the system of constraints cannot be fully solved, the unification algorithm we develop localizes errors arising from inconsistent constraints to individual type holes, rather than attempting to localize them in an \emph{ad hoc} manner to one particular expression and then failing to recover from there.
-The user can resolve these errors by selecting from suggested partial solutions, or leave the errors unresolved, in which case the type hole recovers gradually.
-We formally specify this type hole inference system and implement it in Hazel.
+Constraint-based type inference notoriously complicates error localization and recovery.
+We approach this problem by deploying constraint solving as a type-hole-filling layer atop our gradual bidirectionally typed core.
+The unification algorithm we contribute localizes errors arising from inconsistent constraints to type holes with traced  provenances, rather than attempting to localize them in an \emph{ad hoc} manner to particular expressions.
+The user can choose how to shift these errors to expressions by selecting from suggested partial solutions, or allow the unresolved type holes to recover gradually.
+We implement this type hole inference system in Hazel.
 \end{abstract}

### 2023-02-27 intro.tex

@@
 
-Type errors are both a boon and a bane to developers.
-At their best, type errors help programmers quickly and accurately identify the location and cause of a genuine problem in their program that is blocking progress on their task. 
-At their worst, type errors mislead the programmer / distract them away from the parts of the program that are of interest, and 
-prevent evaluation and editor services from working, even when the error is benign or located at an irrelevant location in the program. 
-(May be one reason some developers favor dynamic languages, especially when prototyping)
+% When a static type checker locates a type error, the hope is that the type error accurately locates a problem that is blocking progress on the task at hand. 
+% However, the reality is that the type error might incorrectly locate the problem, or locate a problem irrelevant to the task at hand.
+In all cases, the developer must fix all type errors before .
+These gaps in service can  persist for hours or days at a time, such as when a key type definition is
+% modified in a way that causes multiple errors to appear throughout a large program. 
+% Until \emph{all} of these errors are fixed, it may be the case that \emph{none} of the changed code paths can be tested at run-time, 
+% and many helpful language services are not fully functional \cite{HazelnutSNAPL}.
+
+Contemporary programming environments provide 
+a wide variety of semantic services that require statically analyzing the type and binding structure of the program being edited, e.g.
+various semantic navigation services, type hints, semantic code completion, and automated refactoring.
+
+Language designers typically specify static type systems as a collection of rules 
+stating conditions under which an expression can be judged to be well-typed,
+i.e. completely free of type errors. 
+Provisioning these semantic services for well-typed expressions is well-supported 
+by these definitions.
+
+When there is a type error, however, these definitions are inadequate.
+It is therefore left as an implementation decision (1) type error localization: how to localize and explain the type error, and (2) type error recovery: how to recover from the error to find other simultaneous errors and to continue to provision various semantic services, which would often still useful even in the presence of errors.
+
+Consider, for example,
+
+let x = 
+  if f(y) then 1 else true
+in 
+  x + 1
+
+(compare Elm/OCaml/Haskell and Hazel directly?)
+
+Different language servers or typecheckers for the same language can and do attempt to localize, report, and recover from these errors in different ways, with little in the way of unifying theory.
 
@@
 Indeed, essentially all widely-used programming environments (often via an underlying language server) have some support for type error 
-localization and many also attempt recovery in some situations. However, the design of these systems has largely remained a matter of folklore or empirical iteration. 
+localization and many also attempt recovery of certain services in certain situations.
+However, the design of these systems has evolved in some part as folklore, as type system definitions have remained silent.
+
 One reason for this is 
@@
 there is no direct specification for how to reason within an erroneous program.
-Different language servers or typecheckers for the same language can and do attempt to localize, report, and recover from these errors in different ways, with little in the way of unifying theory.
 

### 2023-02-27 intro.tex
Type hole inference future work updated. Pre update to constraint generation rules and narrative

@@
 Contemporary programming environments provide 
-a wide variety of semantic services that require statically analyzing the type and binding structure of the program being edited, e.g.
-various semantic navigation services, type hints, semantic code completion, and automated refactoring.
+a variety of semantic services that require statically analyzing the type and binding structure of the program being edited, e.g.
+semantic navigation, type hints, semantic code completion, and automated refactoring.
+However, during the editing process,
+it is common for programs to be ill-typed.
 
@@
 i.e. completely free of type errors. 
-Provisioning these semantic services for well-typed expressions is well-supported 
+ is well-supported 
 by these definitions.

### 2023-02-28 calculus.tex
Move marking of numbers to front of formal development

@@
 The key contribution of this section is a formal calculus, called the \emph{marking calculus}, for
-type error localization in bidirectional systems, a judgmental framework provides a unified theory
-for the design of typing semantics that supports ill-typed programs. \Cref{sec:calculus-calculus}
-describes its form and metatheory by example on a minimal gradually typed lambda calculus, extending
-with binary product types in \cref{sec:calculus-products}. All rules and theorems may be found in
-the supplementary appendix, alongside a complete mechanization in the Agda proof assistant, which is
-discussed in \cref{sec:calculus-agda}. Finally, in \cref{sec:calculus-poly}, we consider how the
-system might be extended to more complex typing features, such as System F-style polymorphism.
+type error localization in bidirectional systems, a judgmental framework that provides a unified
+theory for the design of typing semantics that supports ill-typed programs.
+\Cref{sec:calculus-calculus} describes its form and metatheory by example on a minimal gradually
+typed lambda calculus. All rules and theorems may be found in the supplementary appendix, alongside
+a complete mechanization in the Agda proof assistant, which is discussed in
+\cref{sec:calculus-agda}. Finally, in \cref{sec:calculus-poly}, we consider how the system might be
+extended to more complex typing features, such as System F-style polymorphism.
 
@@
 We now formally introduce the \emph{marking calculus} through its application to a gradually typed
-lambda calculus extended with numbers and booleans. Consider the syntax for such a language, as
-shown in \cref{fig:calculus-syntax-uexp}. The base types $\TNum$ and $\TBool$ classify number and
+lambda calculus extended with numbers and booleans. Consider the syntax for such a language, given
+as $\EMName$ in \cref{fig:calculus-syntax}. The base types $\TNum$ and $\TBool$ classify number and
 boolean expressions. The number literal corresponding to the mathematical number $n$ is given by
-$\ENumMV$, and there is a single addition operation on numeric expression. $\ETrue$ and $\EFalse$
+$\ENumMV$, and there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$
 correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
-$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives an if-else structure. In addition to these forms, $\TUnknown$
-is the unknown type from gradual type theory.
+$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure. In addition to these forms,
+$\TUnknown$ is the unknown type from gradual type theory.
 
+% TODO include holes
 \begin{figure}[htbp]
   \[\begin{array}{rrcl}
-    \textsf{Type} & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
-    \textsf{Exp}  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \\
-         &       & \mid         & \ENumMV \mid \EPlus{\EMV}{\EMV}
-                   \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
+    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
+    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV}
+                       \mid           \ENumMV \mid \EPlus{\EMV}{\EMV}
+                       \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV} \\
+    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV}
+                       \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
+                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV} \\
+             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
   \end{array}\]
   \caption{Syntax of the extended gradually typed lambda calculus}
-  \label{fig:calculus-syntax-uexp}
+  \label{fig:calculus-syntax}
 \end{figure}
 
-The language is typed bidirectionally, which employs two mutually defined judgments. \emph{Type
-synthesis}, written $\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under the typing context
-$\ctx$, the expression $e$ synthesises or infers the type $\TMV$. \emph{Type analysis}, written
-$\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear where an expression of
-type $\TMV$ is expected. We may give these typing semantics traditionally, without considering error
-localization and recovery.
+Types classify expressions by a \emph{bidirectional type system}, which employs two mutually defined
+judgments. \emph{Type synthesis}, written $\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under
+the typing context $\ctx$, the expression $e$ synthesises or infers the type $\TMV$. \emph{Type
+analysis}, written $\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear
+where an expression of type $\TMV$ is expected. We may give these typing semantics traditionally,
+without considering error localization and recovery.
 
@@
   \item The \textbf{unmarked language}, which is the original language. Expressions of this language
-    are \emph{unmarked expressions}.
+    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.
 
@@
 
-  \item \textbf{Marking}, which \emph{marks} any unmarked expression into a marked expression,
-    inserting error markings where appropriate.
+  \item \textbf{Marking}, which \emph{marks}, i.e. type checks and localizes errors for, any unmarked
+    expression into a marked expression, inserting error marks where appropriate.
 \end{itemize}
@@
 language and ``mark'' unmarked programs into marked ones. This corresponds to a type checking
-process with error localization and recovery.
+process with error localization and recovery. The marked language may then serve as a foundation for
+other semantic services. % such as THI.
 
@@
 
-We give, for now, only part of the marked expression syntax in \cref{fig:calculus-syntax-mexp-part}.
-Throughout the remainder of this section, as we consider the formulation of marking, we will add the
-necessary error mark forms.
+\Cref{fig:calculus-syntax} furthermore gives the syntax for marked expressions (as $\ECMName$),
+which replicates all the forms of unmarked expressions. There are additionally three kinds of
+\emph{error marks}, which represent the different type errors that may arise.
+%
+\begin{itemize}
+  \item $\ECUnbound{x}$, the \emph{unbound variable mark}, indicates the use of an unbound variable
+    $x$, as in \cref{fig:calculus-examples-unbound}.
 
-% TODO include holes
-\begin{figure}[htbp]
-  \[\begin{array}{rrcl}
-    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV} \\
-             &       & \mid         & \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
-                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV} \\
-             &       & \mid         & \cdots
-             % &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
-  \end{array}\]
-  \caption{Partial syntax of marked expressions}
-  \label{fig:calculus-syntax-mexp-part}
-\end{figure}
+  \item $\ECInconType{\ECMV}$, the \emph{inconsistent type mark}, indicates that the type of an
+    expression does not match an expected type.
+
+  \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, indicates
+    that the branches of a conditional have inconsistent types.
+\end{itemize}
+%
+Throughout the remainder of this section, as we consider the formulation of marking, we will
+motivate and give precise semantics for these error marks. Note that this is a non-exhaustive list;
+depending on the syntactic forms and typing features found in the language, other kinds of marks may
+be necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of marks that may be
+required in an language with explicit polymorphism.
 
@@
 %
-That is, we may mark, i.e. type check and localize errors for \emph{any} syntactically well-formed
-program. Furthermore, marking should preserve the syntactic structure of programs, modulo error
-marks. Thus, we must define a notion of \emph{mark erasure}, which converts marked expressions back
-into unmarked expression by removing error marks. \Cref{fig:calculus-mark-erasure-part} gives the
-definition of mark erasure, written $\erase{\ECMV}$, for our initial marked expression syntax. Then,
-we may define a \emph{well-formedness} criterion for marking:
+That is, we may mark \emph{any} syntactically well-formed program.
+
+Furthermore, marking should preserve the syntactic structure of programs, modulo error marks. Thus,
+we define a notion of \emph{mark erasure}, whose definition is given
+\cref{fig:calculus-mark-erasure}, which converts marked expressions back into unmarked expression by
+removing error marks. Then, \cref{thm:calculus-marking-well-formedness} provides a
+\emph{well-formedness} criterion marking.
 %
@@
 from the marked expression gives the original unmarked expression. We also ensure that marking gives
-marked expressions that are well-typed; indeed, their types should match the types synthesised or
-analysed against in marking. % TODO more intuition for this theorem
+well-typed marked expressions; indeed, their types should match the types synthesised or analysed
+against in marking. % TODO more intuition for this theorem
 
-\input{figs/calculus-mark-erasure-part}
+\input{figs/calculus-mark-erasure}
 
@@
 
+\subsubsection{Numbers}
+\label{sec:calculus-numbers}
+
+% TODO need to say something about subsumption and how it will be handled later.
+To start, let us consider the simple case of numbers. Because of subsumption, we need only define a
+synthesis rule for unmarked numbers, in which numbers synthesize the type $\TNum$:
+%
+\begin{mathpar}
+  \judgment{ }{
+    \ctxSynType{\ctx}{\ENumMV}{\TNum}
+  }{USNum}
+\end{mathpar}
+
+Into what should numbers be marked? Quite straightforwardly, we simply give the same number as a
+marked expression, synthesizing again $\TNum$. No type errors may occur in a single number, so we
+only need the following rules for marking unmarked numbers and typing marked numbers:
+%
+\begin{mathpar}
+  \judgment{
+  }{
+    \ctxSynFixedInto{\ctx}{\ENumMV}{\ECNumMV}{\TNum}
+  }{ISNum}
+
+  \judgment{ }{
+    \ctxSynType{\ctx}{\ECNumMV}{\TNum}
+  }{MSNum}
+\end{mathpar}
+
+In addition expressions, the type of both operands should be $\TNum$. In a bidirectional system, we
+thus analyse the operands against $\TNum$, giving the following rule:
+%
+\begin{mathpar}
+  \judgment{
+    \ctxAnaType{\ctx}{\EMV_1}{\TNum} \\
+    \ctxAnaType{\ctx}{\EMV_2}{\TNum}
+  }{
+    \ctxSynType{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
+  }{USPlus}
+\end{mathpar}
+
+The marking rule mirrors \textsc{USPlus} closely. Since a known, expected type for each operand is
+known, we shift the ``blame'' for any type errors to them. Hence, we recursively mark each operand
+in analytic mode and rebuild the marked addition expression. The marked typing rule then mirrors
+\textsc{USPlus} exactly.
+%
+\begin{mathpar}
+  \judgment{
+    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\\\
+    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
+  }{
+    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
+  }{ISPlus}
+
+  \judgment{
+    \ctxAnaType{\ctx}{\ECMV_1}{\TNum} \\\\
+    \ctxAnaType{\ctx}{\ECMV_2}{\TNum}
+  }{
+    \ctxSynType{\ctx}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
+  }{MSPlus}
+\end{mathpar}
+
 \subsubsection{Variables}
@@
 
-% TODO subsumption is discussed later...
-The typing of unmarked variables in the gradually typed lambda calculus is standard. Because of
-subsumption, we need only define a synthesis rule:
+Let us now consider the more interesting case of variables. The typing of unmarked variables in the
+gradually typed lambda calculus is standard. Again, because of subsumption, there is only a
+synthetic rule:
 \[%
@@
 If $x$ is bound to a type in the typing context, it synthesizes that type. Straightforwardly,
-marking would convert the unmarked $x$ into the same marked $x$ via the following marking and typing
+marking converts the unmarked $x$ into the same marked $x$ via the following marking and typing
 rules:
@@
 But what happens a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
-bound? Then, \textsc{USVar} does not apply, and the typing semantics say nothing about such
-expressions. To satisfy totality, such a possibility motivates an \emph{unbound variable mark}:
-%
-\[\begin{array}{rrcl}
-  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECUnbound{x}
-\end{array}\]
-%
-accompanied by the following rules:
+bound, i.e. that the premise fails? Then, \textsc{USVar} does not apply, and the typing semantics
+say nothing about such expressions. When devising marking, to satisfy totality, such a possibility
+motivates an \emph{unbound variable mark} $\ECUnbound{x}$, accompanied by the following rules:
 %
@@
 Since nothing may be said about their types, we synthesize the unknown type, which allows type
-checking to continue---instead of simply failing.
-
-To prove \cref{thm:calculus-marking-well-formedness}, we extend the definition of mark erasure to
-handle the new mark:
-%
-\[\begin{array}{rcl}
-  & \cdots & \\
-  \erasesToRow{\ECUnbound{x}}{x}
-\end{array}\]
+checking to continue---instead of simply failing---and the unbound variable to be used in any
+expression. % TODO clarify why this is correct
 
@@
 
-There are explicit synthesis and analysis rules for lambda abstractions:
+There are explicit synthesis and analysis rules for unmarked lambda abstractions:
 %
@@
 $\TNum$, for example, and the lambda abstraction would be inconsistent with the expected type. Thus,
-to the syntax we add an \emph{inconsistent type mark}, which indicates that the expression does not
-synthesize a type consistent with the expected type:
-%
-\[\begin{array}{rrcl}
-  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECInconType{\ECMV}
-\end{array}\]
-%
-along with the following rules, where the lambda abstraction is placed inside the inconsistent type
-mark.
+to the syntax we add an \emph{inconsistent type mark} $\ECInconType{\ECMV}$, which indicates that
+the expression does not synthesize a type consistent with the expected type. We then have the
+following rules, in which the lambda abstraction is placed inside the inconsistent type mark.
 %
@@
     \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
-  }{MALam2*}
+  }{MALam2}
 \end{mathpar}
@@
 \end{mathpar}
-
-Finally, as with the unbound variable mark, we extend mark erasure:
-%
-\[\begin{array}{rcl}
-  & \cdots & \\
-  \erasesToRow{\ECInconType{\ECMV}}{\erase{\ECMV}}
-\end{array}\]
 
@@
 It is natural to extend this approach to other elimination forms that require the handling of
-unmatched types, such as binary products. Indeed, in \cref{sec:calculus-products}, we opt for this
-approach. Another approach is to mark an error around the entire application expression.
-
-\subsubsection{Numbers}
-\label{sec:calculus-numbers}
-
-The marking and typing of numbers is straightforward:
-%
-\begin{mathpar}
-  \judgment{ }{
-    \ctxSynType{\ctx}{\ENumMV}{\TNum}
-  }{USNum}
-
-  \judgment{
-  }{
-    \ctxSynFixedInto{\ctx}{\ENumMV}{\ECNumMV}{\TNum}
-  }{ISNum}
-
-  \judgment{ }{
-    \ctxSynType{\ctx}{\ECNumMV}{\TNum}
-  }{MSNum}
-\end{mathpar}
-
-For addition expressions, we simply recursively mark each operand, which must analyse against
-$\TNum$, and build the marked addition expression. In this way, blame for type errors is shift into
-the operands:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxAnaType{\ctx}{\EMV_1}{\TNum} \\\\
-    \ctxAnaType{\ctx}{\EMV_2}{\TNum}
-  }{
-    \ctxSynType{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
-  }{USPlus}
-
-  \judgment{
-    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\\\
-    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
-  }{
-    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
-  }{ISPlus}
-
-  \judgment{
-    \ctxAnaType{\ctx}{\ECMV_1}{\TNum} \\\\
-    \ctxAnaType{\ctx}{\ECMV_2}{\TNum}
-  }{
-    \ctxSynType{\ctx}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
-  }{MSPlus}
-\end{mathpar}
-%
-Again, we need only synthetic rules; subsumption handles the analytic case.
+unmatched types, such as products. Another approach is to mark an error around the entire
+application expression.
 
@@
 Similar to previous cases, we simply recursively mark the guard and branch expressions and construct
-the marked if-else expression from the results.
+the marked conditional from the results.
 %
@@
 However, in synthetic mode, it may be the case that the two branch types have no join. This occurs,
-in fact, when they are inconsistent. This motivates \emph{the inconsistent branches mark}:
-%
-\[\begin{array}{rrcl}
-  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
-\end{array}\]
-
-Then, adding the following rules ensures marking totality on conditionals:
+in fact, when they are inconsistent. This motivates \emph{the inconsistent branches mark}
+$\ECInconBr{\ECMV}{\ECMV}{\ECMV}$. Then, adding the following rules ensures marking totality on
+conditionals:
 %
@@
 \end{mathpar}
-
-The extension of mark erasure is again straightforward:
-%
-\[\begin{array}{rcl}
-  & \cdots & \\
-  \erasesToRow{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\EIf{(\erase{\ECMV_1})}{(\erase{\ECMV_2})}{(\erase{\ECMV_3})}}
-\end{array}\]
 
@@
 \subsection{Reified Hazelnut Action Semantics}
+\label{sec:calculus-hazelnut}

### 2023-02-28 intro.tex
Updates from Overleaf

@@
 
-Contemporary programming environments provide 
+Modern programming environments provide 
 a variety of semantic services that require statically analyzing the type and binding structure of the program being edited, e.g.
 semantic navigation, type hints, semantic code completion, and automated refactoring.
-However, during the editing process,
-it is common for programs to be ill-typed.
+These semantic services are often packaged into language servers that front-end
+editors and editor extensions can interface with\todo{cites}.
 
-Language designers typically specify static type systems as a collection of rules 
-stating conditions under which an expression can be judged to be well-typed,
-i.e. completely free of type errors. 
- is well-supported 
-by these definitions.
+The problem that animates this paper is that during program editing, 
+it is common for the program being edited to be ill-typed.
+These ill-typed states sometimes persist over long periods of time, e.g.
+a change to a type definition can cause errors to arise at all usage sites in a large program, which can take hours or days to address.
+Language servers often exhibit gaps in service during these periods, because in the conventional formal account of type systems, only well-typed programs have a well-defined type and binding structure.
+As soon as a type error occurs \emph{anywhere}, the program is formally meaningless \emph{everywhere}.
 
-When there is a type error, however, these definitions are inadequate.
-It is therefore left as an implementation decision (1) type error localization: how to localize and explain the type error, and (2) type error recovery: how to recover from the error to find other simultaneous errors and to continue to provision various semantic services, which would often still useful even in the presence of errors.
+Developers demand that these gaps in service be minimized or eliminated, so there has been considerable practical interest in (1)~\textbf{type error localization}: mechanisms for directing the developer to
+the location(s) in a program that may be erroneous, and (2) \textbf{type error recovery}: 
+mechanisms that allow the language server to continue past the first localized error and provision downstream semantic services even when there in the presence of localized type errors.
+Indeed, essentially all widely-used programming systems have some support for type error 
+localization, e.g. via compiler error messages or the language server.
+Many also attempt recovery of certain services in certain situations. 
+However, the design of these mechanisms has evolved in some part as folklore, because type system definitions have remained silent about how to reason in the presence of errors.
+Different type checkers for the same language can and do attempt to localize, report, and recover from these errors in different ways, with little in the way of unifying theory.
 
-Consider, for example,
-
+Consider, for example, the following program in a typed functional language with only local type inference, for simplicity, 
+\begin{lstlisting}[numbers=none]
 let x = 
@@
   x + 1
+\end{lstlisting}
+There are several certain errors here: \li{f} and \li{y} are unbound, and the two branches of the conditional expression have inconsistent types.
+There may or may not be an error in the use of \li{x} on Line 4, depending on which branch type is intended.
+
+A type checker with no support for error localization or recovery would simply report that this program is ill-typed, and indeed this behavior is enough to satisfy the language specification, if given in the conventional style (e.g.
+following the simply typed lambda calculus and its derivatives).
+A more helpful and common approach is to localize the error that causes type checking to initially fail, e.g.
+here the system might report that the variable \li{f} is unbound and highlight its location in the error message or with an editor decoration (i.e.
+without attempting recovery).
+This is the approach taken in many production type checkers, e.g. XXX.\todo{examples}
+
+A first attempt at error recovery might continue past this first error to also report that \li{y} is unbound, then ignore the fact that guard of the conditional cannot therefore be confirmed to be a boolean expression and nevertheless continue into its branches.
+Here, the system might observe that the two branches have inconsistent types, \li{int} and \li{bool}.
+How should this error be localized?
+There are many possible approaches here.
+One approach would be to assume that the first branch is correct, so the error can be localized to the second branch.
+This is \emph{ad hoc}, and the system might just as well assume the second branch is correct, localizing the error to the first branch.
+A third approach might be to report the inconsistency between the two types as an error localized to the conditional expression itself.
+
+Which choice is made affects how error recovery will proceed into the body of the let expression, \li{x + 1}.
+In particular, if the system assumes that the conditional should have \li{int} type, then \li{x : int} and no error needs to be reported here.
+If the system assumes it should have \li{bool} type, then an error does need to be reported. 
+If the system reports only the inconsistency between the two branches, then it is not clear what type \li{x} should have, and so it becomes difficult to continue to recover.
 
@@
 
-Different language servers or typecheckers for the same language can and do attempt to localize, report, and recover from these errors in different ways, with little in the way of unifying theory.
-
 For these reasons, accurate type error localization and robust type error recovery have been a topic of considerable practical interest.
-Indeed, essentially all widely-used programming environments (often via an underlying language server) have some support for type error 
-localization and many also attempt recovery of certain services in certain situations.
-However, the design of these systems has evolved in some part as folklore, as type system definitions have remained silent.
 

### 2023-02-28 let.tex

--- 2023-02-27 calculus.tex
@@
-% TODO standardize naming: marking vs mark insertion, marks vs error marks vs markings
-% TODO uniform style for referencing rules
-
-\section{Marked Lambda Calculus}
-\label{sec:calculus}
-
-Programmers often work with program states that are syntactically well-formed but not well-typed.
-\Cref{fig:calculus-examples} shows common type errors as they appear in the Hazel programming
-environment. For example, a simple error that may occur is the use of an unbound variable. As in
-\cref{fig:calculus-examples-unbound}, editor services indicate the usage of unbound $y$ is invalid
-by highlighting it in red.
-
-\input{figs/calculus-examples}
-
-Sometimes, the type an expression \emph{should} be is known. Consider
-\cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
-$+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
-indicates a mismatch between the expected and actual types of the operand. The application in
-\cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
-this case, however, the expectation is not any singular type, but any member of a family of function
-types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error of inconsistent
-branches: the branches of the if-else expression are of different types. Hazel opts to highlight the
-entire expression in such a situation, while other tools may choose one branch to be ``correct'' and
-indicate a type mismatch error on the other.
-
-% TODO elaborate on this part
-
-Unfortunately, formal typing semantics usually only give meaning to well-typed programs, leaving
-such ill-typed ones meaningless. This is insufficient for editor services and language servers,
-which must provide users feedback on where and why a program is ill-typed. Moreover, they must
-continue providing editor services even when an error is encountered. Indeed, similarly to
-error-recovering parsers, real-world type checkers must recover from type errors. Traditionally,
-this has been performed in an \emph{ad hoc} manner [citations?], with different languages and tools
-adopting varying approaches in the localization, reportage, and recovery from errors.
-
-% TODO examples of this ad hoc stuff
-
-The key contribution of this section is a formal calculus, called the \emph{marking calculus}, for
-type error localization in bidirectional systems, a judgmental framework provides a unified theory
-for the design of typing semantics that supports ill-typed programs. \Cref{sec:calculus-calculus}
-describes its form and metatheory by example on a minimal gradually typed lambda calculus, extending
-with binary product types in \cref{sec:calculus-products}. All rules and theorems may be found in
-the supplementary appendix, alongside a complete mechanization in the Agda proof assistant, which is
-discussed in \cref{sec:calculus-agda}. Finally, in \cref{sec:calculus-poly}, we consider how the
-system might be extended to more complex typing features, such as System F-style polymorphism.
-
-% TODO not really sure when to start talking about bidirectional typing
-
-\emph{Bidirectional typing} combines type checking, which determines if a program satisfies a given
-type, and type synthesis, which generates a type from the program. This gives a simple algorithmic
-system in which type information is recursively propagated through terms and allows for local type
-inference.
-
-% TODO talk about idea that bidirectional typing makes error localization easy? i.e. why choose
-% bidirectional?
-
-\subsection{The Calculus}
-\label{sec:calculus-calculus}
-
-% TODO outline the structure of this section at the beginning
-% TODO make it clear unmarked/marked languages share same types
-
-We now formally introduce the \emph{marking calculus} through its application to a gradually typed
-lambda calculus extended with numbers and booleans. Consider the syntax for such a language, as
-shown in \cref{fig:calculus-syntax-uexp}. The base types $\TNum$ and $\TBool$ classify number and
-boolean expressions. The number literal corresponding to the mathematical number $n$ is given by
-$\ENumMV$, and there is a single addition operation on numeric expression. $\ETrue$ and $\EFalse$
-correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
-$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives an if-else structure. In addition to these forms, $\TUnknown$
-is the unknown type from gradual type theory.
-
-\begin{figure}[htbp]
-  \[\begin{array}{rrcl}
-    \textsf{Type} & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
-    \textsf{Exp}  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \\
-         &       & \mid         & \ENumMV \mid \EPlus{\EMV}{\EMV}
-                   \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
-  \end{array}\]
-  \caption{Syntax of the extended gradually typed lambda calculus}
-  \label{fig:calculus-syntax-uexp}
-\end{figure}
-
-The language is typed bidirectionally, which employs two mutually defined judgments. \emph{Type
-synthesis}, written $\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under the typing context
-$\ctx$, the expression $e$ synthesises or infers the type $\TMV$. \emph{Type analysis}, written
-$\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear where an expression of
-type $\TMV$ is expected. We may give these typing semantics traditionally, without considering error
-localization and recovery.
-
-The \emph{marking calculus} builds atop these typing semantics into a two-layer system:
-%
-\begin{itemize}
-  \item The \textbf{unmarked language}, which is the original language. Expressions of this language
-    are \emph{unmarked expressions}.
-
-  \item The \textbf{marked language}, which mirrors the structure of the unmarked language but is
-    extended with explicit \emph{error marks}. We call expressions of this language to be
-    \emph{marked expressions}, denoted $\ECMV$.
-    % TODO more intuition for what marked expressions are?
-
-  \item \textbf{Marking}, which \emph{marks} any unmarked expression into a marked expression,
-    inserting error markings where appropriate.
-\end{itemize}
-%
-In other words, we extend the ordinary typing semantics of a language with a secondary ``marked''
-language and ``mark'' unmarked programs into marked ones. This corresponds to a type checking
-process with error localization and recovery.
-
-% TODO more intuition
-
-We give, for now, only part of the marked expression syntax in \cref{fig:calculus-syntax-mexp-part}.
-Throughout the remainder of this section, as we consider the formulation of marking, we will add the
-necessary error mark forms.
-
-% TODO include holes
-\begin{figure}[htbp]
-  \[\begin{array}{rrcl}
-    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV} \\
-             &       & \mid         & \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
-                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV} \\
-             &       & \mid         & \cdots
-             % &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
-  \end{array}\]
-  \caption{Partial syntax of marked expressions}
-  \label{fig:calculus-syntax-mexp-part}
-\end{figure}
-
-The marked language possesses its own typing semantics, also formulated bidirectionally. We write
-$\ctxSynType{\ctx}{\ECMV}{\TMV}$ for synthesis and $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ for analysis.
-Finally, marking is bidirectional as well. The synthetic marking judgment
-$\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
-unmarked expression $\EMV$ is ``marked into'' the marked expression $\ECMV$, which synthesizes type
-$\TMV$. Analogously, the analytic marking judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
-states that $\EMV$ is marked into $\ECMV$, which analyses against $\TMV$.
-
-How can we ensure that the marking procedure is correctly defined? There are two critical
-metatheorems to keep in mind as we continue. The first is a \emph{totality} of marking:
-%
-\begin{theorem}[name=Marking Totality] \
-  \label{thm:calculus-marking-totality}
-  \begin{enumerate}
-    \item For all $\ctx$ and $\EMV$, there exist $\ECMV$ and $\TMV$ such that
-      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
-  \item For all $\ctx$, $\EMV$, and $\TMV$, there exists $\ECMV$ such that
-    $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
-  \end{enumerate}
-\end{theorem}
-%
-That is, we may mark, i.e. type check and localize errors for \emph{any} syntactically well-formed
-program. Furthermore, marking should preserve the syntactic structure of programs, modulo error
-marks. Thus, we must define a notion of \emph{mark erasure}, which converts marked expressions back
-into unmarked expression by removing error marks. \Cref{fig:calculus-mark-erasure-part} gives the
-definition of mark erasure, written $\erase{\ECMV}$, for our initial marked expression syntax. Then,
-we may define a \emph{well-formedness} criterion for marking:
-%
-\begin{theorem}[name=Marking Well-Formedness] \
-  \label{thm:calculus-marking-well-formedness}
-  \begin{enumerate}
-    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynType{\ctx}{\ECMV}{\TMV}$ and
-      $\erasesTo{\ECMV}{\EMV}$.
-    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ and
-      $\erasesTo{\ECMV}{\EMV}$.
-  \end{enumerate}
-  % TODO split this into two theorems?
-  % TODO not really sure how to intuitively describe the part about types matching up
-\end{theorem}
-%
-In other words, if an unmarked expression is marked into some marked expression, erasing error marks
-from the marked expression gives the original unmarked expression. We also ensure that marking gives
-marked expressions that are well-typed; indeed, their types should match the types synthesised or
-analysed against in marking. % TODO more intuition for this theorem
-
-\input{figs/calculus-mark-erasure-part}
-
-In the remainder of this subsection, we consider the different type errors that may arise in the
-unmarked language; in this way, the typing semantics of the unmarked language guide the construction
-of the marking process.
-
-\subsubsection{Variables}
-\label{sec:calculus-variables}
-
-% TODO subsumption is discussed later...
-The typing of unmarked variables in the gradually typed lambda calculus is standard. Because of
-subsumption, we need only define a synthesis rule:
-\[%
-  \judgment{
-    \inCtx{\ctx}{x}{\TMV}
-  }{
-    \ctxSynType{\ctx}{x}{\TMV}
-  }{USVar}
-\]%
-If $x$ is bound to a type in the typing context, it synthesizes that type. Straightforwardly,
-marking would convert the unmarked $x$ into the same marked $x$ via the following marking and typing
-rules:
-%
-\begin{mathpar}
-  \judgment{
-    \inCtx{\ctx}{x}{\TMV}
-  }{
-    \ctxSynFixedInto{\ctx}{x}{x}{\TMV}
-  }{ISVar}
-
-  \judgment{
-    \inCtx{\ctx}{x}{\TMV}
-  }{
-    \ctxSynType{\ctx}{x}{\TMV}
-  }{MSVar}
-\end{mathpar}
-% TODO more intuition on why this makes sense?
-
-But what happens a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
-bound? Then, \textsc{USVar} does not apply, and the typing semantics say nothing about such
-expressions. To satisfy totality, such a possibility motivates an \emph{unbound variable mark}:
-%
-\[\begin{array}{rrcl}
-  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECUnbound{x}
-\end{array}\]
-%
-accompanied by the following rules:
-%
-\begin{mathpar}
-  \judgment{
-    \notInCtx{\ctx}{x}
-  }{
-    \ctxSynFixedInto{\ctx}{x}{\ECUnbound{x}}{\TUnknown}
-  }{ISUnbound}
-
-  \judgment{
-    \notInCtx{\ctx}{x}
-  }{
-    \ctxSynType{\ctx}{\ECUnbound{x}}{\TUnknown}
-  }{MSUnbound}
-\end{mathpar}
-%
-\textsc{ISUnbound} states that an variable unbound in the context is marked into that variable
-inside an unbound variable mark. \textsc{MSUnbound} then gives the typing of unbound variable marks.
-Since nothing may be said about their types, we synthesize the unknown type, which allows type
-checking to continue---instead of simply failing.
-
-To prove \cref{thm:calculus-marking-well-formedness}, we extend the definition of mark erasure to
-handle the new mark:
-%
-\[\begin{array}{rcl}
-  & \cdots & \\
-  \erasesToRow{\ECUnbound{x}}{x}
-\end{array}\]
-
-\subsubsection{Lambda Abstractions}
-\label{sec:calculus-lambda-abstractions}
-
-There are explicit synthesis and analysis rules for lambda abstractions:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\TArrow{\TMV_1}{\TMV_2}}
-  }{USLam}
-
-  \judgment{
-    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
-    \consistent{\TMV}{\TMV_1} \\
-    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
-  }{
-    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
-  }{UALam}
-\end{mathpar}
-%
-The synthesis rule is standard. In the premises of \textsc{UALam}, we use the notion of
-\emph{type consistency} from gradual type theory, which defines a reflexive and symmetric (but not
-transitive) relation between types, writing $\consistent{\TMV_1}{\TMV_2}$ to mean that $\TMV_1$ is
-consistent with $\TMV_2$. This replaces the notion of \emph{type equality} to relate the unknown
-type to all other types. The analysis rule also employs the judgment
-$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which establishes that $\TMV$ is a \emph{matched arrow type},
-i.e. it may be considered an arrow type. This notion is purely a technical mechanism to avoid
-duplication of rules related to arrow types. \Cref{fig:calculus-consistency-matched-arrow} gives the
-rules for both judgments.
-
-% TODO mention explicitly that by inductive hypothesis, the marking of body cannot fail?
-% TODO more intuition on these two judgments
-% TODO more intuition on USLam and UALam
+\subsection{Let with Composite Patterns}
+\label{sec:calculus-let}
 
@@
   \raggedright
-  \judgbox{\ensuremath{\consistent{\TMV_1}{\TMV_2}}} $\TMV_1$ is consistent with $\TMV_2$
+  \judgbox{\ensuremath{\ctxSynType{\Gamma_{in}}{p}{\tau ; \Gamma_{out}}}} $\TMV$ has matched binary product type $\TPair{\TMV_1}{\TMV_2}$
   %
@@
     \judgment{ }{
-      \consistent{\TUnknown}{\TMV}
-    }{TCUnknown1}
+      \matchedPair{\TUnknown}{\TUnknown}{\TUnknown}
+    }{TMPHole}
 
     \judgment{ }{
-      \consistent{\TMV}{\TUnknown}
-    }{TCUnknown2}
-
-    \judgment{ }{
-      \consistent{\TMV}{\TMV}
-    }{TCRefl}
-
-    \judgment{
-      \consistent{\TMV_1}{\TMV_1'} \\
-      \consistent{\TMV_2}{\TMV_2'} \\
-    }{
-      \consistent{\TArrow{\TMV_1}{\TMV_2}}{\TArrow{\TMV_1'}{\TMV_2'}}
-    }{TCArr} \\
-  \end{mathpar}
-
-  \judgbox{\ensuremath{\matchedArrow{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched arrow type $\TArrow{\TMV_1}{\TMV_2}$
-  %
-  \begin{mathpar}
-    \judgment{ }{
-      \matchedArrow{\TUnknown}{\TUnknown}{\TUnknown}
-    }{TMAHole}
-
-    \judgment{ }{
-      \matchedArrow{\TArrow{\TMV}{\TMV}}{\TMV}{\TMV}
-    }{TMAArr}
+      \matchedPair{\TPair{\TMV_1}{\TMV_2}}{\TMV_1}{\TMV_2}
+    }{TMPProd}
   \end{mathpar}
   %
-  \caption{Type consistency and matched arrow types.}
-  \label{fig:calculus-consistency-matched-arrow}
+  \caption{Matched product type.}
+  \label{fig:calculus-matched-product}
 \end{figure}
 
-The synthesis rule for marking intuitively follows \textsc{USLam} closely. We should recursively
-mark the body with an extended context binding $x$ to $\TMV$ and construct a new marked lambda
-abstraction:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxSynFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
-  }{
-    \ctxSynFixedInto{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\ELam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
-  }{ISLam}
-
-  \judgment{
-    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\ECLam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
-  }{MSLam}
-\end{mathpar}
-%
-Similarly, we construct corresponding analytic rules:
-%
-\begin{mathpar}
-  \judgment{
-    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
-    \consistent{\TMV}{\TMV_1} \\\\
-    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
-  }{
-    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
-  }{IALam1}
-
-  \judgment{
-    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
-    \consistent{\TMV}{\TMV_1} \\\\
-    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
-  }{
-    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
-  }{MALam1}
-\end{mathpar}
-%
-However, recalling totality, we look again to the premises of \textsc{UALam} to see what type errors
-may arise. First, what if $\TMV_3$ is not a matched arrow type? This would be the case if it were
-$\TNum$, for example, and the lambda abstraction would be inconsistent with the expected type. Thus,
-to the syntax we add an \emph{inconsistent type mark}, which indicates that the expression does not
-synthesize a type consistent with the expected type:
-%
-\[\begin{array}{rrcl}
-  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECInconType{\ECMV}
-\end{array}\]
-%
-along with the following rules, where the lambda abstraction is placed inside the inconsistent type
-mark.
-%
-\begin{mathpar}
-  \judgment{
-    \notMatchedArrow{\TMV_3} \\
-    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TUnknown}
-  }{
-    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
-  }{IALam2}
-
-  \judgment{
-    \notMatchedArrow{\TMV_3} \\
-    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
-  }{
-    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
-  }{MALam2*}
-\end{mathpar}
-%
-Though no expected output type is known, we still need to check and mark the body; it is thus
-analysed against the unknown type in \textsc{IALam2}.
-
-A similar situation arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
-$\inconsistent{\TMV}{\TMV_1}$, i.e. the expected input type is inconsistent with the actual input
-type, giving the last pair of analytic rules for lambda abstractions:
-%
-\begin{mathpar}
-  \judgment{
-    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
-    \inconsistent{\TMV}{\TMV_1} \\\\
-    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
-  }{
-    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
-  }{IALam3}
-
-  \judgment{
-    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
-    \inconsistent{\TMV}{\TMV_1} \\\\
-    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
-  }{
-    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
-  }{MALam3}
-\end{mathpar}
-
-Finally, as with the unbound variable mark, we extend mark erasure:
-%
-\[\begin{array}{rcl}
-  & \cdots & \\
-  \erasesToRow{\ECInconType{\ECMV}}{\erase{\ECMV}}
-\end{array}\]
-
-\subsubsection{Applications}
-\label{sec:calculus-applications}
-
-The case of application is somewhat interesting. In the unmarked language, only a synthesis rule is
-necessary:
-\[%
-  \judgment{
-    \ctxSynType{\ctx}{\EMV_1}{\TMV} \\
-    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\
-    \ctxAnaType{\ctx}{\EMV_2}{\TMV_1}
-  }{
-    \ctxSynType{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\TMV_2}
-  }{USAp}
-\]%
-Following the same methodology, we can straightforwardly give the following marking and typing
-rules:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
-    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\\\
-    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
-  }{
-    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
-  }{ISAp1}
-
-  \judgment{
-    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
-    \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\\\
-    \ctxAnaType{\ctx}{\ECMV_2}{\TMV_1}
-  }{
-    \ctxSynType{\ctx}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
-  }{MSAp1}
-\end{mathpar}
-
-Again, to satisfy totality, we must consider the case when $\TMV$ is not a matched arrow type. There
-is no expected type for the argument, so we perform analytic marking on $\EMV_2$ against the
-unknown type. The applied expression is placed inside the inconsistent type mark, which is in
-accordance with the meaning of that mark.
-%
-\begin{mathpar}
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
-    \notMatchedArrow{\TMV} \\\\
-    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
-  }{
-    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApNonMatched{\ECMV_1}{\ECMV_2}}{\TUnknown}
-  }{ISAp2}
-
-  \judgment{
-    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
-    \notMatchedArrow{\TMV} \\\\
-    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
-  }{
-    \ctxSynType{\ctx}{\ECApNonMatched{\ECMV_1}{\ECMV_2}}{\TUnknown}
-  }{MSAp2}
-\end{mathpar}
-
-It is worth noting, however, that $\EMV_1$ is not actually in analytic position, but in a synthetic
-position where any arrow type is expected. Treating such ``constrained synthetic'' cases specially,
-we may instead designate a specialized mark, written $\ECApNonMatchedAltAlt{\ECMV}$. This yields the
-following alternative rules:
-%
-\[\begin{array}{rrcl}
-  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECApNonMatchedAltAlt{\ECMV}
-\end{array}\]
-%
-\begin{mathpar}
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
-    \notMatchedArrow{\TMV} \\\\
-    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
-  }{
-    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECApNonMatchedAltAlt{\ECMV_1}}{\ECMV_2}}{\TUnknown}
-  }{ISAp2'}
-
-  \judgment{
-    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
-    \notMatchedArrow{\TMV} \\\\
-    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
-  }{
-    \ctxSynType{\ctx}{\ECApNonMatchedAltAlt{\ECMV_1}{\ECMV_2}}{\TUnknown}
-  }{MSAp2'}
-\end{mathpar}
-
-It is natural to extend this approach to other elimination forms that require the handling of
-unmatched types, such as binary products. Indeed, in \cref{sec:calculus-products}, we opt for this
-approach. Another approach is to mark an error around the entire application expression.
-
-\subsubsection{Numbers}
-\label{sec:calculus-numbers}
-
-The marking and typing of numbers is straightforward:
-%
-\begin{mathpar}
-  \judgment{ }{
-    \ctxSynType{\ctx}{\ENumMV}{\TNum}
-  }{USNum}
-
-  \judgment{
-  }{
-    \ctxSynFixedInto{\ctx}{\ENumMV}{\ECNumMV}{\TNum}
-  }{ISNum}
-
-  \judgment{ }{
-    \ctxSynType{\ctx}{\ECNumMV}{\TNum}
-  }{MSNum}
-\end{mathpar}
-
-For addition expressions, we simply recursively mark each operand, which must analyse against
-$\TNum$, and build the marked addition expression. In this way, blame for type errors is shift into
-the operands:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxAnaType{\ctx}{\EMV_1}{\TNum} \\\\
-    \ctxAnaType{\ctx}{\EMV_2}{\TNum}
-  }{
-    \ctxSynType{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
-  }{USPlus}
-
-  \judgment{
-    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\\\
-    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
-  }{
-    \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
-  }{ISPlus}
-
-  \judgment{
-    \ctxAnaType{\ctx}{\ECMV_1}{\TNum} \\\\
-    \ctxAnaType{\ctx}{\ECMV_2}{\TNum}
-  }{
-    \ctxSynType{\ctx}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
-  }{MSPlus}
-\end{mathpar}
-%
-Again, we need only synthetic rules; subsumption handles the analytic case.
-
-\subsubsection{Booleans}
-\label{sec:calculus-booleans}
-
-Boolean values are similarly straightforward:
-%
-\begin{mathpar}
-  \judgment{ }{
-    \ctxSynType{\ctx}{\ETrue}{\TBool}
-  }{USTrue}
-
-  \judgment{
-  }{
-    \ctxSynFixedInto{\ctx}{\ETrue}{\ECTrue}{\TBool}
-  }{ISTrue}
-
-  \judgment{ }{
-    \ctxSynType{\ctx}{\ECTrue}{\TBool}
-  }{MSTrue} \\
-
-  \judgment{ }{
-    \ctxSynType{\ctx}{\EFalse}{\TBool}
-  }{USFalse}
-
-  \judgment{
-  }{
-    \ctxSynFixedInto{\ctx}{\EFalse}{\ECFalse}{\TBool}
-  }{ISFalse}
-
-  \judgment{ }{
-    \ctxSynType{\ctx}{\ECFalse}{\TBool}
-  }{MSFalse}
-\end{mathpar}
-
-If-else expressions, however, offer another interesting case. In the unmarked language, we have
-both explicit synthetic and analytic rules:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxAnaType{\ctx}{\EMV_1}{\TBool} \\\\
-    \ctxSynType{\ctx}{\EMV_2}{\TMV_1} \\
-    \ctxSynType{\ctx}{\EMV_3}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
-  }{USIf}
-
-  \judgment{
-    \ctxAnaType{\ctx}{\EMV_1}{\TBool} \\\\
-    \ctxAnaType{\ctx}{\EMV_1}{\TMV} \\
-    \ctxAnaType{\ctx}{\EMV_2}{\TMV}
-  }{
-    \ctxAnaType{\ctx}{\ECIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TMV}
-  }{UAIf}
-\end{mathpar}
-%
-In synthetic position, conditionals synthesize the \emph{join} of the branch types $\TMV_1$ and
-$\TMV_2$, which we define inductively in \cref{fig:calculus-type-join}. In analytic position, since
-there is an expected type for both branches, we shift the blame for any errors to them.
-
-% TODO intuition why we use join
-
-\begin{figure}[htbp]
-  \raggedright
-  \judgbox{\ensuremath{\TJoin{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
-  %
-  \newcommand{\joinsTo}[3]{\ensuremath{\TJoin{#1}{#2} & = & #3}}
-  \[\begin{array}{rcl}
-    \joinsTo{\TUnknown}{\TMV}{\TMV} \\
-    \joinsTo{\TMV}{\TUnknown}{\TMV} \\
-    \joinsTo{\TNum}{\TNum}{\TNum} \\
-    \joinsTo{\TBool}{\TBool}{\TBool} \\
-    \joinsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}} \\
-  \end{array}\]
-  %
-  \caption{Type join.}
-  \label{fig:calculus-type-join}
-\end{figure}
-% TODO intuition for type join?
-
-Following \textsc{USIf} and \textsc{UAIf}, we may derive the following marking and typing rules.
-Similar to previous cases, we simply recursively mark the guard and branch expressions and construct
-the marked if-else expression from the results.
-%
-\begin{mathpar}
-  \judgment{
-    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
-    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
-    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2}
-  }{
-    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
-  }{ISIf}
-
-  \judgment{
-    \ctxAnaType{\ctx}{\ECMV_1}{\TBool} \\\\
-    \ctxSynType{\ctx}{\ECMV_2}{\TMV_1} \\
-    \ctxSynType{\ctx}{\ECMV_3}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
-  }{MSIf} \\
-
-  \judgment{
-    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
-    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV} \\
-    \ctxAnaFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV} \\
-  }{
-    \ctxAnaFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
-  }{IAIf}
-
-  \judgment{
-    \ctxAnaType{\ctx}{\ECMV_1}{\TBool} \\\\
-    \ctxAnaType{\ctx}{\ECMV_1}{\TMV} \\
-    \ctxAnaType{\ctx}{\ECMV_2}{\TMV}
-  }{
-    \ctxAnaType{\ctx}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
-  }{MAIf}
-\end{mathpar}
-
-However, in synthetic mode, it may be the case that the two branch types have no join. This occurs,
-in fact, when they are inconsistent. This motivates \emph{the inconsistent branches mark}:
-%
-\[\begin{array}{rrcl}
-  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
-\end{array}\]
-
-Then, adding the following rules ensures marking totality on conditionals:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
-    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\\\
-    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2} \\
-    \inconsistent{\TMV_1}{\TMV_2}
-  }{
-    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TUnknown}
-  }{ISInconsistentBranches}
-
-  \judgment{
-    \ctxAnaType{\ctx}{\ECMV_1}{\TBool} \\
-    \ctxSynType{\ctx}{\ECMV_2}{\TMV_1} \\\\
-    \ctxSynType{\ctx}{\ECMV_3}{\TMV_2} \\
-    \inconsistent{\TMV_1}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TUnknown}
-  }{MSInconsistentBranches}
-\end{mathpar}
-
-The extension of mark erasure is again straightforward:
-%
-\[\begin{array}{rcl}
-  & \cdots & \\
-  \erasesToRow{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\EIf{(\erase{\ECMV_1})}{(\erase{\ECMV_2})}{(\erase{\ECMV_3})}}
-\end{array}\]
-
-\subsubsection{Subsumption}
-\label{sec:calculus-subsumption}
-
-Finally, we consider the subsumption rule in typing unmarked expressions, which states that if an
-expression synthesizes a type, it may also be analysed against that type or any consistent type.
-\[%
-  \judgment{
-    \ctxSynType{\ctx}{\EMV}{\TMV'} \\
-    \consistent{\TMV}{\TMV'} \\
-    \subsumable{\EMV}
-  }{
-    \ctxAnaType{\ctx}{\EMV}{\TMV}
-  }{UASubsume}
-\]%
-We also restrict the usage of subsumption to ``subsumable'' syntactic forms, written
-$\subsumable{\EMV}$. This judgment is defined on all syntactic forms except lambda abstractions and
-conditionals, the only ones with explicit analysis rules. In other words, we restrict
-subsumption to be the analytic rule of ``last resort''. This is critical in establishing that
-marking and typing are deterministic (discussed later in\cref{thm:calculus-marking-unicity}).
-
-Since we did not provide explicit analytic rules for marking expressions that are not lambda
-abstractions or conditionals (or typing their marked versions) either, we give corresponding
-subsumption rules:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
-    \consistent{\TMV}{\TMV'} \\
-    \subsumable{\EMV}
-  }{
-    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}
-  }{IASubsume}
-
-  \judgment{
-    \ctxSynType{\ctx}{\ECMV}{\TMV'} \\
-    \consistent{\TMV}{\TMV'} \\
-    \subsumable{\ECMV}
-  }{
-    \ctxAnaType{\ctx}{\ECMV}{\TMV}
-  }{MASubsume}
-\end{mathpar}
-%
-Note that we define a analogous notion of ``subsumability'' for marked expressions, written
-$\subsumable{\ECMV}$.
-
-However, it may be the case that the expression is inconsistent with the expected type. Recalling
-the example of \cref{fig:calculus-examples-inconsistent-types}, subsumption would used to analyse
-$\ETrue$ against $\TNum$ in $\EPlus{\ETrue}{5}$, which fails since $\inconsistent{\TNum}{\TBool}$,
-and $\ETrue$ should be marked with an inconsistent type mark. Hence, the following final set of
-rules is required to handle such cases:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
-    \inconsistent{\TMV}{\TMV'} \\\\
-    \subsumable{\EMV}
-  }{
-    \ctxAnaFixedInto{\ctx}{\EMV}{\ECInconType{\ECMV}}{\TMV}
-  }{IAInconsistentTypes}
-
-  \judgment{
-    \ctxSynType{\ctx}{\ECMV}{\TMV'} \\
-    \inconsistent{\TMV}{\TMV'} \\\\
-    \subsumable{\ECMV}
-  }{
-    \ctxAnaType{\ctx}{\ECInconType{\ECMV}}{\TMV}
-  }{MAInconsistentTypes}
-\end{mathpar}
-%
-These rules complete the development of the marking calculus for the gradually typed lambda
-calculus.
-
-% TODO mini conclusion on methodology
-
-To conclude this section, we present two more metatheorems that help ensure correctness of the
-system. Though \cref{thm:calculus-marking-well-formedness} guarantees that marking does not change
-the syntactic structure of a program, it makes no statement about the presence of error marks in the
-resulting marked program. Following the logic of constructing marking rules, we have
-\cref{thm:calculus-marking-well-ill-typed}.
-
-\begin{theorem}[name=Marking of Well-Typed/Ill-Typed Expressions] \ % TODO better name for this
-  \label{thm:calculus-marking-well-ill-typed}
-  \begin{enumerate}
-    \item \begin{enumerate}
-        \item If $\ctxSynType{\ctx}{\EMV}{\TMV}$, then $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
-          such that $\imageIs{\EMV}{\ECMV}$.
-        \item If $\ctxAnaType{\ctx}{\EMV}{\TMV}$, then $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
-          such that $\imageIs{\EMV}{\ECMV}$.
-      \end{enumerate}
-
-    \item \begin{enumerate}
-        \item If there does not exist $\TMV$ such that $\ctxSynType{\ctx}{\EMV}{\TMV}$, then for all
-          $\ECMV$ and $\TMV'$ such that $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, we have
-          $\imageIsNot{\EMV}{\ECMV}$.
-        \item If there does not exist $\TMV$ such that $\ctxAnaType{\ctx}{\EMV}{\TMV}$, then for all
-          $\ECMV$ and $\TMV'$ such that $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, we have
-          $\imageIsNot{\EMV}{\ECMV}$.
-      \end{enumerate}
-  \end{enumerate}
-\end{theorem}
-%
-That is, since error marks indicate type errors in the program, no marks should be inserted
-into a well-typed program. The converse---that at least one mark is inserted into ill-typed
-programs---is also true. We have chosen to formulate the absence of error marks through an inclusion
-mapping of unmarked expressions into marked expressions; alternatively, a predicate over marked
-expressions may be used. % TODO latter is easier to do in Agda, but maybe it is weaker?
-
-Finally, no less importantly, marking is deterministic. This is given by
-\cref{thm:calculus-marking-unicity}.
-%
-\begin{theorem}[name=Marking Unicity] \
-  \label{thm:calculus-marking-unicity}
-  \begin{enumerate}
-    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV_1}$ and
-      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV_2}$, then $\ECMV_1 = \ECMV_2$ and $\TMV_1 =
-      \TMV_2$.
-    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV}$ and
-      $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV}$, then $\ECMV_1 = \ECMV_2$.
-  \end{enumerate}
-\end{theorem}
-%
-Together with totality, marking may be considered and implemented as a total function.
-
-% TODO non-gradual semantics?
-
-% Note that while the semantics that follow concern a gradually typed language, a non-gradual system
-% may be recovered straightforwardly by removing the unknown type and replacing type consistency with
-% equality.
-
-\subsection{Adding Binary Products}
-\label{sec:calculus-products}
-
-We now demonstrate how one might extend or adapt the marking calculus for new typing features by
-adding binary product types. First, we extend in \cref{fig:calculus-products-judgments} the syntax
-of types and notions of consistency and join. As with arrow types, we additionally define an
-auxiliary notion of \emph{matched binary product types}: the judgment
-$\matchedPair{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be considered the binary product
-type $\TPair{\TMV_1}{\TMV_2}$.
-
-\input{figs/calculus-products-judgments}
-
-The unmarked language with a pair constructor and projection operators.
-\Cref{fig:calculus-products-uexp} gives their syntax and typing rules. Similarly to application,
-\textsc{USProjL} and \textsc{USProjR} employ the matched binary product type; a projection
-$\EProjL{\EMV}$ or $\EProjR{\EMV}$ is then well-typed if the $\EMV$ may be considered to be a pair.
-
-\input{figs/calculus-products-uexp}
-
-\input{figs/calculus-products-mexp}
-
-\input{figs/calculus-products-marking}
-
-\subsection{Mechanization}
-\label{sec:calculus-agda}
-
-The metatheory of the calculus has been fully mechanized in the Agda proof assistant [citation]. The
-mechanization's documentation contains more details regarding technical decisions made therein.
-
-\subsection{Polymorphism and Other Typing Features}
-\label{sec:calculus-poly}
-
-\subsection{Implementation in Hazel}
-
-\subsection{Reified Hazelnut Action Semantics}
+%TODO:andrew

### 2023-02-28 intro.tex
Add new constraint rules using marked expression syntax and narrative surrounding them. Removes old rules (some errors due to stray references to them persist)

--- 2023-02-27 intro.tex
@@
 
-Contemporary programming environments provide 
-a variety of semantic services that require statically analyzing the type and binding structure of the program being edited, e.g.
-semantic navigation, type hints, semantic code completion, and automated refactoring.
-However, during the editing process,
-it is common for programs to be ill-typed.
+Modern programming environments provide developers with 
+various semantic services, such as type hints, semantic navigation, semantic code completion, and automated refactoring, that require statically reasoning about the type and binding structure of the program as it is being edited.
+% These semantic services are often packaged into language servers\todo{cites}.
+The problem is that the program being edited can be ill-typed, sometimes for long periods of time. 
+For example, when the developer changes an important type definition, errors might arise at dozens of usage sites in a large program, which could take hours or days to address.
+Semantic services often exhibit gaps in service during these periods \cite{HazelnutSNAPL}, because in the conventional formal account of type systems, only well-typed programs have a well-defined type and binding structure.
+As soon as a type error occurs \emph{anywhere}, the program is formally meaningless \emph{everywhere}.
 
-Language designers typically specify static type systems as a collection of rules 
-stating conditions under which an expression can be judged to be well-typed,
-i.e. completely free of type errors. 
- is well-supported 
-by these definitions.
+Developers benefit when these gaps in service are minimized or eliminated, so there has long been considerable practical interest in (1)~\textbf{type error localization}: mechanisms for directing the developer to
+the location(s) in a program that may be erroneous, and (2) \textbf{type error recovery}: 
+mechanisms that allow the language server to optimistically recover from localized errors and continue to provide downstream semantic services, ideally throughout the program.
 
-When there is a type error, however, these definitions are inadequate.
-It is therefore left as an implementation decision (1) type error localization: how to localize and explain the type error, and (2) type error recovery: how to recover from the error to find other simultaneous errors and to continue to provision various semantic services, which would often still useful even in the presence of errors.
+Essentially all widely-used programming systems have some support for type error 
+localization, e.g. via compiler error messages or user interface affordances fed by data from a language server.\todo{cites}
+Many also attempt recovery of certain services in certain situations, discussed below. 
+However, type error localization and recovery mechanisms has developed idiosyncratically, often as folklore amongst tool designers.
+Different type checkers or language servers, even for the same language, localize and recover from type errors in different ways, with little in the way of unifying theory of the sort that grounds the design of modern type systems themselves.
 
-Consider, for example,
-
+Consider, for example, the following program written in a typed functional language with, for now, only local type inference \cite{pierce}: 
+\begin{lstlisting}[numbers=none]
 let x = 
@@
   x + 1
+\end{lstlisting}
+A type checker with no support for error localization or recovery at all would simply report that this program is ill-typed, because a type cannot be derived for it according to the type system specification, if given in the conventional style (e.g.
+following the simply typed lambda calculus and its derivatives).
 
-(compare Elm/OCaml/Haskell and Hazel directly?)
+The more common and practical approach is to localize the first error that causes type checking to fail and emit an error message\todo{cites} (which can help, though studies suggest that developers primarily use location information).
+In this example, such a system might report that the variable \li{f} is unbound and communicate its location, e.g.
+using editor decorations.
+This is the approach taken in many production type checkers, e.g. XXX.\todo{examples}
 
-Different language servers or typecheckers for the same language can and do attempt to localize, report, and recover from these errors in different ways, with little in the way of unifying theory.
+A language server with support for type error recovery might continue past this initial error to also report that \li{y} is unbound.
+To recover further, the system might optimistically ignore the fact that guard of the conditional cannot be confirmed to be a boolean expression and continue into its branches, observing that they have inconsistent types, \li{int} and \li{bool}.
 
-For these reasons, accurate type error localization and robust type error recovery have been a topic of considerable practical interest. 
-Indeed, essentially all widely-used programming environments (often via an underlying language server) have some support for type error 
-localization and many also attempt recovery of certain services in certain situations.
-However, the design of these systems has evolved in some part as folklore, as type system definitions have remained silent.
+There are several potential approaches to localizing this inconsistency. 
+One approach would be to assume, arbitrarily, that the first branch is correct, so the error can be localized to the second branch, or \emph{vice versa}. 
+Heuristic approaches have been developed to make choices like these less arbitrary, e.g. by inspecting the recent editing history \todo{cite}, or training a machine learning model \todo{cite}. 
+Another approach is to localize the inconsistency between the two branches to the conditional expression itself. This entirely avoids needing to heuristically decide which branch is correct.
 
-One reason for this is 
-that type system definitions typically only define the conditions under which an expression is well-typed, e.g.
-in the style of the simply typed lambda calculus and its derivatives.
-When there is a type error anywhere in the program, it is everywhere meaningless according to definitions in this style, i.e.
-there is no direct specification for how to reason within an erroneous program.
+This localization decision affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
+If localization assumes that the \li{then} branch is correct, then \li{x : int} and there is no error to report here.
+If the \li{else} branch is chosen, then \li{x : bool} and an error can be reported (though this error might mislead the programmer if this earlier localization step was incorrect). 
+If instead the inconsistency is localized to the conditional expression as a whole, then it is not obvious which type to assign to \li{x} at all. (Mention gradual typing here?)
 
-The primary contribution of this paper is to provide these missing theoretical foundations for type error localization and recovery. 
+This exercise demonstrates that type error localization and recovery mechanisms engage in non-trivial reasoning and make choices that are not fully specified by the type system itself. 
+The first contribution of this paper is to address this problem by developing formal foundations for type error localization and recovery. 
 Start assuming a system with local type inference / bidirectional typing, marked lambda calculus(Give judgement forms right off the bat??) 

### 2023-02-28 let.tex
Add new constraint rules using marked expression syntax and narrative surrounding them. Removes old rules (some errors due to stray references to them persist)

@@
 
+%TODO(andrew): justification, examples
+
+We augment our type system with a new variant on the unknown type which can trigger a swtich from analysis to synthesis.
+The new type $\TUnknownSwitch$ behaves idenitically to $\TUnknown$ with respect to type consistency, the marched arrow and product judgements, and the join metafunction:
+%
+\[\begin{array}{rrcl}
+  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TUnknownSwitch
+\end{array}\]
+
+The unmarked language is extended by let expressions with composite patterns:
+%
+\[\begin{array}{rrcl}
+  \EMName  & \EMV  & \Coloneqq & \cdots
+                               \mid \ELet{\PMV}{\EMV}{\EMV}
+\end{array}\]
+\[\begin{array}{rrcl}
+  \PMName  & \PMV  & \Coloneqq x 
+                               \mid \_
+                               \mid (p_1,p_2)
+                               \mid p : \TMV
+
+\end{array}\]
+%
+Let expression have the following semantics:
+\begin{mathpar}
+    \judgment{
+     \ctxSynPat{\ctx}{\PMV}{\TMV_p}
+     \\
+     \ctxAnaType{\ctx}{\EMV_{def}}{\TMV_p}
+     \\
+     \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
+     \\\\
+     \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
+     \\
+     \ctxSynType{\ctx_{b}}{\EMV_{b}}{\TMV}
+   }{
+     \ctxSynType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
+   }{USLet}
+   \\
+   \judgment{
+     \ctxSynPat{\ctx}{\PMV}{\TMV_p}
+     \\
+     \ctxAnaType{\ctx}{\EMV_{def}}{\TMV_p}
+     \\
+     \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
+     \\\\
+     \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
+     \\
+     \ctxAnaType{\ctx_{b}}{\EMV_{b}}{\TMV}
+   }{
+     \ctxAnaType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
+   }{UALet}
+\end{mathpar}
+Additionally, we add a rule governing the switch from analysis to synthesis, triggered by analysis against the new SynSwitch type:
+\begin{mathpar}
+    \judgment{
+     \ctxSynType{\ctx}{\EMV}{_}
+   }{
+     \ctxAnaType{\ctx}{\EMV}{\TUnknownSwitch}
+   }{SynSwitch}
+
+\end{mathpar}
+
 \begin{figure}[htbp]
   \raggedright
-  \judgbox{\ensuremath{\ctxSynType{\Gamma_{in}}{p}{\tau ; \Gamma_{out}}}} $\TMV$ has matched binary product type $\TPair{\TMV_1}{\TMV_2}$
+  \judgbox{\ensuremath{\ctxSynPat{\Gamma}{\PMV}{\TMV}}} Pattern $\PMV$ synthesizes type $\TMV$
   %
   \begin{mathpar}
-    \judgment{ }{
-      \matchedPair{\TUnknown}{\TUnknown}{\TUnknown}
-    }{TMPHole}
 
-    \judgment{ }{
-      \matchedPair{\TPair{\TMV_1}{\TMV_2}}{\TMV_1}{\TMV_2}
-    }{TMPProd}
+   \judgment{ }{\ctxSynPat{\ctx}{\_}{\TUnknownSwitch}}{USPatWild}
+
+   \judgment{ }{\ctxSynPat{\ctx}{x}{\TUnknownSwitch}}{USPatVar}
+
+   %\judgment{ }{\ctxSynPat{\Gamma}{\hehole{}}{\TUnknownSwitch}}{USPatHole}
+   \judgment{
+     \ctxAnaPat{\ctx}{\PMV{}}{\tau}{\_}
+   }{
+        \ctxSynPat{\ctx}{\PMV : \tau}{}{\tau}
+    }{USPatAnn}
+
+   \\\\
+   \judgment{
+     \ctxSynPat{\ctx}{\PMV_1}{\TMV_1} \\
+     \ctxSynPat{\ctx}{\PMV_2}{\TMV_2}
+   }{
+     \ctxSynPat{\ctx}{\EPair{\PMV_1}{\PMV_2}}{\TPair{\TMV_1}{\TMV_2}}
+   }{USPatPair}
+
+   \judgment{
+     \ctxSynPat{\ctx}{\PMV{}}{\_}
+   }{
+        \ctxSynPat{\ctx}{\ECMarkedInconType{\PMV}{}}{\TUnknownSwitch}
+    }{USPatErr}
+
   \end{mathpar}
   %
-  \caption{Matched product type.}
-  \label{fig:calculus-matched-product}
+  \caption{Pattern Synthesis}
+  \label{fig:calculus-pattern-synthesis}
+\end{figure}
+
+\begin{figure}[htbp]
+  \raggedright
+  \judgbox{\ensuremath{\ctxAnaPat{\ctx_{in}}{\PMV}{\TMV}{\ctx_{out}}}} Pattern $\PMV$ analyzes against type $\TMV$ producing context $\Gamma_{out}$
+  %
+  \begin{mathpar}
+   \judgment{ }{\ctxAnaPat{\ctx}{\_}{\TMV}{\ctx}}{UAPatWild}
+
+   \judgment{ }{\ctxAnaPat{\Gamma}{x}{\TMV}{\ctx, x:\TMV}}{UAPatVar}
+
+   %\judgment{ }{\ctxAnaPat{\ctx}{\hehole{}}{\tau}}{UAPatHole}{\ctx}
+   \judgment{
+     \ctxAnaPat{\ctx}{\PMV{}}{\tau'}{\ctx'}
+     \\
+     \consistent{\tau}{\tau'}
+   }{
+        \ctxAnaPat{\ctx}{\PMV : \tau'}{\tau}{\ctx'}
+    }{UAPatAnn}
+
+   \\\\
+   \judgment{
+      \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
+      \ctxAnaPat{\ctx}{\PMV_1}{\TMV_1}{\ctx_1} \\
+      \ctxAnaPat{\ctx_1}{\PMV_2}{\TMV_2}{\ctx_2}
+    }{
+      \ctxAnaPat{\ctx}{\ECPair{\PMV_1}{\PMV_2}}{\TMV}{\ctx_2}
+    }{UAPatPair}
+
+   \judgment{
+     \ctxSynPat{\ctx}{\PMV{}}{\_}
+   }{
+        \ctxAnaPat{\ctx}{\ECMarkedInconType{\PMV}{}}{\tau}{\ctx}
+    }{UAPatErr}
+  \end{mathpar}
+  %
+  \caption{Pattern Analysis}
+  \label{fig:calculus-pattern-analysis}
 \end{figure}

### 2023-02-28 calculus.tex
Updates from Overleaf
Finish marking binary products section

@@
 \cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
-this case, however, the expectation is not any singular type, but any member of a family of function
+this case, however, the expectation is not a singular type, but any member of a family of function
 types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error of inconsistent
@@
 We now demonstrate how one might extend or adapt the marking calculus for new typing features by
-adding binary product types. First, we extend the syntax of types and the accompanying notions of
-consistency and join, which closely mirror the definitions for arrow types.
+adding binary product types. First, in \cref{fig:calculus-product-judgments}, we extend the syntax
+of types and the accompanying notions of consistency and join, which closely mirror the definitions
+for arrow types.
 %
@@
 
-We additionally define an auxiliary notion of \emph{matched binary product types} in
-\cref{fig:calculus-matched-product}: the judgment $\matchedPair{\TMV}{\TMV_1}{\TMV_2}$ establishes
-that $\TMV$ may be considered the binary product type $\TPair{\TMV_1}{\TMV_2}$.
+We additionally define an auxiliary notion of \emph{matched binary product types}: the judgment
+$\matchedPair{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be considered the binary product
+type $\TPair{\TMV_1}{\TMV_2}$.
 
@@
   \raggedright
+  \judgbox{\ensuremath{\consistent{\TMV_1}{\TMV_2}}} $\TMV_1$ is consistent with $\TMV_2$
+  %
+  \begin{mathpar}
+    \judgment{
+      \consistent{\TMV_1}{\TMV_1'} \\
+      \consistent{\TMV_2}{\TMV_2'} \\
+    }{
+      \consistent{\TPair{\TMV_1}{\TMV_2}}{\TPair{\TMV_1'}{\TMV_2'}}
+    }{TCProd} \\
+  \end{mathpar}
+
+  \judgbox{\ensuremath{\TJoin{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
+  %
+  \newcommand{\joinsTo}[3]{\ensuremath{\TJoin{#1}{#2} & = & #3}}
+  \[\begin{array}{rcl}
+    & \cdots & \\
+    \joinsTo{(\TPair{\TMV_1}{\TMV_2})}{(\TPair{\TMV_1'}{\TMV_2'})}{\TPair{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}} \\
+  \end{array}\]
+
   \judgbox{\ensuremath{\matchedPair{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched binary product type $\TPair{\TMV_1}{\TMV_2}$
@@
   %
-  \caption{Matched product type.}
-  \label{fig:calculus-matched-product}
+  \caption{Binary product consistency, join, and matched product type.}
+  \label{fig:calculus-product-judgments}
 \end{figure}
 
-The unmarked language is extended with a pair constructor and projection operators.
+The unmarked language is extended with a pair constructor and projection operators. In addition to
+the synthesis rules, we have explicit analysis rules. If an expected type is known for a pair
+expression, the responsibility for type errors is given to the components. Likewise, we shift blame
+to the projected expression when analysing projections.
 %
@@
   }{USProjR} \\
+
+  \judgment{
+    \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
+    \ctxAnaType{\ctx}{\EMV_1}{\TMV_1} \\
+    \ctxAnaType{\ctx}{\EMV_2}{\TMV_2}
+  }{
+    \ctxAnaType{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TMV}
+  }{UAPair}
+
+  \judgment{
+    \ctxAnaType{\ctx}{\EMV}{\TPair{\TMV}{\TUnknown}}
+  }{
+    \ctxAnaType{\ctx}{\EProjL{\EMV}}{\TMV}
+  }{UAProjL}
+
+  \judgment{
+    \ctxAnaType{\ctx}{\EMV}{\TPair{\TUnknown}{\TMV}}
+  }{
+    \ctxAnaType{\ctx}{\EProjR{\EMV}}{\TMV}
+  }{UAProjR}
 \end{mathpar}
@@
 Similarly to application, \textsc{USProjL} and \textsc{USProjR} employ the matched binary product
-type; a projection $\EProjL{\EMV}$ or $\EProjR{\EMV}$ is then well-typed if the $\EMV$ may be
-considered to be a pair.
-
-\input{figs/calculus-products-mexp}
-
-\input{figs/calculus-products-marking}
+type judgment; a projection $\EProjL{\EMV}$ or $\EProjR{\EMV}$ is then well-typed if the $\EMV$ may
+be considered to be a pair.
+
+The marked language is extended with the same forms, and we devise the following marking rules,
+again modelling off the above unmarked typing rules:
+%
+\[\begin{array}{rrcl}
+  \ECMName & \ECMV & \Coloneqq & \cdots
+                               \mid \ECPair{\ECMV}{\ECMV}
+                               \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
+                               \mid \ECInconMatchedPair{\ECMV}
+\end{array}\]
+%
+\begin{mathpar}
+  \judgment{
+    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
+    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
+  }{
+    \ctxSynFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TPair{\TMV_1}{\TMV_2}}
+  }{ISPair}
+
+  \judgment{
+    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
+    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV_1}
+  }{ISProjL1}
+
+  \judgment{
+    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
+    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV_2}
+  }{ISProjR1} \\
+
+  \judgment{
+    \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
+    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
+    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
+  }{
+    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
+  }{IAPair1}
+
+  \judgment{
+    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TPair{\TMV}{\TUnknown}}
+  }{
+    \ctxAnaFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV}
+  }{IAProjL}
+
+  \judgment{
+    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TPair{\TMV}{\TUnknown}}
+  }{
+    \ctxAnaFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV}
+  }{IAProjR}
+\end{mathpar}
+%
+accompanied by corresponding typing rules for marked expressions:
+%
+\begin{mathpar}
+  \judgment{
+    \ctxSynType{\ctx}{\ECMV_1}{\TMV_1} \\
+    \ctxSynType{\ctx}{\ECMV_2}{\TMV_2}
+  }{
+    \ctxSynType{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TPair{\TMV_1}{\TMV_2}}
+  }{MSPair}
+
+  \judgment{
+    \ctxSynType{\ctx}{\ECMV}{\TMV} \\
+    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynType{\ctx}{\ECProjL{\ECMV}}{\TMV_1}
+  }{MSProjL1}
+
+  \judgment{
+    \ctxSynType{\ctx}{\ECMV}{\TMV} \\
+    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynType{\ctx}{\ECProjR{\ECMV}}{\TMV_2}
+  }{MSProjR1}
+
+  \judgment{
+    \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
+    \ctxAnaType{\ctx}{\ECMV_1}{\TMV_1} \\
+    \ctxAnaType{\ctx}{\ECMV_2}{\TMV_2}
+  }{
+    \ctxAnaType{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
+  }{MAPair1}
+
+  \judgment{
+    \ctxAnaType{\ctx}{\ECMV}{\TPair{\TMV}{\TUnknown}}
+  }{
+    \ctxAnaType{\ctx}{\ECProjL{\ECMV}}{\TMV}
+  }{MAProjL}
+
+  \judgment{
+    \ctxAnaType{\ctx}{\ECMV}{\TPair{\TUnknown}{\TMV}}
+  }{
+    \ctxAnaType{\ctx}{\ECProjR{\ECMV}}{\TMV}
+  }{MAProjR}
+\end{mathpar}
+
+However, as with lambda abstractions and application seen previously, two sets of rules are further
+necessary to satisfy totality. Firstly, the premise in \textsc{UAPair} that $\TMV$ is a matched
+binary product may fail, requiring marking and typing rules to handle that case:
+%
+\begin{mathpar}
+  \judgment{
+    \notMatchedPair{\TMV} \\
+    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TUnknown} \\
+    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
+  }{
+    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECInconType{\ECPair{\ECMV_1}{\ECMV_2}}}{\TMV}
+  }{IAPair2}
+
+  \judgment{
+    \notMatchedPair{\TMV} \\
+    \ctxAnaType{\ctx}{\ECMV_1}{\TUnknown} \\
+    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
+  }{
+    \ctxAnaType{\ctx}{\ECInconType{\ECPair{\ECMV_1}{\ECMV_2}}}{\TMV}
+  }{MAPair2}
+\end{mathpar}
+
+The other kind of error may arise when the subject of a projection does not synthesize a matched
+product type. As in our development of marking for applications, we may place the subject in an
+inconsistent type mark, or we may treat it as a special ``constrained synthesis'' mode, utilising a
+new error mark $\ECInconMatchedPair{\ECMV}$. The latter approach leads to the following rules:
+%
+\begin{mathpar}
+  \judgment{
+    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
+    \notMatchedPair{\TMV}
+  }{
+    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECInconMatchedPair{\ECMV}}}{\TMV_1}
+  }{ISProjL2}
+
+  \judgment{
+    \ctxSynType{\ctx}{\ECMV}{\TMV} \\
+    \notMatchedPair{\TMV}
+  }{
+    \ctxSynType{\ctx}{\ECProjL{\ECInconMatchedPair{\ECMV}}}{\TUnknown}
+  }{MSProjL2} \\
+
+  \judgment{
+    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
+    \notMatchedPair{\TMV}
+  }{
+    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECInconMatchedPair{\ECMV}}}{\TMV_2}
+  }{ISProjR2}
+
+  \judgment{
+    \ctxSynType{\ctx}{\ECMV}{\TMV} \\
+    \notMatchedPair{\TMV}
+  }{
+    \ctxSynType{\ctx}{\ECProjR{\ECInconMatchedPair{\ECMV}}}{\TUnknown}
+  }{MSProjR2}
+\end{mathpar}
+
+\input{let}
 

### 2023-02-28 calculus.tex
Brief section on hazelnut action calculus

@@
 entire expression in such a situation, while other tools may choose one branch to be ``correct'' and
-indicate a type mismatch error on the other.
+``blame'' the type mismatch error on the other.
 
@@
 which must provide users feedback on where and why a program is ill-typed. Moreover, they must
-continue providing editor services even when an error is encountered. Indeed, similarly to
-error-recovering parsers, real-world type checkers must recover from type errors. Traditionally,
-this has been performed in an \emph{ad hoc} manner [citations?], with different languages and tools
-adopting varying approaches in the localization, reportage, and recovery from errors.
+continue providing semantic services even when an error is encountered. Indeed, similarly to
+error-recovering parsers, real-world type checkers must robustly recover from type errors.
+Traditionally, this has been performed in an \emph{ad hoc} manner [citations?], with different
+languages and tools adopting varying approaches in the localization, reportage, and recovery of
+errors.
 
@@
 
-The key contribution of this section is a formal calculus, called the \emph{marking calculus}, for
-type error localization in bidirectional systems, a judgmental framework that provides a unified
-theory for the design of typing semantics that supports ill-typed programs.
-\Cref{sec:calculus-calculus} describes its form and metatheory by example on a minimal gradually
-typed lambda calculus. All rules and theorems may be found in the supplementary appendix, alongside
-a complete mechanization in the Agda proof assistant, which is discussed in
-\cref{sec:calculus-agda}. Finally, in \cref{sec:calculus-poly}, we consider how the system might be
-extended to more complex typing features, such as System F-style polymorphism.
+The key contribution of this section is the \emph{marking calculus}, a formal calculus for type
+error localization in bidirectional systems. Namely, the calculus provides a judgmental framework
+with a unified theory for the design of typing semantics that support ill-typed programs.
+\Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
+lambda calculus, extending it with binary product types in \cref{sec:calculus-products}. All rules
+and theorems may be found in the supplementary appendix, alongside a complete mechanization in the
+Agda proof assistant, which is discussed briefly in \cref{sec:calculus-agda}. With products, we then
+consider how error localization is complicated by patterned let expressions in
+\cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how the system might be extended
+to more complex typing features, such as System F-style polymorphism.
 
@@
 
-\emph{Bidirectional typing} combines type checking, which determines if a program satisfies a given
-type, and type synthesis, which generates a type from the program. This gives a simple algorithmic
-system in which type information is recursively propagated through terms and allows for local type
-inference.
+% \emph{Bidirectional typing} combines type checking, which determines if a program satisfies a given
+% type, and type synthesis, which generates a type from the program. This gives a simple algorithmic
+% system in which type information is recursively propagated through terms and allows for local type
+% inference.
 
@@
 $\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure. In addition to these forms,
-$\TUnknown$ is the unknown type from gradual type theory.
-
-% TODO include holes
+$\TUnknown$ is the unknown type from gradual type theory. $\EEHole$ denotes an \emph{expression
+hole}, which are used to model syntactically incomplete programs---but are not critical within the
+calculus.
+
 \begin{figure}[htbp]
@@
                        \mid           \ENumMV \mid \EPlus{\EMV}{\EMV}
-                       \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV} \\
+                       \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
+                       \mid           \EEHole \\
     \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV}
                        \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
-                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV} \\
+                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV}
+                       \mid           \EEHole \\
              &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
   \end{array}\]
-  \caption{Syntax of the extended gradually typed lambda calculus}
+  \caption{Syntax of an extended gradually typed lambda calculus.}
   \label{fig:calculus-syntax}
@@
 
-\section{Marking Calculus in Structured Editing}
+\section{The Marking Calculus in Structured Editing}
 \label{sec:calculus-structured-editing}
@@
 \subsection{Implementation in Hazel}
+\label{sec:calculus-hazel}
 
@@
 \label{sec:calculus-hazelnut}
+
+The Hazelnut action calculus introduced by \citet{HazelnutPOPL} give a structure editing scheme that
+ensures programs are statically meaningful.
+% TODO say something more about Hazelnut, especially in relation to error holes
+
+Hazelnut, however, is unable to cope with actions that require non-local hole fixes. % TODO example of this
+Such actions remain undefined, i.e. unperformable.
+
+By applying the marking calculus to Hazelnut, the action calculus may be \emph{wholly decoupled}
+from the typing semantics. Under this new system, the action calculus is concerned only with the
+manipulation of syntax, and the marking procedure yields statically meaningful terms with error
+holes inserted at the appropriate positions. Furthermore, marking solves the problem of non-local
+hole fixes; actions may be performed anywhere (as long as they are of the correct sort), and marking
+totality ensures that they give meaningful terms.
+% TODO this is not strictly true, only for 'wrapping' actions

### 2023-02-28 calculus.tex
Use American spelling for synthesize and analyze

@@
 judgments. \emph{Type synthesis}, written $\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under
-the typing context $\ctx$, the expression $e$ synthesises or infers the type $\TMV$. \emph{Type
+the typing context $\ctx$, the expression $e$ synthesizes or infers the type $\TMV$. \emph{Type
 analysis}, written $\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear
@@
 $\TMV$. Analogously, the analytic marking judgment $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$
-states that $\EMV$ is marked into $\ECMV$, which analyses against $\TMV$.
+states that $\EMV$ is marked into $\ECMV$, which analyzes against $\TMV$.
 
@@
 from the marked expression gives the original unmarked expression. We also ensure that marking gives
-well-typed marked expressions; indeed, their types should match the types synthesised or analysed
+well-typed marked expressions; indeed, their types should match the types synthesized or analyzed
 against in marking. % TODO more intuition for this theorem
@@
 In addition expressions, the type of both operands should be $\TNum$. In a bidirectional system, we
-thus analyse the operands against $\TNum$, giving the following rule:
+thus analyze the operands against $\TNum$, giving the following rule:
 %
@@
 Though no expected output type is known, we still need to check and mark the body; it is thus
-analysed against the unknown type in \textsc{IALam2}.
+analyzed against the unknown type in \textsc{IALam2}.
 
@@
 Finally, we consider the subsumption rule in typing unmarked expressions, which states that if an
-expression synthesizes a type, it may also be analysed against that type or any consistent type.
+expression synthesizes a type, it may also be analyzed against that type or any consistent type.
 \[%
@@
 However, it may be the case that the expression is inconsistent with the expected type. Recalling
-the example of \cref{fig:calculus-examples-inconsistent-types}, subsumption would used to analyse
+the example of \cref{fig:calculus-examples-inconsistent-types}, subsumption would used to analyze
 $\ETrue$ against $\TNum$ in $\EPlus{\ETrue}{5}$, which fails since $\inconsistent{\TNum}{\TBool}$,

### 2023-02-28 calculus.tex
Prefix marking rules with MK

@@
     \ctxSynFixedInto{\ctx}{\ENumMV}{\ECNumMV}{\TNum}
-  }{ISNum}
+  }{MKSNum}
 
@@
     \ctxSynFixedInto{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
-  }{ISPlus}
+  }{MKSPlus}
 
@@
     \ctxSynFixedInto{\ctx}{x}{x}{\TMV}
-  }{ISVar}
+  }{MKSVar}
 
@@
     \ctxSynFixedInto{\ctx}{x}{\ECUnbound{x}}{\TUnknown}
-  }{ISUnbound}
+  }{MKSUnbound}
 
@@
 %
-\textsc{ISUnbound} states that an variable unbound in the context is marked into that variable
+\textsc{MKSUnbound} states that an variable unbound in the context is marked into that variable
 inside an unbound variable mark. \textsc{MSUnbound} then gives the typing of unbound variable marks.
@@
     \ctxSynFixedInto{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\ELam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
-  }{ISLam}
+  }{MKSLam}
 
@@
     \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
-  }{IALam1}
+  }{MKALam1}
 
@@
     \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
-  }{IALam2}
+  }{MKALam2}
 
@@
 Though no expected output type is known, we still need to check and mark the body; it is thus
-analyzed against the unknown type in \textsc{IALam2}.
+analyzed against the unknown type in \textsc{MKALam2}.
 
@@
     \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
-  }{IALam3}
+  }{MKALam3}
 
@@
     \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
-  }{ISAp1}
+  }{MKSAp1}
 
@@
     \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApNonMatched{\ECMV_1}{\ECMV_2}}{\TUnknown}
-  }{ISAp2}
+  }{MKSAp2}
 
@@
     \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECApNonMatchedAltAlt{\ECMV_1}}{\ECMV_2}}{\TUnknown}
-  }{ISAp2'}
+  }{MKSAp2'}
 
@@
     \ctxSynFixedInto{\ctx}{\ETrue}{\ECTrue}{\TBool}
-  }{ISTrue}
+  }{MKSTrue}
 
@@
     \ctxSynFixedInto{\ctx}{\EFalse}{\ECFalse}{\TBool}
-  }{ISFalse}
+  }{MKSFalse}
 
@@
     \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
-  }{ISIf}
+  }{MKSIf}
 
@@
     \ctxAnaFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
-  }{IAIf}
+  }{MKAIf}
 
@@
     \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TUnknown}
-  }{ISInconsistentBranches}
+  }{MKSInconsistentBranches}
 
@@
     \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}
-  }{IASubsume}
+  }{MKASubsume}
 
@@
     \ctxAnaFixedInto{\ctx}{\EMV}{\ECInconType{\ECMV}}{\TMV}
-  }{IAInconsistentTypes}
+  }{MKAInconsistentTypes}
 
@@
     \ctxSynFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TPair{\TMV_1}{\TMV_2}}
-  }{ISPair}
+  }{MKSPair}
 
@@
     \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV_1}
-  }{ISProjL1}
+  }{MKSProjL1}
 
@@
     \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV_2}
-  }{ISProjR1} \\
+  }{MKSProjR1} \\
 
@@
     \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
-  }{IAPair1}
+  }{MKAPair1}
 
@@
     \ctxAnaFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV}
-  }{IAProjL}
+  }{MKAProjL}
 
@@
     \ctxAnaFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV}
-  }{IAProjR}
+  }{MKAProjR}
 \end{mathpar}
@@
     \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECInconType{\ECPair{\ECMV_1}{\ECMV_2}}}{\TMV}
-  }{IAPair2}
+  }{MKAPair2}
 
@@
     \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECInconMatchedPair{\ECMV}}}{\TMV_1}
-  }{ISProjL2}
+  }{MKSProjL2}
 
@@
     \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECInconMatchedPair{\ECMV}}}{\TMV_2}
-  }{ISProjR2}
+  }{MKSProjR2}
 

### 2023-02-28 appendixC.tex

\newpage
\section{Appendix C: An extended syntax of PossibleTypeSets}

\begin{figure}[h!]
\centering
\hrule
$\arraycolsep=4pt\begin{array}{lll}
BinaryOperator~~ \square & ::=~~~~~
\rightarrow  ~\vert~ 
\times
\\
PotentialTypeSet~~ s & ::=
single(t) ~\vert~
cons(t, s)
\\
PotentialType~~ t & ::= 
  \tnum ~\vert~
  \tbool ~\vert~
  \TVar ~\vert~
  \tehole^p ~\vert~
  s ~\square~ s
  \\
\end{array}$
\label{fig:syntax_possible_type_sets}
\caption{Extended syntax of PotentialTypeSets and PotentialTypes}
\vspace{5px}
\hrule
\[\begin{array}{rcl}
    single(s_1 ~\square_{\alpha}~ s_2) ~\amalg~ single(s_3 ~\square_{\alpha}~ s_4) & = & single((s_1 ~\amalg~ s_3) ~\square_{\alpha}~ (s_2 ~\amalg~ s_4)) \\
    single(t) ~\amalg~ single(t') & = & cons(t, single(t')) \\
    cons(t,s) ~\amalg~ single(t) & = & cons(t, s) \\
    cons(s_1 ~\square_{\alpha}~ s_2, s) ~\amalg~ single(s_3 ~\square_{\alpha}~ s_4) & = & cons((s_1 ~\amalg~ s_3) ~\square_{\alpha}~ (s_2 ~\amalg~ s_4), s) \\
    cons(t,s) ~\amalg~ cons(t',s') & = & cons(t,s) ~\amalg~ single(t') ~\amalg~ s' \\
\end{array}\] 
\caption{Extended rules for merging PotentialTypeSets}
\vspace{5px} 
\hrule
\label{fig:extended_merging_possible_type_sets}
\vspace{-5px}
\end{figure}

Here, we take a more general approach to representing binary types. Namely, a \emph{PotentialType} may be the combination of two \emph{PotentialType}s through any \emph{BinaryOperator} $\square$. The rules for extension follow the same logic as before, where we enforce that extension of a \emph{PotentialTypeSet} containing $s_1 ~\square_{\alpha}~ s_2$ with $s_3 ~\square_{\alpha}~ s_4$ simply merges the left and right \emph{PotentialTypeSet}s, thus yielding $(s_1 ~\amalg~ s_3) ~\square_{\alpha}~ (s_2 ~\amalg~ s_4)$. Additionally, we choose to represent type variables as a ground type, thus enforcing that type variables are inconsistent with other ground types and consistent with type holes and each other.

### 2023-02-28 let.tex

@@
 
+In a traditional bidirectional presentation (CITE), let expressions without composite patterns usually have two rules: A type-annotated version which places the definition in analytic position, and an unannotated synthetic version.
+With composite patterns, this straightforward division is no longer possible, as whether or not there are type constraints on the definition depends on the (arbitrarily nested) sub-components of the pattern.
+The most obvious approach is to instead use a single rule which synthesizes a type from the pattern, then analyzes the definition against that type.
+However, this runs afoul of user expectation.
+Consider the expression $\ELet{(a,b)}{(1,2)}{...}$.
+The semantics of pattern matching are often (CITE) given by expansion to nested lets under appropriate projections, beta-reduced as far as possible.
+Here this would give $\ELet{a}{1}{\ELet{b}{2}{...}}$.
+Notice that in the un-expanded expression, $1$ and $2$ are in analytic against the unknown type, whereas in the latter, they are in synthetic position.
+At this point one should ask: Is there a meaningful difference between these two situations?
+There is at least a subtle semantic distinction: Synthetic means 'no type constraints', whereas analytic against unknown imposes a trivial type constraint.
+This distinction behaviorally manifests in the case of expressions which can have internal type inconsistencies.
+As per the previous section on booleans, in the expression $\ELet{a}{\EIf{\ETrue}{1}{\EFalse}}{...}$, the conditional will be marked with an inconsistent branches error if in synthetic position, but if analyzed against unknown---with which both branches are consistent---no mark will be placed.
+This raises the question: Do we need this distinction?
+Perhaps we can simply erase it, for example adding a rule to switch to synthetic mode whenever we analyze against hole, or vice-versa.
+This is similar to the path we take, but simply collapsing the distinction removes an important expressive dimension of user intent in gradual typing.
+To better illustrate, we additionally add a compositional pattern annotation form, which can be applied to sub-patterns: $\ELet{(a, b:\TUnknown)}{(\EIf{\ETrue}{1}{\EFalse},\EIf{\ETrue}{1}{\EFalse})}{...}$.
+Here, we annotate the sub-pattern $b$ with an unknown type annotation.
+Here, the user is expressing their intent that the corresponding sub-definition should have some definite, as-yet unspecified, type.
+It is reasonable that the conditional should no longer be marked, as this action passes the buck, attribution-wise, to the annotation.
+However, it is also reasonable that the annotated unknown type itself should be considered in an error state, since there are no possible fillings for this hole.
+Indeed, precisely this will be accomplished in section 4.
+
 We augment our type system with a new variant on the unknown type which can trigger a swtich from analysis to synthesis.
-The new type $\TUnknownSwitch$ behaves idenitically to $\TUnknown$ with respect to type consistency, the marched arrow and product judgements, and the join metafunction:
+The new type $\TUnknownSwitch$ behaves identically to $\TUnknown$ with respect to type consistency, the marched arrow and product judgements, and the join metafunction:
 %
@@
 %
-Let expression have the following semantics:
+Let expressions are typed according to the following $\mathtt{USLet}$ and $\mathtt{UALet}$ rules.
+Additionally, we add a rule governing the switch from analysis to synthesis, triggered by analysis against the new Unknown-SynSwitch type:
 \begin{mathpar}
@@
      \\
-     \ctxSynType{\ctx_{b}}{\EMV_{b}}{\TMV}
+     \ctxSynOrAnaType{\ctx_{b}}{\EMV_{b}}{\TMV}
    }{
-     \ctxSynType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
-   }{USLet}
-   \\
-   \judgment{
-     \ctxSynPat{\ctx}{\PMV}{\TMV_p}
-     \\
-     \ctxAnaType{\ctx}{\EMV_{def}}{\TMV_p}
-     \\
-     \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
-     \\\\
-     \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
-     \\
-     \ctxAnaType{\ctx_{b}}{\EMV_{b}}{\TMV}
-   }{
-     \ctxAnaType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
-   }{UALet}
-\end{mathpar}
-Additionally, we add a rule governing the switch from analysis to synthesis, triggered by analysis against the new SynSwitch type:
-\begin{mathpar}
+     \ctxSynOrAnaType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
+   }{U(S/A)Let}
+
    \judgment{
@@
    }{SynSwitch}
-
+    %\judgment{
+    % \ctxSynPat{\ctx}{\PMV}{\TMV_p}
+    % \\
+    % \ctxAnaType{\ctx}{\EMV_{def}}{\TMV_p}
+    % \\
+    % \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
+    % \\\\
+    % \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
+    % \\
+    % \ctxSynType{\ctx_{b}}{\EMV_{b}}{\TMV}
+   %}{
+   %  \ctxSynType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
+   %}{USLet}
+   %\\
+   %\judgment{
+   %  \ctxSynPat{\ctx}{\PMV}{\TMV_p}
+   %  \\
+   %  \ctxAnaType{\ctx}{\EMV_{def}}{\TMV_p}
+   %  \\
+   %  \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
+   %  \\\\
+   %  \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
+   %  \\
+   %  \ctxAnaType{\ctx_{b}}{\EMV_{b}}{\TMV}
+   %}{
+   %  \ctxAnaType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
+   %}{UALet}
 \end{mathpar}
+Finally, pattern synthesis and analysis are governed by the following rules.
+Note how variables and wildcards synthesize the Unknown-SynSwitch type, but note also that this type is never added to the context:
 
@@
 
-   \\\\
    \judgment{
@@
    }{USPatPair}
-
-   \judgment{
-     \ctxSynPat{\ctx}{\PMV{}}{\_}
-   }{
-        \ctxSynPat{\ctx}{\ECMarkedInconType{\PMV}{}}{\TUnknownSwitch}
-    }{USPatErr}
-
   \end{mathpar}
   %
-  \caption{Pattern Synthesis}
   \label{fig:calculus-pattern-synthesis}
@@
   %
-  \caption{Pattern Analysis}
   \label{fig:calculus-pattern-analysis}

### 2023-03-01 intro.tex
Finish constraint generation and updates to the type system section (Raef)
Most marked rules for lets and patterns done --andrew

@@
 Modern programming environments provide developers with 
-various semantic services, such as type hints, semantic navigation, semantic code completion, and automated refactoring, that require statically reasoning about the type and binding structure of the program as it is being edited.
+various semantic services---e.g.
+type hints, semantic navigation, semantic code completion, and automated refactoring---that require statically reasoning about the type and binding structure of a program as it is being edited.
 % These semantic services are often packaged into language servers\todo{cites}.
-The problem is that the program being edited can be ill-typed, sometimes for long periods of time. 
-For example, when the developer changes an important type definition, errors might arise at dozens of usage sites in a large program, which could take hours or days to address.
-Semantic services often exhibit gaps in service during these periods \cite{HazelnutSNAPL}, because in the conventional formal account of type systems, only well-typed programs have a well-defined type and binding structure.
+The problem is that the program being edited can be ill-typed, sometimes for a long period of time. 
+For example, after changing a type definition, errors might arise at dozens of usage sites in a large program, which can take hours or days to resolve.
+Semantic services often exhibit gaps in service during these periods \cite{HazelnutSNAPL}, because type systems defined in the conventional way specify the type and binding structure only for well-typed programs.
 As soon as a type error occurs \emph{anywhere}, the program is formally meaningless \emph{everywhere}.
 
-Developers benefit when these gaps in service are minimized or eliminated, so there has long been considerable practical interest in (1)~\textbf{type error localization}: mechanisms for directing the developer to
-the location(s) in a program that may be erroneous, and (2) \textbf{type error recovery}: 
-mechanisms that allow the language server to optimistically recover from localized errors and continue to provide downstream semantic services, ideally throughout the program.
+Developers benefit when these gaps in service are minimized or eliminated, so there has long been considerable practical interest in (1)~\textbf{type error localization}: mechanisms for identifying the location(s) in a program at fault for a type error, and (2) \textbf{type error recovery}:
+mechanisms that allow the language server to optimistically recover from a localized error and continue on to locate other errors and provide downstream semantic services, ideally throughout the program and with minimal degradation in quality.
 
@@
 In this example, such a system might report that the variable \li{f} is unbound and communicate its location, e.g.
-using editor decorations.
+using a decorative mark in the editor.
 This is the approach taken in many production type checkers, e.g. XXX.\todo{examples}

### 2023-03-01 let.tex
Most marked rules for lets and patterns done --andrew
THI: Constraint solving section mostly updated; graph examples included

--- 2023-02-28 let.tex
@@
-\subsection{Let with Composite Patterns}
+\subsection{Let-Expressions with Composite Destructuring Patterns}
 \label{sec:calculus-let}
@@
 
-We augment our type system with a new variant on the unknown type which can trigger a swtich from analysis to synthesis.
-The new type $\TUnknownSwitch$ behaves idenitically to $\TUnknown$ with respect to type consistency, the marched arrow and product judgements, and the join metafunction:
+In a traditional bidirectional presentation (CITE 'Strict Bidirectional Type Checking' Chlipala et al), let-expressions without composite patterns have two rules: A type-annotated version placing the definition in analytic position, and an unannotated synthetic version.
+With composite patterns, this straightforward division is no longer possible, as type constraints on the definition may depend on the (arbitrarily nested) sub-components of the pattern.
 %
+\[\begin{array}{rrcl}
+  \EMName & \EMV & \Coloneqq & \cdots
+                               \mid \ELet{\PMV}{\EMV}{\EMV} \\
+  \PMName & \PMV & \Coloneqq & x 
+                               \mid \_
+                               \mid (p_1, p_2)
+                               \mid p : \TMV \\
+  \PCMName & \PCMV & \Coloneqq & x 
+                               \mid \_
+                               \mid (p_1, p_2)
+                               \mid p : \TMV
+                               \mid \ECInconType{\PCMV}
+                               \mid \ECInconMatchedPair{\PCMV}
+\end{array}\]
+%
+The most obvious approach is to instead use a single rule which synthesizes a type from the let-expression's pattern, then analyzes the definition against that type.
+However, this may run afoul of user expectation.
+Consider the expression $\ELet{(a,b)}{(1,2)}{e}$, which a user would reasonably expect to be semantically equivalent to the expanded expression $\ELet{a}{1}{\ELet{b}{2}{e}}$.
+Notice that in the un-expanded expression, $1$ and $2$ are in analytic against the unknown type, whereas in the latter, they are in synthetic position.
+
+One may ask if there a meaningful difference between these two situations.
+There is at least a subtle semantic distinction: Synthetic means 'no type constraints', whereas analysis against unknown imposes a trivial type constraint.
+
+This distinction manifests behaviorally in the case of expressions that can have internal type inconsistencies.
+As per the previous section on booleans, given the expression $\ELet{a}{\EIf{\ETrue}{1}{\EFalse}}{...}$, the conditional will be marked with an inconsistent branches error when in synthetic position, but if analyzed against the unknown type---with which both branches are consistent---no mark will be placed.
+
+One option would be to erase the distinction entirely, adding a rule to switch to synthetic mode whenever we analyze against unknown, or to simply eliminate synthetic mode altogether.
+We consider these options too drastic; we wish to preserve the above behavioral distinction for conditionals, both indicating an inconsistency error in the unconstrained case, and permitting the user to defer that error via annotation by the unknown type.
+To this effect, we add a compositional pattern annotation form, which can be applied to sub-patterns:
+$\ELet{(a, b:\TUnknown)}{(\EIf{\ETrue}{1}{\EFalse},\EIf{\ETrue}{1}{\EFalse})}{e}$. Here we want the first condition to be in synthetic and hence marked, and the second to be in analytic and unmarked.
 \[\begin{array}{rrcl}
@@
 \end{array}\]
+To accomplish this, we use a variation on the first distinction-erasure option, but with a twist.
+We do not unconditionally switch from analysis against unknown to synthesis, but instead augment our type system with a new variant on the unknown type which triggers the switch.
+The new type $\TUnknownSwitch$ behaves identically to $\TUnknown$ with respect to type consistency, the matched arrow and product judgements, and the join metafunction.
+\begin{mathpar}
+   \judgment{
+     \ctxSynType{\ctx}{\EMV}{\_}
+   }{
+     \ctxAnaType{\ctx}{\EMV}{\TUnknownSwitch}
+   }{SynSwitch}
+\end{mathpar}
 
-The unmarked language is extended by let expressions with composite patterns:
-%
-\[\begin{array}{rrcl}
-  \EMName  & \EMV  & \Coloneqq & \cdots
-                               \mid \ELet{\PMV}{\EMV}{\EMV}
-\end{array}\]
-\[\begin{array}{rrcl}
-  \PMName  & \PMV  & \Coloneqq x 
-                               \mid \_
-                               \mid (p_1,p_2)
-                               \mid p : \TMV
+However, it is also reasonable that the annotated unknown type itself should be considered in an error state, since there are no possible fillings for this hole.
+Indeed, precisely this will be accomplished in section 4.
 
-\end{array}\]
-%
-Let expression have the following semantics:
+Let expressions are typed according to the following $\mathtt{USLet}$ and $\mathtt{UALet}$ rules.
+Additionally, we add a rule governing the switch from analysis to synthesis, triggered by analysis against the new Unknown-SynSwitch type:
 \begin{mathpar}
@@
      \\
-     \ctxSynType{\ctx_{b}}{\EMV_{b}}{\TMV}
+     \ctxSynOrAnaType{\ctx_{b}}{\EMV_{b}}{\TMV}
    }{
-     \ctxSynType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
-   }{USLet}
-   \\
+     \ctxSynOrAnaType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
+   }{U(S/A)Let}
+
    \judgment{
-     \ctxSynPat{\ctx}{\PMV}{\TMV_p}
-     \\
-     \ctxAnaType{\ctx}{\EMV_{def}}{\TMV_p}
-     \\
-     \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
-     \\\\
-     \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
-     \\
-     \ctxAnaType{\ctx_{b}}{\EMV_{b}}{\TMV}
-   }{
-     \ctxAnaType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
-   }{UALet}
-\end{mathpar}
-Additionally, we add a rule governing the switch from analysis to synthesis, triggered by analysis against the new SynSwitch type:
-\begin{mathpar}
-    \judgment{
-     \ctxSynType{\ctx}{\EMV}{_}
+     \ctxSynType{\ctx}{\EMV}{\_}
    }{
@@
   \begin{mathpar}
-
    \judgment{ }{\ctxSynPat{\ctx}{\_}{\TUnknownSwitch}}{USPatWild}
@@
 
-   \\\\
    \judgment{
-      \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
-      \ctxAnaPat{\ctx}{\PMV_1}{\TMV_1}{\ctx_1} \\
+      \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\\\
+      \ctxAnaPat{\ctx}{\PMV_1}{\TMV_1}{\ctx_1} \\\\
       \ctxAnaPat{\ctx_1}{\PMV_2}{\TMV_2}{\ctx_2}
@@
     }{UAPatPair}
-
-   \judgment{
-     \ctxSynPat{\ctx}{\PMV{}}{\_}
-   }{
-        \ctxAnaPat{\ctx}{\ECMarkedInconType{\PMV}{}}{\tau}{\ctx}
-    }{UAPatErr}
   \end{mathpar}
   %
-  \caption{Pattern Analysis}
   \label{fig:calculus-pattern-analysis}
@@
 
-%TODO:andrew
+\begin{figure}[htbp]
+  \raggedright
+  \judgbox{\ensuremath{\ctxSynFixedInto{\ctx_{in}}{\PMV}{\PCMV}{\TMV}}} Pattern $\PMV$ synthesizes $\TMV$ producing marked $\PCMV$ \\
+  \judgbox{\ensuremath{\ctxAnaFixedIntoPat{\ctx_{in}}{\PMV}{\PCMV}{\TMV}{\ctx_{out}}}} Pattern $\PMV$ analyzes against $\TMV$ producing marked $\PCMV$ and context $\Gamma_{out}$
+   %
+  \begin{mathpar}
+   \judgment{ }{\ctxSynFixedInto{\ctx}{\_}{\_}{\TUnknownSwitch}}{ISPatWild}
+
+   \judgment{ }{\ctxSynFixedInto{\ctx}{x}{x}{\TUnknownSwitch}}{ISPatVar}
+
+   \judgment{
+     \ctxSynFixedInto{\ctx}{\PMV_1}{\PCMV_1}{\TMV_1} \\\\
+     \ctxSynFixedInto{\ctx}{\PMV_2}{\PCMV_2}{\TMV_2}
+   }{
+      \ctxSynFixedInto{\ctx}{\EPair{\PMV_1}{\PMV_2}}{\ECPair{\PCMV_1}{\PCMV_2}}{\TPair{\TMV_1}{\TMV_2}}
+   }{ISPatPair}
+
+   \judgment{
+     \ctxAnaPat{\ctx}{\PMV{}}{\TMV}{\_} \\
+     \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\TMV}{\_}
+   }{
+        \ctxSynFixedInto{\ctx}{\PMV : \tau}{\PCMV : \tau}{\tau}
+    }{ISPatAnn}
+
+   \judgment{
+     \ctxNotAnaPat{\ctx}{\PMV{}}{\TMV}{\_} \\
+     \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\TUnknown}{\_}
+   }{
+        \ctxSynFixedInto{\ctx}{\PMV : \TMV}{\ECInconType{\PCMV} : \TMV}{\TMV}
+    }{ISPatAnnFail}
+   \\\\
+   \judgment{ }{\ctxAnaFixedIntoPat{\ctx}{\_}{\_}{\TMV}{\ctx}}{IAPatWild}
+
+   \judgment{ }{\ctxAnaFixedIntoPat{\ctx}{x}{x}{\TMV}{\ctx, x:\TMV}}{IAPatVar}
+   \\\\
+   \judgment{
+     \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\\\
+     \ctxAnaFixedIntoPat{\ctx}{\PMV_1}{\PCMV_1}{\TMV_1}{\ctx_1} \\\\
+     \ctxAnaFixedIntoPat{\ctx_1}{\PMV_2}{\PCMV_2}{\TMV_2}{\ctx_2}
+   }{
+      \ctxAnaFixedIntoPat{\ctx}{\EPair{\PMV_1}{\PMV_2}}{\ECPair{\PCMV_1}{\PCMV_2}}{\TMV}{\ctx_2}
+   }{IAPatPair}
+
+   \judgment{
+     \notMatchedPair{\TMV} \\\\
+     \ctxAnaFixedIntoPat{\ctx}{\PMV_1}{\PCMV_1}{\TUnknown}{\ctx_1} \\\\
+     \ctxAnaFixedIntoPat{\ctx_1}{\PMV_2}{\PCMV_2}{\TUnknown}{\ctx_2}
+   }{
+      \ctxAnaFixedIntoPat{\ctx}{\EPair{\PMV_1}{\PMV_2}}{\ECInconMatchedPair{\ECPair{\PCMV_1}{\PCMV_2}}}{\TMV}{\ctx_2}
+   }{IAPatPairFail}
+
+   \judgment{
+     \ctxAnaPat{\ctx}{\PMV{}}{\tau'}{\ctx'} \\
+     \consistent{\tau}{\tau'} \\
+     \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\tau'}{\_}
+   }{
+        \ctxAnaFixedIntoPat{\ctx}{\PMV : \tau'}{\PCMV : \tau'}{\tau}{\ctx'}
+   }{IAPatAnn}
+
+   \judgment{
+     \ctxNotAnaPat{\ctx}{\PMV{}}{\tau}{\ctx'} \\
+     \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\tau}{\_}
+   }{
+        \ctxAnaFixedIntoPat{\ctx}{\PMV : \tau}{\ECInconType{\PCMV} : \tau}{\tau}{\ctx'}
+   }{IAPatAnnFail1}
+
+   \judgment{
+     \ctxAnaPat{\ctx}{\PMV{}}{\tau}{\ctx'} \\
+     \inconsistent{\tau}{\tau'} \\
+     \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\tau}{\_}
+   }{
+        \ctxAnaFixedIntoPat{\ctx}{\PMV : \tau'}{\ECInconType{\PCMV : \tau'}}{\tau}{\ctx'}
+   }{IAPatAnnFail2}
+
+   \judgment{
+     \ctxNotAnaPat{\ctx}{\PMV{}}{\tau}{\ctx'} \\
+     \inconsistent{\tau}{\tau'} \\
+     \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\tau}{\_}
+   }{
+        \ctxAnaFixedIntoPat{\ctx}{\PMV : \tau'}{\ECInconType{\ECInconType{\PCMV} : \tau'}}{\tau}{\ctx'}
+   }{IAPatAnnFail3}
+
+  \end{mathpar}
+  %
+  \caption{Pattern Marking}
+  \label{fig:calculus-pattern-synthesis-marking}
+\end{figure}
+
+\begin{figure}[htbp]
+  \begin{mathpar}
+
+   \judgment{ }{\ctxSynPat{\ctx}{\_}{\TUnknownSwitch}}{MSPatWild}
+
+   \judgment{ }{\ctxSynPat{\ctx}{x}{\TUnknownSwitch}}{MSPatVar}
+
+   \judgment{
+     \ctxAnaPat{\ctx}{\PCMV{}}{\tau}{\_}
+   }{
+        \ctxSynPat{\ctx}{\PCMV : \tau}{}{\tau}
+    }{MSPatAnn}
+
+   \judgment{
+     \ctxSynPat{\ctx}{\PCMV_1}{\TMV_1} \\
+     \ctxSynPat{\ctx}{\PCMV_2}{\TMV_2}
+   }{
+     \ctxSynPat{\ctx}{\EPair{\PCMV_1}{\PCMV_2}}{\TPair{\TMV_1}{\TMV_2}}
+   }{MSPatPair}
+  \end{mathpar}
+  %
+  \caption{Marked Pattern Synthesis}
+  \label{fig:calculus-pattern-synthesis-marked}
+\end{figure}
+
+\begin{figure}[htbp]
+  \raggedright
+  \begin{mathpar}
+   \judgment{ }{\ctxAnaPat{\ctx}{\_}{\TMV}{\ctx}}{MAPatWild}
+
+   \judgment{ }{\ctxAnaPat{\Gamma}{x}{\TMV}{\ctx, x:\TMV}}{MAPatVar}
+
+   \judgment{
+     \ctxAnaPat{\ctx}{\PCMV{}}{\tau'}{\ctx'}
+     \\
+     \consistent{\tau}{\tau'}
+   }{
+        \ctxAnaPat{\ctx}{\PCMV : \tau'}{\tau}{\ctx'}
+    }{MAPatAnn}
+
+    \judgment{
+     \ctxAnaPat{\ctx}{\PCMV{}}{\tau'}{\ctx'}
+     \\
+     \inconsistent{\tau}{\tau'}
+   }{
+        \ctxAnaPat{\ctx}{\ECInconType{\PCMV}}{\tau}{\ctx'}
+    }{MAInconsistentTypes}
+
+   \judgment{
+      \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\\\
+      \ctxAnaPat{\ctx}{\PCMV_1}{\TMV_1}{\ctx_1} \\\\
+      \ctxAnaPat{\ctx_1}{\PCMV_2}{\TMV_2}{\ctx_2}
+    }{
+      \ctxAnaPat{\ctx}{\ECPair{\PCMV_1}{\PCMV_2}}{\TMV}{\ctx_2}
+    }{MAPatPair}
+  \end{mathpar}
+  %
+  \caption{Marked Pattern Analysis}
+  \label{fig:calculus-pattern-analysis-marked}
+\end{figure}

### 2023-03-01 let.tex
THI: updates to polymorphism section complete. pre first overarching edit pass

@@
   \EMName & \EMV & \Coloneqq & \cdots
+                               \mid \ELet{\PMV}{\EMV}{\EMV} \\
+  \ECMName & \EMV & \Coloneqq & \cdots
                                \mid \ELet{\PMV}{\EMV}{\EMV} \\
@@
 One option would be to erase the distinction entirely, adding a rule to switch to synthetic mode whenever we analyze against unknown, or to simply eliminate synthetic mode altogether.
-We consider these options too drastic; we wish to preserve the above behavioral distinction for conditionals, both indicating an inconsistency error in the unconstrained case, and permitting the user to defer that error via annotation by the unknown type.
+We consider these options too drastic; we wish to preserve the above behavioral distinction for conditionals, both indicating an inconsistency error in the unconstrained synthetic case, and permitting the user to defer that error via annotation by the unknown type.
 To this effect, we add a compositional pattern annotation form, which can be applied to sub-patterns:
-$\ELet{(a, b:\TUnknown)}{(\EIf{\ETrue}{1}{\EFalse},\EIf{\ETrue}{1}{\EFalse})}{e}$. Here we want the first condition to be in synthetic and hence marked, and the second to be in analytic and unmarked.
+$\ELet{(a, b:\TUnknown)}{(\ECInconBr{\ETrue}{1}{\EFalse},\EIf{\ETrue}{2}{\EFalse})}{e}$.
+Here we want the first condition to be in synthetic and hence marked as inconsistent, and the second to be in analytic against unknown and thus unmarked.
 \[\begin{array}{rrcl}
@@
 \end{array}\]
+
 To accomplish this, we use a variation on the first distinction-erasure option, but with a twist.
@@
 The new type $\TUnknownSwitch$ behaves identically to $\TUnknown$ with respect to type consistency, the matched arrow and product judgements, and the join metafunction.
+In our previous example, the pattern variable $a$ will synthesize the $\TUnknownSwitch$ type, whereas the annotation $b: \TUnknown$ synthesizes $\TUnknown$.
+Thus $1$ will end up in synthetic position and $2$ in analytic, achieving the desired distinction.
 \begin{mathpar}
@@
 
-However, it is also reasonable that the annotated unknown type itself should be considered in an error state, since there are no possible fillings for this hole.
-Indeed, precisely this will be accomplished in section 4.
+(It is worth noting that, despite the lack of marking for the $\TUnknown$-annotated conditional, there is no possible filling of the type hole, so it is also reasonable that the annotated unknown type itself should be considered as an error state.
+Indeed, precisely this will be accomplished in section 4 via type hole inference.)
 
-Let expressions are typed according to the following $\mathtt{USLet}$ and $\mathtt{UALet}$ rules.
-Additionally, we add a rule governing the switch from analysis to synthesis, triggered by analysis against the new Unknown-SynSwitch type:
 \begin{mathpar}
@@
 
-   \judgment{
-     \ctxSynType{\ctx}{\EMV}{\_}
-   }{
-     \ctxAnaType{\ctx}{\EMV}{\TUnknownSwitch}
-   }{SynSwitch}
     %\judgment{
@@
 \end{mathpar}
-Finally, pattern synthesis and analysis are governed by the following rules.
-Note how variables and wildcards synthesize the Unknown-SynSwitch type, but note also that this type is never added to the context:
+In order to ensure propagation of type information between the pattern and definition, we require a bit more machinery than a typical let rule.
+First, we synthesize the pattern type and analyze the definition against that type.
+These two steps are needed to ensure that the pattern and definition types are consistent.
+Then we synthesize the definition's type, and analyze the pattern against that type.
+This analysis of the pattern is guaranteed to succeed; however is is required, as we want the context produced by the pattern $\ctx_b$ to take into account the definition's type.
+Finally, the let body is synthesized or analyzed (depending on whether the let itself is being synthesized or analyzed) within that context.
 
-\begin{figure}[htbp]
-  \raggedright
+The marking rules are directly analogous:
+\begin{mathpar}
+   \judgment{
+     \ctxSynFixedInto{\ctx}{\PMV}{\PCMV}{\TMV_p}
+     \\
+     \ctxAnaFixedInto{\ctx}{\EMV_{def}}{\ECMV_{def}}{\TMV_{p}}
+     \\\\
+     \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
+     \\
+     \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
+     \\
+     \ctxSynFixedInto{\ctx_{b}}{\EMV_{b}}{\ECMV_{b}}{\TMV}
+   }{
+     \ctxSynFixedInto{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\ELet{\PCMV}{\ECMV_{def}}{\ECMV_{b}}}{\TMV}
+   }{ISLet}
+
+   \judgment{
+     \ctxSynFixedInto{\ctx}{\PMV}{\PCMV}{\TMV_p}
+     \\
+     \ctxAnaFixedInto{\ctx}{\EMV_{def}}{\ECMV_{def}}{\TMV_{p}}
+     \\\\
+     \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
+     \\
+     \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
+     \\
+     \ctxAnaFixedInto{\ctx_{b}}{\EMV_{b}}{\ECMV_{b}}{\TMV}
+   }{
+     \ctxAnaFixedInto{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\ELet{\PCMV}{\ECMV_{def}}{\ECMV_{b}}}{\TMV}
+   }{IALet}
+
+   \judgment{
+     \ctxSynPat{\ctx}{\PCMV}{\TMV_p}
+     \\
+     \ctxAnaType{\ctx}{\ECMV_{def}}{\TMV_p}
+     \\
+     \ctxSynType{\ctx}{\ECMV_{def}}{\TMV_{def}}
+     \\\\
+     \ctxAnaPat{\ctx}{\PCMV}{\TMV_{def}}{\ctx_{b}}
+     \\
+     \ctxSynOrAnaType{\ctx_{b}}{\ECMV_{b}}{\TMV}
+   }{
+     \ctxSynOrAnaType{\ctx}{\ELet{\PCMV}{\ECMV_{def}}{\ECMV_{b}}}{\TMV}
+   }{M(S/A)Let}
+\end{mathpar}
+
+Note that the $\TUnknownSwitch$ exists entirely to insure that sub-expressions of the definition are assigned an appropriate mode.
+As per the pattern rules below, it is never added to the context, which might result in spooky synthesis-at-a-distance:
+
   \judgbox{\ensuremath{\ctxSynPat{\Gamma}{\PMV}{\TMV}}} Pattern $\PMV$ synthesizes type $\TMV$
@@
   \label{fig:calculus-pattern-synthesis}
-\end{figure}
 
-\begin{figure}[htbp]
-  \raggedright
+For our analytic pattern judgement, out output context is produced, consisting of the input context augmented with any bindings made within the pattern: \\
+\\
 \judgbox{\ensuremath{\ctxAnaPat{\ctx_{in}}{\PMV}{\TMV}{\ctx_{out}}}} Pattern $\PMV$ analyzes against type $\TMV$ producing context $\Gamma_{out}$
@@
   \end{mathpar}
-  %
-  \label{fig:calculus-pattern-analysis}
-\end{figure}
 
-\begin{figure}[htbp]
-  \raggedright
+Marking judgements for patterns introduce two error types which parallel those for expressions:\\
+\\
   \judgbox{\ensuremath{\ctxSynFixedInto{\ctx_{in}}{\PMV}{\PCMV}{\TMV}}} Pattern $\PMV$ synthesizes $\TMV$ producing marked $\PCMV$ \\
@@
    \judgment{
-     \ctxNotAnaPat{\ctx}{\PMV{}}{\tau}{\ctx'} \\
-     \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\tau}{\_}
+     \ctxSynFixedInto{\ctx}{\PMV}{\PCMV}{\tau'}
+     \\
+     \inconsistent{\tau}{\tau'}
    }{
-        \ctxAnaFixedIntoPat{\ctx}{\PMV : \tau}{\ECInconType{\PCMV} : \tau}{\tau}{\ctx'}
-   }{IAPatAnnFail1}
+        \ctxAnaFixedIntoPat{\ctx}{\PMV}{\ECInconType{\PCMV}}{\tau}{\ctx}
+    }{IAInconsistentTypesPat}
 
-   \judgment{
-     \ctxAnaPat{\ctx}{\PMV{}}{\tau}{\ctx'} \\
-     \inconsistent{\tau}{\tau'} \\
-     \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\tau}{\_}
-   }{
-        \ctxAnaFixedIntoPat{\ctx}{\PMV : \tau'}{\ECInconType{\PCMV : \tau'}}{\tau}{\ctx'}
-   }{IAPatAnnFail2}
+   %\judgment{
+   %  \ctxNotAnaPat{\ctx}{\PMV{}}{\tau}{\ctx'} \\
+   %  \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\tau}{\_}
+   %}{
+   %     \ctxAnaFixedIntoPat{\ctx}{\PMV : \tau}{\ECInconType{\PCMV} : %\tau}{\tau}{\ctx'}
+   %}{IAPatAnnFail1}
 
-   \judgment{
-     \ctxNotAnaPat{\ctx}{\PMV{}}{\tau}{\ctx'} \\
-     \inconsistent{\tau}{\tau'} \\
-     \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\tau}{\_}
-   }{
-        \ctxAnaFixedIntoPat{\ctx}{\PMV : \tau'}{\ECInconType{\ECInconType{\PCMV} : \tau'}}{\tau}{\ctx'}
-   }{IAPatAnnFail3}
+   %\judgment{
+   %  \ctxAnaPat{\ctx}{\PMV{}}{\tau}{\ctx'} \\
+   %  \inconsistent{\tau}{\tau'} \\
+   %  \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\tau}{\_}
+   %}{
+   %     \ctxAnaFixedIntoPat{\ctx}{\PMV : \tau'}{\ECInconType{\PCMV : %\tau'}}{\tau}{\ctx'}
+   %}{IAPatAnnFail2}
+
+   %\judgment{
+   %  \ctxNotAnaPat{\ctx}{\PMV{}}{\tau}{\ctx'} \\
+   %  \inconsistent{\tau}{\tau'} \\
+   %  \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\tau}{\_}
+   %}{
+   %     \ctxAnaFixedIntoPat{\ctx}{\PMV : \tau'}%{\ECInconType{\ECInconType{\PCMV} : \tau'}}{\tau}{\ctx'}
+   %}{IAPatAnnFail3}
 
@@
   \label{fig:calculus-pattern-synthesis-marking}
-\end{figure}
 
-\begin{figure}[htbp]
+And finally, the typing rules for marked expression precisely parallel the unmarked ones:
+
   \begin{mathpar}
@@
    }{MSPatPair}
-  \end{mathpar}
-  %
-  \caption{Marked Pattern Synthesis}
-  \label{fig:calculus-pattern-synthesis-marked}
-\end{figure}
 
-\begin{figure}[htbp]
-  \raggedright
-  \begin{mathpar}
    \judgment{ }{\ctxAnaPat{\ctx}{\_}{\TMV}{\ctx}}{MAPatWild}
@@
     \judgment{
-     \ctxAnaPat{\ctx}{\PCMV{}}{\tau'}{\ctx'}
+     \ctxSynPat{\ctx}{\PCMV{}}{\tau'}
      \\
@@
    }{
-        \ctxAnaPat{\ctx}{\ECInconType{\PCMV}}{\tau}{\ctx'}
+        \ctxAnaPat{\ctx}{\ECInconType{\PCMV}}{\tau}{\ctx}
     }{MAInconsistentTypes}
@@
    \judgment{
-      \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\\\
-      \ctxAnaPat{\ctx}{\PCMV_1}{\TMV_1}{\ctx_1} \\\\
+      \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
+      \ctxAnaPat{\ctx}{\PCMV_1}{\TMV_1}{\ctx_1} \\
       \ctxAnaPat{\ctx_1}{\PCMV_2}{\TMV_2}{\ctx_2}
@@
     }{MAPatPair}
+
+    \judgment{
+       \notMatchedPair{\TMV} \\
+      \ctxAnaPat{\ctx}{\PCMV_1}{\TUnknown}{\ctx_1} \\
+      \ctxAnaPat{\ctx_1}{\PCMV_2}{\TUnknown}{\ctx_2}
+    }{
+      \ctxAnaPat{\ctx}{\ECInconMatchedPair{\ECPair{\PCMV_1}{\PCMV_2}}}{\TMV}{\ctx_2}
+    }{MAPatPairFail}
   \end{mathpar}
   %
-  \caption{Marked Pattern Analysis}
-  \label{fig:calculus-pattern-analysis-marked}
-\end{figure}
+  \caption{Marked Pattern Typing}
+  \label{fig:calculus-pattern-marked-typing}

### 2023-03-01 related.tex
Update related work; minor edits to various sections in type hole inference (Raef)

@@
 \label{sec:related}
+\textbf{Type Inference and Unification} In \emph{Gradual Typing with Unification Based Inference}, a version of unification for gradually typed languages based on Huet's unification algorithm is explored ~\cite{GradualInfer} ~\cite{Huet}.
+Other common approaches to unification are generally based on Hinley-Milner type inference ~\cite{MilnerInfer}.
+These algorithms are efficient, but can fail to solve constraints in the face of inconsistencies.
 
-Jeremy Siek Gradual Typing with Inference
-- Siek's type inference halts when it can't solve, ours doesn't
+\textbf{Correcting Sets and Assigning Blame} The generation of correcting sets is a common approach to error localization.
+The correcting set is assigned blame for failure in type inference \cite{sherrloc} \cite{typeinferDif} \cite{Pavlinovic2015}.
+Other approaches to assigning blame include statistically derived approaches that leverage machine learning to predict the best candidates for blame \cite{SeidelBlame}.
+
+\textbf{Dynamic Witnesses} Explanations for type errors can arguably be best illustrated by providing inputs that elicit runtime errors.
+This procedure can be used to generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with 70\% accuracy \cite{Seidel2016}.

### 2023-03-01 intro.tex

--- 2023-02-28 intro.tex
@@
 
-Modern programming environments provide developers with 
-various semantic services, such as type hints, semantic navigation, semantic code completion, and automated refactoring, that require statically reasoning about the type and binding structure of the program as it is being edited.
-% These semantic services are often packaged into language servers\todo{cites}.
-The problem is that the program being edited can be ill-typed, sometimes for long periods of time. 
-For example, when the developer changes an important type definition, errors might arise at dozens of usage sites in a large program, which could take hours or days to address.
-Semantic services often exhibit gaps in service during these periods \cite{HazelnutSNAPL}, because in the conventional formal account of type systems, only well-typed programs have a well-defined type and binding structure.
-As soon as a type error occurs \emph{anywhere}, the program is formally meaningless \emph{everywhere}.
+Modern programming environments provide developers with a collection of semantic services%
+---for example, type hints, semantic navigation, semantic code completion, and automated refactorings---%
+that require static reasoning about the type and binding structure of a program as it is being edited. 
+The problem is that when the program being edited is ill-typed, 
+these semantic services can become degraded or unavailable \cite{HazelnutSNAPL}. 
+These gaps in service are not always transient. 
+For example, a change to a type definition might result in type errors at dozens of use sites in a large program, which might take hours or days to resolve.
 
-Developers benefit when these gaps in service are minimized or eliminated, so there has long been considerable practical interest in (1)~\textbf{type error localization}: mechanisms for directing the developer to
-the location(s) in a program that may be erroneous, and (2) \textbf{type error recovery}: 
-mechanisms that allow the language server to optimistically recover from localized errors and continue to provide downstream semantic services, ideally throughout the program.
+These gaps in service are fundamentally rooted in a definitional gap: a type system defined in a conventional way, 
+e.g. in the tradition of the typed lambda calculus and its derivatives \cite{TaplBook,pfpl},
+assigns meaning only to well-typed programs. 
+If a type error appears \emph{anywhere}, the program is formally meaningless \emph{everywhere}.
 
-Essentially all widely-used programming systems have some support for type error 
-localization, e.g. via compiler error messages or user interface affordances fed by data from a language server.\todo{cites}
-Many also attempt recovery of certain services in certain situations, discussed below. 
-However, type error localization and recovery mechanisms has developed idiosyncratically, often as folklore amongst tool designers.
-Different type checkers or language servers, even for the same language, localize and recover from type errors in different ways, with little in the way of unifying theory of the sort that grounds the design of modern type systems themselves.
+This gap problem has prompted considerable practical interest in 
+(1)~\textbf{type error localization}: mechanisms for identifying the location(s) in a program that may be at fault for a type error, and 
+(2) \textbf{type error recovery}: mechanisms that allow the language server to optimistically recover from a localized type error 
+and continue on to locate other errors and provide downstream semantic services, 
+ideally at every location in the program and with minimal degradation in service.
+Essentially all widely-used programming systems have some support for type error localization. 
+When debugging type errors, developers have been observed to attend in particular to error locations presented in compiler error messages 
+or directly in their editor, via error marks decorating the localized code.\todo{cite}
+Many systems also attempt recovery of certain services in certain situations, discussed below.
 
-Consider, for example, the following program written in a typed functional language with, for now, only local type inference \cite{pierce}: 
+However, type error localization and recovery mechanisms have developed idiosyncratically, 
+often as folklore amongst language and tool implementors. 
+Different type checkers or language servers \cite{langauge-servers,merlin}\todo{cite}, even for the same language, localize and recover from type errors in different ways, 
+with little in the way of unifying theory of the sort that grounds the design of modern type systems themselves.
+
+Consider, for example, the following program written in Hazel, a typed functional dialect of Elm \cite{elm} with 
+support for local type inference, specified in the well-established bidirectional style \cite{pierce,hazelnut,BidirTyping}: 
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
-A type checker with no support for error localization or recovery at all would simply report that this program is ill-typed, because a type cannot be derived for it according to the type system specification, if given in the conventional style (e.g.
-following the simply typed lambda calculus and its derivatives).
+A type checker with no support for error localization or recovery---%
+the sort often written by students in undergraduate courses---%
+would simply report that this program is ill-typed, 
+i.e. a type cannot be derived for it according to Hazel's type system.
 
-The more common and practical approach is to localize the first error that causes type checking to fail and emit an error message\todo{cites} (which can help, though studies suggest that developers primarily use location information).
-In this example, such a system might report that the variable \li{f} is unbound and communicate its location, e.g.
-using editor decorations.
-This is the approach taken in many production type checkers, e.g. XXX.\todo{examples}
+A more practical approach, and a common one even in production type checkers, 
+is to localize the first error that causes local type inference to fail and emit an explanatory error message.
+In this example, the system might report that the variable \li{f} located on Line 2 is unbound, then stop.
 
-A language server with support for type error recovery might continue past this initial error to also report that \li{y} is unbound.
-To recover further, the system might optimistically ignore the fact that guard of the conditional cannot be confirmed to be a boolean expression and continue into its branches, observing that they have inconsistent types, \li{int} and \li{bool}.
+A language server with support for type error recovery would 
+ be tasked with continuing past this first error.
+ The general difficulty here is that there is now missing semantic information, namely the type of \li{f}, that 
+ the bidirectional type system as specified 
+ would appear to demand in order to proceed, here in order to determine which type the argument, \li{y}, is to be analyzed against.
+ Intuitively, however, it seems that this type is actually \emph{unnecessary} to make an error localization decision about \li{y}: 
+it is unbound, so a second error can be simultaneously localized despite the missing information.
 
-There are several potential approaches to localizing this inconsistency. 
-One approach would be to assume, arbitrarily, that the first branch is correct, so the error can be localized to the second branch, or \emph{vice versa}. 
-Heuristic approaches have been developed to make choices like these less arbitrary, e.g. by inspecting the recent editing history \todo{cite}, or training a machine learning model \todo{cite}. 
-Another approach is to localize the inconsistency between the two branches to the conditional expression itself. This entirely avoids needing to heuristically decide which branch is correct.
+To recover further, we might choose to also ignore the bidirectional type system's demand that the guard of the conditional be confirmed to be a boolean expression 
+(because the type of \li{f(y)} is also unknown) 
+and continue into its branches, observing that they have inconsistent types, \li{Int} and \li{Bool}. 
+There are several ways to localize this inconsistency. 
+One approach would be to assume, arbitrarily, that the first branch is correct, localizing the error to the second branch, or \emph{vice versa}.
+(TODO: what does Elm do? what does OCaml do?)
+Heuristics have been developed to help make choices like these less arbitrary, e.g. 
+by observing recent editing history \todo{cite}, 
+or training a machine learning model \todo{cite}. 
+The other approach, which we take in this paper's language server for Hazel, is to localize the inconsistency to the conditional expression itself, as indicated in
+the screenshot in Fig.~\ref{fig:hazel-intro-example}(b)\todo{figure}. 
+This entirely avoids needing to heuristically decide which branch is correct: the problem is actually that the branch types differ.
 
-This localization decision affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
-If localization assumes that the \li{then} branch is correct, then \li{x : int} and there is no error to report here.
-If the \li{else} branch is chosen, then \li{x : bool} and an error can be reported (though this error might mislead the programmer if this earlier localization step was incorrect). 
-If instead the inconsistency is localized to the conditional expression as a whole, then it is not obvious which type to assign to \li{x} at all. (Mention gradual typing here?)
+This localization decision cannot be made in a vacuum, however; it affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
+If localization assumes that the \li{then} branch is correct, then \li{x : Int} and there is no error to report here.
+If the \li{else} branch is chosen, then \li{x : Bool} and an error can be reported (though this error might mislead the programmer if this earlier localization guess was incorrect). 
+If instead the inconsistency is localized to the conditional expression as a whole, as in our language server, then we again confront the problem of missing type information: 
+\li{x} does not have a known type,
+though it is known to be bound. 
+There is no definitive problem here, based on the incomplete information available about \li{x}'s type, 
+so we choose not to report an error in Hazel.
 
-This exercise demonstrates that type error localization and recovery mechanisms engage in non-trivial reasoning and make choices that are not fully specified by the type system itself. 
-The first contribution of this paper is to address this problem by developing formal foundations for type error localization and recovery. 
+This informal exercise demonstrates that (1) localization choices can vary, particularly in regard to the extent to which they make \emph{ad hoc} guesses about intent;  
+(2) when combined with error recovery, localization decisions have implications on downstream localization decisions; and 
+(3) type error recovery mechanisms must be able to reason without complete knowledge about types and binding.
+
+This paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus} (Sec.~\ref{sec:calculus}).
+We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
+type annotations to expressions, (2) gradual typing, which offers a principled approach for handling missing type information during recovery, and 
+(3) non-empty holes, which operationalize type error localization by functioning as syntactic error marks, of the sort that a program editor would display on-screen.
+These ideas were previously brought together in the Hazelnut calculus, but its account of type error localization and recovery was partial and tied integrally into its structure editor. 
+We formally establish, with accompanying metatheory mechanized in Agda, that the marked lambda calculus has \emph{total error recovery}, i.e. that every syntactically well-formed program 
+can be marked, i.e. errors can be localized.
+
+the key ideas behind bidirectional type error localization with gradual recovery by specifying 
+the \emph{marked lambda calculus} (Sec.~\ref{sec:calculus}).
+
+Taking a formal approach has several benefits: it allows us to mechanically \emph{prove} that the system exhibits total error recovery, i.e. that recovery can never fail, 
+it gives us the ability to precisely specify the type and binding structure of
+
 Start assuming a system with local type inference / bidirectional typing, marked lambda calculus(Give judgement forms right off the bat??) 

### 2023-03-01 abstract.tex
Edits supplemental material for type hole inference stuff. Spacing is still a little funky in section B and I'm not quite sure why (Raef)

@@
 This paper addresses this problem, contributing the first comprehensive formal account of type error localization and recovery: the marked lambda calculus.
-In particular, we define a gradual type system for expressions with marked errors and holes, together with a total procedure for marking unmarked expressions.
+In particular, we define a gradual type system for expressions with marked errors, together with a total procedure for marking unmarked expressions with holes.
 We mechanize the metatheory of the marked lambda calculus in Agda and implement it, scaled up, as the new basis for Hazel, a full-scale live functional programming environment.

### 2023-03-01 intro.tex
Edits supplemental material for type hole inference stuff. Spacing is still a little funky in section B and I'm not quite sure why (Raef)

@@
 
-These gaps in service are fundamentally rooted in a definitional gap: a type system defined in a conventional way, 
+These service gaps are fundamentally rooted in a definitional gap: a type system defined in a conventional way, 
 e.g. in the tradition of the typed lambda calculus and its derivatives \cite{TaplBook,pfpl},
@@
 ideally at every location in the program and with minimal degradation in service.
-Essentially all widely-used programming systems have some support for type error localization. 
-When debugging type errors, developers have been observed to attend in particular to error locations presented in compiler error messages 
-or directly in their editor, via error marks decorating the localized code.\todo{cite}
+Essentially all widely-used programming systems have some support for type error localization, and developers have been observed to attend in particular to error locations as reported in compiler error messages
+or directly in their editor, via error markings decorating the localized code.\todo{cite}
 Many systems also attempt recovery of certain services in certain situations, discussed below.
@@
 in 
-  x + 1
+  x + "abc"
 \end{lstlisting}
@@
 There are several ways to localize this inconsistency. 
-One approach would be to assume, arbitrarily, that the first branch is correct, localizing the error to the second branch, or \emph{vice versa}.
-(TODO: what does Elm do? what does OCaml do?)
-Heuristics have been developed to help make choices like these less arbitrary, e.g. 
+One approach would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
+Heuristics have been developed to help make this choice less arbitrary, e.g. 
 by observing recent editing history \todo{cite}, 
 or training a machine learning model \todo{cite}. 
-The other approach, which we take in this paper's language server for Hazel, is to localize the inconsistency to the conditional expression itself, as indicated in
+The other approach, which we take in this paper's language server for Hazel, is to localize the inconsistency to the conditional expression itself, reporting that the problem is actually that the branch types differ, as indicated in
 the screenshot in Fig.~\ref{fig:hazel-intro-example}(b)\todo{figure}.
-This entirely avoids needing to heuristically decide which branch is correct: the problem is actually that the branch types differ.
 
 This localization decision cannot be made in a vacuum, however; it affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
-If localization assumes that the \li{then} branch is correct, then \li{x : Int} and there is no error to report here.
+If localization assumes that the \li{then} branch is correct, then \li{x : Int} and there is no error to report for the left operand.
 If the \li{else} branch is chosen, then \li{x : Bool} and an error can be reported (though this error might mislead the programmer if this earlier localization guess was incorrect). 
@@
 \li{x} does not have a known type,
-though it is known to be bound. 
-There is no definitive problem here, based on the incomplete information available about \li{x}'s type, 
+though it \emph{is} known to be bound. 
+There is no definitive type or binding error here, based on the incomplete information available about \li{x}'s type, 
 so we choose not to report an error in Hazel.
+No matter how localization handles the left operand, we would like to be able to recover and 
+localize the definitive type inconsistency on the right operand (the \li{+} operator is integer addition in Hazel.)
 

### 2023-03-01 calculus.tex
Rename marking calculus to marked lambda calculus

@@
 
-\section{The Marking Calculus}
+\section{The Marked Lambda Calculus}
 \label{sec:calculus}
@@
 
-The key contribution of this section is the \emph{marking calculus}, a formal calculus for type
-error localization in bidirectional systems. Namely, the calculus provides a judgmental framework
-with a unified theory for the design of typing semantics that support ill-typed programs.
+The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
+type error localization in bidirectional systems. Namely, the calculus provides a judgmental
+framework with a unified theory for the design of typing semantics that support ill-typed programs.
 \Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
@@
 
-We now formally introduce the \emph{marking calculus} through its application to a gradually typed
+We now formally introduce the marked lambda calculus through its application to a gradually typed
 lambda calculus extended with numbers and booleans. Consider the syntax for such a language, given
@@
 
-The \emph{marking calculus} builds atop these typing semantics into a two-layer system:
+The marked lambda calculus builds atop these typing semantics into a two-layer system:
 %
@@
 %
-These rules complete the development of the marking calculus for the gradually typed lambda
+These rules complete the development of the marked lambda calculus for the gradually typed lambda
 calculus.
@@
 
-We now demonstrate how one might extend or adapt the marking calculus for new typing features by
-adding binary product types. First, in \cref{fig:calculus-product-judgments}, we extend the syntax
-of types and the accompanying notions of consistency and join, which closely mirror the definitions
-for arrow types.
+We now demonstrate how one might extend or adapt the marked lambda calculus for new typing features
+by adding binary product types. First, in \cref{fig:calculus-product-judgments}, we extend the
+syntax of types and the accompanying notions of consistency and join, which closely mirror the
+definitions for arrow types.
 %
@@
 
-\section{The Marking Calculus in Structured Editing}
+\section{The Marked Lambda in Structured Editing}
 \label{sec:calculus-structured-editing}
@@
 
-By applying the marking calculus to Hazelnut, the action calculus may be \emph{wholly decoupled}
+By applying the marked lambda calculus to Hazelnut, the action calculus may be \emph{wholly decoupled}
 from the typing semantics. Under this new system, the action calculus is concerned only with the

### 2023-03-02 conclusion.tex
thi: minor edits

\section{Conclusion}
\label{sec:conclusion}
Taking a formal approach has several benefits: it allows us to mechanically \emph{prove} that the system exhibits total error recovery, i.e. that recovery can never fail, 
it gives us the ability to precisely specify the type and binding structure of

### 2023-03-02 intro.tex
thi: minor edits

@@
 
-Consider, for example, the following program written in Hazel, a typed functional dialect of Elm \cite{elm} with 
-support for local type inference, specified in the well-established bidirectional style \cite{pierce,hazelnut,BidirTyping}: 
-\begin{lstlisting}[numbers=none]
-let x = 
-  if f(y) then 1 else true
-in 
-  x + "abc"
-\end{lstlisting}
+Consider, the example program in \autoref{fig:hazel-intro}, which is written in
+this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{elm}. Hazel  
+supports local type inference, specified in the well-established bidirectional style \cite{pierce,hazelnut,BidirTyping}.
+\begin{figure}
+    \includegraphics[scale=0.5]{images/hazel-intro-screenshot.png}
+    \caption{Type error localization and recovery in Hazel}
+    \label{fig:hazel-intro}
+\end{figure}
+
 A type checker with no support for error localization or recovery---%
-the sort often written by students in undergraduate courses---%
+as might be written by students in an undergraduate course---%
 would simply report that this program is ill-typed, 
@@
 or training a machine learning model \todo{cite}. 
-The other approach, which we take in this paper's language server for Hazel, is to localize the inconsistency to the conditional expression itself, reporting that the problem is actually that the branch types differ, as indicated in
-the screenshot in Fig.~\ref{fig:hazel-intro-example}(b)\todo{figure}.
+The other approach, which we take in this paper's language server for Hazel, is to localize the inconsistency to the conditional expression itself, reporting that the problem is that the branch types differ, as indicated in
+the screenshot in \autoref{fig:hazel-intro}.
+Hazel reports typing information, including error messages, about the term at the cursor in its status bar, so when the cursor is on the conditional expression, the message below would be shown:
 
-This localization decision cannot be made in a vacuum, however; it affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
-If localization assumes that the \li{then} branch is correct, then \li{x : Int} and there is no error to report for the left operand.
-If the \li{else} branch is chosen, then \li{x : Bool} and an error can be reported (though this error might mislead the programmer if this earlier localization guess was incorrect). 
+TODO
+
+Whichever localization decision is made affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
+If localization had assumed that the \li{then} branch is correct, as for example the Elm type checker does, then \li{x : Bool} and an error should be reported on the left operand (though this error might mislead the programmer if this earlier localization guess was incorrect).
+If the \li{else} branch is chosen, then \li{x : Int} and an error should not be reported. 
 If instead the inconsistency is localized to the conditional expression as a whole, as in our language server, then we again confront the problem of missing type information: 
@@
 though it \emph{is} known to be bound. 
-There is no definitive type or binding error here, based on the incomplete information available about \li{x}'s type, 
+There is no definitive type or binding error here, based on the incomplete information available in the typing context, 
 so we choose not to report an error in Hazel.
+
 No matter how localization handles the left operand, we would like to be able to recover and 
-localize the definitive type inconsistency on the right operand (the \li{+} operator is integer addition in Hazel.)
+localize the type inconsistency on the right operand (the \li{+} operator is integer addition in Hazel).
 
-This informal exercise demonstrates that (1) localization choices can vary, particularly in regard to the extent to which they make \emph{ad hoc} guesses about intent;  
-(2) when combined with error recovery, localization decisions have implications on downstream localization decisions; and 
-(3) type error recovery mechanisms must be able to reason without complete knowledge about types and binding.
+This informal exercise demonstrates that (1) localization choices can vary, particularly with regard to whether they make \emph{ad hoc} guesses about intent;  
+(2) when combined with error recovery, localization decisions can influence downstream localization decisions; and 
+(3) error recovery mechanisms must be able to reason without complete knowledge about types and binding.
 
@@
 We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
-type annotations to expressions, (2) gradual typing, which offers a principled approach for handling missing type information during recovery, and 
-(3) non-empty holes, which operationalize type error localization by functioning as syntactic error marks, of the sort that a program editor would display on-screen.
-These ideas were previously brought together in the Hazelnut calculus, but its account of type error localization and recovery was partial and tied integrally into its structure editor. 
-We formally establish, with accompanying metatheory mechanized in Agda, that the marked lambda calculus has \emph{total error recovery}, i.e. that every syntactically well-formed program 
-can be marked, i.e. errors can be localized.
+type annotations to expressions \cite{pierce,BidirTyping}, (2) gradual typing, which offers a principled approach for handling missing type information during recovery \cite{GradualTyping,sieksnapl}\todo{siek05?}, and
+(3) non-empty holes, which allow us to operationalize type error localization by functioning as syntactic error marks, which an editor can display directly on-screen \cite{HazelnutPOPL}.
 
-the key ideas behind bidirectional type error localization with gradual recovery by specifying 
-the \emph{marked lambda calculus} (Sec.~\ref{sec:calculus}).
+These ideas were previously brought together in the Hazelnut calculus \cite{HazelnutPOPL}, but its account of type error localization and recovery was partial (see Sec.~\ref{sec:related}) and tied integrally into its structure editor.
+We resolve these problems to formally establish, with accompanying metatheory mechanized in Agda, that the marked lambda calculus has \emph{total error recovery}: \emph{every} syntactically well-formed program sketch (i.e.
+program structure with, optionally, empty holes) can be marked, i.e.
+its errors can be localized, such that the resulting program has a well-defined type and binding structure.
+We consider a number of situations where error localization decisions are subtle, including conditionals and when combining pattern matching with explicit type annotations.
 
-Taking a formal approach has several benefits: it allows us to mechanically \emph{prove} that the system exhibits total error recovery, i.e. that recovery can never fail, 
-it gives us the ability to precisely specify the type and binding structure of
+This system, scaled up and implemented in Hazel (Sec.~\ref{sec:calculus-hazel}), solves the semantic gap problem: Hazel's semantic services, which include type hints \cite{potter1}, semantic navigation, semantic code completion \cite{potter1,blinn}, contextualized documentation \cite{potter2}, and others,
+are available at all times and at all cursor locations, as long as the program sketch is syntactically well-formed. MENTION HAZELNUT LIVE?
 
-Start assuming a system with local type inference / bidirectional typing, marked lambda calculus(Give judgement forms right off the bat??) 
-(Mention gradual typing here?)
+Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches: manual hole insertion, syntax error recovery, and structure editing. 
+Hazel has experimented with several of these approaches, offering (1) a textual syntax, which allows the programmer to manually insert empty holes as necessary, following similar features available in GHC Haskell, Agda, Idris, and other languages, (2) a text-like structure editor that offers partial syntax error recovery (inserting holes to maintain well-formedness as long as all matching delimiters have been entered), and (3) a term-based structure editor with total syntax error recovery, i.e.
+a guarantee that every edit maintains syntactic well-formedness.
+All of these editing methods, and their analogs in other program editors, can feed into the mechanisms described in Sec.~\ref{sec:calculus}.
+However, as a secondary contribution,
+we demonstrate in Sec.~\ref{sec:calculus-structured-editing} that total type error recovery can resolve the problem of undefined editing behavior in the Hazelnut term-based structure editor calculus, namely that it does not allow edits that would require marking anywhere other than the cursor.
+We first reformulate Hazelnut as a purely syntactic structure editor, shifting re-marking to a separate whole-program pass after each edit.
+We also consider a more incremental approach, where remarking is driven by knowledge of the edit location, lexical scoping, and previous types.
+In either case, the result is a typed structure editor calculus that solves both the syntactic and semantic gap problem, i.e.
+every editor state is meaningful.
 
-We build on the foundation of the Hazelnut type system, which allows for local type inconsistencies to be placed within a non-empty hole, 
-e.g.
-$1 + false$ is meaningless but $1 + \{false\}$ can be given type Num.
-In other words, the non-empty holes serves as a membrane around type inconsistencies, here between the expected type, Num, and the synthesized type, Bool.
-Problems:
-
-1. does not handle every kind of error that can arise, e.g. inconsistent branch types, unbound variables, etc.
-
-2.
-does not provide a general account for how to correctly place the non-empty holes -- in that paper, they assume that a structure editor will be used and show how the structure editor can determine whether a non-empty hole is needed at the cursor after an edit action, but edits that might cause non-local errors to appear are undefined.
-users can also manually mark these locations, similar to Agda's non-empty holes.
-(Examples?)
-
-To solve both of these problems, we establish totality of marking -- we prove it to be total (and to insert only marks, without making any other changes), i.e. it must handle 
-every possible syntactically well-formed program and thus every error. Mechanized in Agda.
-
-Marking does not rely on a structure editor, so it is suitable as a foundational calculus for language servers supporting editors of any 
-design, both text editors (with syntax error recovery, a well-studied topic that we consider outside of the scope of this paper) and structure editors.
-To support this claim, we have implemented the calculus in a version of Hazel built atop tylr, which features both a text editor with explicit syntax for holes (a la Agda, GHC, etc.) and a textual structure editor with syntax error recovery via automatic hole insertion.
-Maximal liveness guarantee!
-
-Additional foundational contribution, we have also defined two versions of Hazelnut that eliminate the undefined behavior just mentioned.
-Composed with marking, this solves the problem.
-Also briefly consider incremental remarking in the style of Hazelnut.
-
-As we go, we'll discuss various localization design decisions that the language designer should consider when specifying the marked type system for their language.
-Overall, though, our work validates the folklore that systems that rely on local type inference tend to have an easier time deciding how to localize type errors (Pierce paper).
-Many modern languages feature local type inference, including ....
-However, non-local constraint-based type inference is also a common language feature, and it is known to substantially complicate error localization and recovery. 
-Heuristic approaches based on ML-based approaches, weighting, etc.
-In this paper, we consider an alternative approach to incorporating non-local constraints. Rather than ...
+Local type inference makes a good foundation for type error localization and comprehensible error message generation because it specifies clear, local information flows.
+Languages with non-local, constraint-based type inference gather constraints from all locations in a program, which is often helpful, but these systems are also notorious for complicating type error localization and recovery, because inconsistencies can flow from any number of locations in the program,
+and it is not clear how to recover when conflicting constraints are found. 
+Complex techniques, e.g. based on machine learning or weighting, are needed to heuristically localize errors to expressions. 
+In Hazel, we have taken a hybrid approach: 
+we generate constraints for each unknown type that appears in the program explicitly (i.e.
+type hole) and each unknown type generated internally during \emph{after} bidirectional type error localization and recovery.
+We develop a ?
+unification algorithm that can then generate possible hole fillings
+for each of thes...
+if the hole can be uniquely solved, then we suggest the solution to the programmer. 
+if conflicting constraints arise, then we localize this problem directly to type holes, rather 
+than attempting to localize the problem to expressions. 
+The user can select from different partial fillings -- the bidirectional system can then shift the error to relevant expressions in a more principled way.
+Until a choice is made, the system continues to recover gradually as explained in sec 2.
+Sec. 4.\todo{clean up text}

### 2023-03-02 related.tex
thi: minor edits

--- 2023-02-25 related.tex
@@
 \label{sec:related}
+\textbf{Type Inference and Unification} \emph{Gradual Typing with Unification Based Inference} explores a version of unification for gradually typed languages based on Huet's unification algorithm ~\cite{GradualInfer} ~\cite{Huet}.
+Other common approaches to unification are generally based on Hindley-Milner type inference ~\cite{MilnerInfer}.
+These algorithms are efficient, but can fail to solve constraints in the face of inconsistencies.
 
-Jeremy Siek Gradual Typing with Inference
-- Siek's type inference halts when it can't solve, ours doesn't
+\textbf{Correcting Sets and Assigning Blame} The generation of correcting sets is a common approach to error localization.
+The correcting set is assigned blame for failure in type inference \cite{sherrloc} \cite{typeinferDif} \cite{Pavlinovic2015}.
+Other approaches to assigning blame include statistically derived approaches that leverage machine learning to predict the best candidates for blame \cite{SeidelBlame}.
+
+\textbf{Dynamic Witnesses} Explanations for type errors can be well illustrated by providing sample inputs (dynamic witnesses) that elicit runtime errors.
+With this approach, one can generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with around 70\% accuracy \cite{Seidel2016}.

### 2023-03-02 calculus.tex
Edit calculus section and update rules

@@
   \[\begin{array}{rrcl}
-    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
-    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV}
-                       \mid           \ENumMV \mid \EPlus{\EMV}{\EMV}
-                       \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
+    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \mid \TProd{\TMV}{\TMV} \\
+    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \mid \ELet{x}{\EMV}{\EMV}
+                       \mid           \ENumMV \mid \EPlus{\EMV}{\EMV} \\
+             &       & \mid         & \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
+                       \mid           \EPair{\EMV}{\EMV}
+                       \mid           \EProjL{\EMV} \mid \EProjR{\EMV}
                        \mid           \EEHole \\
-    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV}
-                       \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV}
-                       \mid           \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV}
-                       \mid           \EEHole \\
-             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV}
+    \ECMName & \ECMV & \Coloneqq & x \mid \ECLam{x}{\TMV}{\ECMV} \mid \ECAp{\ECMV}{\ECMV} \mid \ECLet{x}{\ECMV}{\ECMV}
+                       \mid           \ECNumMV \mid \ECPlus{\ECMV}{\ECMV} \\
+             &       & \mid         & \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV}
+                       \mid           \ECPair{\ECMV}{\ECMV} \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
+                       \mid           \ECEHole \\
+             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV} \mid \ECInconAsc{\ECMV} \\
+             &       & \mid         & \ECSynNonMatchedArrow{\ECMV} \mid \ECAnaNonMatchedArrow{\ECMV}
+                       \mid           \ECSynNonMatchedProd{\ECMV} \mid \ECAnaNonMatchedProd{\ECMV}
   \end{array}\]
@@
 Types classify expressions by a \emph{bidirectional type system}, which employs two mutually defined
-judgments. \emph{Type synthesis}, written $\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under
+judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$, establishes that, under
 the typing context $\ctx$, the expression $e$ synthesizes or infers the type $\TMV$. \emph{Type
-analysis}, written $\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear
+analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear
 where an expression of type $\TMV$ is expected. We may give these typing semantics traditionally,
@@
 \Cref{fig:calculus-syntax} furthermore gives the syntax for marked expressions (as $\ECMName$),
-which replicates all the forms of unmarked expressions. There are additionally three kinds of
-\emph{error marks}, which represent the different type errors that may arise.
-%
-\begin{itemize}
-  \item $\ECUnbound{x}$, the \emph{unbound variable mark}, indicates the use of an unbound variable
-    $x$, as in \cref{fig:calculus-examples-unbound}.
-
-  \item $\ECInconType{\ECMV}$, the \emph{inconsistent type mark}, indicates that the type of an
-    expression does not match an expected type.
-
-  \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, indicates
-    that the branches of a conditional have inconsistent types.
-\end{itemize}
-%
-Throughout the remainder of this section, as we consider the formulation of marking, we will
-motivate and give precise semantics for these error marks. Note that this is a non-exhaustive list;
-depending on the syntactic forms and typing features found in the language, other kinds of marks may
-be necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of marks that may be
-required in an language with explicit polymorphism.
+which replicates all the forms of unmarked expressions. There are additionally several kinds of
+\emph{error marks}, which represent the different type errors that may arise. Intuitively, they each
+denote a different error message that might be shown in an editor or emitted by a compiler.
+%
+% \begin{itemize}
+  % \item $\ECUnbound{x}$, the \emph{unbound variable mark}, indicates the use of an unbound variable
+    % $x$, as in \cref{fig:calculus-examples-unbound}.
+%
+  % \item $\ECInconType{\ECMV}$, the \emph{inconsistent type mark}, indicates that the type of an
+    % expression does not match an expected type.
+%
+  % \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, indicates
+    % that the branches of a conditional have mismatched types.
+%
+  % \item $\ECLamInconAsc{x}{\TMV}{\ECMV}$, the \emph{inconsistent ascription mark}, indicates that
+    % the type ascription $\TMV$ is mismatches with an expected type.
+%
+  % \item $\ECSynNonMatchedArrow{\ECMV}$
+% \end{itemize}
+%
+Throughout the remainder of this section, as we formulate marking for the gradually typed lambda
+calculus, we will motivate and give precise semantics for these error marks. Note that this is a
+non-exhaustive list; depending on the syntactic forms and typing features found in the language,
+other kinds of marks may be necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of
+marks that may be required in an language with explicit polymorphism.
 
 The marked language possesses its own typing semantics, also formulated bidirectionally. We write
-$\ctxSynType{\ctx}{\ECMV}{\TMV}$ for synthesis and $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ for analysis.
+$\ctxSynTypeM{\ctx}{\ECMV}{\TMV}$ for synthesis and $\ctxAnaTypeM{\ctx}{\ECMV}{\TMV}$ for analysis.
 Finally, marking is bidirectional as well. The synthetic marking judgment
@@
   \begin{enumerate}
-    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynType{\ctx}{\ECMV}{\TMV}$ and
+    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynTypeM{\ctx}{\ECMV}{\TMV}$ and
       $\erasesTo{\ECMV}{\EMV}$.
-    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaType{\ctx}{\ECMV}{\TMV}$ and
+    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaTypeM{\ctx}{\ECMV}{\TMV}$ and
       $\erasesTo{\ECMV}{\EMV}$.
@@
 %
-In other words, if an unmarked expression is marked into some marked expression, erasing error marks
-from the marked expression gives the original unmarked expression. We also ensure that marking gives
-well-typed marked expressions; indeed, their types should match the types synthesized or analyzed
-against in marking. % TODO more intuition for this theorem
+In other words, marking and then erasing error marks from the result gives the original unmarked
+expression. We also ensure that marking gives well-typed marked expressions; indeed, their types
+should match the types synthesized or analyzed against in marking.
+% TODO more intuition for this theorem
 
@@
 In the remainder of this subsection, we consider the different type errors that may arise in the
-unmarked language; in this way, the typing semantics of the unmarked language guide the construction
-of the marking process.
+unmarked language while keeping in mind these two theorems; in this way, the typing semantics of the
+unmarked language guide the construction of the marking process.
 
@@
 
-% TODO need to say something about subsumption and how it will be handled later.
-To start, let us consider the simple case of numbers. Because of subsumption, we need only define a
-synthesis rule for unmarked numbers, in which numbers synthesize the type $\TNum$:
-%
-\begin{mathpar}
-  \judgment{ }{
-    \ctxSynType{\ctx}{\ENumMV}{\TNum}
+To start, let us consider the simple case of numbers. Because of \emph{subsumption}, which we
+discuss next, we need only define a synthesis rule for unmarked numbers, in which numbers synthesize
+the type $\TNum$:
+%
+\begin{mathpar}
+  \judgment{ }{
+    \ctxSynTypeU{\ctx}{\ENumMV}{\TNum}
   }{USNum}
@@
 
-Into what should numbers be marked? Quite straightforwardly, we simply give the same number as a
-marked expression, synthesizing again $\TNum$. No type errors may occur in a single number, so we
-only need the following rules for marking unmarked numbers and typing marked numbers:
+How should numbers be marked? Quite straightforwardly, we simply give the same number as a marked
+expression, synthesizing again $\TNum$. No type errors may occur in a single number, so we only need
+the following rules for marking numbers and typing the results: 
 %
@@
   \judgment{ }{
-    \ctxSynType{\ctx}{\ECNumMV}{\TNum}
+    \ctxSynTypeM{\ctx}{\ECNumMV}{\TNum}
   }{MSNum}
@@
 
-In addition expressions, the type of both operands should be $\TNum$. In a bidirectional system, we
-thus analyze the operands against $\TNum$, giving the following rule:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxAnaType{\ctx}{\EMV_1}{\TNum} \\
-    \ctxAnaType{\ctx}{\EMV_2}{\TNum}
-  }{
-    \ctxSynType{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
+For addition expressions, the type of both operands should be $\TNum$. To denote this in a
+bidirectional system, the operands are analyzed against $\TNum$, giving the following typing rule
+for unmarked addition expressions:
+%
+\begin{mathpar}
+  \judgment{
+    \ctxAnaTypeU{\ctx}{\EMV_1}{\TNum} \\
+    \ctxAnaTypeU{\ctx}{\EMV_2}{\TNum}
+  }{
+    \ctxSynTypeU{\ctx}{\EPlus{\EMV_1}{\EMV_2}}{\TNum}
   }{USPlus}
@@
 
-The marking rule mirrors \textsc{USPlus} closely. Since a known, expected type for each operand is
-known, we shift the ``blame'' for any type errors to them. Hence, we recursively mark each operand
-in analytic mode and rebuild the marked addition expression. The marked typing rule then mirrors
+The marking rule mirrors \textsc{USPlus} closely. Since an expected type for each operand is known,
+we shift the responsibility for any type errors to them. Hence, we recursively mark each operand in
+analytic mode and rebuild the marked addition expression. The marked typing rule then mirrors
 \textsc{USPlus} exactly.
@@
   \judgment{
-    \ctxAnaType{\ctx}{\ECMV_1}{\TNum} \\\\
-    \ctxAnaType{\ctx}{\ECMV_2}{\TNum}
-  }{
-    \ctxSynType{\ctx}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
+    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TNum} \\\\
+    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TNum}
+  }{
+    \ctxSynTypeM{\ctx}{\ECPlus{\ECMV_1}{\ECMV_2}}{\TNum}
   }{MSPlus}
 \end{mathpar}
+
+\subsubsection{Subsumption}
+\label{sec:calculus-subsumption}
+
+Only synthetic rules were necessary for typing numbers and addition because of \emph{subsumption},
+given below, which states that if an expression synthesizes a type, it may also be analyzed against
+that type or any \emph{consistent} type.
+\[%
+  \judgment{
+    \ctxSynTypeU{\ctx}{\EMV}{\TMV'} \\
+    \consistent{\TMV}{\TMV'} \\
+    \subsumable{\EMV}
+  }{
+    \ctxAnaTypeU{\ctx}{\EMV}{\TMV}
+  }{UASubsume}
+\]%
+We rely on the notion of \emph{type consistency} from gradual type theory, which defines a reflexive
+and symmetric (but not transitive) relation between types, writing $\consistent{\TMV_1}{\TMV_2}$ to
+mean that $\TMV_1$ is consistent with $\TMV_2$. Defined in \cref{fig:calculus-consistency}, this
+replaces the notion of \emph{type equality} to relate the unknown type to all other types.
+
+\begin{figure}[htbp]
+  \raggedright
+  \judgbox{\ensuremath{\consistent{\TMV_1}{\TMV_2}}} $\TMV_1$ is consistent with $\TMV_2$
+  %
+  \begin{mathpar}
+    \judgment{ }{
+      \consistent{\TUnknown}{\TMV}
+    }{TCUnknown1}
+
+    \judgment{ }{
+      \consistent{\TMV}{\TUnknown}
+    }{TCUnknown2}
+
+    \judgment{ }{
+      \consistent{\TMV}{\TMV}
+    }{TCRefl}
+
+    \judgment{
+      \consistent{\TMV_1}{\TMV_1'} \\
+      \consistent{\TMV_2}{\TMV_2'} \\
+    }{
+      \consistent{\TArrow{\TMV_1}{\TMV_2}}{\TArrow{\TMV_1'}{\TMV_2'}}
+    }{TCArr} \\
+  \end{mathpar}
+  %
+  \caption{Type consistency.}
+  \label{fig:calculus-consistency}
+\end{figure}
+
+Hence, when checking $\ENumMV$ against a known expected type, subsumption checks that the type is
+consistent with $\TNum$, the type that $\ENumMV$ synthesizes. This succeeds for $\TUnknown$ and
+$\TNum$.
+
+We also restrict the usage of subsumption to ``subsumable'' syntactic forms, written
+$\subsumable{\EMV}$. This judgment is defined on all syntactic forms except lambda abstractions and
+conditionals, the only ones with explicit analysis rules. In other words, we restrict subsumption to
+be the rule of ``last resort'', which is necessary to establish that marking and typing are
+deterministic (discussed later in \cref{thm:calculus-marking-unicity}).
+
+Now, to define analytic marking on forms without explicit analytic typing rules, we also need
+subsumption rules for marking and marked expression typing. Note that we define a analogous notion
+of ``subsumability'' for marked expressions, written $\subsumable{\ECMV}$.
+%
+\begin{mathpar}
+  \judgment{
+    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
+    \consistent{\TMV}{\TMV'} \\
+    \subsumable{\EMV}
+  }{
+    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}
+  }{MKASubsume}
+
+  \judgment{
+    \ctxSynTypeM{\ctx}{\ECMV}{\TMV'} \\
+    \consistent{\TMV}{\TMV'} \\
+    \subsumable{\ECMV}
+  }{
+    \ctxAnaTypeM{\ctx}{\ECMV}{\TMV}
+  }{MASubsume}
+\end{mathpar}
+
+But what happens when the synthesized type of an expression is \emph{not} consistent with the
+expected type, i.e. that the premise $\consistent{\TMV}{\TMV'}$ fails? Recalling the example of
+\cref{fig:calculus-examples-inconsistent-types}, subsumption would be used to analyze $\ETrue$
+against $\TNum$ in $\EPlus{\ETrue}{5}$, which fails since $\inconsistent{\TNum}{\TBool}$, i.e.
+\textsc{UASubsume} does not apply, and traditional typing semantics would simply fail the type
+checking process at this point. \textsc{MKASubsume} would also not be applicable, leaving marking
+undefined in those cases.
+
+% TODO use example more closely related to num here since we haven't talked about booleans yet?
+
+To satisfy marking totality, such a possibility motivates an \emph{inconsistent type mark}
+$\ECInconType{\ECMV}$, which states exactly that the synthesized type of $\ECMV$ is inconsistent
+with the expected type. It is then accompanied by the following rules:
+%
+\begin{mathpar}
+  \judgment{
+    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
+    \inconsistent{\TMV}{\TMV'} \\
+    \subsumable{\EMV}
+  }{
+    \ctxAnaFixedInto{\ctx}{\EMV}{\ECInconType{\ECMV}}{\TMV}
+  }{MKAInconsistentTypes}
+
+  \judgment{
+    \ctxSynTypeM{\ctx}{\ECMV}{\TMV'} \\
+    \inconsistent{\TMV}{\TMV'} \\
+    \subsumable{\ECMV}
+  }{
+    \ctxAnaTypeM{\ctx}{\ECInconType{\ECMV}}{\TMV}
+  }{MAInconsistentTypes}
+\end{mathpar}
+%
+Observe that the premises of \textsc{MKAInconsistentTypes} are identical to those of
+\textsc{MKASubsume}, except that $\inconsistent{\TMV}{\TMV'}$. By marking an error, type-checking
+may continue---instead of simply failing.
 
@@
 
-Let us now consider the more interesting case of variables. The typing of unmarked variables in the
-gradually typed lambda calculus is standard. Again, because of subsumption, there is only a
-synthetic rule:
+Let us now consider the case of variables. Typing in the unmarked language is standard, and because
+of subsumption, only a synthetic rule is required:
 \[%
@@
   }{
-    \ctxSynType{\ctx}{x}{\TMV}
+    \ctxSynTypeU{\ctx}{x}{\TMV}
   }{USVar}
 \]%
-If $x$ is bound to a type in the typing context, it synthesizes that type. Straightforwardly,
-marking converts the unmarked $x$ into the same marked $x$ via the following marking and typing
+That is, if $x$ is bound to a type in the typing context, it synthesizes that type.
+Straightforwardly, marking converts an unmarked variable into the same variable via the following
 rules:
@@
   }{
-    \ctxSynType{\ctx}{x}{\TMV}
+    \ctxSynTypeM{\ctx}{x}{\TMV}
   }{MSVar}
@@
 
-But what happens a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
-bound, i.e. that the premise fails? Then, \textsc{USVar} does not apply, and the typing semantics
-say nothing about such expressions. When devising marking, to satisfy totality, such a possibility
-motivates an \emph{unbound variable mark} $\ECUnbound{x}$, accompanied by the following rules:
+However, consider the case that a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
+bound. Again, \textsc{USVar} would not apply. A total marking should, however, report the error and
+continue, motivating an \emph{unbound variable mark} $\ECUnbound{x}$ and the accompanying rules:
 %
@@
   }{
-    \ctxSynType{\ctx}{\ECUnbound{x}}{\TUnknown}
+    \ctxSynTypeM{\ctx}{\ECUnbound{x}}{\TUnknown}
   }{MSUnbound}
@@
 %
-\textsc{MKSUnbound} states that an variable unbound in the context is marked into that variable
-inside an unbound variable mark. \textsc{MSUnbound} then gives the typing of unbound variable marks.
-Since nothing may be said about their types, we synthesize the unknown type, which allows type
-checking to continue---instead of simply failing---and the unbound variable to be used in any
-expression. % TODO clarify why this is correct
+An unbound variable is marked as such, and since nothing may be said about their types, we
+synthesize the unknown type. Similarly to the inconsistent type mark, this allows type checking to
+proceed, with usage of the unbound variable permitted in any expression.
+% TODO clarify/give intuition as to why this is correct
 
@@
 
-There are explicit synthesis and analysis rules for unmarked lambda abstractions:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\TArrow{\TMV_1}{\TMV_2}}
+Unlike numbers and variables, there are explicit synthesis and analysis rules for unmarked lambda
+abstractions: % TODO intuition why
+%
+\begin{mathpar}
+  \judgment{
+    \ctxSynTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
+  }{
+    \ctxSynTypeU{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\TArrow{\TMV_1}{\TMV_2}}
   }{USLam}
@@
     \consistent{\TMV}{\TMV_1} \\
-    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
-  }{
-    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
+    \ctxAnaTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
+  }{
+    \ctxAnaTypeU{\ctx}{\ECLam{x}{\TMV}{\EMV}}{\TMV_3}
   }{UALam}
@@
 %
-The synthesis rule is standard. In the premises of \textsc{UALam}, we use the notion of
-\emph{type consistency} from gradual type theory, which defines a reflexive and symmetric (but not
-transitive) relation between types, writing $\consistent{\TMV_1}{\TMV_2}$ to mean that $\TMV_1$ is
-consistent with $\TMV_2$. This replaces the notion of \emph{type equality} to relate the unknown
-type to all other types. The analysis rule also employs the judgment
-$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which establishes that $\TMV$ is a \emph{matched arrow type},
-i.e. it may be considered an arrow type. This notion is purely a technical mechanism to avoid
-duplication of rules related to arrow types. \Cref{fig:calculus-consistency-matched-arrow} gives the
-rules for both judgments.
+The synthesis rule is standard. The analysis rule employs the judgment
+$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which establishes that $\TMV$ is a \emph{matched arrow type}
+[citation], i.e. it may be considered an arrow type. Defined in \cref{fig:calculus-matched-arrow},
+this notion is purely a technical mechanism to avoid duplication of rules related to arrow types.
 
@@
   \raggedright
-  \judgbox{\ensuremath{\consistent{\TMV_1}{\TMV_2}}} $\TMV_1$ is consistent with $\TMV_2$
-  %
-  \begin{mathpar}
-    \judgment{ }{
-      \consistent{\TUnknown}{\TMV}
-    }{TCUnknown1}
-
-    \judgment{ }{
-      \consistent{\TMV}{\TUnknown}
-    }{TCUnknown2}
-
-    \judgment{ }{
-      \consistent{\TMV}{\TMV}
-    }{TCRefl}
-
-    \judgment{
-      \consistent{\TMV_1}{\TMV_1'} \\
-      \consistent{\TMV_2}{\TMV_2'} \\
-    }{
-      \consistent{\TArrow{\TMV_1}{\TMV_2}}{\TArrow{\TMV_1'}{\TMV_2'}}
-    }{TCArr} \\
-  \end{mathpar}
-
   \judgbox{\ensuremath{\matchedArrow{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched arrow type $\TArrow{\TMV_1}{\TMV_2}$
@@
   %
-  \caption{Type consistency and matched arrow types.}
-  \label{fig:calculus-consistency-matched-arrow}
+  \caption{Matched arrow types.}
+  \label{fig:calculus-matched-arrow}
 \end{figure}
@@
   \judgment{
-    \ctxSynType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\ECLam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
+    \ctxSynTypeM{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
+  }{
+    \ctxSynTypeM{\ctx}{\ECLam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
   }{MSLam}
@@
     \consistent{\TMV}{\TMV_1} \\\\
-    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
-  }{
-    \ctxAnaType{\ctx}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
+    \ctxAnaTypeM{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
+  }{
+    \ctxAnaTypeM{\ctx}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
   }{MALam1}
@@
 may arise. First, what if $\TMV_3$ is not a matched arrow type? This would be the case if it were
-$\TNum$, for example, and the lambda abstraction would be inconsistent with the expected type. Thus,
+$\TNum$, for example. The lambda abstraction's synthesized type would be inconsistent with the
+expected type, but this case is slightly distinct from that of the inconsistent types mark;
+
+Thus,
 to the syntax we add an \emph{inconsistent type mark} $\ECInconType{\ECMV}$, which indicates that
@@
   }{
-    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
+    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLamAnaNonMatchedArrow{x}{\TMV}{\ECMV}}{\TMV_3}
   }{MKALam2}
@@
     \notMatchedArrow{\TMV_3} \\
-    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
-  }{
-    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
+    \ctxAnaTypeM{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TUnknown}
+  }{
+    \ctxAnaTypeM{\ctx}{\ECLamAnaNonMatchedArrow{x}{\TMV}{\ECMV}}{\TMV_3}
   }{MALam2}
@@
 A similar situation arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
-$\inconsistent{\TMV}{\TMV_1}$, i.e. the expected input type is inconsistent with the actual input
-type, giving the last pair of analytic rules for lambda abstractions:
+$\inconsistent{\TMV}{\TMV_1}$, i.e. the actual type annotation for $x$ is inconsistent with the
+expected input type. The \emph{inconsistent ascription mark} $\ECInconAsc{x}$ indicates exactly this
+error, and we add a final pair of analytic rules:
 %
@@
     \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
-    \inconsistent{\TMV}{\TMV_1} \\\\
-    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
-  }{
-    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
+    \inconsistent{\TMV}{\TMV_1} \\
+    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
+  }{
+    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLamInconAsc{x}{\TMV}{\ECMV}}{\TMV_3}
   }{MKALam3}
@@
     \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
-    \inconsistent{\TMV}{\TMV_1} \\\\
-    \ctxAnaType{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
-  }{
-    \ctxAnaType{\ctx}{\ECInconType{\ECLam{x}{\TMV}{\ECMV}}}{\TMV_3}
+    \inconsistent{\TMV}{\TMV_1} \\
+    \ctxAnaTypeM{\extendCtx{\ctx}{x}{\TMV_1}}{\ECMV}{\TMV_2}
+  }{
+    \ctxAnaTypeM{\ctx}{\ECLamInconAsc{x}{\TMV}{\ECMV}}{\TMV_3}
   }{MALam3}
@@
 
-The case of application is somewhat interesting. In the unmarked language, only a synthesis rule is
-necessary:
+In the unmarked language, only a synthesis rule is necessary for applications:
 \[%
   \judgment{
-    \ctxSynType{\ctx}{\EMV_1}{\TMV} \\
+    \ctxSynTypeU{\ctx}{\EMV_1}{\TMV} \\
     \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\
-    \ctxAnaType{\ctx}{\EMV_2}{\TMV_1}
-  }{
-    \ctxSynType{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\TMV_2}
+    \ctxAnaTypeU{\ctx}{\EMV_2}{\TMV_1}
+  }{
+    \ctxSynTypeU{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\TMV_2}
   }{USAp}
@@
   \judgment{
-    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
+    \ctxSynTypeM{\ctx}{\ECMV_1}{\TMV} \\
     \matchedArrow{\TMV}{\TMV_1}{\TMV_2} \\\\
-    \ctxAnaType{\ctx}{\ECMV_2}{\TMV_1}
-  }{
-    \ctxSynType{\ctx}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
+    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TMV_1}
+  }{
+    \ctxSynTypeM{\ctx}{\ECAp{\ECMV_1}{\ECMV_2}}{\TMV_2}
   }{MSAp1}
@@
 as in the example of \cref{fig:calculus-examples-app-non-lambda}. There is no expected type for the
-argument, so we perform analytic marking on $\EMV_2$ against the unknown type. The applied
-expression is placed inside the inconsistent type mark, which is in accordance with the meaning of
-that mark.
+argument, so we perform analytic marking on $\EMV_2$ against the unknown type. In this case, it is
+not quite right to mark $\EMV_1$ with the inconsistent type mark; rather than any single type, it is
+a member of the family of function types that is expected. For such a ``constrained'' synthetic
+mode, we use a specialized mark $\ECSynNonMatchedArrow{\ECMV}$, which indicates that $\ECMV$ was
+expected to be a function---but was not. The output type is unknown, so the entire expression
+synthesizes the unknown type.
 %
@@
   }{
-    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApNonMatched{\ECMV_1}{\ECMV_2}}{\TUnknown}
+    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECApSynNonMatchedArrow{\ECMV_1}{\ECMV_2}}{\TUnknown}
   }{MKSAp2}
@@
   \judgment{
-    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
+    \ctxSynTypeM{\ctx}{\ECMV_1}{\TMV} \\
     \notMatchedArrow{\TMV} \\\\
-    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
-  }{
-    \ctxSynType{\ctx}{\ECApNonMatched{\ECMV_1}{\ECMV_2}}{\TUnknown}
+    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TUnknown}
+  }{
+    \ctxSynTypeM{\ctx}{\ECApSynNonMatchedArrow{\ECMV}{\ECMV}}{\TUnknown}
   }{MSAp2}
@@
 
-It is worth noting, however, that $\EMV_1$ is not actually in analytic position, but in a synthetic
-position where any arrow type is expected. Treating such ``constrained synthetic'' cases specially,
-we may instead designate a specialized mark, written $\ECApNonMatchedAltAlt{\ECMV}$. This yields the
-following alternative rules:
-%
-\[\begin{array}{rrcl}
-  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECApNonMatchedAltAlt{\ECMV}
-\end{array}\]
-%
-\begin{mathpar}
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
-    \notMatchedArrow{\TMV} \\\\
-    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
-  }{
-    \ctxSynFixedInto{\ctx}{\EAp{\EMV_1}{\EMV_2}}{\ECAp{\ECApNonMatchedAltAlt{\ECMV_1}}{\ECMV_2}}{\TUnknown}
-  }{MKSAp2'}
-
-  \judgment{
-    \ctxSynType{\ctx}{\ECMV_1}{\TMV} \\
-    \notMatchedArrow{\TMV} \\\\
-    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
-  }{
-    \ctxSynType{\ctx}{\ECApNonMatchedAltAlt{\ECMV_1}{\ECMV_2}}{\TUnknown}
-  }{MSAp2'}
-\end{mathpar}
-
 It is natural to extend this approach to other elimination forms that require the handling of
-unmatched types, such as products. Another approach is to mark an error around the entire
-application expression.
+unmatched types, such as products. Indeed, the same approach is taken for projections in
+\cref{sec:calculus-products}. Another approach might indicate the same kind of error but mark the
+entire application.
 
@@
 
-Boolean values are similarly straightforward:
-%
-\begin{mathpar}
-  \judgment{ }{
-    \ctxSynType{\ctx}{\ETrue}{\TBool}
+The boolean values are straightforward to handle:
+%
+\begin{mathpar}
+  \judgment{ }{
+    \ctxSynTypeU{\ctx}{\ETrue}{\TBool}
   }{USTrue}
@@
   \judgment{ }{
-    \ctxSynType{\ctx}{\ECTrue}{\TBool}
+    \ctxSynTypeM{\ctx}{\ECTrue}{\TBool}
   }{MSTrue} \\
@@
   \judgment{ }{
-    \ctxSynType{\ctx}{\EFalse}{\TBool}
+    \ctxSynTypeU{\ctx}{\EFalse}{\TBool}
   }{USFalse}
@@
   \judgment{ }{
-    \ctxSynType{\ctx}{\ECFalse}{\TBool}
+    \ctxSynTypeM{\ctx}{\ECFalse}{\TBool}
   }{MSFalse}
@@
 
-If-else expressions, however, offer another interesting case. In the unmarked language, we have
-both explicit synthetic and analytic rules:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxAnaType{\ctx}{\EMV_1}{\TBool} \\\\
-    \ctxSynType{\ctx}{\EMV_2}{\TMV_1} \\
-    \ctxSynType{\ctx}{\EMV_3}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
+Conditionals, however, offer an interesting case. In the unmarked language, we have both explicit
+synthetic and analytic rules: % TODO intuition why
+%
+\begin{mathpar}
+  \judgment{
+    \ctxAnaTypeU{\ctx}{\EMV_1}{\TBool} \\\\
+    \ctxSynTypeU{\ctx}{\EMV_2}{\TMV_1} \\
+    \ctxSynTypeU{\ctx}{\EMV_3}{\TMV_2}
+  }{
+    \ctxSynTypeU{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
   }{USIf}
@@
   \judgment{
-    \ctxAnaType{\ctx}{\EMV_1}{\TBool} \\\\
-    \ctxAnaType{\ctx}{\EMV_1}{\TMV} \\
-    \ctxAnaType{\ctx}{\EMV_2}{\TMV}
-  }{
-    \ctxAnaType{\ctx}{\ECIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TMV}
+    \ctxAnaTypeU{\ctx}{\EMV_1}{\TBool} \\\\
+    \ctxAnaTypeU{\ctx}{\EMV_1}{\TMV} \\
+    \ctxAnaTypeU{\ctx}{\EMV_2}{\TMV}
+  }{
+    \ctxAnaTypeU{\ctx}{\ECIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TMV}
   }{UAIf}
@@
 Following \textsc{USIf} and \textsc{UAIf}, we may derive the following marking and typing rules.
-Similar to previous cases, we simply recursively mark the guard and branch expressions and construct
-the marked conditional from the results.
+Similar to previous cases, we recursively mark the guard and branches, constructing the marked
+conditional from the results.
 %
@@
   \judgment{
-    \ctxAnaType{\ctx}{\ECMV_1}{\TBool} \\\\
-    \ctxSynType{\ctx}{\ECMV_2}{\TMV_1} \\
-    \ctxSynType{\ctx}{\ECMV_3}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
+    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TBool} \\\\
+    \ctxSynTypeM{\ctx}{\ECMV_2}{\TMV_1} \\
+    \ctxSynTypeM{\ctx}{\ECMV_3}{\TMV_2}
+  }{
+    \ctxSynTypeM{\ctx}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
   }{MSIf} \\
@@
   \judgment{
-    \ctxAnaType{\ctx}{\ECMV_1}{\TBool} \\\\
-    \ctxAnaType{\ctx}{\ECMV_1}{\TMV} \\
-    \ctxAnaType{\ctx}{\ECMV_2}{\TMV}
-  }{
-    \ctxAnaType{\ctx}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
+    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TBool} \\\\
+    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TMV} \\
+    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TMV}
+  }{
+    \ctxAnaTypeM{\ctx}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
   }{MAIf}
@@
 
-However, in synthetic mode, it may be the case that the two branch types have no join. This occurs,
-in fact, when they are inconsistent. This motivates \emph{the inconsistent branches mark}
-$\ECInconBr{\ECMV}{\ECMV}{\ECMV}$. Then, adding the following rules ensures marking totality on
+However, in synthetic position, it may be the case that the two branch types have no join. This
+occurs, in fact, when they are inconsistent, motivating the \emph{the inconsistent branches mark}
+$\ECInconBr{\ECMV}{\ECMV}{\ECMV}$. Then, adding the following rules ensures totality on
 conditionals:
@@
   \judgment{
-    \ctxAnaType{\ctx}{\ECMV_1}{\TBool} \\
-    \ctxSynType{\ctx}{\ECMV_2}{\TMV_1} \\\\
-    \ctxSynType{\ctx}{\ECMV_3}{\TMV_2} \\
+    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TBool} \\
+    \ctxSynTypeM{\ctx}{\ECMV_2}{\TMV_1} \\\\
+    \ctxSynTypeM{\ctx}{\ECMV_3}{\TMV_2} \\
     \inconsistent{\TMV_1}{\TMV_2}
   }{
-    \ctxSynType{\ctx}{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TUnknown}
+    \ctxSynTypeM{\ctx}{\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TUnknown}
   }{MSInconsistentBranches}
@@
 
-\subsubsection{Subsumption}
-\label{sec:calculus-subsumption}
-
-Finally, we consider the subsumption rule in typing unmarked expressions, which states that if an
-expression synthesizes a type, it may also be analyzed against that type or any consistent type.
-\[%
-  \judgment{
-    \ctxSynType{\ctx}{\EMV}{\TMV'} \\
-    \consistent{\TMV}{\TMV'} \\
-    \subsumable{\EMV}
-  }{
-    \ctxAnaType{\ctx}{\EMV}{\TMV}
-  }{UASubsume}
-\]%
-We also restrict the usage of subsumption to ``subsumable'' syntactic forms, written
-$\subsumable{\EMV}$. This judgment is defined on all syntactic forms except lambda abstractions and
-conditionals, the only ones with explicit analysis rules. In other words, we restrict
-subsumption to be the analytic rule of ``last resort''. This is critical in establishing that
-marking and typing are deterministic (discussed later in \cref{thm:calculus-marking-unicity}).
-
-Since we did not provide explicit analytic rules for marking expressions that are not lambda
-abstractions or conditionals (or typing their marked versions) either, we give corresponding
-subsumption rules:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
-    \consistent{\TMV}{\TMV'} \\
-    \subsumable{\EMV}
-  }{
-    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}
-  }{MKASubsume}
-
-  \judgment{
-    \ctxSynType{\ctx}{\ECMV}{\TMV'} \\
-    \consistent{\TMV}{\TMV'} \\
-    \subsumable{\ECMV}
-  }{
-    \ctxAnaType{\ctx}{\ECMV}{\TMV}
-  }{MASubsume}
-\end{mathpar}
-%
-Note that we define a analogous notion of ``subsumability'' for marked expressions, written
-$\subsumable{\ECMV}$.
-
-However, it may be the case that the expression is inconsistent with the expected type. Recalling
-the example of \cref{fig:calculus-examples-inconsistent-types}, subsumption would used to analyze
-$\ETrue$ against $\TNum$ in $\EPlus{\ETrue}{5}$, which fails since $\inconsistent{\TNum}{\TBool}$,
-and $\ETrue$ should be marked with an inconsistent type mark. Hence, the following final set of
-rules is required to handle such cases:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'} \\
-    \inconsistent{\TMV}{\TMV'} \\\\
-    \subsumable{\EMV}
-  }{
-    \ctxAnaFixedInto{\ctx}{\EMV}{\ECInconType{\ECMV}}{\TMV}
-  }{MKAInconsistentTypes}
-
-  \judgment{
-    \ctxSynType{\ctx}{\ECMV}{\TMV'} \\
-    \inconsistent{\TMV}{\TMV'} \\\\
-    \subsumable{\ECMV}
-  }{
-    \ctxAnaType{\ctx}{\ECInconType{\ECMV}}{\TMV}
-  }{MAInconsistentTypes}
-\end{mathpar}
-%
 These rules complete the development of the marked lambda calculus for the gradually typed lambda
@@
     \item \begin{enumerate}
-        \item If $\ctxSynType{\ctx}{\EMV}{\TMV}$ and $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
-          then $\imageIs{\EMV}{\ECMV}$.
-        \item If $\ctxAnaType{\ctx}{\EMV}{\TMV}$ and $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
-          then $\imageIs{\EMV}{\ECMV}$.
+        \item If $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$ and $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
+          then $\markless{\ECMV}$.
+        \item If $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$ and $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
+          then $\markless{\ECMV}$.
       \end{enumerate}
@@
     \item \begin{enumerate}
-        \item If there does not exist $\TMV$ such that $\ctxSynType{\ctx}{\EMV}{\TMV}$, then for all
-          $\ECMV$ and $\TMV'$ such that $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, we have
-          $\imageIsNot{\EMV}{\ECMV}$.
-        \item If there does not exist $\TMV$ such that $\ctxAnaType{\ctx}{\EMV}{\TMV}$, then for all
-          $\ECMV$ and $\TMV'$ such that $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, we have
-          $\imageIsNot{\EMV}{\ECMV}$.
+        \item If there does not exist $\TMV$ such that $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$, then for
+          all $\ECMV$ and $\TMV'$ such that $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, it is not
+          the case that $\markless{\ECMV}$.
+        \item If there does not exist $\TMV$ such that $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$, then for
+          all $\ECMV$ and $\TMV'$ such that $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, it is not
+          the case that $\markless{\ECMV}$.
       \end{enumerate}
@@
 %
-That is, since error marks indicate type errors in the program, no marks should be inserted
-into a well-typed program. The converse---that at least one mark is inserted into ill-typed
-programs---is also true. We have chosen to formulate the absence of error marks through an inclusion
-mapping of unmarked expressions into marked expressions; alternatively, a predicate over marked
-expressions may be used. % TODO latter is easier to do in Agda, but maybe it is weaker?
+That is, since error marks indicate type errors in the program, no marks should be introduced
+when marking a well-typed program, a property given by the judgment $\markless{\ECMV}$. The
+converse---that at least one mark is inserted into ill-typed programs---is also true.
+% TODO latter is easier to do in Agda, but maybe it is weaker?
 
@@
 %
-Together with totality, marking may be considered and implemented as a total function.
+Together, totality and unicity give that marking may be considered and implemented as a total
+function. Indeed, given the algorithmic nature of bidirectional typing, it is fairly easy to extract
+an implementation from the rules.
 
@@
 \[\begin{array}{rrcl}
-  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TPair{\TMV}{\TMV}
+  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TProd{\TMV}{\TMV}
 \end{array}\]
@@
 We additionally define an auxiliary notion of \emph{matched binary product types}: the judgment
-$\matchedPair{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be considered the binary product
-type $\TPair{\TMV_1}{\TMV_2}$.
+$\matchedProd{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be considered the binary product
+type $\TProd{\TMV_1}{\TMV_2}$.
 
@@
     }{
-      \consistent{\TPair{\TMV_1}{\TMV_2}}{\TPair{\TMV_1'}{\TMV_2'}}
+      \consistent{\TProd{\TMV_1}{\TMV_2}}{\TProd{\TMV_1'}{\TMV_2'}}
     }{TCProd} \\
@@
     & \cdots & \\
-    \joinsTo{(\TPair{\TMV_1}{\TMV_2})}{(\TPair{\TMV_1'}{\TMV_2'})}{\TPair{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}} \\
+    \joinsTo{(\TProd{\TMV_1}{\TMV_2})}{(\TProd{\TMV_1'}{\TMV_2'})}{\TProd{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}} \\
   \end{array}\]
 
-  \judgbox{\ensuremath{\matchedPair{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched binary product type $\TPair{\TMV_1}{\TMV_2}$
-  %
-  \begin{mathpar}
-    \judgment{ }{
-      \matchedPair{\TUnknown}{\TUnknown}{\TUnknown}
+  \judgbox{\ensuremath{\matchedProd{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched binary product type $\TProd{\TMV_1}{\TMV_2}$
+  %
+  \begin{mathpar}
+    \judgment{ }{
+      \matchedProd{\TUnknown}{\TUnknown}{\TUnknown}
     }{TMPHole}
@@
     \judgment{ }{
-      \matchedPair{\TPair{\TMV_1}{\TMV_2}}{\TMV_1}{\TMV_2}
+      \matchedProd{\TProd{\TMV_1}{\TMV_2}}{\TMV_1}{\TMV_2}
     }{TMPProd}
@@
   \judgment{
-    \ctxSynType{\ctx}{\EMV_1}{\TMV_1} \\
-    \ctxSynType{\ctx}{\EMV_2}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TPair{\TMV_1}{\TMV_2}}
+    \ctxSynTypeU{\ctx}{\EMV_1}{\TMV_1} \\
+    \ctxSynTypeU{\ctx}{\EMV_2}{\TMV_2}
+  }{
+    \ctxSynTypeU{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TProd{\TMV_1}{\TMV_2}}
   }{USPair}
@@
   \judgment{
-    \ctxSynType{\ctx}{\EMV}{\TMV} \\
-    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\EProjL{\EMV}}{\TMV_1}
+    \ctxSynTypeU{\ctx}{\EMV}{\TMV} \\
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynTypeU{\ctx}{\EProjL{\EMV}}{\TMV_1}
   }{USProjL}
@@
   \judgment{
-    \ctxSynType{\ctx}{\EMV}{\TMV} \\
-    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\EProjR{\EMV}}{\TMV_2}
+    \ctxSynTypeU{\ctx}{\EMV}{\TMV} \\
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynTypeU{\ctx}{\EProjR{\EMV}}{\TMV_2}
   }{USProjR} \\
@@
   \judgment{
-    \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
-    \ctxAnaType{\ctx}{\EMV_1}{\TMV_1} \\
-    \ctxAnaType{\ctx}{\EMV_2}{\TMV_2}
-  }{
-    \ctxAnaType{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TMV}
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
+    \ctxAnaTypeU{\ctx}{\EMV_1}{\TMV_1} \\
+    \ctxAnaTypeU{\ctx}{\EMV_2}{\TMV_2}
+  }{
+    \ctxAnaTypeU{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TMV}
   }{UAPair}
@@
   \judgment{
-    \ctxAnaType{\ctx}{\EMV}{\TPair{\TMV}{\TUnknown}}
-  }{
-    \ctxAnaType{\ctx}{\EProjL{\EMV}}{\TMV}
+    \ctxAnaTypeU{\ctx}{\EMV}{\TProd{\TMV}{\TUnknown}}
+  }{
+    \ctxAnaTypeU{\ctx}{\EProjL{\EMV}}{\TMV}
   }{UAProjL}
@@
   \judgment{
-    \ctxAnaType{\ctx}{\EMV}{\TPair{\TUnknown}{\TMV}}
-  }{
-    \ctxAnaType{\ctx}{\EProjR{\EMV}}{\TMV}
+    \ctxAnaTypeU{\ctx}{\EMV}{\TProd{\TUnknown}{\TMV}}
+  }{
+    \ctxAnaTypeU{\ctx}{\EProjR{\EMV}}{\TMV}
   }{UAProjR}
@@
                                \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
-                               \mid \ECInconMatchedPair{\ECMV}
+                               \mid \ECSynNonMatchedProd{\ECMV}
 \end{array}\]
@@
   }{
-    \ctxSynFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TPair{\TMV_1}{\TMV_2}}
+    \ctxSynFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TProd{\TMV_1}{\TMV_2}}
   }{MKSPair}
@@
     \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
-    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
   }{
@@
   \judgment{
-    \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
     \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
@@
   \judgment{
-    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TPair{\TMV}{\TUnknown}}
+    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TProd{\TMV}{\TUnknown}}
   }{
@@
   \judgment{
-    \ctxSynType{\ctx}{\ECMV_1}{\TMV_1} \\
-    \ctxSynType{\ctx}{\ECMV_2}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TPair{\TMV_1}{\TMV_2}}
+    \ctxSynTypeM{\ctx}{\ECMV_1}{\TMV_1} \\
+    \ctxSynTypeM{\ctx}{\ECMV_2}{\TMV_2}
+  }{
+    \ctxSynTypeM{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TProd{\TMV_1}{\TMV_2}}
   }{MSPair}
@@
   \judgment{
-    \ctxSynType{\ctx}{\ECMV}{\TMV} \\
-    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\ECProjL{\ECMV}}{\TMV_1}
+    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynTypeM{\ctx}{\ECProjL{\ECMV}}{\TMV_1}
   }{MSProjL1}
@@
   \judgment{
-    \ctxSynType{\ctx}{\ECMV}{\TMV} \\
-    \matchedPair{\TMV}{\TMV_1}{\TMV_2}
-  }{
-    \ctxSynType{\ctx}{\ECProjR{\ECMV}}{\TMV_2}
+    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynTypeM{\ctx}{\ECProjR{\ECMV}}{\TMV_2}
   }{MSProjR1}
@@
   \judgment{
-    \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
-    \ctxAnaType{\ctx}{\ECMV_1}{\TMV_1} \\
-    \ctxAnaType{\ctx}{\ECMV_2}{\TMV_2}
-  }{
-    \ctxAnaType{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
+    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TMV_1} \\
+    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TMV_2}
+  }{
+    \ctxAnaTypeM{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
   }{MAPair1}
@@
   \judgment{
-    \ctxAnaType{\ctx}{\ECMV}{\TPair{\TMV}{\TUnknown}}
-  }{
-    \ctxAnaType{\ctx}{\ECProjL{\ECMV}}{\TMV}
+    \ctxAnaTypeM{\ctx}{\ECMV}{\TProd{\TMV}{\TUnknown}}
+  }{
+    \ctxAnaTypeM{\ctx}{\ECProjL{\ECMV}}{\TMV}
   }{MAProjL}
@@
   \judgment{
-    \ctxAnaType{\ctx}{\ECMV}{\TPair{\TUnknown}{\TMV}}
-  }{
-    \ctxAnaType{\ctx}{\ECProjR{\ECMV}}{\TMV}
+    \ctxAnaTypeM{\ctx}{\ECMV}{\TProd{\TUnknown}{\TMV}}
+  }{
+    \ctxAnaTypeM{\ctx}{\ECProjR{\ECMV}}{\TMV}
   }{MAProjR}
@@
 
-However, as with lambda abstractions and application seen previously, two sets of rules are further
+However, as previously seen with lambda abstractions and applications, two sets of rules are further
 necessary to satisfy totality. Firstly, the premise in \textsc{UAPair} that $\TMV$ is a matched
@@
   \judgment{
-    \notMatchedPair{\TMV} \\
+    \notMatchedProd{\TMV} \\
     \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TUnknown} \\
@@
   }{
-    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECInconType{\ECPair{\ECMV_1}{\ECMV_2}}}{\TMV}
+    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPairAnaNonMatchedProd{\ECMV_1}{\ECMV_2}}{\TMV}
   }{MKAPair2}
@@
   \judgment{
-    \notMatchedPair{\TMV} \\
-    \ctxAnaType{\ctx}{\ECMV_1}{\TUnknown} \\
-    \ctxAnaType{\ctx}{\ECMV_2}{\TUnknown}
-  }{
-    \ctxAnaType{\ctx}{\ECInconType{\ECPair{\ECMV_1}{\ECMV_2}}}{\TMV}
+    \notMatchedProd{\TMV} \\
+    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TUnknown} \\
+    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TUnknown}
+  }{
+    \ctxAnaTypeM{\ctx}{\ECPairAnaNonMatchedProd{\ECMV_1}{\ECMV_2}}{\TMV}
   }{MAPair2}
@@
 
-The other kind of error may arise when the subject of a projection does not synthesize a matched
-product type. As in our development of marking for applications, we may place the subject in an
-inconsistent type mark, or we may treat it as a special ``constrained synthesis'' mode, utilising a
-new error mark $\ECInconMatchedPair{\ECMV}$. The latter approach leads to the following rules:
+Another kind of error may arise when the subject of a projection does not synthesize a matched
+product type. As in our development of marking for applications, we utilize a new error mark
+$\ECSynNonMatchedProd{\ECMV}$, leading to the following rules:
 %
@@
     \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
-    \notMatchedPair{\TMV}
-  }{
-    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECInconMatchedPair{\ECMV}}}{\TMV_1}
+    \notMatchedProd{\TMV}
+  }{
+    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjLSynNonMatchedProd{\ECMV}}{\TMV_1}
   }{MKSProjL2}
@@
   \judgment{
-    \ctxSynType{\ctx}{\ECMV}{\TMV} \\
-    \notMatchedPair{\TMV}
-  }{
-    \ctxSynType{\ctx}{\ECProjL{\ECInconMatchedPair{\ECMV}}}{\TUnknown}
+    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\
+    \notMatchedProd{\TMV}
+  }{
+    \ctxSynTypeM{\ctx}{\ECProjLSynNonMatchedProd{\ECMV}}{\TUnknown}
   }{MSProjL2} \\
@@
     \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
-    \notMatchedPair{\TMV}
-  }{
-    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECInconMatchedPair{\ECMV}}}{\TMV_2}
+    \notMatchedProd{\TMV}
+  }{
+    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjRSynNonMatchedProd{\ECMV}}{\TMV_2}
   }{MKSProjR2}
@@
   \judgment{
-    \ctxSynType{\ctx}{\ECMV}{\TMV} \\
-    \notMatchedPair{\TMV}
-  }{
-    \ctxSynType{\ctx}{\ECProjR{\ECInconMatchedPair{\ECMV}}}{\TUnknown}
+    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\
+    \notMatchedProd{\TMV}
+  }{
+    \ctxSynTypeM{\ctx}{\ECProjRSynNonMatchedProd{\ECMV}}{\TUnknown}
   }{MSProjR2}
@@
 
+[Something about structure editors]
+
 The previous section presented a formal calculus for error localization and recovery. We now discuss
-its usage in a structured editing setting.
-
-In particular, the calculus has been integrated into the Hazel programming environment, a live
-programming environment with support for typed holes.
+its usage in a structured editing setting. In particular, the calculus has been integrated into the
+Hazel programming environment, a live programming environment with support for typed holes.
 
@@
 
-The Hazelnut action calculus introduced by \citet{HazelnutPOPL} give a structure editing scheme that
-ensures programs are statically meaningful.
-% TODO say something more about Hazelnut, especially in relation to error holes
+The Hazelnut action calculus introduced by \citet{HazelnutPOPL} gives a structure editing scheme
+that employs \emph{expression holes}, denoted $\EEHole$, which represent syntactically incomplete
+part of the program. Not only are terms always syntactically well-formed, they are always
+\emph{well-typed}, ensuring that semantic services are always available. To accomplish this, the
+action calculus inserts on-demand \emph{error holes}, denoted $\hhole{\EMV}{}$, which indicate type
+errors.
+
+% TODO example
 

### 2023-03-02 let.tex
Edit calculus section and update rules

@@
 \[\begin{array}{rrcl}
-  \EMName & \EMV & \Coloneqq & \cdots
-                               \mid \ELet{\PMV}{\EMV}{\EMV} \\
-  \ECMName & \EMV & \Coloneqq & \cdots
-                               \mid \ELet{\PMV}{\EMV}{\EMV} \\
-  \PMName & \PMV & \Coloneqq & x 
-                               \mid \_
-                               \mid (p_1, p_2)
-                               \mid p : \TMV \\
-  \PCMName & \PCMV & \Coloneqq & x 
-                               \mid \_
-                               \mid (p_1, p_2)
-                               \mid p : \TMV
-                               \mid \ECInconType{\PCMV}
-                               \mid \ECInconMatchedPair{\PCMV}
+  \EMName  & \EMV  & \Coloneqq & \cdots \mid \ELet{\PMV}{\EMV}{\EMV} \\
+  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECLet{\PMV}{\ECMV}{\ECMV} \\
+  \PMName  & \PMV  & \Coloneqq & \PWild \mid \PVar{x} \mid \PPair{\PMV}{\PMV} \mid \PAsc{\PMV}{\TMV} \\
+  \PCMName & \PCMV & \Coloneqq & \PCWild \mid \PCVar{x} \mid \PCPair{\PCMV}{\PCMV} \mid \PCAsc{\PCMV}{\TMV}
+                               \mid \PCInconType{\PCMV} \mid \PCAnaNonMatchedProd{\PCMV}
 \end{array}\]
@@
 \[\begin{array}{rrcl}
-  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TUnknownSwitch
+  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TUnknownSwitch \\
 \end{array}\]
@@
   \judgment{
-     \ctxSynType{\ctx}{\EMV}{\_}
+    \ctxSynType{\ctx}{\EMV}{\ctx'}
   }{
     \ctxAnaType{\ctx}{\EMV}{\TUnknownSwitch}
-   }{SynSwitch}
+  }{UASynSwitch}
 \end{mathpar}
@@
   \judgment{
-     \ctxSynPat{\ctx}{\PMV}{\TMV_p}
-     \\
-     \ctxAnaType{\ctx}{\EMV_{def}}{\TMV_p}
-     \\
-     \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
-     \\\\
-     \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
-     \\
-     \ctxSynOrAnaType{\ctx_{b}}{\EMV_{b}}{\TMV}
+    \ctxSynPatU{\ctx}{\PMV}{\TMV_p} \\
+    \ctxAnaTypeU{\ctx}{\EMV_{1}}{\TMV_p} \\
+    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\\\
+    \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
+    \ctxSynTypeU{\ctx'}{\EMV_{2}}{\TMV_2}
   }{
-     \ctxSynOrAnaType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
-   }{U(S/A)Let}
+    \ctxSynType{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\TMV_2}
+  }{USLetPat}
 
-    %\judgment{
-    % \ctxSynPat{\ctx}{\PMV}{\TMV_p}
-    % \\
-    % \ctxAnaType{\ctx}{\EMV_{def}}{\TMV_p}
-    % \\
-    % \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
-    % \\\\
-    % \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
-    % \\
-    % \ctxSynType{\ctx_{b}}{\EMV_{b}}{\TMV}
-   %}{
-   %  \ctxSynType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
-   %}{USLet}
-   %\\
-   %\judgment{
-   %  \ctxSynPat{\ctx}{\PMV}{\TMV_p}
-   %  \\
-   %  \ctxAnaType{\ctx}{\EMV_{def}}{\TMV_p}
-   %  \\
-   %  \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
-   %  \\\\
-   %  \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
-   %  \\
-   %  \ctxAnaType{\ctx_{b}}{\EMV_{b}}{\TMV}
-   %}{
-   %  \ctxAnaType{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\TMV}
-   %}{UALet}
+  \judgment{
+    \ctxSynPatU{\ctx}{\PMV}{\TMV_p} \\
+    \ctxAnaTypeU{\ctx}{\EMV_{1}}{\TMV_p} \\
+    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\\\
+    \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
+    \ctxAnaTypeU{\ctx'}{\EMV_{2}}{\TMV_2}
+  }{
+    \ctxAnaTypeU{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\TMV_2}
+  }{UALetPat}
 \end{mathpar}
@@
   \judgment{
-     \ctxSynFixedInto{\ctx}{\PMV}{\PCMV}{\TMV_p}
-     \\
-     \ctxAnaFixedInto{\ctx}{\EMV_{def}}{\ECMV_{def}}{\TMV_{p}}
-     \\\\
-     \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
-     \\
-     \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
-     \\
-     \ctxSynFixedInto{\ctx_{b}}{\EMV_{b}}{\ECMV_{b}}{\TMV}
+    \ctxSynFixedInto{\ctx}{\PMV}{\PCMV}{\TMV_p} \\
+    \ctxAnaFixedInto{\ctx}{\EMV_{1}}{\ECMV_{1}}{\TMV_{p}} \\\\
+    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\
+    \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
+    \ctxSynFixedInto{\ctx'}{\EMV_{2}}{\ECMV_{2}}{\TMV_2}
   }{
-     \ctxSynFixedInto{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\ELet{\PCMV}{\ECMV_{def}}{\ECMV_{b}}}{\TMV}
-   }{ISLet}
+    \ctxSynFixedInto{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\ELet{\PCMV}{\ECMV_{1}}{\ECMV_{2}}}{\TMV_2}
+  }{ISLetPat}
 
   \judgment{
-     \ctxSynFixedInto{\ctx}{\PMV}{\PCMV}{\TMV_p}
-     \\
-     \ctxAnaFixedInto{\ctx}{\EMV_{def}}{\ECMV_{def}}{\TMV_{p}}
-     \\\\
-     \ctxSynType{\ctx}{\EMV_{def}}{\TMV_{def}}
-     \\
-     \ctxAnaPat{\ctx}{\PMV}{\TMV_{def}}{\ctx_{b}}
-     \\
-     \ctxAnaFixedInto{\ctx_{b}}{\EMV_{b}}{\ECMV_{b}}{\TMV}
+    \ctxSynFixedInto{\ctx}{\PMV}{\PCMV}{\TMV_p} \\
+    \ctxAnaFixedInto{\ctx}{\EMV_{1}}{\ECMV_{1}}{\TMV_{p}} \\\\
+    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\
+    \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
+    \ctxAnaFixedInto{\ctx'}{\EMV_{2}}{\ECMV_{2}}{\TMV_2}
   }{
-     \ctxAnaFixedInto{\ctx}{\ELet{\PMV}{\EMV_{def}}{\EMV_{b}}}{\ELet{\PCMV}{\ECMV_{def}}{\ECMV_{b}}}{\TMV}
-   }{IALet}
+    \ctxAnaFixedInto{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\ELet{\PCMV}{\ECMV_{1}}{\ECMV_{2}}}{\TMV_2}
+  }{IALetPat}
 
   \judgment{
-     \ctxSynPat{\ctx}{\PCMV}{\TMV_p}
-     \\
-     \ctxAnaType{\ctx}{\ECMV_{def}}{\TMV_p}
-     \\
-     \ctxSynType{\ctx}{\ECMV_{def}}{\TMV_{def}}
-     \\\\
-     \ctxAnaPat{\ctx}{\PCMV}{\TMV_{def}}{\ctx_{b}}
-     \\
-     \ctxSynOrAnaType{\ctx_{b}}{\ECMV_{b}}{\TMV}
+    \ctxSynPatM{\ctx}{\PCMV}{\TMV_p} \\
+    \ctxAnaTypeM{\ctx}{\ECMV_{1}}{\TMV_p} \\
+    \ctxSynTypeM{\ctx}{\ECMV_{1}}{\TMV_{1}} \\\\
+    \ctxAnaPatM{\ctx}{\PCMV}{\TMV_{1}}{\ctx'} \\
+    \ctxSynTypeM{\ctx'}{\ECMV_{2}}{\TMV_2}
   }{
-     \ctxSynOrAnaType{\ctx}{\ELet{\PCMV}{\ECMV_{def}}{\ECMV_{b}}}{\TMV}
-   }{M(S/A)Let}
+    \ctxSynTypeM{\ctx}{\ELet{\PCMV}{\ECMV_{1}}{\ECMV_{2}}}{\TMV_2}
+  }{MSLetPat}
+
+  \judgment{
+    \ctxSynPatM{\ctx}{\PCMV}{\TMV_p} \\
+    \ctxAnaTypeM{\ctx}{\ECMV_{1}}{\TMV_p} \\
+    \ctxSynTypeM{\ctx}{\ECMV_{1}}{\TMV_{1}} \\\\
+    \ctxAnaPatM{\ctx}{\PCMV}{\TMV_{1}}{\ctx'} \\
+    \ctxAnaTypeM{\ctx'}{\ECMV_{2}}{\TMV_2}
+  }{
+    \ctxAnaTypeM{\ctx}{\ELet{\PCMV}{\ECMV_{1}}{\ECMV_{2}}}{\TMV_2}
+  }{MALetPat}
 \end{mathpar}
@@
 
-  \judgbox{\ensuremath{\ctxSynPat{\Gamma}{\PMV}{\TMV}}} Pattern $\PMV$ synthesizes type $\TMV$
+\judgbox{\ensuremath{\ctxSynPatU{\ctx}{\PMV}{\TMV}}} $\PMV$ synthesizes type $\TMV$
 %
 \begin{mathpar}
-   \judgment{ }{\ctxSynPat{\ctx}{\_}{\TUnknownSwitch}}{USPatWild}
+  \judgment{ }{
+    \ctxSynPatU{\ctx}{\PWild}{\TUnknownSwitch}
+  }{USPWild}
 
-   \judgment{ }{\ctxSynPat{\ctx}{x}{\TUnknownSwitch}}{USPatVar}
-
-   %\judgment{ }{\ctxSynPat{\Gamma}{\hehole{}}{\TUnknownSwitch}}{USPatHole}
-   \judgment{
-     \ctxAnaPat{\ctx}{\PMV{}}{\tau}{\_}
-   }{
-        \ctxSynPat{\ctx}{\PMV : \tau}{}{\tau}
-    }{USPatAnn}
+  \judgment{ }{
+    \ctxSynPatU{\ctx}{\PVar{x}}{\TUnknownSwitch}
+  }{USPVar}
 
   \judgment{
-     \ctxSynPat{\ctx}{\PMV_1}{\TMV_1} \\
-     \ctxSynPat{\ctx}{\PMV_2}{\TMV_2}
+    \ctxSynPatU{\ctx}{\PMV_1}{\TMV_1} \\
+    \ctxSynPatU{\ctx}{\PMV_2}{\TMV_2}
   }{
-     \ctxSynPat{\ctx}{\EPair{\PMV_1}{\PMV_2}}{\TPair{\TMV_1}{\TMV_2}}
-   }{USPatPair}
+    \ctxSynPatU{\ctx}{\PPair{\PMV_1}{\PMV_2}}{\TProd{\TMV_1}{\TMV_2}}
+  }{USPPair}
+
+  \judgment{
+    \ctxAnaPatU{\ctx}{\PMV}{\TMV}{\ctx'}
+  }{
+    \ctxSynPatU{\ctx}{\PAsc{\PMV}{\TMV}}{\TMV}
+  }{USPAnn}
 \end{mathpar}
@@
 \\
-\judgbox{\ensuremath{\ctxAnaPat{\ctx_{in}}{\PMV}{\TMV}{\ctx_{out}}}} Pattern $\PMV$ analyzes against type $\TMV$ producing context $\Gamma_{out}$
+\judgbox{\ensuremath{\ctxAnaPatU{\ctx_1}{\PMV}{\TMV}{\ctx_2}}} $\PMV$ analyzes against type $\TMV$ producing context $\ctx_2$
 %
 \begin{mathpar}
-   \judgment{ }{\ctxAnaPat{\ctx}{\_}{\TMV}{\ctx}}{UAPatWild}
+  \judgment{ }{
+    \ctxAnaPatU{\ctx}{\PWild}{\TMV}{\ctx}
+  }{UAPWild}
 
-   \judgment{ }{\ctxAnaPat{\Gamma}{x}{\TMV}{\ctx, x:\TMV}}{UAPatVar}
-
-   %\judgment{ }{\ctxAnaPat{\ctx}{\hehole{}}{\tau}}{UAPatHole}{\ctx}
-   \judgment{
-     \ctxAnaPat{\ctx}{\PMV{}}{\tau'}{\ctx'}
-     \\
-     \consistent{\tau}{\tau'}
-   }{
-        \ctxAnaPat{\ctx}{\PMV : \tau'}{\tau}{\ctx'}
-    }{UAPatAnn}
+  \judgment{ }{
+    \ctxAnaPatU{\ctx}{\PVar{x}}{\TMV}{\extendCtx{\ctx}{x}{\TMV}}
+  }{UAPVar}
 
   \judgment{
-      \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\\\
-      \ctxAnaPat{\ctx}{\PMV_1}{\TMV_1}{\ctx_1} \\\\
-      \ctxAnaPat{\ctx_1}{\PMV_2}{\TMV_2}{\ctx_2}
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
+    \ctxAnaPatU{\ctx}{\PMV_1}{\TMV_1}{\ctx_1} \\
+    \ctxAnaPatU{\ctx_1}{\PMV_2}{\TMV_2}{\ctx_2}
   }{
-      \ctxAnaPat{\ctx}{\ECPair{\PMV_1}{\PMV_2}}{\TMV}{\ctx_2}
-    }{UAPatPair}
+    \ctxAnaPatU{\ctx}{\PPair{\PMV_1}{\PMV_2}}{\TMV}{\ctx_2}
+  }{UAPPair}
+
+  \judgment{
+    \ctxAnaPatU{\ctx}{\PMV}{\TMV'}{\ctx'} \\
+    \consistent{\TMV}{\TMV'}
+  }{
+    \ctxAnaPatU{\ctx}{\PAsc{\PMV}{\TMV'}}{\TMV}{\ctx'}
+  }{UAPAnn}
 \end{mathpar}
@@
 \\
-  \judgbox{\ensuremath{\ctxSynFixedInto{\ctx_{in}}{\PMV}{\PCMV}{\TMV}}} Pattern $\PMV$ synthesizes $\TMV$ producing marked $\PCMV$ \\
-  \judgbox{\ensuremath{\ctxAnaFixedIntoPat{\ctx_{in}}{\PMV}{\PCMV}{\TMV}{\ctx_{out}}}} Pattern $\PMV$ analyzes against $\TMV$ producing marked $\PCMV$ and context $\Gamma_{out}$
+\judgbox{\ensuremath{\ctxSynFixedIntoPat{\ctx}{\PMV}{\PCMV}{\TMV}}} $\PMV$ is marked into $\PCMV$ and synthesizes $\TMV$
+\judgbox{\ensuremath{\ctxAnaFixedIntoPat{\ctx_1}{\PMV}{\PCMV}{\TMV}{\ctx_2}}} $\PMV$ is marked into $\PCMV$ and analyzes against $\TMV$ producing $\ctx_2$
    %
 \begin{mathpar}
-   \judgment{ }{\ctxSynFixedInto{\ctx}{\_}{\_}{\TUnknownSwitch}}{ISPatWild}
+  \judgment{ }{
+    \ctxSynFixedInto{\ctx}{\PWild}{\PCWild}{\TUnknownSwitch}
+  }{ISPWild}
 
-   \judgment{ }{\ctxSynFixedInto{\ctx}{x}{x}{\TUnknownSwitch}}{ISPatVar}
+  \judgment{ }{
+    \ctxSynFixedInto{\ctx}{\PVar{x}}{\PCVar{x}}{\TUnknownSwitch}
+  }{ISPVar}
 
@@
   }{
-      \ctxSynFixedInto{\ctx}{\EPair{\PMV_1}{\PMV_2}}{\ECPair{\PCMV_1}{\PCMV_2}}{\TPair{\TMV_1}{\TMV_2}}
-   }{ISPatPair}
+     \ctxSynFixedInto{\ctx}{\EPair{\PMV_1}{\PMV_2}}{\ECPair{\PCMV_1}{\PCMV_2}}{\TProd{\TMV_1}{\TMV_2}}
+  }{ISPPair}
 
   \judgment{
-     \ctxAnaPat{\ctx}{\PMV{}}{\TMV}{\_} \\
-     \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\TMV}{\_}
+    \ctxAnaPatU{\ctx}{\PMV{}}{\TMV}{\ctx'} \\
+    \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\TMV}{\ctx''}
   }{
-        \ctxSynFixedInto{\ctx}{\PMV : \tau}{\PCMV : \tau}{\tau}
-    }{ISPatAnn}
+    \ctxSynFixedInto{\ctx}{\PAsc{\PMV}{\TMV}}{\PCAsc{\PCMV}{\TMV}}{\tau}
+  }{ISPAnn1}
 
   \judgment{
-     \ctxNotAnaPat{\ctx}{\PMV{}}{\TMV}{\_} \\
-     \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\TUnknown}{\_}
+    \ctxNotAnaPatU{\ctx}{\PMV{}}{\TMV}{\ctx'} \\
+    \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\TUnknown}{\ctx''}
   }{
-        \ctxSynFixedInto{\ctx}{\PMV : \TMV}{\ECInconType{\PCMV} : \TMV}{\TMV}
-    }{ISPatAnnFail}
-   \\\\
-   \judgment{ }{\ctxAnaFixedIntoPat{\ctx}{\_}{\_}{\TMV}{\ctx}}{IAPatWild}
+    \ctxSynFixedInto{\ctx}{\PAsc{\PMV}{\TMV}}{\PCAsc{\PCInconType{\PCMV}}{\TMV}}{\TMV}
+  }{ISPAnn2}
 
-   \judgment{ }{\ctxAnaFixedIntoPat{\ctx}{x}{x}{\TMV}{\ctx, x:\TMV}}{IAPatVar}
-   \\\\
+  \judgment{ }{
+    \ctxAnaFixedIntoPat{\ctx}{\PWild}{\PCWild}{\TMV}{\ctx}
+  }{IAPWild}
+
+  \judgment{ }{
+    \ctxAnaFixedIntoPat{\ctx}{\PVar{x}}{\PCVar{x}}{\TMV}{\extendCtx{\ctx}{x}{\TMV}}
+  }{IAPVar}
+
   \judgment{
-     \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\\\
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
     \ctxAnaFixedIntoPat{\ctx}{\PMV_1}{\PCMV_1}{\TMV_1}{\ctx_1} \\\\
@@
   }{
-      \ctxAnaFixedIntoPat{\ctx}{\EPair{\PMV_1}{\PMV_2}}{\ECPair{\PCMV_1}{\PCMV_2}}{\TMV}{\ctx_2}
-   }{IAPatPair}
+    \ctxAnaFixedIntoPat{\ctx}{\PPair{\PMV_1}{\PMV_2}}{\PCPair{\PCMV_1}{\PCMV_2}}{\TMV}{\ctx_2}
+  }{IAPPair1}
 
   \judgment{
-     \notMatchedPair{\TMV} \\\\
+    \notMatchedProd{\TMV} \\
     \ctxAnaFixedIntoPat{\ctx}{\PMV_1}{\PCMV_1}{\TUnknown}{\ctx_1} \\\\
@@
   }{
-      \ctxAnaFixedIntoPat{\ctx}{\EPair{\PMV_1}{\PMV_2}}{\ECInconMatchedPair{\ECPair{\PCMV_1}{\PCMV_2}}}{\TMV}{\ctx_2}
-   }{IAPatPairFail}
+    \ctxAnaFixedIntoPat{\ctx}{\PPair{\PMV_1}{\PMV_2}}{\PCAnaNonMatchedProd{\PCPair{\PCMV_1}{\PCMV_2}}}{\TMV}{\ctx_2}
+  }{IAPPair2}
 
   \judgment{
-     \ctxAnaPat{\ctx}{\PMV{}}{\tau'}{\ctx'} \\
-     \consistent{\tau}{\tau'} \\
-     \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\tau'}{\_}
+    \consistent{\TMV}{\TMV'} \\
+    \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\TMV'}{\ctx'}
   }{
-        \ctxAnaFixedIntoPat{\ctx}{\PMV : \tau'}{\PCMV : \tau'}{\tau}{\ctx'}
-   }{IAPatAnn}
+    \ctxAnaFixedIntoPat{\ctx}{\PAsc{\PMV}{\TMV'}}{\PCAsc{\PCMV}{\TMV'}}{\TMV}{\ctx'}
+  }{IAPAnn1}
 
   \judgment{
-     \ctxSynFixedInto{\ctx}{\PMV}{\PCMV}{\tau'}
-     \\
-     \inconsistent{\tau}{\tau'}
+    \inconsistent{\TMV}{\TMV'} \\
+    \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\TMV'}{\ctx'}
   }{
-        \ctxAnaFixedIntoPat{\ctx}{\PMV}{\ECInconType{\PCMV}}{\tau}{\ctx}
-    }{IAInconsistentTypesPat}
-
-   %\judgment{
-   %  \ctxNotAnaPat{\ctx}{\PMV{}}{\tau}{\ctx'} \\
-   %  \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\tau}{\_}
-   %}{
-   %     \ctxAnaFixedIntoPat{\ctx}{\PMV : \tau}{\ECInconType{\PCMV} : %\tau}{\tau}{\ctx'}
-   %}{IAPatAnnFail1}
-
-   %\judgment{
-   %  \ctxAnaPat{\ctx}{\PMV{}}{\tau}{\ctx'} \\
-   %  \inconsistent{\tau}{\tau'} \\
-   %  \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\tau}{\_}
-   %}{
-   %     \ctxAnaFixedIntoPat{\ctx}{\PMV : \tau'}{\ECInconType{\PCMV : %\tau'}}{\tau}{\ctx'}
-   %}{IAPatAnnFail2}
-
-   %\judgment{
-   %  \ctxNotAnaPat{\ctx}{\PMV{}}{\tau}{\ctx'} \\
-   %  \inconsistent{\tau}{\tau'} \\
-   %  \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\tau}{\_}
-   %}{
-   %     \ctxAnaFixedIntoPat{\ctx}{\PMV : \tau'}%{\ECInconType{\ECInconType{\PCMV} : \tau'}}{\tau}{\ctx'}
-   %}{IAPatAnnFail3}
-
+    \ctxAnaFixedIntoPat{\ctx}{\PAsc{\PMV}{\TMV'}}{\PCInconType{\PCAsc{\PCMV}{\TMV'}}}{\TMV}{\ctx'}
+  }{IAPAnn2}
 \end{mathpar}
   %
-  \caption{Pattern Marking}
-  \label{fig:calculus-pattern-synthesis-marking}
+  % \caption{Pattern Marking}
+  % \label{fig:calculus-pattern-synthesis-marking}
 
@@
 
+\judgbox{\ensuremath{\ctxSynPatM{\ctx}{\PCMV}{\TMV}}} $\PCMV$ synthesizes type $\TMV$
+%
 \begin{mathpar}
+  \judgment{ }{
+    \ctxSynPatM{\ctx}{\PCWild}{\TUnknownSwitch}
+  }{MSPWild}
 
-   \judgment{ }{\ctxSynPat{\ctx}{\_}{\TUnknownSwitch}}{MSPatWild}
-
-   \judgment{ }{\ctxSynPat{\ctx}{x}{\TUnknownSwitch}}{MSPatVar}
+  \judgment{ }{
+    \ctxSynPatU{\ctx}{\PCVar{x}}{\TUnknownSwitch}
+  }{MSPVar}
 
   \judgment{
-     \ctxAnaPat{\ctx}{\PCMV{}}{\tau}{\_}
+    \ctxSynPatM{\ctx}{\PCMV_1}{\TMV_1} \\
+    \ctxSynPatM{\ctx}{\PCMV_2}{\TMV_2}
   }{
-        \ctxSynPat{\ctx}{\PCMV : \tau}{}{\tau}
-    }{MSPatAnn}
+    \ctxSynPatM{\ctx}{\PCPair{\PCMV_1}{\PCMV_2}}{\TProd{\TMV_1}{\TMV_2}}
+  }{MSPPair}
 
   \judgment{
-     \ctxSynPat{\ctx}{\PCMV_1}{\TMV_1} \\
-     \ctxSynPat{\ctx}{\PCMV_2}{\TMV_2}
+    \ctxAnaPatM{\ctx}{\PCMV}{\TMV}{\ctx'}
   }{
-     \ctxSynPat{\ctx}{\EPair{\PCMV_1}{\PCMV_2}}{\TPair{\TMV_1}{\TMV_2}}
-   }{MSPatPair}
+    \ctxSynPatM{\ctx}{\PCAsc{\PCMV}{\TMV}}{\TMV}
+  }{MSPAnn}
+\end{mathpar}
 
-   \judgment{ }{\ctxAnaPat{\ctx}{\_}{\TMV}{\ctx}}{MAPatWild}
+\judgbox{\ensuremath{\ctxAnaPatM{\ctx_1}{\PCMV}{\TMV}{\ctx_2}}} $\PCMV$ analyzes against type $\TMV$ producing context $\ctx_2$
+%
+\begin{mathpar}
+  \judgment{ }{
+    \ctxAnaPatU{\ctx}{\PCWild}{\TMV}{\ctx}
+  }{MAPWild}
 
-   \judgment{ }{\ctxAnaPat{\Gamma}{x}{\TMV}{\ctx, x:\TMV}}{MAPatVar}
+  \judgment{ }{
+    \ctxAnaPatU{\ctx}{\PCVar{x}}{\TMV}{\extendCtx{\ctx}{x}{\TMV}}
+  }{MAPVar}
 
   \judgment{
-     \ctxAnaPat{\ctx}{\PCMV{}}{\tau'}{\ctx'}
-     \\
-     \consistent{\tau}{\tau'}
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
+    \ctxAnaPatM{\ctx}{\PCMV_1}{\TMV_1}{\ctx_1} \\\\
+    \ctxAnaPatM{\ctx_1}{\PCMV_2}{\TMV_2}{\ctx_2}
   }{
-        \ctxAnaPat{\ctx}{\PCMV : \tau'}{\tau}{\ctx'}
-    }{MAPatAnn}
+    \ctxAnaPatM{\ctx}{\PCPair{\PCMV_1}{\PCMV_2}}{\TMV}{\ctx_2}
+  }{MAPPair1}
 
   \judgment{
-     \ctxSynPat{\ctx}{\PCMV{}}{\tau'}
-     \\
-     \inconsistent{\tau}{\tau'}
+    \notMatchedProd{\TMV} \\
+    \ctxAnaPatM{\ctx}{\PCMV_1}{\TUnknown}{\ctx_1} \\\\
+    \ctxAnaPatM{\ctx_1}{\PCMV_2}{\TUnknown}{\ctx_2}
   }{
-        \ctxAnaPat{\ctx}{\ECInconType{\PCMV}}{\tau}{\ctx}
-    }{MAInconsistentTypes}
+    \ctxAnaPatM{\ctx}{\PCAnaNonMatchedProd{\PCPair{\PCMV_1}{\PCMV_2}}}{\TMV}{\ctx_2}
+  }{MAPPair2}
 
   \judgment{
-      \matchedPair{\TMV}{\TMV_1}{\TMV_2} \\
-      \ctxAnaPat{\ctx}{\PCMV_1}{\TMV_1}{\ctx_1} \\
-      \ctxAnaPat{\ctx_1}{\PCMV_2}{\TMV_2}{\ctx_2}
+    \consistent{\TMV}{\TMV'} \\
+    \ctxAnaPatM{\ctx}{\PCMV}{\TMV'}{\ctx'}
   }{
-      \ctxAnaPat{\ctx}{\ECPair{\PCMV_1}{\PCMV_2}}{\TMV}{\ctx_2}
-    }{MAPatPair}
+    \ctxAnaPatM{\ctx}{\PCAsc{\PCMV}{\TMV'}}{\TMV}{\ctx'}
+  }{MAPAnn1}
 
   \judgment{
-       \notMatchedPair{\TMV} \\
-      \ctxAnaPat{\ctx}{\PCMV_1}{\TUnknown}{\ctx_1} \\
-      \ctxAnaPat{\ctx_1}{\PCMV_2}{\TUnknown}{\ctx_2}
+    \inconsistent{\TMV}{\TMV'} \\
+    \ctxAnaPatM{\ctx}{\PCMV}{\TMV'}{\ctx'}
   }{
-      \ctxAnaPat{\ctx}{\ECInconMatchedPair{\ECPair{\PCMV_1}{\PCMV_2}}}{\TMV}{\ctx_2}
-    }{MAPatPairFail}
+    \ctxAnaPatM{\ctx}{\PCInconType{\PCAsc{\PCMV}{\TMV'}}}{\TMV}{\ctx'}
+  }{MAPAnn2}
 \end{mathpar}
   %
-  \caption{Marked Pattern Typing}
-  \label{fig:calculus-pattern-marked-typing}
+  % \caption{Marked Pattern Typing}
+  % \label{fig:calculus-pattern-marked-typing}

### 2023-03-02 calculus.tex
Describe reified hazelnut action semantics

@@
   \judgment{
-    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\\\
+    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TNum} \\
     \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TNum}
@@
   \judgment{
-    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TNum} \\\\
+    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TNum} \\
     \ctxAnaTypeM{\ctx}{\ECMV_2}{\TNum}
@@
     \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV} \\
-    \notMatchedArrow{\TMV} \\\\
+    \notMatchedArrow{\TMV} \\
     \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
@@
     \ctxSynTypeM{\ctx}{\ECMV_1}{\TMV} \\
-    \notMatchedArrow{\TMV} \\\\
+    \notMatchedArrow{\TMV} \\
     \ctxAnaTypeM{\ctx}{\ECMV_2}{\TUnknown}
@@
   \judgment{
-    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\
+    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\\\
     \matchedProd{\TMV}{\TMV_1}{\TMV_2}
@@
     \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
-    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TMV_1} \\
+    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TMV_1} \\\\
     \ctxAnaTypeM{\ctx}{\ECMV_2}{\TMV_2}
@@
 
-Hazelnut, however, is unable to cope with actions that require non-local hole fixes. % TODO example of this
-Such actions remain undefined, i.e. unperformable.
+Hazelnut, however, is unable to cope with actions that require \emph{non-local} hole fixes. For
+example, though the system can wrap an arbitrary expression in an arithmetic operation, it cannot
+wrap an arbitrary expression $\EMV$ in a lambda abstraction, i.e. construct
+$\ELam{x}{\TUnknown}{\EMV}$ directly from $\EMV$. The binding of $x$ with the unknown type may
+shadow a previous declaration and require the removal of error holes anywhere in body, but Hazelnut
+may only perform such fixes for immediately local holes. Such actions remain undefined, i.e.
+unperformable.
 
@@
 manipulation of syntax, and the marking procedure yields statically meaningful terms with error
-holes inserted at the appropriate positions. Furthermore, marking solves the problem of non-local
-hole fixes; actions may be performed anywhere (as long as they are of the correct sort), and marking
-totality ensures that they give meaningful terms.
-% TODO this is not strictly true, only for 'wrapping' actions
+marks, now error holes, inserted at the appropriate positions as a kind of editor decoration. This
+solves the problem outlined above; since marking operates on the entire program after each action,
+arbitrary constructions are permitted.
+
+Instead of a wholly untyped action calculus, one may re-adapt the existing Hazelnut calculus with
+scope-based marking. Action operate over \emph{marked expressions}, and error marks operate as error
+holes. When wrapping $\ECMV$ into $\ECLam{x}{\TUnknown}{\ECMV}$, we may re-mark the body under the
+context with $x$ binding type $\TUnknown$:
+\[%
+  \judgment{
+    \ctxSynFixedInto{\extendCtx{\ctx}{x}{\TUnknown}}{\erase{\ECMV}}{\ECMV'}{\TMV'}
+  }{
+    \ASEConLam{\ctx}{\ECMV}{\TMV}{\ECMV'}{\TMV'}{x}
+  }{ASEConLam}
+\]%
+This system optimizes over the untyped version by performing marking only when necessary. In
+particular, whenever binders are introduced around an expression, marking redraws error holes. The
+supplemental material contains a complete formal description of both variants, and the untyped
+action semantics and related metatheory are mechanized in Agda.
+% TODO note that we have not verified the metatheory for the typed system

### 2023-03-02 calculus.tex
Add discussion of alternative conditional rules

@@
 
-The boolean values are straightforward to handle:
+The boolean values are straightforward to handle, behaving in a similar manner as numbers: 
 %
@@
 Conditionals, however, offer an interesting case. In the unmarked language, we have both explicit
-synthetic and analytic rules: % TODO intuition why
+synthetic and analytic rules:
 %
@@
 
-These rules complete the development of the marked lambda calculus for the gradually typed lambda
-calculus.
+The framework also allows for other localization approaches. For example, we may regard the first
+branch as ``correct'' and localize any errors to the second. The following rule formalizes such a
+design:
+\[%
+  \judgment{
+    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
+    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV} \\
+    \ctxAnaFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV}
+  }{
+    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV}
+  }{MKSIf'}
+\]%
+Of course, corresponding unmarked and marked typing rules would follow similar structure, and other
+heuristics may be devised for different kinds of branching structures.
 
@@
 
+\subsubsection{Additional Theorems}
 To conclude this section, we present two more metatheorems that help ensure correctness of the

### 2023-03-02 calculus.tex
Add section on marking holes

@@
 \Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
-lambda calculus, extending it with binary product types in \cref{sec:calculus-products}. All rules
-and theorems may be found in the supplementary appendix, alongside a complete mechanization in the
-Agda proof assistant, which is discussed briefly in \cref{sec:calculus-agda}. With products, we then
-consider how error localization is complicated by patterned let expressions in
+lambda calculus (GTLC), extending it with binary product types in \cref{sec:calculus-products}. All
+rules and theorems may be found in the supplementary appendix, alongside a complete mechanization in
+the Agda proof assistant, which is discussed briefly in \cref{sec:calculus-agda}. With products, we
+then consider how error localization is complicated by patterned let expressions in
 \cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how the system might be extended
@@
 
-We now formally introduce the marked lambda calculus through its application to a gradually typed
-lambda calculus extended with numbers and booleans. Consider the syntax for such a language, given
-as $\EMName$ in \cref{fig:calculus-syntax}. The base types $\TNum$ and $\TBool$ classify number and
-boolean expressions. The number literal corresponding to the mathematical number $n$ is given by
-$\ENumMV$, and there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$
-correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
+We now formally introduce the marked lambda calculus through its application to the GTLC extended
+with numbers and booleans. Consider the syntax for such a language, given as $\EMName$ in
+\cref{fig:calculus-syntax}. The base types $\TNum$ and $\TBool$ classify number and boolean
+expressions. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$,
+and there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$ correspond
+to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
 $\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure. In addition to these forms,
-$\TUnknown$ is the unknown type from gradual type theory. $\EEHole$ denotes an \emph{expression
-hole}, which are used to model syntactically incomplete programs---but are not critical within the
-calculus.
+$\TUnknown$ is the unknown type from gradual type theory. Finally, $\EEHole$ denotes an
+\emph{expression hole}, which are used to model syntactically incomplete programs. Note, however,
+that these are not critical to the calculus.
 
@@
   \end{array}\]
-  \caption{Syntax of an extended gradually typed lambda calculus.}
+  \caption{Syntax of an extended GTLC.}
   \label{fig:calculus-syntax}
@@
 %
-Throughout the remainder of this section, as we formulate marking for the gradually typed lambda
-calculus, we will motivate and give precise semantics for these error marks. Note that this is a
-non-exhaustive list; depending on the syntactic forms and typing features found in the language,
-other kinds of marks may be necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of
-marks that may be required in an language with explicit polymorphism.
+Throughout the remainder of this section, as we formulate marking for the GTLC, we will motivate and
+give precise semantics for these error marks. Note that this is a non-exhaustive list; depending on
+the syntactic forms and typing features found in the language, other kinds of marks may be
+necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of marks that may be required
+in an language with explicit polymorphism.
 
@@
 
-To start, let us consider the simple case of numbers. Because of \emph{subsumption}, which we
-discuss next, we need only define a synthesis rule for unmarked numbers, in which numbers synthesize
-the type $\TNum$:
+To start, let us consider the simple case of numbers. Because of \emph{subsumption}, which is
+discussed in more detail next, we need only define a synthesis rule for unmarked numbers, in which
+number literals synthesize the type $\TNum$:
 %
@@
       \matchedArrow{\TUnknown}{\TUnknown}{\TUnknown}
-    }{TMAHole}
+    }{TMAUnknown}
 
@@
 heuristics may be devised for different kinds of branching structures.
+
+\subsubsection{Holes}
+For completeness, we finish the development of the marked lambda calculus on the GTLC with the
+marking of holes. Like number literals and booleans, these are governed by straightforward rules.
+%
+\begin{mathpar}
+  \judgment{ }{
+    \ctxSynTypeU{\ctx}{\EEHole}{\TUnknown}
+  }{USHole}
+
+  \judgment{ }{
+    \ctxSynFixedInto{\ctx}{\EEHole}{\ECEHole}{\TUnknown}
+  }{MKSHole}
+
+  \judgment{ }{
+    \ctxSynTypeM{\ctx}{\ECEHole}{\TUnknown}
+  }{MSHole}
+\end{mathpar}
 
@@
 
-% TODO non-gradual semantics?
-
-% Note that while the semantics that follow concern a gradually typed language, a non-gradual system
-% may be recovered straightforwardly by removing the unknown type and replacing type consistency with
-% equality.
-
 \subsection{Adding Binary Products}
@@
       \matchedProd{\TUnknown}{\TUnknown}{\TUnknown}
-    }{TMPHole}
+    }{TMPUnknown}
 

### 2023-03-02 calculus.tex
Finish discussion of lambda error cases

@@
 may arise. First, what if $\TMV_3$ is not a matched arrow type? This would be the case if it were
-$\TNum$, for example. The lambda abstraction's synthesized type would be inconsistent with the
-expected type, but this case is slightly distinct from that of the inconsistent types mark;
-
-Thus,
-to the syntax we add an \emph{inconsistent type mark} $\ECInconType{\ECMV}$, which indicates that
-the expression does not synthesize a type consistent with the expected type. We then have the
-following rules, in which the lambda abstraction is placed inside the inconsistent type mark.
+$\TNum$, for example. The lambda abstraction's synthesized type would indeed be inconsistent with
+the expected type, but this case is slightly distinct from that of the inconsistent types mark: the
+lambda expression is in analytic position. Instead, we mark $\ECAnaNonMatchedArrow{\ECMV}$ to
+indicate that an expression of the non-matched arrow type $\TMV_3$ was expected, but a lambda
+abstraction was encountered.
 %
@@
 
-A similar situation arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
+In this example language, the distinction from the inconsistent type mark is of reduced
+significance---all expressions synthesize a type. However, the addition of unannotated lambda
+abstractions, for example, would necessitate such a disctinction.
+
+Another error arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
 $\inconsistent{\TMV}{\TMV_1}$, i.e. the actual type annotation for $x$ is inconsistent with the

### 2023-03-02 intro.tex
Updates from Overleaf

@@
 These gaps in service are not always transient. 
-For example, a change to a type definition might result in type errors at dozens of use sites in a large program, which might take hours or days to resolve.
+For example, a change to a type definition might result in type errors at dozens of use sites in a large program, which might take hours or days to resolve, all without the full aid of these services.
 
-These service gaps are fundamentally rooted in a definitional gap: a type system defined in a conventional way, 
+These service gaps are fundamentally rooted in a definitional gap: a type system defined in the conventional way, 
 e.g. in the tradition of the typed lambda calculus and its derivatives \cite{TaplBook,pfpl},
@@
 This gap problem has prompted considerable practical interest in 
-(1)~\textbf{type error localization}: mechanisms for identifying the location(s) in a program that may be at fault for a type error, and 
+(1)~\textbf{type error localization}: mechanisms for identifying the location(s) in a program that explain a type error, and 
 (2) \textbf{type error recovery}: mechanisms that allow the language server to optimistically recover from a localized type error 
@@
 ideally at every location in the program and with minimal degradation in service.
-Essentially all widely-used programming systems have some support for type error localization, and developers have been observed to attend in particular to error locations as reported in compiler error messages
-or directly in their editor, via error markings decorating the localized code.\todo{cite}
+Essentially all widely-used programming systems have some support for type error localization, 
+reporting error locations in compiler error messages 
+or directly in the editor via markings decorating the localized errors. Developers attend to these error locations when debugging type errors.\todo{cite}
 Many systems also attempt recovery of certain services in certain situations, discussed below.
@@
 
-Consider, the example program in \autoref{fig:hazel-intro}, which is written in
+Consider, the example program in \autoref{fig:hazel-intro}, which is shown as presented to the user in
 this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{elm}. Hazel  
@@
 A type checker with no support for error localization or recovery---%
-as might be written by students in an undergraduate course---%
+as students in an undergraduate course might write---%
 would simply report that this program is ill-typed, 
@@
 
-A more practical approach, and a common one even in production type checkers, 
+A more practical approach, and one common even in production type checkers, 
 is to localize the first error that causes local type inference to fail and emit an explanatory error message.
@@
 To recover further, we might choose to also ignore the bidirectional type system's demand that the guard of the conditional be confirmed to be a boolean expression 
-(because the type of \li{f(y)} is also unknown) 
+(because the type of \li{f(y)} is also not well-defined) 
 and continue into its branches, observing that they have inconsistent types, \li{Int} and \li{Bool}. 
-There are several ways to localize this inconsistency. 
+There are several ways to localize this error. 
 One approach would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
@@
 or training a machine learning model \todo{cite}. 
-The other approach, which we take in this paper's language server for Hazel, is to localize the inconsistency to the conditional expression itself, reporting that the problem is that the branch types differ, as indicated in
+The other approach, which we take in Hazel, is to localize the inconsistency to the conditional expression itself, reporting that the problem is that the branch types differ, as indicated in
 the screenshot in \autoref{fig:hazel-intro}.
-Hazel reports typing information, including error messages, about the term at the cursor in its status bar, so when the cursor is on the conditional expression, the message below would be shown:
+When the cursor is on the conditional expression, the Hazel type inspector---a status bar that reports typing information, including type error messages, about the term at the cursor \cite{hannah1}---shows the following error:
 
-TODO
+TODO\todo{tpye inspector}
 
-Whichever localization decision is made affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
-If localization had assumed that the \li{then} branch is correct, as for example the Elm type checker does, then \li{x : Bool} and an error should be reported on the left operand (though this error might mislead the programmer if this earlier localization guess was incorrect).
-If the \li{else} branch is chosen, then \li{x : Int} and an error should not be reported. 
-If instead the inconsistency is localized to the conditional expression as a whole, as in our language server, then we again confront the problem of missing type information: 
+This localization decision affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
+If localization had assumed that the \li{then} branch correct, as for example in the Elm type checker, then \li{x : Bool} and an error should be reported on the left operand (though this error might mislead the programmer if this earlier localization guess was incorrect).
+If the \li{else} branch is chosen, then \li{x : Int} and an error would not be reported. 
+If the inconsistency is localized to the conditional expression as a whole, as in our language server, then we again confront the problem of missing type information: 
 \li{x} does not have a known type,
@@
 (2) when combined with error recovery, localization decisions can influence downstream localization decisions; and 
-(3) error recovery mechanisms must be able to reason without complete knowledge about types and binding.
+(3) error recovery mechanisms must be able to reason without complete knowledge about types and binding. We argue that such semantic subtleties call for a rigorous theoretical treatment of the topic.
 
@@
 We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
-type annotations to expressions \cite{pierce,BidirTyping}, (2) gradual typing, which offers a principled approach for handling missing type information during recovery \cite{GradualTyping,sieksnapl}\todo{siek05?}, and
-(3) non-empty holes, which allow us to operationalize type error localization by functioning as syntactic error marks, which an editor can display directly on-screen \cite{HazelnutPOPL}.
+type annotations to expressions \cite{pierce,BidirTyping}, (2) gradual typing, which offers a principled approach for recovering from missing type information  \cite{GradualTyping,sieksnapl}\todo{siek05?}, and
+(3) non-empty holes, which function as syntactic membranes marking erroneous terms, which the editor can display directly on-screen \cite{HazelnutPOPL}.
 
-These ideas were previously brought together in the Hazelnut calculus \cite{HazelnutPOPL}, but its account of type error localization and recovery was partial (see Sec.~\ref{sec:related}) and tied integrally into its structure editor.
-We resolve these problems to formally establish, with accompanying metatheory mechanized in Agda, that the marked lambda calculus has \emph{total error recovery}: \emph{every} syntactically well-formed program sketch (i.e.
-program structure with, optionally, empty holes) can be marked, i.e.
+The Hazelnut calculus \cite{HazelnutPOPL} previously brought these ideas together, but its account of type error localization and recovery was partial and tied integrally into its structure editor (see Sec.~\ref{sec:calculus-hazel}).
+We resolve these problems to formally establish, with accompanying metatheory mechanized in Agda, that the marked lambda calculus features \emph{total error recovery}: {every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{sketching}) can be marked, i.e.
 its errors can be localized, such that the resulting program has a well-defined type and binding structure.
-We consider a number of situations where error localization decisions are subtle, including conditionals and when combining pattern matching with explicit type annotations.
+As we define the calculus, we consider a number of situations where error localization decisions are subtle, including conditionals and destructuring let with granular type annotations.
 
-This system, scaled up and implemented in Hazel (Sec.~\ref{sec:calculus-hazel}), solves the semantic gap problem: Hazel's semantic services, which include type hints \cite{potter1}, semantic navigation, semantic code completion \cite{potter1,blinn}, contextualized documentation \cite{potter2}, and others,
-are available at all times and at all cursor locations, as long as the program sketch is syntactically well-formed. MENTION HAZELNUT LIVE?
+This system, scaled up and implemented in Hazel (Sec.~\ref{sec:calculus-hazel}), solves the semantic gap problem: Hazel's semantic services, which include type hints \cite{potter1}, semantic navigation, semantic code completion \cite{potter1,blinn}, contextualized documentation \cite{potter2}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{HazelnutLive}, even semantic services that require run-time information, like live testing,
+are available at all times and at all cursor locations, as long as the program sketch is syntactically well-formed.
 
-Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches: manual hole insertion, syntax error recovery, and structure editing. 
-Hazel has experimented with several of these approaches, offering (1) a textual syntax, which allows the programmer to manually insert empty holes as necessary, following similar features available in GHC Haskell, Agda, Idris, and other languages, (2) a text-like structure editor that offers partial syntax error recovery (inserting holes to maintain well-formedness as long as all matching delimiters have been entered), and (3) a term-based structure editor with total syntax error recovery, i.e.
-a guarantee that every edit maintains syntactic well-formedness.
-All of these editing methods, and their analogs in other program editors, can feed into the mechanisms described in Sec.~\ref{sec:calculus}.
+Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as is featured in Agda \cite{agda}, GHC Haskell \cite{haskell-holes}, Idris \cite{idris-holes}, and other languages), syntax error recovery \cite{error-reovery}, and structure editing \cite{HazelnutPOPL}.
+Hazel has experimented with several of these approaches, offering (1) a textual syntax, which allows the programmer to manually insert empty holes as necessary, (2) a text-like structure editor that offers partial syntax error recovery (inserting holes to maintain well-formedness as long as all matching delimiters have been placed) \cite{tylr}, and (3) a term-based structure editor with total syntax error recovery, i.e.
+a guarantee that every edit maintains syntactic well-formedness \cite{HazelnutPOPL}\todo{popl17}.
+The choice of editing method is independent of the mechanisms described in Sec.~\ref{sec:calculus}.
 However, as a secondary contribution,
-we demonstrate in Sec.~\ref{sec:calculus-structured-editing} that total type error recovery can resolve the problem of undefined editing behavior in the Hazelnut term-based structure editor calculus, namely that it does not allow edits that would require marking anywhere other than the cursor.
-We first reformulate Hazelnut as a purely syntactic structure editor, shifting re-marking to a separate whole-program pass after each edit.
-We also consider a more incremental approach, where remarking is driven by knowledge of the edit location, lexical scoping, and previous types.
-In either case, the result is a typed structure editor calculus that solves both the syntactic and semantic gap problem, i.e.
-every editor state is meaningful.
+we demonstrate in Sec.~\ref{sec:calculus-structured-editing} that total type error recovery can resolve the problem of undefined editing behavior in the Hazelnut term-based structure editor calculus, which arise because the system as formulated cannot handle edits that would require marking anywhere other than at the cursor.
+We discuss two reformulations of Hazelnut, first as an untyped structure editor where full re-marking occurs after each edit, then as a bidirectionally typed structure editor where re-marking occurs only where necessary based on the edit location and the type and binding structure.
 
-Local type inference makes a good foundation for type error localization and comprehensible error message generation because it specifies clear, local information flows.
-Languages with non-local, constraint-based type inference gather constraints from all locations in a program, which is often helpful, but these systems are also notorious for complicating type error localization and recovery, because inconsistencies can flow from any number of locations in the program,
-and it is not clear how to recover when conflicting constraints are found. 
+Local type inference pairs well with type error localization and recovery because information flows systematically down the tree, leading to predictable decisions (as observed also by \cite{pierce}).
+Languages with constraint-based type inference gather constraints from all relevant locations in a program.
+This approach is powerful, but it is also notorious for complicating type error localization and recovery, because inconsistencies can arise through a confluence of constraints originating from any number of locations in the program.
 Complex techniques, e.g. based on machine learning or weighting, are needed to heuristically localize errors to expressions. 
-In Hazel, we have taken a hybrid approach: 
-we generate constraints for each unknown type that appears in the program explicitly (i.e.
-type hole) and each unknown type generated internally during \emph{after} bidirectional type error localization and recovery.
-We develop a ?
-unification algorithm that can then generate possible hole fillings
-for each of thes...
-if the hole can be uniquely solved, then we suggest the solution to the programmer. 
-if conflicting constraints arise, then we localize this problem directly to type holes, rather 
-than attempting to localize the problem to expressions. 
-The user can select from different partial fillings -- the bidirectional system can then shift the error to relevant expressions in a more principled way.
-Until a choice is made, the system continues to recover gradually as explained in sec 2.
-Sec. 4.\todo{clean up text}
+In Sec.~\ref{sec:thi}\todo{sec}, we describe how we blend local and constraint-based type inference via type holes in Hazel. In particular, 
+we generate constraints for each type hole that appears in the program explicitly or that is  generated internally during the bidirectional type error localization and recovery process. 
+We develop a unification algorithm that can recover from inconsistent constraints by maintaining possible type hole fillings. For holes that can be uniquely solved, we suggest the solution. 
+If conflicting constraints arise, we localize the error directly to the relevant type holes, rather 
+than attempting to localize it to expressions. 
+The user can select from different partial fillings, which fills the type hole. From there, the bidirectional system shifts the error to relevant expressions.
+Until a choice is made, the system continues to recover gradually, as covered in Sec.~\ref{sec:calculus}.

### 2023-03-02 intro.tex
intro words finished maybe

@@
 
-Local type inference pairs well with type error localization and recovery because information flows systematically down the tree, leading to predictable decisions (as observed also by \cite{pierce}).
+Local type inference pairs well with type error localization and recovery because information flows systematically down the tree, leading to predictable and systematic localization decisions (as informally observed also by \cite{pierce}).
 Languages with constraint-based type inference gather constraints from all relevant locations in a program.
 This approach is powerful, but it is also notorious for complicating type error localization and recovery, because inconsistencies can arise through a confluence of constraints originating from any number of locations in the program.
-Complex techniques, e.g. based on machine learning or weighting, are needed to heuristically localize errors to expressions. 
+Complex techniques, e.g. based on machine learning or manual weighting, have been proposed to heuristically localize errors to expressions. 
 In Sec.~\ref{sec:thi}\todo{sec}, we describe how we blend local and constraint-based type inference via type holes in Hazel. In particular, 

### 2023-03-02 calculus.tex
Add some references through section 2

@@
 rules and theorems may be found in the supplementary appendix, alongside a complete mechanization in
-the Agda proof assistant, which is discussed briefly in \cref{sec:calculus-agda}. With products, we
-then consider how error localization is complicated by patterned let expressions in
+the Agda proof assistant \cite{norell:thesis} (discussed in \cref{sec:calculus-agda}). With
+products, we then consider how error localization is complicated by patterned let expressions in
 \cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how the system might be extended
@@
 judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$, establishes that, under
-the typing context $\ctx$, the expression $e$ synthesizes or infers the type $\TMV$. \emph{Type
-analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear
-where an expression of type $\TMV$ is expected. We may give these typing semantics traditionally,
-without considering error localization and recovery.
+the typing context $\ctx$, the expression $e$ synthesizes or locally infers \cite{Localinf} the type
+$\TMV$. \emph{Type analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$, states that the expression
+$\EMV$ may appear where an expression of type $\TMV$ is expected. We may give these typing semantics
+traditionally, without considering error localization and recovery.
 
@@
 $\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which establishes that $\TMV$ is a \emph{matched arrow type}
-[citation], i.e. it may be considered an arrow type. Defined in \cref{fig:calculus-matched-arrow},
-this notion is purely a technical mechanism to avoid duplication of rules related to arrow types.
+\cite{cimini:2016}, i.e. it may be considered an arrow type. Defined in
+\cref{fig:calculus-matched-arrow}, this notion is purely a technical mechanism to avoid duplication
+of rules related to arrow types.
 
@@
 converse---that at least one mark is inserted into ill-typed programs---is also true.
-% TODO latter is easier to do in Agda, but maybe it is weaker?
 
@@
 The semantics and metatheory presented above, including the extension with binary products, have
-been fully mechanized in the Agda proof assistant [citation]. Though the mechanization's
-documentation contains more detailed discussion regarding technical decisions made therein, we
-highlight some important aspects here.
+been fully mechanized in the Agda proof assistant. Though the mechanization's documentation contains
+more detailed discussion regarding technical decisions made therein, we highlight some important
+aspects here.
 

### 2023-03-02 calculus.tex
Include judgmental forms in hazelnut section

@@
 \emph{well-typed}, ensuring that semantic services are always available. To accomplish this, the
-action calculus inserts on-demand \emph{error holes}, denoted $\hhole{\EMV}{}$, which indicate type
-errors.
+bidirectional action calculus---given by the synthetic action judgment, written
+$\ASEAction{\ctx}{\EMV}{\TMV}{\EMV'}{\TMV'}{\AMV}$, and the analytic action judgment, written
+$\AAEAction{\ctx}{\EMV}{\EMV'}{\TMV}{\AMV}$---inserts on-demand \emph{error holes}, denoted
+$\hhole{\EMV}{}$, which indicate type errors.
 
@@
 By applying the marked lambda calculus to Hazelnut, the action calculus may be \emph{wholly decoupled}
-from the typing semantics. Under this new system, the action calculus is concerned only with the
-manipulation of syntax, and the marking procedure yields statically meaningful terms with error
-marks, now error holes, inserted at the appropriate positions as a kind of editor decoration. This
-solves the problem outlined above; since marking operates on the entire program after each action,
-arbitrary constructions are permitted.
-
-Instead of a wholly untyped action calculus, one may re-adapt the existing Hazelnut calculus with
-scope-based marking. Action operate over \emph{marked expressions}, and error marks operate as error
-holes. When wrapping $\ECMV$ into $\ECLam{x}{\TUnknown}{\ECMV}$, we may re-mark the body under the
-context with $x$ binding type $\TUnknown$:
+from the typing semantics. Under this new system, the action calculus, given by the singular action
+judgment, written $\AUEAction{\EMV}{\EMV'}{\AMV}$, is concerned only with the manipulation of
+syntax, and the marking procedure yields statically meaningful terms with error marks, now error
+holes, inserted at the appropriate positions as a kind of editor decoration. This solves the problem
+outlined above; since marking operates on the entire program after each action, arbitrary
+constructions are permitted.
+
+Alternatively, instead of a wholly untyped action calculus, one may re-adapt the existing Hazelnut
+calculus with scope-based marking. Action operate over \emph{marked expressions}, and error marks
+operate as error holes. The system is once again bidirectional, writing (note $\ECMV$ instead of
+$\EMV$) the judgments $\ASEAction{\ctx}{\ECMV}{\TMV}{\ECMV'}{\TMV'}{\AMV}$ and
+$\AAEAction{\ctx}{\ECMV}{\ECMV'}{\TMV}{\AMV}$. Then, when wrapping $\ECMV$ into
+$\ECLam{x}{\TUnknown}{\ECMV}$, we may simply re-mark the body under the context with $x$ binding
+the unknown type:
 \[%
@@
 action semantics and related metatheory are mechanized in Agda.
+
 % TODO note that we have not verified the metatheory for the typed system

### 2023-03-02 intro.tex

@@
 Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as is featured in Agda \cite{agda}, GHC Haskell \cite{haskell-holes}, Idris \cite{idris-holes}, and other languages), syntax error recovery \cite{error-reovery}, and structure editing \cite{HazelnutPOPL}.
-Hazel has experimented with several of these approaches, offering (1) a textual syntax, which allows the programmer to manually insert empty holes as necessary, (2) a text-like structure editor that offers partial syntax error recovery (inserting holes to maintain well-formedness as long as all matching delimiters have been placed) \cite{tylr}, and (3) a term-based structure editor with total syntax error recovery, i.e.
+Hazel has both a textual syntax and a structure editor, which are briefly described in Sec.~\ref{sec:calculus-hazel}.\todo{sec} 
+Marking is not integrated with structure editing, as it was in Hazelnut. 
+%offering (1) a textual syntax, which allows the programmer to manually insert empty holes as necessary, (2) a text-like structure editor that offers partial syntax error recovery (inserting holes to maintain well-formedness as long as all matching delimiters have been placed) \cite{tylr}, and (3) a term-based structure editor with total syntax error recovery, i.e.
 a guarantee that every edit maintains syntactic well-formedness \cite{HazelnutPOPL}\todo{popl17}.
-The choice of editing method is independent of the mechanisms described in Sec.~\ref{sec:calculus}.
 However, as a secondary contribution, 
-we demonstrate in Sec.~\ref{sec:calculus-structured-editing} that total type error recovery can resolve the problem of undefined editing behavior in the Hazelnut term-based structure editor calculus, which arise because the system as formulated cannot handle edits that would require marking anywhere other than at the cursor.
-We discuss two reformulations of Hazelnut, first as an untyped structure editor where full re-marking occurs after each edit, then as a bidirectionally typed structure editor where re-marking occurs only where necessary based on the edit location and the type and binding structure.
+we briefly discuss in Sec.~\ref{sec:calculus-structured-editing} how (1)  total marking can resolve the problem of undefined editing behavior in the Hazelnut structure editor calculus, 
+and (2) integrating marking with typed structure editing as specified by Hazelnut allows us to re-mark only where necessary based on the edit location and the type and binding structure.
 
@@
 This approach is powerful, but it is also notorious for complicating type error localization and recovery, because inconsistencies can arise through a confluence of constraints originating from any number of locations in the program.
-Complex techniques, e.g. based on machine learning or manual weighting, have been proposed to heuristically localize errors to expressions. 
+Complex techniques, e.g. based on machine learning or manual weighting, have been proposed to heuristically localize these errors to expressions. 
 In Sec.~\ref{sec:thi}\todo{sec}, we describe how we blend local and constraint-based type inference via type holes in Hazel. In particular,  
-we generate constraints for each type hole that appears in the program explicitly or that is  generated internally during the bidirectional type error localization and recovery process. 
+we generate constraints for each type hole that appears in the program explicitly or that is  generated internally during the  bidirectional type error localization and recovery process from Sec.~\ref{sec:calculus}.
 We develop a unification algorithm that can recover from inconsistent constraints by maintaining possible type hole fillings. For holes that can be uniquely solved, we suggest the solution. 
-If conflicting constraints arise, we localize the error directly to the relevant type holes, rather 
-than attempting to localize it to expressions. 
-The user can select from different partial fillings, which fills the type hole. From there, the bidirectional system shifts the error to relevant expressions.
-Until a choice is made, the system continues to recover gradually, as covered in Sec.~\ref{sec:calculus}.
+If conflicting constraints arise, we localize the errors to the relevant type holes, rather 
+than attempting to localize to expressions. 
+The user can explore different partial type hole fillings. Making a choice allows the bidirectional system to shift the error to relevant expressions.
+Until a choice is made, the system continues to recover gradually.

### 2023-03-02 calculus.tex
Edit text in relation to type error mosaic figure

@@
 environment. For example, a simple error that may occur is the use of an unbound variable. As in
-\cref{fig:calculus-examples-unbound}, editor services indicate the usage of unbound $y$ is invalid
-by highlighting it in red.
+\cref{fig:calculus-examples-unbound}, the editor indicates that the usage of an unbound $y$ is
+invalid by highlighting it in red.
 
@@
 types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error of inconsistent
-branches: the branches of the if-else expression are of different types. Hazel opts to highlight the
-entire expression in such a situation, while other tools may choose one branch to be ``correct'' and
-``blame'' the type mismatch error on the other.
+branches: the branches of the conditional are of different types: $5$ is a number, while
+$\textsf{false}$ is a boolean. Hazel opts to highlight the entire expression in such a situation,
+while other tools may choose one branch to be ``correct'' and ``blame'' the type mismatch error on
+the other.
 
@@
   \end{array}\]
-  \caption{Syntax of an extended GTLC.}
+  \caption{Syntax of an extended GTLC and its marked version.}
   \label{fig:calculus-syntax}
@@
 not quite right to mark $\EMV_1$ with the inconsistent type mark; rather than any single type, it is
-a member of the family of function types that is expected. For such a ``constrained'' synthetic
+any type within the family of arrow types that is expected. For such a ``constrained'' synthetic
 mode, we use a specialized mark $\ECSynNonMatchedArrow{\ECMV}$, which indicates that $\ECMV$ was
@@
 
-The framework also allows for other localization approaches. For example, we may regard the first
-branch as ``correct'' and localize any errors to the second. The following rule formalizes such a
-design:
+The framework also allows for other localization approaches. For example, as discussed at the
+beginning of the section, we may regard the first branch as ``correct'' and localize any errors to
+the second. The following rule formalizes such a design:
 \[%

### 2023-03-02 intro.tex

@@
 
-Consider, the example program in \autoref{fig:hazel-intro}, which is shown as presented to the user in
+Consider, the example program below, which is shown as presented to the user in
 this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{elm}. Hazel  
-supports local type inference, specified in the well-established bidirectional style \cite{pierce,hazelnut,BidirTyping}.
-\begin{figure}
+supports local type inference, specified in the well-established bidirectional style \cite{pierce,hazelnut,BidirTyping}:
+
+\begin{center}
     \includegraphics[scale=0.5]{images/hazel-intro-screenshot.png}
-    \caption{Type error localization and recovery in Hazel}
-    \label{fig:hazel-intro}
-\end{figure}
+\end{center}
 
@@
 as students in an undergraduate course might write---%
-would simply report that this program is ill-typed, 
-i.e. a type cannot be derived for it according to Hazel's type system.
-
+would simply report that this program is ill-typed. 
 A more practical approach, and one common even in production type checkers, 
@@
 
-To recover further, we might choose to also ignore the bidirectional type system's demand that the guard of the conditional be confirmed to be a boolean expression 
+To recover further, we might choose to also ignore the bidirectional type system's demand that the guard be confirmed to be a boolean expression 
 (because the type of \li{f(y)} is also not well-defined) 
@@
 
-This paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus} (Sec.~\ref{sec:calculus}).
+This paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus} (Sec.~\ref{sec:actions}).\todo{sec}
 We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
@@
 we briefly discuss in Sec.~\ref{sec:calculus-structured-editing} how (1)  total marking can resolve the problem of undefined editing behavior in the Hazelnut structure editor calculus, 
-and (2) integrating marking with typed structure editing as specified by Hazelnut allows us to re-mark only where necessary based on the edit location and the type and binding structure.
+and (2) integrating marking with typed structure editing allows us to re-mark only where necessary based on the edit location and the type and binding structure.
 
-Local type inference pairs well with type error localization and recovery because information flows systematically down the tree, leading to predictable and systematic localization decisions (as informally observed also by \cite{pierce}).
+Local type inference pairs well with type error localization and recovery because information flows systematically through the tree, leading to predictable and systematic localization decisions (as informally observed also by \cite{pierce}).
 Languages with constraint-based type inference gather constraints from all relevant locations in a program.
@@
 Complex techniques, e.g. based on machine learning or manual weighting, have been proposed to heuristically localize these errors to expressions. 
-In Sec.~\ref{sec:thi}\todo{sec}, we describe how we blend local and constraint-based type inference via type holes in Hazel. In particular,  
+In Sec.~\ref{sec:thi}\todo{sec}, we describe how we blend local and constraint-based type inference in Hazel. In particular,  
 we generate constraints for each type hole that appears in the program explicitly or that is  generated internally during the  bidirectional type error localization and recovery process from Sec.~\ref{sec:calculus}.
@@
 than attempting to localize to expressions. 
-The user can explore different partial type hole fillings. Making a choice allows the bidirectional system to shift the error to relevant expressions.
-Until a choice is made, the system continues to recover gradually.
+The user can explore the different partial type hole fillings. Making a choice then allows the bidirectional system to localize the error to relevant expressions.
+Until the choice is finalized, the system continues to recover gradually.

### 2023-03-02 intro.tex

@@
 or training a machine learning model \todo{cite}. 
-The other approach, which we take in Hazel, is to localize the inconsistency to the conditional expression itself, reporting that the problem is that the branch types differ, as indicated in
+The other approach, which Hazel takes, is to localize the inconsistency to the conditional expression itself, reporting that the problem is that the branch types differ, as indicated in
 the screenshot in \autoref{fig:hazel-intro}.
@@
 Until the choice is finalized, the system continues to recover gradually.
+
+let x : ? = ? in 
+if x then x(x + 1) else

### 2023-03-02 calculus.tex
Remove analytic projection rules
Cut down products section

@@
-% TODO standardize naming: marking vs mark insertion, marks vs error marks vs markings
-% TODO uniform style for referencing rules
-
 \section{The Marked Lambda Calculus}
@@
 \subsubsection{Holes}
-For completeness, we finish the development of the marked lambda calculus on the GTLC with the
-marking of holes. Like number literals and booleans, these are governed by straightforward rules.
+For completeness, we finish the development of the marked lambda calculus on the extended GTLC with
+the marking of holes. Like number literals and booleans, these are governed by straightforward
+rules.
 %
@@
 We now demonstrate how one might extend or adapt the marked lambda calculus for new typing features
-by adding binary product types. First, in \cref{fig:calculus-product-judgments}, we extend the
-syntax of types and the accompanying notions of consistency and join, which closely mirror the
-definitions for arrow types.
-%
-\[\begin{array}{rrcl}
-  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TProd{\TMV}{\TMV}
-\end{array}\]
-
-We additionally define an auxiliary notion of \emph{matched binary product types}: the judgment
-$\matchedProd{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be considered the binary product
-type $\TProd{\TMV_1}{\TMV_2}$.
-
-\begin{figure}[htbp]
-  \raggedright
-  \judgbox{\ensuremath{\consistent{\TMV_1}{\TMV_2}}} $\TMV_1$ is consistent with $\TMV_2$
-  %
-  \begin{mathpar}
-    \judgment{
-      \consistent{\TMV_1}{\TMV_1'} \\
-      \consistent{\TMV_2}{\TMV_2'} \\
-    }{
-      \consistent{\TProd{\TMV_1}{\TMV_2}}{\TProd{\TMV_1'}{\TMV_2'}}
-    }{TCProd} \\
-  \end{mathpar}
-
-  \judgbox{\ensuremath{\TJoin{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
-  %
-  \[\begin{array}{rcl}
-    & \cdots & \\
-    \joinsTo{(\TProd{\TMV_1}{\TMV_2})}{(\TProd{\TMV_1'}{\TMV_2'})}{\TProd{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}} \\
-  \end{array}\]
-
-  \judgbox{\ensuremath{\matchedProd{\TMV}{\TMV_1}{\TMV_2}}} $\TMV$ has matched binary product type $\TProd{\TMV_1}{\TMV_2}$
-  %
-  \begin{mathpar}
-    \judgment{ }{
-      \matchedProd{\TUnknown}{\TUnknown}{\TUnknown}
-    }{TMPUnknown}
-
-    \judgment{ }{
-      \matchedProd{\TProd{\TMV_1}{\TMV_2}}{\TMV_1}{\TMV_2}
-    }{TMPProd}
-  \end{mathpar}
-  %
-  \caption{Binary product consistency, join, and matched product type.}
-  \label{fig:calculus-product-judgments}
-\end{figure}
-
-The unmarked language is extended with a pair constructor and projection operators. In addition to
-the synthesis rules, we have explicit analysis rules. If an expected type is known for a pair
-expression, the responsibility for type errors is given to the components. Likewise, we shift blame
-to the projected expression when analysing projections.
-%
-\[\begin{array}{rrcl}
-  \EMName  & \EMV  & \Coloneqq & \cdots
-                               \mid \EPair{\EMV}{\EMV}
-                               \mid \EProjL{\EMV} \mid \EProjR{\EMV}
-\end{array}\]
-%
-\begin{mathpar}
-  \judgment{
-    \ctxSynTypeU{\ctx}{\EMV_1}{\TMV_1} \\
-    \ctxSynTypeU{\ctx}{\EMV_2}{\TMV_2}
-  }{
-    \ctxSynTypeU{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TProd{\TMV_1}{\TMV_2}}
-  }{USPair}
-
-  \judgment{
-    \ctxSynTypeU{\ctx}{\EMV}{\TMV} \\
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
-  }{
-    \ctxSynTypeU{\ctx}{\EProjL{\EMV}}{\TMV_1}
-  }{USProjL}
-
-  \judgment{
-    \ctxSynTypeU{\ctx}{\EMV}{\TMV} \\
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
-  }{
-    \ctxSynTypeU{\ctx}{\EProjR{\EMV}}{\TMV_2}
-  }{USProjR} \\
-
-  \judgment{
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
-    \ctxAnaTypeU{\ctx}{\EMV_1}{\TMV_1} \\
-    \ctxAnaTypeU{\ctx}{\EMV_2}{\TMV_2}
-  }{
-    \ctxAnaTypeU{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TMV}
-  }{UAPair}
-
-  \judgment{
-    \ctxAnaTypeU{\ctx}{\EMV}{\TProd{\TMV}{\TUnknown}}
-  }{
-    \ctxAnaTypeU{\ctx}{\EProjL{\EMV}}{\TMV}
-  }{UAProjL}
-
-  \judgment{
-    \ctxAnaTypeU{\ctx}{\EMV}{\TProd{\TUnknown}{\TMV}}
-  }{
-    \ctxAnaTypeU{\ctx}{\EProjR{\EMV}}{\TMV}
-  }{UAProjR}
-\end{mathpar}
-%
-Similarly to application, \textsc{USProjL} and \textsc{USProjR} employ the matched binary product
-type judgment; a projection $\EProjL{\EMV}$ or $\EProjR{\EMV}$ is then well-typed if the $\EMV$ may
-be considered to be a pair.
-
-The marked language is extended with the same forms, and we devise the following marking rules,
-again modelling off the above unmarked typing rules:
-%
-\[\begin{array}{rrcl}
-  \ECMName & \ECMV & \Coloneqq & \cdots
-                               \mid \ECPair{\ECMV}{\ECMV}
-                               \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
-                               \mid \ECSynNonMatchedProd{\ECMV}
-\end{array}\]
+by adding binary product types, written $\TProd{\TMV}{\TMV}$. In the interest of space, we focus on
+aspects related to marking; a comprehensive set of rules is found in the supplemental material.
+
+First, we extend the syntax of types and the accompanying notions of consistency and join, which
+closely mirror the definitions for arrow types. Additionally, analogous to the notion of matched
+arrow types, the judgment $\matchedProd{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be
+considered the binary product type $\TProd{\TMV_1}{\TMV_2}$ [cite?].
+
+The unmarked language is extended with a pair constructor $\EPair{\EMV}{\EMV}$ and projection
+operators $\EProjL{\EMV}$ and $\EProjR{\EMV}$, governed by standard bidirectional rules. In addition
+to the synthesis rules, an explicit analysis rules is defined for the former: if an expected type is
+known, the responsibility for type errors is shift to the components.
+
+The marked language is extended with the same forms, and we again devise marking rules by modelling
+off the unmarked typing rules.
 %
@@
   }{MKAPair1}
-
-  \judgment{
-    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TProd{\TMV}{\TUnknown}}
-  }{
-    \ctxAnaFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV}
-  }{MKAProjL}
-
-  \judgment{
-    \ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TProd{\TMV}{\TUnknown}}
-  }{
-    \ctxAnaFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV}
-  }{MKAProjR}
-\end{mathpar}
-%
-accompanied by corresponding typing rules for marked expressions:
-%
-\begin{mathpar}
-  \judgment{
-    \ctxSynTypeM{\ctx}{\ECMV_1}{\TMV_1} \\
-    \ctxSynTypeM{\ctx}{\ECMV_2}{\TMV_2}
-  }{
-    \ctxSynTypeM{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TProd{\TMV_1}{\TMV_2}}
-  }{MSPair}
-
-  \judgment{
-    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\\\
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
-  }{
-    \ctxSynTypeM{\ctx}{\ECProjL{\ECMV}}{\TMV_1}
-  }{MSProjL1}
-
-  \judgment{
-    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\\\
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
-  }{
-    \ctxSynTypeM{\ctx}{\ECProjR{\ECMV}}{\TMV_2}
-  }{MSProjR1}
-
-  \judgment{
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
-    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TMV_1} \\\\
-    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TMV_2}
-  }{
-    \ctxAnaTypeM{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
-  }{MAPair1}
-
-  \judgment{
-    \ctxAnaTypeM{\ctx}{\ECMV}{\TProd{\TMV}{\TUnknown}}
-  }{
-    \ctxAnaTypeM{\ctx}{\ECProjL{\ECMV}}{\TMV}
-  }{MAProjL}
-
-  \judgment{
-    \ctxAnaTypeM{\ctx}{\ECMV}{\TProd{\TUnknown}{\TMV}}
-  }{
-    \ctxAnaTypeM{\ctx}{\ECProjR{\ECMV}}{\TMV}
-  }{MAProjR}
 \end{mathpar}
@@
 necessary to satisfy totality. Firstly, the premise in \textsc{UAPair} that $\TMV$ is a matched
-binary product may fail, requiring marking and typing rules to handle that case:
+binary product may fail, requiring the introduction of a new error mark,
+$\ECAnaNonMatchedProd{\ECMV}$, which is analogous to $\ECAnaNonMatchedArrow{\ECMV}$ for lambda
+abstractions. Another kind of error may arise when the subject of a projection does not synthesize a
+matched product type. As in our development of marking for applications, we utilize a new mark,
+written $\ECSynNonMatchedProd{\ECMV}$. These cases give the final marking rules for products:
 %
@@
     \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPairAnaNonMatchedProd{\ECMV_1}{\ECMV_2}}{\TMV}
-  }{MKAPair2}
-
-  \judgment{
-    \notMatchedProd{\TMV} \\
-    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TUnknown} \\
-    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TUnknown}
-  }{
-    \ctxAnaTypeM{\ctx}{\ECPairAnaNonMatchedProd{\ECMV_1}{\ECMV_2}}{\TMV}
-  }{MAPair2}
-\end{mathpar}
-
-Another kind of error may arise when the subject of a projection does not synthesize a matched
-product type. As in our development of marking for applications, we utilize a new error mark
-$\ECSynNonMatchedProd{\ECMV}$, leading to the following rules:
-%
-\begin{mathpar}
+  }{MKAPair2} \\
+
   \judgment{
@@
   \judgment{
-    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\
-    \notMatchedProd{\TMV}
-  }{
-    \ctxSynTypeM{\ctx}{\ECProjLSynNonMatchedProd{\ECMV}}{\TUnknown}
-  }{MSProjL2} \\
-
-  \judgment{
     \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
@@
   }{MKSProjR2}
-
-  \judgment{
-    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\
-    \notMatchedProd{\TMV}
-  }{
-    \ctxSynTypeM{\ctx}{\ECProjRSynNonMatchedProd{\ECMV}}{\TUnknown}
-  }{MSProjR2}
 \end{mathpar}

### 2023-03-02 let.tex
Cut down products section
Cut down/clean up patterned let section

@@
-\subsection{Let-Expressions with Composite Destructuring Patterns}
+\subsection{Destructuring Let with Composite Patterns}
 \label{sec:calculus-let}
 
-%TODO(andrew): justification, examples
-
-In a traditional bidirectional presentation (CITE 'Strict Bidirectional Type Checking' Chlipala et al), let-expressions without composite patterns have two rules: A type-annotated version placing the definition in analytic position, and an unannotated synthetic version.
-With composite patterns, this straightforward division is no longer possible, as type constraints on the definition may depend on the (arbitrarily nested) sub-components of the pattern.
+To ease the use of products, many languages feature destructuring bindings [cite]. Let us add let
+expressions and simple patterns, writing $\PWild$ for the wildcard pattern, $\PVar{x}$ for a
+variable pattern, $\PPair{\PMV_1}{\PMV_2}$ for a pair. In a bidirectional setting, the pattern
+itself synthesizes a type which constrains the definition.
 %
@@
   \EMName  & \EMV  & \Coloneqq & \cdots \mid \ELet{\PMV}{\EMV}{\EMV} \\
-  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECLet{\PMV}{\ECMV}{\ECMV} \\
-  \PMName  & \PMV  & \Coloneqq & \PWild \mid \PVar{x} \mid \PPair{\PMV}{\PMV} \mid \PAsc{\PMV}{\TMV} \\
-  \PCMName & \PCMV & \Coloneqq & \PCWild \mid \PCVar{x} \mid \PCPair{\PCMV}{\PCMV} \mid \PCAsc{\PCMV}{\TMV}
-                               \mid \PCInconType{\PCMV} \mid \PCAnaNonMatchedProd{\PCMV}
+  \PMName  & \PMV  & \Coloneqq & \PWild \mid \PVar{x} \mid \PPair{\PMV}{\PMV} \mid \PAsc{\PMV}{\TMV}
 \end{array}\]
 %
-The most obvious approach is to instead use a single rule which synthesizes a type from the let-expression's pattern, then analyzes the definition against that type.
-However, this may run afoul of user expectation.
-Consider the expression $\ELet{(a,b)}{(1,2)}{e}$, which a user would reasonably expect to be semantically equivalent to the expanded expression $\ELet{a}{1}{\ELet{b}{2}{e}}$.
-Notice that in the un-expanded expression, $1$ and $2$ are in analytic against the unknown type, whereas in the latter, they are in synthetic position.
+It turns out, however, that these composite patterns with type annotations can complicate type error
+localization in subtle ways.
 
-One may ask if there a meaningful difference between these two situations.
-There is at least a subtle semantic distinction: Synthetic means 'no type constraints', whereas analysis against unknown imposes a trivial type constraint.
+Consider the following program:
+\[%
+  \ELet{\PPair{\PVar{a}}{\PVar{b}}}{\EPair{1}{2}}{\EMV}
+\]%
 
-This distinction manifests behaviorally in the case of expressions that can have internal type inconsistencies.
-As per the previous section on booleans, given the expression $\ELet{a}{\EIf{\ETrue}{1}{\EFalse}}{...}$, the conditional will be marked with an inconsistent branches error when in synthetic position, but if analyzed against the unknown type---with which both branches are consistent---no mark will be placed.
+To type this, the most obvious approach is to synthesize a type from the pattern and analyze the
+definition against that type. However, this may run afoul of user expectation, which might
+reasonably suppose it to be equivalent to the expanded expression
+$\ELet{\PVar{a}}{1}{\ELet{\PVar{b}}{2}{\EMV}}$. In the original, the pattern synthesizes the type
+$\TProd{\TUnknown}{\TUnknown}$, and $\EPair{1}{2}$ is analyzed against it, hence $1$ and $2$ are
+each analyzed against $\TUnknown$. In the expanded version, however, they are typed in synthetic
+mode.
 
-One option would be to erase the distinction entirely, adding a rule to switch to synthetic mode whenever we analyze against unknown, or to simply eliminate synthetic mode altogether.
-We consider these options too drastic; we wish to preserve the above behavioral distinction for conditionals, both indicating an inconsistency error in the unconstrained synthetic case, and permitting the user to defer that error via annotation by the unknown type.
-To this effect, we add a compositional pattern annotation form, which can be applied to sub-patterns:
-$\ELet{(a, b:\TUnknown)}{(\ECInconBr{\ETrue}{1}{\EFalse},\EIf{\ETrue}{2}{\EFalse})}{e}$.
-Here we want the first condition to be in synthetic and hence marked as inconsistent, and the second to be in analytic against unknown and thus unmarked.
-\[\begin{array}{rrcl}
-  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TUnknownSwitch \\
-\end{array}\]
+Though the distinction is benign in this example, there is a subtle semantic distinction: synthetic
+mode imposes \emph{no} type constraints, whereas analytic mode imposes a \emph{trivial} type
+constraint. This manifests when expressions may have internal type inconsistencies, such as in the
+following program:
+\[%
+  \ELet{\PVar{a}}{\EIf{\ETrue}{1}{\EFalse}}{\EMV}
+\]%
+If the pattern $\PVar{a}$ suggests that the conditional is in synthetic position, it will be marked
+with an inconsistent branches error following our development of marking above. If instead it is in
+analytic position against the unknown type, no mark will be produced.
 
-To accomplish this, we use a variation on the first distinction-erasure option, but with a twist.
-We do not unconditionally switch from analysis against unknown to synthesis, but instead augment our type system with a new variant on the unknown type which triggers the switch.
-The new type $\TUnknownSwitch$ behaves identically to $\TUnknown$ with respect to type consistency, the matched arrow and product judgements, and the join metafunction.
-In our previous example, the pattern variable $a$ will synthesize the $\TUnknownSwitch$ type, whereas the annotation $b: \TUnknown$ synthesizes $\TUnknown$.
-Thus $1$ will end up in synthetic position and $2$ in analytic, achieving the desired distinction.
+To remedy this situation and preserve the semantic distinction between synthesis and analysis
+against the unknown type, we introduce a pattern annotation form, written $\PAsc{\PMV}{\TMV}$, which
+allows the explicit imposition of typing constraints. That is, the absence of any type annotation on
+a variable pattern places the corresponding definition in synthetic mode, while the programmer may
+impose typing constraints---the trivial constraint, if they wish---on the definition.
+
+As illustration, consider the following program:
+\[%
+  \ELet{\PPair{\PVar{a}}{\PAsc{\PVar{b}}{\TUnknown}}}{\EPair{\EIf{\ETrue}{1}{\EFalse}}{\EIf{\ETrue}{2}{\EFalse}}}{\EMV}
+\]%
+Since $\PVar{a}$ has no constraint, the left component is in synthetic position, leading to an
+inconsistent branches error. The type annotation on $\PVar{b}$ puts the right component in analytic
+mode against the unknown type, leading to no error marks.
+
+This system is achieved by augmenting the type system with a variant of the unknown type, written
+$\TUnknownSwitch$, that triggers a ``switch'' to synthesis and otherwise behaves identically to
+$\TUnknown$. In the previous example, the unannotated pattern variable $\PVar{a}$ synthesizes
+$\TUnknownSwitch$, and the annotated pattern $\PAsc{\PVar{b}}{\TUnknown}$ synthesizes $\TUnknown$.
+We write $\ctxSynPatU{\ctx}{\PMV}{\TMV}$ to say that the pattern $\PMV$ synthesizes type $\TMV$.
+%
 \begin{mathpar}
@@
   }{UASynSwitch}
+
+  \judgment{ }{
+    \ctxSynPatU{\ctx}{\PVar{x}}{\TUnknownSwitch}
+  }{USPVar}
+
+  \judgment{
+    \ctxAnaPatU{\ctx}{\PMV}{\TMV}{\ctx'}
+  }{
+    \ctxSynPatU{\ctx}{\PAsc{\PMV}{\TMV}}{\TMV}
+  }{USPAnn}
 \end{mathpar}
 
-(It is worth noting that, despite the lack of marking for the $\TUnknown$-annotated conditional, there is no possible filling of the type hole, so it is also reasonable that the annotated unknown type itself should be considered as an error state.
-Indeed, precisely this will be accomplished in section 4 via type hole inference.)
+Given by the judgment $\ctxAnaPatU{\ctx}{\PMV}{\TMV}{\ctx'}$, patterns are also typed analytically,
+in which case they produce an output context $\ctx'$, which extends $\ctx$ with bindings introduced
+by the pattern. Note that $\TUnknownSwitch$ exists entirely to ensure that sub-expressions of the
+definition are assigned an appropriate typing mode---it is never added to the context, i.e. it
+cannot escape into the rest of the program and cause body expressions to synthesize accidentally.
+%
+\begin{mathpar}
+  \judgment{ }{
+    \ctxAnaPatU{\ctx}{\PVar{x}}{\TMV}{\extendCtx{\ctx}{x}{\TMV}}
+  }{UAPVar}
 
-\begin{mathpar}
+  \judgment{
+    \ctxAnaPatU{\ctx}{\PMV}{\TMV'}{\ctx'} \\
+    \consistent{\TMV}{\TMV'}
+  }{
+    \ctxAnaPatU{\ctx}{\PAsc{\PMV}{\TMV'}}{\TMV}{\ctx'}
+  }{UAPAnn}
+\end{mathpar}
+
+To ensure propagation of type information between pattern, definition, and body, the destructuring
+let requires some extra machinery.
+\[%
   \judgment{
@@
   }{USLetPat}
+\]%
+First, to ensure that the pattern and definition types are consistent, the pattern synthesizes a
+type, against which the definition is analyzed. Then, the definition synthesizes a type, against
+which the pattern is analyzed. This analysis is guaranteed to succeed; we need the context it
+produces for use in the synthesis of the body's type. The analytic rule is similar, with the final
+premise and conclusion changed to analysis.
 
-  \judgment{
-    \ctxSynPatU{\ctx}{\PMV}{\TMV_p} \\
-    \ctxAnaTypeU{\ctx}{\EMV_{1}}{\TMV_p} \\
-    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\\\
-    \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
-    \ctxAnaTypeU{\ctx'}{\EMV_{2}}{\TMV_2}
-  }{
-    \ctxAnaTypeU{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\TMV_2}
-  }{UALetPat}
-\end{mathpar}
-In order to ensure propagation of type information between the pattern and definition, we require a bit more machinery than a typical let rule.
-First, we synthesize the pattern type and analyze the definition against that type.
-These two steps are needed to ensure that the pattern and definition types are consistent.
-Then we synthesize the definition's type, and analyze the pattern against that type.
-This analysis of the pattern is guaranteed to succeed; however is is required, as we want the context produced by the pattern $\ctx_b$ to take into account the definition's type.
-Finally, the let body is synthesized or analyzed (depending on whether the let itself is being synthesized or analyzed) within that context.
-
-The marking rules are directly analogous:
-\begin{mathpar}
+The marking rule is directly analogous and utilizes the marking of patterns, which we omit---they
+may be derived in the same way as those for expressions. In the analytic cases, we introduce error
+marks that parallel to $\ECInconType{\ECMV}$ and $\ECAnaNonMatchedProd{\ECMV}$.
+\[%
   \judgment{
     \ctxSynFixedInto{\ctx}{\PMV}{\PCMV}{\TMV_p} \\
-    \ctxAnaFixedInto{\ctx}{\EMV_{1}}{\ECMV_{1}}{\TMV_{p}} \\\\
-    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\
+    \ctxAnaFixedInto{\ctx}{\EMV_{1}}{\ECMV_{1}}{\TMV_{p}} \\
+    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\\\
     \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
@@
   }{ISLetPat}
+\]%
 
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\PMV}{\PCMV}{\TMV_p} \\
-    \ctxAnaFixedInto{\ctx}{\EMV_{1}}{\ECMV_{1}}{\TMV_{p}} \\\\
-    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\
-    \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
-    \ctxAnaFixedInto{\ctx'}{\EMV_{2}}{\ECMV_{2}}{\TMV_2}
-  }{
-    \ctxAnaFixedInto{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\ELet{\PCMV}{\ECMV_{1}}{\ECMV_{2}}}{\TMV_2}
-  }{IALetPat}
-
-  \judgment{
-    \ctxSynPatM{\ctx}{\PCMV}{\TMV_p} \\
-    \ctxAnaTypeM{\ctx}{\ECMV_{1}}{\TMV_p} \\
-    \ctxSynTypeM{\ctx}{\ECMV_{1}}{\TMV_{1}} \\\\
-    \ctxAnaPatM{\ctx}{\PCMV}{\TMV_{1}}{\ctx'} \\
-    \ctxSynTypeM{\ctx'}{\ECMV_{2}}{\TMV_2}
-  }{
-    \ctxSynTypeM{\ctx}{\ELet{\PCMV}{\ECMV_{1}}{\ECMV_{2}}}{\TMV_2}
-  }{MSLetPat}
-
-  \judgment{
-    \ctxSynPatM{\ctx}{\PCMV}{\TMV_p} \\
-    \ctxAnaTypeM{\ctx}{\ECMV_{1}}{\TMV_p} \\
-    \ctxSynTypeM{\ctx}{\ECMV_{1}}{\TMV_{1}} \\\\
-    \ctxAnaPatM{\ctx}{\PCMV}{\TMV_{1}}{\ctx'} \\
-    \ctxAnaTypeM{\ctx'}{\ECMV_{2}}{\TMV_2}
-  }{
-    \ctxAnaTypeM{\ctx}{\ELet{\PCMV}{\ECMV_{1}}{\ECMV_{2}}}{\TMV_2}
-  }{MALetPat}
-\end{mathpar}
-
-Note that the $\TUnknownSwitch$ exists entirely to insure that sub-expressions of the definition are assigned an appropriate mode.
-As per the pattern rules below, it is never added to the context, which might result in spooky synthesis-at-a-distance:
-
-\judgbox{\ensuremath{\ctxSynPatU{\ctx}{\PMV}{\TMV}}} $\PMV$ synthesizes type $\TMV$
-%
-\begin{mathpar}
-  \judgment{ }{
-    \ctxSynPatU{\ctx}{\PWild}{\TUnknownSwitch}
-  }{USPWild}
-
-  \judgment{ }{
-    \ctxSynPatU{\ctx}{\PVar{x}}{\TUnknownSwitch}
-  }{USPVar}
-
-  \judgment{
-    \ctxSynPatU{\ctx}{\PMV_1}{\TMV_1} \\
-    \ctxSynPatU{\ctx}{\PMV_2}{\TMV_2}
-  }{
-    \ctxSynPatU{\ctx}{\PPair{\PMV_1}{\PMV_2}}{\TProd{\TMV_1}{\TMV_2}}
-  }{USPPair}
-
-  \judgment{
-    \ctxAnaPatU{\ctx}{\PMV}{\TMV}{\ctx'}
-  }{
-    \ctxSynPatU{\ctx}{\PAsc{\PMV}{\TMV}}{\TMV}
-  }{USPAnn}
-\end{mathpar}
-  %
-  \label{fig:calculus-pattern-synthesis}
-
-For our analytic pattern judgement, out output context is produced, consisting of the input context augmented with any bindings made within the pattern: \\
-\\
-\judgbox{\ensuremath{\ctxAnaPatU{\ctx_1}{\PMV}{\TMV}{\ctx_2}}} $\PMV$ analyzes against type $\TMV$ producing context $\ctx_2$
-%
-\begin{mathpar}
-  \judgment{ }{
-    \ctxAnaPatU{\ctx}{\PWild}{\TMV}{\ctx}
-  }{UAPWild}
-
-  \judgment{ }{
-    \ctxAnaPatU{\ctx}{\PVar{x}}{\TMV}{\extendCtx{\ctx}{x}{\TMV}}
-  }{UAPVar}
-
-  \judgment{
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
-    \ctxAnaPatU{\ctx}{\PMV_1}{\TMV_1}{\ctx_1} \\
-    \ctxAnaPatU{\ctx_1}{\PMV_2}{\TMV_2}{\ctx_2}
-  }{
-    \ctxAnaPatU{\ctx}{\PPair{\PMV_1}{\PMV_2}}{\TMV}{\ctx_2}
-  }{UAPPair}
-
-  \judgment{
-    \ctxAnaPatU{\ctx}{\PMV}{\TMV'}{\ctx'} \\
-    \consistent{\TMV}{\TMV'}
-  }{
-    \ctxAnaPatU{\ctx}{\PAsc{\PMV}{\TMV'}}{\TMV}{\ctx'}
-  }{UAPAnn}
-\end{mathpar}
-
-Marking judgements for patterns introduce two error types which parallel those for expressions:\\
-\\
-\judgbox{\ensuremath{\ctxSynFixedIntoPat{\ctx}{\PMV}{\PCMV}{\TMV}}} $\PMV$ is marked into $\PCMV$ and synthesizes $\TMV$
-\judgbox{\ensuremath{\ctxAnaFixedIntoPat{\ctx_1}{\PMV}{\PCMV}{\TMV}{\ctx_2}}} $\PMV$ is marked into $\PCMV$ and analyzes against $\TMV$ producing $\ctx_2$
-   %
-\begin{mathpar}
-  \judgment{ }{
-    \ctxSynFixedInto{\ctx}{\PWild}{\PCWild}{\TUnknownSwitch}
-  }{ISPWild}
-
-  \judgment{ }{
-    \ctxSynFixedInto{\ctx}{\PVar{x}}{\PCVar{x}}{\TUnknownSwitch}
-  }{ISPVar}
-
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\PMV_1}{\PCMV_1}{\TMV_1} \\\\
-    \ctxSynFixedInto{\ctx}{\PMV_2}{\PCMV_2}{\TMV_2}
-  }{
-     \ctxSynFixedInto{\ctx}{\EPair{\PMV_1}{\PMV_2}}{\ECPair{\PCMV_1}{\PCMV_2}}{\TProd{\TMV_1}{\TMV_2}}
-  }{ISPPair}
-
-  \judgment{
-    \ctxAnaPatU{\ctx}{\PMV{}}{\TMV}{\ctx'} \\
-    \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\TMV}{\ctx''}
-  }{
-    \ctxSynFixedInto{\ctx}{\PAsc{\PMV}{\TMV}}{\PCAsc{\PCMV}{\TMV}}{\tau}
-  }{ISPAnn1}
-
-  \judgment{
-    \ctxNotAnaPatU{\ctx}{\PMV{}}{\TMV}{\ctx'} \\
-    \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\TUnknown}{\ctx''}
-  }{
-    \ctxSynFixedInto{\ctx}{\PAsc{\PMV}{\TMV}}{\PCAsc{\PCInconType{\PCMV}}{\TMV}}{\TMV}
-  }{ISPAnn2}
-
-  \judgment{ }{
-    \ctxAnaFixedIntoPat{\ctx}{\PWild}{\PCWild}{\TMV}{\ctx}
-  }{IAPWild}
-
-  \judgment{ }{
-    \ctxAnaFixedIntoPat{\ctx}{\PVar{x}}{\PCVar{x}}{\TMV}{\extendCtx{\ctx}{x}{\TMV}}
-  }{IAPVar}
-
-  \judgment{
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
-    \ctxAnaFixedIntoPat{\ctx}{\PMV_1}{\PCMV_1}{\TMV_1}{\ctx_1} \\\\
-    \ctxAnaFixedIntoPat{\ctx_1}{\PMV_2}{\PCMV_2}{\TMV_2}{\ctx_2}
-  }{
-    \ctxAnaFixedIntoPat{\ctx}{\PPair{\PMV_1}{\PMV_2}}{\PCPair{\PCMV_1}{\PCMV_2}}{\TMV}{\ctx_2}
-  }{IAPPair1}
-
-  \judgment{
-    \notMatchedProd{\TMV} \\
-    \ctxAnaFixedIntoPat{\ctx}{\PMV_1}{\PCMV_1}{\TUnknown}{\ctx_1} \\\\
-    \ctxAnaFixedIntoPat{\ctx_1}{\PMV_2}{\PCMV_2}{\TUnknown}{\ctx_2}
-  }{
-    \ctxAnaFixedIntoPat{\ctx}{\PPair{\PMV_1}{\PMV_2}}{\PCAnaNonMatchedProd{\PCPair{\PCMV_1}{\PCMV_2}}}{\TMV}{\ctx_2}
-  }{IAPPair2}
-
-  \judgment{
-    \consistent{\TMV}{\TMV'} \\
-    \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\TMV'}{\ctx'}
-  }{
-    \ctxAnaFixedIntoPat{\ctx}{\PAsc{\PMV}{\TMV'}}{\PCAsc{\PCMV}{\TMV'}}{\TMV}{\ctx'}
-  }{IAPAnn1}
-
-  \judgment{
-    \inconsistent{\TMV}{\TMV'} \\
-    \ctxAnaFixedIntoPat{\ctx}{\PMV}{\PCMV}{\TMV'}{\ctx'}
-  }{
-    \ctxAnaFixedIntoPat{\ctx}{\PAsc{\PMV}{\TMV'}}{\PCInconType{\PCAsc{\PCMV}{\TMV'}}}{\TMV}{\ctx'}
-  }{IAPAnn2}
-\end{mathpar}
-  %
-  % \caption{Pattern Marking}
-  % \label{fig:calculus-pattern-synthesis-marking}
-
-And finally, the typing rules for marked expression precisely parallel the unmarked ones:
-
-\judgbox{\ensuremath{\ctxSynPatM{\ctx}{\PCMV}{\TMV}}} $\PCMV$ synthesizes type $\TMV$
-%
-\begin{mathpar}
-  \judgment{ }{
-    \ctxSynPatM{\ctx}{\PCWild}{\TUnknownSwitch}
-  }{MSPWild}
-
-  \judgment{ }{
-    \ctxSynPatU{\ctx}{\PCVar{x}}{\TUnknownSwitch}
-  }{MSPVar}
-
-  \judgment{
-    \ctxSynPatM{\ctx}{\PCMV_1}{\TMV_1} \\
-    \ctxSynPatM{\ctx}{\PCMV_2}{\TMV_2}
-  }{
-    \ctxSynPatM{\ctx}{\PCPair{\PCMV_1}{\PCMV_2}}{\TProd{\TMV_1}{\TMV_2}}
-  }{MSPPair}
-
-  \judgment{
-    \ctxAnaPatM{\ctx}{\PCMV}{\TMV}{\ctx'}
-  }{
-    \ctxSynPatM{\ctx}{\PCAsc{\PCMV}{\TMV}}{\TMV}
-  }{MSPAnn}
-\end{mathpar}
-
-\judgbox{\ensuremath{\ctxAnaPatM{\ctx_1}{\PCMV}{\TMV}{\ctx_2}}} $\PCMV$ analyzes against type $\TMV$ producing context $\ctx_2$
-%
-\begin{mathpar}
-  \judgment{ }{
-    \ctxAnaPatU{\ctx}{\PCWild}{\TMV}{\ctx}
-  }{MAPWild}
-
-  \judgment{ }{
-    \ctxAnaPatU{\ctx}{\PCVar{x}}{\TMV}{\extendCtx{\ctx}{x}{\TMV}}
-  }{MAPVar}
-
-  \judgment{
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
-    \ctxAnaPatM{\ctx}{\PCMV_1}{\TMV_1}{\ctx_1} \\\\
-    \ctxAnaPatM{\ctx_1}{\PCMV_2}{\TMV_2}{\ctx_2}
-  }{
-    \ctxAnaPatM{\ctx}{\PCPair{\PCMV_1}{\PCMV_2}}{\TMV}{\ctx_2}
-  }{MAPPair1}
-
-  \judgment{
-    \notMatchedProd{\TMV} \\
-    \ctxAnaPatM{\ctx}{\PCMV_1}{\TUnknown}{\ctx_1} \\\\
-    \ctxAnaPatM{\ctx_1}{\PCMV_2}{\TUnknown}{\ctx_2}
-  }{
-    \ctxAnaPatM{\ctx}{\PCAnaNonMatchedProd{\PCPair{\PCMV_1}{\PCMV_2}}}{\TMV}{\ctx_2}
-  }{MAPPair2}
-
-  \judgment{
-    \consistent{\TMV}{\TMV'} \\
-    \ctxAnaPatM{\ctx}{\PCMV}{\TMV'}{\ctx'}
-  }{
-    \ctxAnaPatM{\ctx}{\PCAsc{\PCMV}{\TMV'}}{\TMV}{\ctx'}
-  }{MAPAnn1}
-
-  \judgment{
-    \inconsistent{\TMV}{\TMV'} \\
-    \ctxAnaPatM{\ctx}{\PCMV}{\TMV'}{\ctx'}
-  }{
-    \ctxAnaPatM{\ctx}{\PCInconType{\PCAsc{\PCMV}{\TMV'}}}{\TMV}{\ctx'}
-  }{MAPAnn2}
-\end{mathpar}
-  %
-  % \caption{Marked Pattern Typing}
-  % \label{fig:calculus-pattern-marked-typing}
+% TODO conclude briefly?

### 2023-03-02 intro.tex
Updates from Overleaf

@@
 supports local type inference, specified in the well-established bidirectional style \cite{pierce,hazelnut,BidirTyping}:
-
+\vspace{-3px}
 \begin{center}
@@
 \end{center}
-
+\vspace{-3px}
 A type checker with no support for error localization or recovery---%
@@
 If localization had assumed that the \li{then} branch correct, as for example in the Elm type checker, then \li{x : Bool} and an error should be reported on the left operand (though this error might mislead the programmer if this earlier localization guess was incorrect).
-If the \li{else} branch is chosen, then \li{x : Int} and an error would not be reported. 
+If the \li{else} branch was chosen, then \li{x : Int} and an error would not be reported. 
 If the inconsistency is localized to the conditional expression as a whole, as in our language server, then we again confront the problem of missing type information: 
@@
 though it \emph{is} known to be bound. 
-There is no definitive type or binding error here, based on the incomplete information available in the typing context, 
+There is no definitive type or binding error here, 
 so we choose not to report an error in Hazel.
-
 No matter how localization handles the left operand, we would like to be able to recover and 
@@
 
-This system, scaled up and implemented in Hazel (Sec.~\ref{sec:calculus-hazel}), solves the semantic gap problem: Hazel's semantic services, which include type hints \cite{potter1}, semantic navigation, semantic code completion \cite{potter1,blinn}, contextualized documentation \cite{potter2}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{HazelnutLive}, even semantic services that require run-time information, like live testing,
-are available at all times and at all cursor locations, as long as the program sketch is syntactically well-formed.
+This system, scaled up and implemented in Hazel (Sec.~\ref{sec:calculus-hazel}), solves the semantic gap problem: Hazel's semantic services are available as long as the program sketch is syntactically well-formed.
+% which include type hints \cite{potter1}, semantic navigation, semantic code completion \cite{potter1,blinn}, contextualized documentation \cite{potter2}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{HazelnutLive}, even semantic services that require run-time information, like live testing,
+%
 
@@
 This approach is powerful, but it is also notorious for complicating type error localization and recovery, because inconsistencies can arise through a confluence of constraints originating from any number of locations in the program.
-Complex techniques, e.g. based on machine learning or manual weighting, have been proposed to heuristically localize these errors to expressions. 
+
 In Sec.~\ref{sec:thi}\todo{sec}, we describe how we blend local and constraint-based type inference in Hazel. In particular,  
-we generate constraints for each type hole that appears in the program explicitly or that is  generated internally during the  bidirectional type error localization and recovery process from Sec.~\ref{sec:calculus}.
-We develop a unification algorithm that can recover from inconsistent constraints by maintaining possible type hole fillings. For holes that can be uniquely solved, we suggest the solution. 
+we generate constraints for each type hole that arises (either explicitly in the program or during type error localization and recovery as defined in  Sec.~\ref{sec:calculus}). 
+We develop a unification algorithm that can recover from inconsistent constraints by maintaining possible type hole solutions. For holes that can be uniquely solved, we suggest the solution. 
 If conflicting constraints arise, we localize the errors to the relevant type holes, rather 
-than attempting to localize to expressions. 
-The user can explore the different partial type hole fillings. Making a choice then allows the bidirectional system to localize the error to relevant expressions.
+than to expressions.
+
+For example, in the following Hazel expression, bidirectional error localization 
+does not mark any expressions as erroneous because a type for \li{x} cannot locally be inferred (the bound expression is a hole). The type hole at the cursor therefore behaves gradually. 
+However, by gathering constraints from the three uses of \li{x} and attempting to unify, we see that this type hole is unfillable -- it is being used at three different types.
+This error is localized, with a \li{!}, to the type hole itself.
+\begin{center}
+\vspace{-3px}
+    \includegraphics[scale=0.55]{images/intro-thi.png}
+    \includegraphics[scale=0.4]{images/example_CI.png}
+\end{center}
+The user can investigate by choosing a partial solution, which (temporarily) fills the type hole (not shown) and thus allows the local inference system to systematically localize the error.
+Here, each choice causes different set of two uses of \li{x} to be marked, e.g.
+the first and third use for \li{Int -> Int} (not shown).
 Until the choice is finalized, the system continues to recover gradually.
-
-let x : ? = ? in 
-if x then x(x + 1) else

### 2023-03-02 supplement/unionFindPTS.tex
Edits supplemental material for type hole inference stuff. Spacing is still a little funky in section B and I'm not quite sure why (Raef)
Updates from Overleaf

--- 2023-02-26 appendixA.tex
@@
 \newpage
-\section{Appendix A: An implementation of union-findable Possible Type Sets}
+\section{An implementation of union-findable Possible Type Sets}
 
@@
 
-let union => if (UnionFind.eq t1 t2) then () else merge_sets_UF t1 t2;
+let (*$\amalg_{UF}$*) => if (UnionFind.eq t1 t2) then () else merge_sets_UF t1 t2;
+
 \end{lstlisting}
-\vspace{-4px}
  \hrule
@@
 \end{figure}
+
+\begin{figure}[htb!]
+\begin{lstlisting}[escapeinside={(*}{*)}]
+let create_traversable_edge (graph, t1, t2) =
+    let (rep1, rep2) = (UnionFind.find(t1), UnionFind.find(t2)) in
+    let (s1, s2) = (Map.find(graph, rep1), Map.find(graph, rep2)) in
+    let union_rep = UnionFind.union(rep1, rep2) in
+    let s_merged = s1 (*$\amalg_{UF}$*) s2 in
+    Map.update(graph, union_rep, s_merged);;
+let create_solution_edge (graph, t1, t2) =
+    let rep1 = UnionFind.find(t1) in
+    let s = Map.find(graph, rep1) in
+    let s' = s (*$\amalg_{UF}$*) UnionFind.elem(single(t2)) in
+    Map.update(graph, t1, s');;
+\end{lstlisting}
+\hrule
+\caption{Constraint solving algorithm helper functions, using Union-Find.}
+\label{fig:algcode_construct_graph_helpers}
+\end{figure}
+
+\begin{figure}[htb!]
+\begin{lstlisting}[escapeinside={(*}{*)}]
+let handle_occurs_check (graph) =
+    let annotate_with_error (key) =
+        let s = Map.find(graph, UnionFind.find(key) in
+        let pass_occurs = is_acyclic(s)
+        (key, s, pass_occurs)
+    in
+    List.map(Map.keys(graph), annotate_with_error);;
+\end{lstlisting}
+ \hrule
+\caption{Occurs check assessment}
+\label{fig:occurs_check}
+\end{figure}

### 2023-03-02 calculus.tex
Some edits on section 2

@@
 
-% TODO examples of this ad hoc stuff
-
 The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
-type error localization in bidirectional systems. Namely, the calculus provides a judgmental
-framework with a unified theory for the design of typing semantics that support ill-typed programs.
-\Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
-lambda calculus (GTLC), extending it with binary product types in \cref{sec:calculus-products}. All
-rules and theorems may be found in the supplementary appendix, alongside a complete mechanization in
-the Agda proof assistant \cite{norell:thesis} (discussed in \cref{sec:calculus-agda}). With
-products, we then consider how error localization is complicated by patterned let expressions in
-\cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how the system might be extended
-to more complex typing features, such as System F-style polymorphism.
+bidirectional type error localization. Namely, the calculus employs bidirectional and gradual typing
+to provide a judgmental framework with a unified theory for the design of typing semantics that
+support ill-typed programs. \Cref{sec:calculus-calculus} describes its form and metatheory by
+example on a small gradually typed lambda calculus (GTLC) \cite{Siek06a}, extending it with binary
+product types in \cref{sec:calculus-products}. All rules and theorems may be found in the
+supplementary appendix, alongside a complete mechanization in the Agda proof assistant
+\cite{norell:thesis} (discussed in \cref{sec:calculus-agda}). With products, we then consider how
+error localization is complicated by patterned let expressions in \cref{sec:calculus-let}. Finally,
+\cref{sec:calculus-poly} explores how the system might be extended to more complex typing features,
+such as System F-style polymorphism.
 
@@
 expressions. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$,
-and there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$ correspond
-to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
-$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure. In addition to these forms,
-$\TUnknown$ is the unknown type from gradual type theory. Finally, $\EEHole$ denotes an
-\emph{expression hole}, which are used to model syntactically incomplete programs. Note, however,
-that these are not critical to the calculus.
+and there is a single arithmetic operation of addition on numeric expressions. $\ETrue$ and
+$\EFalse$ correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
+$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure.
+
+In addition to these forms, $\TUnknown$ is the unknown type from gradual type theory \cite{Siek06a}.
+Modelling the edit state of a program in development, $\EEHole$ denotes an \emph{expression hole},
+used to represent syntactically incomplete portions of the program à la Hazelnut
+\cite{HazelnutPOPL}. Note, however, that these are not critical to the calculus.
 
@@
 
-Types classify expressions by a \emph{bidirectional type system}, which employs two mutually defined
-judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$, establishes that, under
-the typing context $\ctx$, the expression $e$ synthesizes or locally infers \cite{Localinf} the type
-$\TMV$. \emph{Type analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$, states that the expression
-$\EMV$ may appear where an expression of type $\TMV$ is expected. We may give these typing semantics
-traditionally, without considering error localization and recovery.
+Types classify expressions by a \emph{bidirectional type system} \cite{BidirTyping}, which employs
+two mutually defined judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$,
+establishes that, under the typing context $\ctx$, the expression $e$ synthesizes or locally infers
+\cite{Localinf} the type $\TMV$. \emph{Type analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$,
+states that the expression $\EMV$ may appear where an expression of type $\TMV$ is expected. We may
+give these typing semantics traditionally, without consideration for error localization and
+recovery.
 
@@
 
-  \item \textbf{Marking}, which \emph{marks}, i.e. type checks and localizes errors for, any unmarked
-    expression into a marked expression, inserting error marks where appropriate.
+  \item \textbf{Marking}, which \emph{marks} any unmarked expression into a marked expression,
+    inserting error marks where appropriate.
 \end{itemize}
@@
 process with error localization and recovery. The marked language may then serve as a foundation for
-other semantic services. % such as THI.
-
-% TODO more intuition
+other semantic services. % TODO such as THI.
 
@@
 \emph{error marks}, which represent the different type errors that may arise. Intuitively, they each
-denote a different error message that might be shown in an editor or emitted by a compiler.
-%
-% \begin{itemize}
-  % \item $\ECUnbound{x}$, the \emph{unbound variable mark}, indicates the use of an unbound variable
-    % $x$, as in \cref{fig:calculus-examples-unbound}.
-%
-  % \item $\ECInconType{\ECMV}$, the \emph{inconsistent type mark}, indicates that the type of an
-    % expression does not match an expected type.
-%
-  % \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, indicates
-    % that the branches of a conditional have mismatched types.
-%
-  % \item $\ECLamInconAsc{x}{\TMV}{\ECMV}$, the \emph{inconsistent ascription mark}, indicates that
-    % the type ascription $\TMV$ is mismatches with an expected type.
-%
-  % \item $\ECSynNonMatchedArrow{\ECMV}$
-% \end{itemize}
-%
+denote a different error message that might be shown by an editor or emitted by a compiler.
+
 Throughout the remainder of this section, as we formulate marking for the GTLC, we will motivate and
-give precise semantics for these error marks. Note that this is a non-exhaustive list; depending on
-the syntactic forms and typing features found in the language, other kinds of marks may be
-necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of marks that may be required
-in an language with explicit polymorphism.
+give precise semantics for these error marks. Note that these constitute a non-exhaustive list;
+depending on the syntactic forms and typing features found in a given language, other kinds of marks
+may be necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of marks that may be
+required in an language with explicit polymorphism.
 
@@
 
-Furthermore, marking should preserve the syntactic structure of programs, modulo error marks. Thus,
-we define a notion of \emph{mark erasure}, whose definition is given
-\cref{fig:calculus-mark-erasure}, which converts marked expressions back into unmarked expression by
+\input{figs/calculus-mark-erasure}
+
+Furthermore, since error marks are effectively annotations on top of the program, marking should
+preserve syntactic structure modulo those marks. \cref{fig:calculus-mark-erasure} thus gives a
+notion of \emph{mark erasure}, which converts marked expressions back into unmarked expression by
 removing error marks. Then, \cref{thm:calculus-marking-well-formedness} provides a
-\emph{well-formedness} criterion marking.
+\emph{well-formedness} criterion for marking.
 %
@@
   \end{enumerate}
-  % TODO split this into two theorems?
-  % TODO not really sure how to intuitively describe the part about types matching up
 \end{theorem}
@@
 expression. We also ensure that marking gives well-typed marked expressions; indeed, their types
-should match the types synthesized or analyzed against in marking.
-% TODO more intuition for this theorem
-
-\input{figs/calculus-mark-erasure}
+should match those synthesized or analyzed against in marking.
 
 In the remainder of this subsection, we consider the different type errors that may arise in the
-unmarked language while keeping in mind these two theorems; in this way, the typing semantics of the
-unmarked language guide the construction of the marking process.
+unmarked language while keeping in mind these two theorems; in this way, the traditional typing
+semantics of the unmarked language guide the construction of the marking process.
 
@@
 For addition expressions, the type of both operands should be $\TNum$. To denote this in a
-bidirectional system, the operands are analyzed against $\TNum$, giving the following typing rule
-for unmarked addition expressions:
+bidirectional system, they are analyzed against $\TNum$, giving the following typing rule for
+unmarked addition expressions:
 %
@@
 we shift the responsibility for any type errors to them. Hence, we recursively mark each operand in
-analytic mode and rebuild the marked addition expression. The marked typing rule then mirrors
-\textsc{USPlus} exactly.
+analytic mode and rebuild the marked addition expression. The typing rule for marked addition
+expressions then mirrors \textsc{USPlus} exactly.
 %

### 2023-03-02 calculus.tex
Updates from Overleaf

--- 2023-03-02 calculus.tex
@@
       \consistent{\TArrow{\TMV_1}{\TMV_2}}{\TArrow{\TMV_1'}{\TMV_2'}}
-    }{TCArr} \\
-  \end{mathpar}
-  %
+    }{TCArr}
+  \end{mathpar}
+  \vspace{-10px}
   \caption{Type consistency.}
@@
     \joinsTo{\TBool}{\TBool}{\TBool} \\
-    \joinsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}} \\
+    \joinsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}}
   \end{array}\]
-  %
+  \vspace{-10px}
   \caption{Type join.}
@@
 
-\subsubsection{Additional Theorems}
+\subsubsection{Additional Metatheorems}
 To conclude this section, we present two more metatheorems that help ensure correctness of the
@@
 Together, totality and unicity give that marking may be considered and implemented as a total
-function. Indeed, given the algorithmic nature of bidirectional typing, it is fairly easy to extract
-an implementation from the rules.
+function. Indeed, given the algorithmic nature of bidirectional typing, it is fairly direct to extract
+an implementation from these rules.
 
@@
 arrow types, the judgment $\matchedProd{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be
-considered the binary product type $\TProd{\TMV_1}{\TMV_2}$ [cite?].
+considered the binary product type $\TProd{\TMV_1}{\TMV_2}$ [cite?].\todo{cite}
 
@@
 We now consider how the calculus might be extended to richer typing features such as parametric
-polymorphism. Specifically, we consider System F-style explicit polymorphism.
-
-\subsection{Mechanization}
+polymorphism. Specifically, we consider System F-style explicit polymorphism.\todo{write me}
+
+\subsection{Agda Mechanization}
 \label{sec:calculus-agda}

### 2023-03-02 calculus.tex
Shorten intro of section 2

@@
 
-% TODO elaborate on this part
-
-Unfortunately, formal typing semantics usually only give meaning to well-typed programs, leaving
-such ill-typed ones meaningless. This is insufficient for editor services and language servers,
-which must provide users feedback on where and why a program is ill-typed. Moreover, they must
-continue providing semantic services even when an error is encountered. Indeed, similarly to
-error-recovering parsers, real-world type checkers must robustly recover from type errors.
-Traditionally, this has been performed in an \emph{ad hoc} manner [citations?], with different
-languages and tools adopting varying approaches in the localization, reportage, and recovery of
-errors.
-
 The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
-bidirectional type error localization. Namely, the calculus employs bidirectional and gradual typing
-to provide a judgmental framework with a unified theory for the design of typing semantics that
-support ill-typed programs. \Cref{sec:calculus-calculus} describes its form and metatheory by
-example on a small gradually typed lambda calculus (GTLC) \cite{Siek06a}, extending it with binary
-product types in \cref{sec:calculus-products}. All rules and theorems may be found in the
-supplementary appendix, alongside a complete mechanization in the Agda proof assistant
-\cite{norell:thesis} (discussed in \cref{sec:calculus-agda}). With products, we then consider how
-error localization is complicated by patterned let expressions in \cref{sec:calculus-let}. Finally,
-\cref{sec:calculus-poly} explores how the system might be extended to more complex typing features,
-such as System F-style polymorphism.
-
-% TODO not really sure when to start talking about bidirectional typing
-
-% \emph{Bidirectional typing} combines type checking, which determines if a program satisfies a given
-% type, and type synthesis, which generates a type from the program. This gives a simple algorithmic
-% system in which type information is recursively propagated through terms and allows for local type
-% inference.
-
-% TODO talk about idea that bidirectional typing makes error localization easy? i.e. why choose
-% bidirectional?
+bidirectional type error localization and recovery. Namely, the calculus provides a judgmental
+framework with a unified theory for the design of typing semantics that support ill-typed programs.
+\Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
+lambda calculus (GTLC) \cite{Siek06a}, extending it with binary product types in
+\cref{sec:calculus-products}. All rules and theorems may be found in the supplementary appendix,
+alongside a complete mechanization in the Agda proof assistant \cite{norell:thesis} (discussed in
+\cref{sec:calculus-agda}). With products, we then consider how error localization is complicated by
+patterned let expressions in \cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how
+the system might be extended to more complex typing features, such as System F-style polymorphism.
 

### 2023-03-02 calculus.tex
Resolve/make explicit some todos

@@
 The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
-bidirectional type error localization and recovery. Namely, the calculus provides a judgmental
-framework with a unified theory for the design of typing semantics that support ill-typed programs.
-\Cref{sec:calculus-calculus} describes its form and metatheory by example on a small gradually typed
-lambda calculus (GTLC) \cite{Siek06a}, extending it with binary product types in
+bidirectional type error localization with gradual recovery. Namely, the calculus provides a
+judgmental framework with a unified theory for the design of typing semantics that support ill-typed
+programs. \Cref{sec:calculus-calculus} describes its form and metatheory by example on a small
+gradually typed lambda calculus (GTLC) \cite{Siek06a}, extending it with binary product types in
 \cref{sec:calculus-products}. All rules and theorems may be found in the supplementary appendix,
@@
 \label{sec:calculus-calculus}
-
-% TODO outline the structure of this section at the beginning
-% TODO make it clear unmarked/marked languages share same types
 
@@
     \emph{marked expressions}, denoted $\ECMV$.
-    % TODO more intuition for what marked expressions are?
 
@@
 process with error localization and recovery. The marked language may then serve as a foundation for
-other semantic services. % TODO such as THI.
+other semantic services, such as constraint-based inference assistance (as discussed in
+\cref{sec:thi}).
 
@@
 \cref{fig:calculus-examples-inconsistent-types}, subsumption would be used to analyze $\ETrue$
-against $\TNum$ in $\EPlus{\ETrue}{5}$, which fails since $\inconsistent{\TNum}{\TBool}$, i.e.
+against $\TNum$ in $\EPlus{\ETrue}{1}$, which fails since $\inconsistent{\TNum}{\TBool}$, i.e.
 \textsc{UASubsume} does not apply, and traditional typing semantics would simply fail the type
@@
 undefined in those cases.
-
-% TODO use example more closely related to num here since we haven't talked about booleans yet?
 
@@
 \end{mathpar}
-% TODO more intuition on why this makes sense?
 
@@
 synthesize the unknown type. Similarly to the inconsistent type mark, this allows type checking to
-proceed, with usage of the unbound variable permitted in any expression.
-% TODO clarify/give intuition as to why this is correct
+proceed, with usage of the unbound variable permitted in any expression. \todo{intution for
+correctness} \todo{relate to gradual recovery}
 
@@
 Unlike numbers and variables, there are explicit synthesis and analysis rules for unmarked lambda
-abstractions: % TODO intuition why
+abstractions. This is because expected input and output types are known, and we may verify that the
+type annotation and body match them.
 %
@@
 %
-The synthesis rule is standard. The analysis rule employs the judgment
-$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$, which establishes that $\TMV$ is a \emph{matched arrow type}
-\cite{cimini:2016}, i.e. it may be considered an arrow type. Defined in
-\cref{fig:calculus-matched-arrow}, this notion is purely a technical mechanism to avoid duplication
-of rules related to arrow types.
-
-% TODO mention explicitly that by inductive hypothesis, the marking of body cannot fail?
-% TODO more intuition on these two judgments
-% TODO more intuition on USLam and UALam
+The synthesis rule is standard. Analysis employs the judgment $\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$,
+which establishes that $\TMV$ is a \emph{matched arrow type} \cite{cimini:2016}, i.e. it may be
+considered an arrow type. Defined in \cref{fig:calculus-matched-arrow}, this notion is purely a
+technical mechanism to avoid duplication of rules related to arrow types.
 
@@
 In synthetic position, conditionals synthesize the \emph{join} of the branch types $\TMV_1$ and
-$\TMV_2$, which we define inductively in \cref{fig:calculus-type-join}. In analytic position, since
+$\TMV_2$, which we define inductively in \cref{fig:calculus-type-join}. We choose the ``more
+specific'' type of the two\todo{intuition for why this is correct}. In analytic position, since
 there is an expected type for both branches, we shift the blame for any errors to them.
-
-% TODO intuition why we use join
 
@@
 \end{figure}
-% TODO intuition for type join?
 
@@
 
-% TODO mini conclusion on methodology
-
 \subsubsection{Additional Metatheorems}
@@
 
-\begin{theorem}[name=Marking of Well-Typed/Ill-Typed Expressions] \ % TODO better name for this
+\begin{theorem}[name=Marking of Well-Typed/Ill-Typed Expressions] \
   \label{thm:calculus-marking-well-ill-typed}
@@
 \label{sec:calculus-agda}
+% TODO clarify exactly what is/isn't mechanized
 
@@
 $\AAEAction{\ctx}{\EMV}{\EMV'}{\TMV}{\AMV}$---inserts on-demand \emph{error holes}, denoted
-$\hhole{\EMV}{}$, which indicate type errors.
-
-% TODO example
+$\hhole{\EMV}{}$, which indicate type errors. \todo{example?}
 
@@
 action semantics and related metatheory are mechanized in Agda.
-
-% TODO note that we have not verified the metatheory for the typed system

### 2023-03-02 calculus.tex
Updates from Overleaf

--- 2023-03-02 calculus.tex
@@
 \Cref{fig:calculus-examples} shows common type errors as they appear in the Hazel programming
-environment. For example, a simple error that may occur is the use of an unbound variable. As in
-\cref{fig:calculus-examples-unbound}, the editor indicates that the usage of an unbound $y$ is
-invalid by highlighting it in red.
+environment. For example, a simple error that may occur is the use of an unbound variable (\cref{fig:calculus-examples-unbound}).
+% As in
+% \cref{fig:calculus-examples-unbound}, the editor indicates that the usage of an unbound $y$ is
+% invalid by highlighting it in red.
 
@@
 
-Sometimes, the type an expression \emph{should} be is known. Consider
+With local type inference, an expression might have a expected type. Consider
 \cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
 $+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
-indicates a mismatch between the expected and actual types of the operand. The application in
+indicates an inconsistency between the expected and actual types of the operand. The application in
 \cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
 this case, however, the expectation is not a singular type, but any member of a family of function
-types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error of inconsistent
-branches: the branches of the conditional are of different types: $5$ is a number, while
-$\textsf{false}$ is a boolean. Hazel opts to highlight the entire expression in such a situation,
+types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error caused by inconsistent
+branch types. As discussed in the introduction, Hazel opts to highlight the entire expression in such a situation,
 while other tools may choose one branch to be ``correct'' and ``blame'' the type mismatch error on
@@
 $\subsumable{\EMV}$. This judgment is defined on all syntactic forms except lambda abstractions and
-conditionals, the only ones with explicit analysis rules. In other words, we restrict subsumption to
+conditionals, the only ones with both synthesis and analysis rules (see below). In other words, we restrict subsumption to
 be the rule of ``last resort'', which is necessary to establish that marking and typing are
@@
 To satisfy marking totality, such a possibility motivates an \emph{inconsistent type mark}
-$\ECInconType{\ECMV}$, which states exactly that the synthesized type of $\ECMV$ is inconsistent
-with the expected type. It is then accompanied by the following rules:
+$\ECInconType{\ECMV}$, which is applied when the synthesized type of $\ECMV$ is inconsistent
+with the expected type:
 %
@@
   \end{mathpar}
-  %
+  \vspace{-10px}
   \caption{Matched arrow types.}
@@
     \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
-    \inconsistent{\TMV}{\TMV_1} \\
+    \inconsistent{\TMV}{\TMV_1} \\\\
     \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
@@
     \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
-    \inconsistent{\TMV}{\TMV_1} \\
+    \inconsistent{\TMV}{\TMV_1} \\\\
     \ctxAnaTypeM{\extendCtx{\ctx}{x}{\TMV_1}}{\ECMV}{\TMV_2}
@@
 \]%
-Following the same methodology, we can straightforwardly give the following marking and typing
+Following the same methodology, we have the following marking and typing
 rules:
@@
 unmatched types, such as products. Indeed, the same approach is taken for projections in
-\cref{sec:calculus-products}. Another approach might indicate the same kind of error but mark the
-entire application.
+\cref{sec:calculus-products}. 
+% Another approach might indicate the same kind of error but mark the
+% entire application.
 
@@
 
-The boolean values are straightforward to handle, behaving in a similar manner as numbers: 
+The boolean values are similar to numbers: 
 %
@@
 
-Conditionals, however, offer an interesting case. In the unmarked language, we have both explicit
+Conditionals, however, are an interesting case. In the unmarked language, we have both explicit
 synthetic and analytic rules:
@@
 Following \textsc{USIf} and \textsc{UAIf}, we may derive the following marking and typing rules.
-Similar to previous cases, we recursively mark the guard and branches, constructing the marked
-conditional from the results.
+% Similar to previous cases, we recursively mark the guard and branches, constructing the marked
+% conditional from the results.
 %
@@
 \]%
-Of course, corresponding unmarked and marked typing rules would follow similar structure, and other
-heuristics may be devised for different kinds of branching structures.
+% Of course, corresponding unmarked and marked typing rules would follow similar structure, and other
+% heuristics may be devised for different kinds of branching structures.
 
@@
 For completeness, we finish the development of the marked lambda calculus on the extended GTLC with
-the marking of holes. Like number literals and booleans, these are governed by straightforward
-rules.
+the marking of empty holes, which are never marked.
 %
@@
 the syntactic structure of a program, it makes no statement about the presence of error marks in the
-resulting marked program. Following the logic of constructing marking rules, we have
-\cref{thm:calculus-marking-well-ill-typed}.
+resulting marked program. 
+\cref{thm:calculus-marking-well-ill-typed} established that well-typed expressions are left unmarked and ill-typed expressions have at least one mark.
 
@@
 %
-That is, since error marks indicate type errors in the program, no marks should be introduced
-when marking a well-typed program, a property given by the judgment $\markless{\ECMV}$. The
-converse---that at least one mark is inserted into ill-typed programs---is also true.
-
 Finally, no less importantly, marking is deterministic. This is given by
@@
 %
-Together, totality and unicity give that marking may be considered and implemented as a total
-function. Indeed, given the algorithmic nature of bidirectional typing, it is fairly direct to extract
-an implementation from these rules.
+Together, totality and unicity give that marking may be implemented as a total
+function. Indeed, given the algorithmic nature of bidirectional typing, it is fairly direct to implement these rules.
 
@@
 
-We now demonstrate how one might extend or adapt the marked lambda calculus for new typing features
-by adding binary product types, written $\TProd{\TMV}{\TMV}$. In the interest of space, we focus on
-aspects related to marking; a comprehensive set of rules is found in the supplemental material.
-
-First, we extend the syntax of types and the accompanying notions of consistency and join, which
-closely mirror the definitions for arrow types. Additionally, analogous to the notion of matched
-arrow types, the judgment $\matchedProd{\TMV}{\TMV_1}{\TMV_2}$ establishes that $\TMV$ may be
-considered the binary product type $\TProd{\TMV_1}{\TMV_2}$ [cite?].\todo{cite}
+The marked lambda calculus is intended as a minimal core.
+The methodology described above can be used to extend the marked lambda calculus with new typing features.
+For example, let us add binary product types, written $\TProd{\TMV_1}{\TMV_2}$.
+In the interest of space, we focus on
+aspects related to marking and omit standard rules; a comprehensive set of rules is found in the supplemental material.
+
+% First, we extend the syntax of types and the accompanying notions of consistency and join, which
+% closely mirror the definitions for arrow types. Additionally, analogous to the notion of matched
+% arrow types, we need a matched product type relation $\matchedProd{\TMV}{\TMV_1}{\TMV_2}$.
 
@@
 operators $\EProjL{\EMV}$ and $\EProjR{\EMV}$, governed by standard bidirectional rules. In addition
-to the synthesis rules, an explicit analysis rules is defined for the former: if an expected type is
-known, the responsibility for type errors is shift to the components.
-
-The marked language is extended with the same forms, and we again devise marking rules by modelling
-off the unmarked typing rules.
+to the synthesis rules, an explicit analysis rules is defined for pairs: if the expected type has a matched product type (cf.
+matched arrow types above), the responsibility for type errors is shifted to the components.
+
+We can devise the basic marking rules from the unmarked typing rules.
 %
@@
   }{MKSPair}
-
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
-  }{
-    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV_1}
-  }{MKSProjL1}
-
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
-  }{
-    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV_2}
-  }{MKSProjR1} \\
 
@@
   }{MKAPair1}
-\end{mathpar}
-
+
+  \judgment{
+    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV_1}
+  }{MKSProjL1}
+
+  \judgment{
+    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV_2}
+  }{MKSProjR1}
+\end{mathpar}
 However, as previously seen with lambda abstractions and applications, two sets of rules are further

### 2023-03-02 let.tex
Updates from Overleaf

@@
-\subsection{Destructuring Let with Composite Patterns}
+\subsection{Destructuring Let with Type Annotated Composite Patterns}
 \label{sec:calculus-let}
 
-To ease the use of products, many languages feature destructuring bindings [cite]. Let us add let
+To ease the use of products, many languages feature destructuring bindings.
+In a typed setting, we may want to granularly add type annotations in patterns as well.
+As it turns out this combination of features is surprisingly tricky to get right!
+Let us add let
 expressions and simple patterns, writing $\PWild$ for the wildcard pattern, $\PVar{x}$ for a
-variable pattern, $\PPair{\PMV_1}{\PMV_2}$ for a pair. In a bidirectional setting, the pattern
-itself synthesizes a type which constrains the definition.
+variable pattern, $\PPair{\PMV_1}{\PMV_2}$ for a pair. 
 %
@@
 %
-It turns out, however, that these composite patterns with type annotations can complicate type error
-localization in subtle ways.
 
@@
 mode imposes \emph{no} type constraints, whereas analytic mode imposes a \emph{trivial} type
-constraint. This manifests when expressions may have internal type inconsistencies, such as in the
-following program:
+constraint. This manifests when expressions may have internal type inconsistencies,  as in the
+following:
 \[%
@@
   \judgment{
-    \ctxSynType{\ctx}{\EMV}{\ctx'}
+    \ctxSynTypeU{\ctx}{\EMV}{\ctx'}
   }{
-    \ctxAnaType{\ctx}{\EMV}{\TUnknownSwitch}
+    \ctxAnaTypeU{\ctx}{\EMV}{\TUnknownSwitch}
   }{UASynSwitch}
@@
   }{
-    \ctxSynType{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\TMV_2}
+    \ctxSynTypeU{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\TMV_2}
   }{USLetPat}

### 2023-03-02 calculus.tex

@@
 
-% TODO elaborate on this part
-
-Unfortunately, formal typing semantics usually only give meaning to well-typed programs, leaving
-such ill-typed ones meaningless. This is insufficient for editor services and language servers,
-which must provide users feedback on where and why a program is ill-typed. Moreover, they must
-continue providing semantic services even when an error is encountered. Indeed, similarly to
-error-recovering parsers, real-world type checkers must robustly recover from type errors.
-Traditionally, this has been performed in an \emph{ad hoc} manner [citations?], with different
-languages and tools adopting varying approaches in the localization, reportage, and recovery of
-errors.
-
 The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
-bidirectional type error localization. Namely, the calculus employs bidirectional and gradual typing
-to provide a judgmental framework with a unified theory for the design of typing semantics that
-support ill-typed programs. \Cref{sec:calculus-calculus} describes its form and metatheory by
-example on a small gradually typed lambda calculus (GTLC) \cite{Siek06a}, extending it with binary
-product types in \cref{sec:calculus-products}. All rules and theorems may be found in the
-supplementary appendix, alongside a complete mechanization in the Agda proof assistant
-\cite{norell:thesis} (discussed in \cref{sec:calculus-agda}). With products, we then consider how
-error localization is complicated by patterned let expressions in \cref{sec:calculus-let}. Finally,
-\cref{sec:calculus-poly} explores how the system might be extended to more complex typing features,
-such as System F-style polymorphism.
-
-% TODO not really sure when to start talking about bidirectional typing
-
-% \emph{Bidirectional typing} combines type checking, which determines if a program satisfies a given
-% type, and type synthesis, which generates a type from the program. This gives a simple algorithmic
-% system in which type information is recursively propagated through terms and allows for local type
-% inference.
-
-% TODO talk about idea that bidirectional typing makes error localization easy? i.e. why choose
-% bidirectional?
+bidirectional type error localization with gradual recovery. Namely, the calculus provides a
+judgmental framework with a unified theory for the design of typing semantics that support ill-typed
+programs. \Cref{sec:calculus-calculus} describes its form and metatheory by example on a small
+gradually typed lambda calculus (GTLC) \cite{Siek06a}, extending it with binary product types in
+\cref{sec:calculus-products}. All rules and theorems may be found in the supplementary appendix,
+alongside a complete mechanization in the Agda proof assistant \cite{norell:thesis} (discussed in
+\cref{sec:calculus-agda}). With products, we then consider how error localization is complicated by
+patterned let expressions in \cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how
+the system might be extended to more complex typing features, such as System F-style polymorphism.
 

### 2023-03-02 calculus.tex
Updates from Overleaf

@@
 \Cref{fig:calculus-examples} shows common type errors as they appear in the Hazel programming
-environment. For example, a simple error that may occur is the use of an unbound variable (\cref{fig:calculus-examples-unbound}).
+environment. 
+%For example, a simple error that may occur is the use of an unbound variable (\cref{fig:calculus-examples-unbound}).
 % As in
@@
 
-With local type inference, an expression might have a expected type. Consider
-\cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
-$+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
-indicates an inconsistency between the expected and actual types of the operand. The application in
-\cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
-this case, however, the expectation is not a singular type, but any member of a family of function
-types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error caused by inconsistent
-branch types. As discussed in the introduction, Hazel opts to highlight the entire expression in such a situation,
-while other tools may choose one branch to be ``correct'' and ``blame'' the type mismatch error on
-the other.
+% With local type inference, an expression might have a expected type. Consider
+% \cref{fig:calculus-examples-inconsistent-types}, in which it is expected that both operands of the
+% $+$ operator are numbers. Since $\textsf{true}$ is not a number, a type error arises, and Hazel
+% indicates an inconsistency between the expected and actual types of the operand. The application in
+% \cref{fig:calculus-examples-app-non-lambda} of a non-lambda leads to a similar kind of error. In
+% this case, however, the expectation is not a singular type, but any member of a family of function
+% types. \Cref{fig:calculus-examples-inconsistent-branches} presents an error caused by inconsistent
+% branch types. As discussed in the introduction, Hazel opts to highlight the entire expression in such a situation,
+% while other tools may choose one branch to be ``correct'' and ``blame'' the type mismatch error on
+% the other.
 
-The key contribution of this section is the \emph{marked lambda calculus}, a formal calculus for
-bidirectional type error localization with gradual recovery. Namely, the calculus provides a
-judgmental framework with a unified theory for the design of typing semantics that support ill-typed
-programs. \Cref{sec:calculus-calculus} describes its form and metatheory by example on a small
-gradually typed lambda calculus (GTLC) \cite{Siek06a}, extending it with binary product types in
-\cref{sec:calculus-products}. All rules and theorems may be found in the supplementary appendix,
+The key contribution of this section is the \emph{marked lambda calculus}, a calculus based on the gradually typed lambda calculus (GTLC) that formalizes the mechanism by which errors like these can be localized, and how to  recover, in all cases, from such errors.
+% In brief, the marked lambda calculus formalizes 
+% bidirectional type error localization with gradual recovery. 
+\Cref{sec:calculus-calculus} introduces the syntax, judgemental structure, guiding metatheory, and then goes through the rules, organized by syntactic form rather than judgement, to build intuition about how the various judgements relate to one another \cite{Siek06a}.
+All rules and theorems may be found organized by judgement form in the supplementary appendix for reference,
 alongside a complete mechanization in the Agda proof assistant \cite{norell:thesis} (discussed in
-\cref{sec:calculus-agda}). With products, we then consider how error localization is complicated by
+\cref{sec:calculus-agda}). We intentionally keep the marked lambda calculus minimal products, we then consider how error localization is complicated by
 patterned let expressions in \cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how

### 2023-03-02 calculus.tex
Updates from Overleaf

@@
 alongside a complete mechanization in the Agda proof assistant \cite{norell:thesis} (discussed in
-\cref{sec:calculus-agda}). We intentionally keep the marked lambda calculus minimal products, we then consider how error localization is complicated by
-patterned let expressions in \cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} explores how
-the system might be extended to more complex typing features, such as System F-style polymorphism.
+\cref{sec:calculus-agda}).
+We intentionally keep the marked lambda calculus minimal, because it is intended to capture the essential idea and introduce a general pattern that language designers can employ to created mark variants of their own type systems.
+As an initial example, we consider product types (in \cref{sec:calculus-products}).
+We then consider a surprisingly subtle combination of features as a more substantial case study: the combination of
+destructing let expressions with granular type annotations, in \cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} briefly explores how
+the system might be extended to more complex typing features, such as polymorphism.
 
-\subsection{The Calculus}
+\subsection{The Core Calculus}
 \label{sec:calculus-calculus}
 
-We now formally introduce the marked lambda calculus through its application to the GTLC extended
-with numbers and booleans. Consider the syntax for such a language, given as $\EMName$ in
-\cref{fig:calculus-syntax}. The base types $\TNum$ and $\TBool$ classify number and boolean
+The marked lambda calculus is based on the gradually typed lambda calculus \cite{Siek06a}\todo{snapl} extended
+with numbers and booleans and empty expression holes. The syntax is given in
+\cref{fig:calculus-syntax}. There are two expression languages:
+\begin{itemize}
+  \item The \textbf{unmarked language}, which is our original language. Expressions of this language
+    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.
+
+  \item The \textbf{marked language}, which mirrors the structure of the unmarked language but is
+    extended with explicit \emph{error marks}. We call expressions of this language
+    \emph{marked expressions}, denoted $\ECMV$.
+\end{itemize}
+In this simple setting, we only need one sort for types, $\tau$. 
+The base types $\TNum$ and $\TBool$ classify number and boolean
 expressions. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$,
-and there is a single arithmetic operation of addition on numeric expressions. $\ETrue$ and
-$\EFalse$ correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, respectively, and
-$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives a conditional structure.
+and there is a single arithmetic operation, addition. $\ETrue$ and
+$\EFalse$ are the boolean values and
+$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ is the boolean conditional. Arrow types classify lambdas in the usual way.
 
-In addition to these forms, $\TUnknown$ is the unknown type from gradual type theory \cite{Siek06a}.
-Modelling the edit state of a program in development, $\EEHole$ denotes an \emph{expression hole},
+$\TUnknown$ is a type hole, which we identify with the unknown type from gradual type theory \cite{Siek06a}.
+Finally, to model the edit state of a program in development, $\EEHole$ denotes an \emph{empty expression hole},
 used to represent syntactically incomplete portions of the program à la Hazelnut
-\cite{HazelnutPOPL}. Note, however, that these are not critical to the calculus.
+\cite{HazelnutPOPL}. Note, however, that empty expression holes are not semantically critical to the calculus---they are included to emphasize that the system can handle program sketches.
 
@@
                        \mid           \EPair{\EMV}{\EMV}
-                       \mid           \EProjL{\EMV} \mid \EProjR{\EMV}
                        \mid           \EEHole \\
@@
              &       & \mid         & \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV}
-                       \mid           \ECPair{\ECMV}{\ECMV} \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
                        \mid           \ECEHole \\
@@
              &       & \mid         & \ECSynNonMatchedArrow{\ECMV} \mid \ECAnaNonMatchedArrow{\ECMV}
-                       \mid           \ECSynNonMatchedProd{\ECMV} \mid \ECAnaNonMatchedProd{\ECMV}
   \end{array}\]
-  \caption{Syntax of an extended GTLC and its marked version.}
+  \caption{Syntax of the marked lambda calculus.}
   \label{fig:calculus-syntax}
@@
 
-Types classify expressions by a \emph{bidirectional type system} \cite{BidirTyping}, which employs
+Before giving any of the rules, let us summarize the overall judgemental structure of the calculus.
+Note that colors in the judgement forms below are entirely redundant reading aids---color has no semantic significance.
+
+As our starting point, types classify unmarked expressions by a completely standard \emph{bidirectional type system} \cite{BidirTyping,LocalInf}, which employs
 two mutually defined judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$,
-establishes that, under the typing context $\ctx$, the expression $e$ synthesizes or locally infers
-\cite{Localinf} the type $\TMV$. \emph{Type analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$,
-states that the expression $\EMV$ may appear where an expression of type $\TMV$ is expected. We may
-give these typing semantics traditionally, without consideration for error localization and
-recovery.
+establishes that, under the typing context $\ctx$, the expression $e$ synthesizes or locally infers the type $\TMV$. \emph{Type analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$,
+states that the expression $\EMV$ may appear where an expression of type $\TMV$ is expected.
 
-The marked lambda calculus builds atop these typing semantics into a two-layer system:
+The key operation is marking, which transforms an unmarked expression into a marked expression,
+    inserting error marks where appropriate.
 %
-\begin{itemize}
-  \item The \textbf{unmarked language}, which is the original language. Expressions of this language
-    are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.
-
-  \item The \textbf{marked language}, which mirrors the structure of the unmarked language but is
-    extended with explicit \emph{error marks}. We call expressions of this language to be
-    \emph{marked expressions}, denoted $\ECMV$.
-
-  \item \textbf{Marking}, which \emph{marks} any unmarked expression into a marked expression,
-    inserting error marks where appropriate.
-\end{itemize}
-%
-In other words, we extend the ordinary typing semantics of a language with a secondary ``marked''
-language and ``mark'' unmarked programs into marked ones. This corresponds to a type checking
+This corresponds to a type checking
 process with error localization and recovery. The marked language may then serve as a foundation for
-other semantic services, such as constraint-based inference assistance (as discussed in
+other semantic services, such as constraint-based inference (as discussed in
 \cref{sec:thi}). 
-
-\Cref{fig:calculus-syntax} furthermore gives the syntax for marked expressions (as $\ECMName$),
-which replicates all the forms of unmarked expressions. There are additionally several kinds of
-\emph{error marks}, which represent the different type errors that may arise. Intuitively, they each
-denote a different error message that might be shown by an editor or emitted by a compiler.
+Intuitively, each of the mark forms corresponds to a different kind of error message that might be shown by an editor or emitted by a compiler.
+(We do not specify what the error message should say in this paper.)
 
 Throughout the remainder of this section, as we formulate marking for the GTLC, we will motivate and
-give precise semantics for these error marks. Note that these constitute a non-exhaustive list;
-depending on the syntactic forms and typing features found in a given language, other kinds of marks
-may be necessary. Indeed, in \cref{sec:calculus-poly}, we consider the kinds of marks that may be
-required in an language with explicit polymorphism.
+give precise semantics for each error mark. When the core is extended, new marks may be needed, as we discuss below.
 
-The marked language possesses its own typing semantics, also formulated bidirectionally. We write
-$\ctxSynTypeM{\ctx}{\ECMV}{\TMV}$ for synthesis and $\ctxAnaTypeM{\ctx}{\ECMV}{\TMV}$ for analysis.
-Finally, marking is bidirectional as well. The synthetic marking judgment
+The marked language possesses its own type system, also formulated bidirectionally. We write
+$\ctxSynTypeM{\ctx}{\ECMV}{\TMV}$ for synthesis and $\ctxAnaTypeM{\ctx}{\ECMV}{\TMV}$ for analysis (in addition to the color difference, note the subscript on the turnstile to distinguish marked from unmarked typing).
+
+Finally, the key judgement, \emph{marking}, is also specified bidirectionally. The synthetic marking judgment
 $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
@@
 How can we ensure that the marking procedure is correctly defined? There are two critical
-metatheorems to keep in mind as we continue. The first is a \emph{totality} of marking:
+metatheorems that guide us as we continue. The first is a \emph{totality} of marking:
 %
@@
     \item For all $\ctx$ and $\EMV$, there exist $\ECMV$ and $\TMV$ such that
-      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
+      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ and  $\ctxSynTypeM{\ctx}{\ECMV}{\TMV}$.
     \item For all $\ctx$, $\EMV$, and $\TMV$, there exists $\ECMV$ such that
-      $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$.
+      $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ and $\ctxAnaTypeM{\ctx}{\ECMV}{\TMV}$.
   \end{enumerate}
@@
 %
-That is, we may mark \emph{any} syntactically well-formed program.
+That is, we may mark \emph{any} syntactically well-formed program in any context, resulting in a \emph{well-typed} marked program.
 
+Furthermore, since error marks are effectively annotations on top of the program, marking should
+preserve syntactic structure modulo those marks. \cref{fig:calculus-mark-erasure} defines \emph{mark erasure}, which converts marked expressions back into unmarked expression by
+removing error marks. 
 \input{figs/calculus-mark-erasure}
 
-Furthermore, since error marks are effectively annotations on top of the program, marking should
-preserve syntactic structure modulo those marks. \cref{fig:calculus-mark-erasure} thus gives a
-notion of \emph{mark erasure}, which converts marked expressions back into unmarked expression by
-removing error marks. Then, \cref{thm:calculus-marking-well-formedness} provides a
+Then, \cref{thm:calculus-marking-well-formedness} provides a
 \emph{well-formedness} criterion for marking.
@@
   \begin{enumerate}
-    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxSynTypeM{\ctx}{\ECMV}{\TMV}$ and
+    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then 
       $\erasesTo{\ECMV}{\EMV}$.
-    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then $\ctxAnaTypeM{\ctx}{\ECMV}{\TMV}$ and
+    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then  and
       $\erasesTo{\ECMV}{\EMV}$.
@@
 %
-In other words, marking and then erasing error marks from the result gives the original unmarked
-expression. We also ensure that marking gives well-typed marked expressions; indeed, their types
-should match those synthesized or analyzed against in marking.
 
-In the remainder of this subsection, we consider the different type errors that may arise in the
-unmarked language while keeping in mind these two theorems; in this way, the traditional typing
-semantics of the unmarked language guide the construction of the marking process.
+These metatheorems imply that to go from a standard bidirectional system for the unmarked language to a marking system, we need to handle all possible failure modes with appropriate marks and marking logic (otherwise totality would be violated) and without otherwise changing the program (otherwise well-formedness would be violated).
+Let us consider each form in turn.
 
@@
 
-\subsection{Polymorphism and Other Typing Features}
-\label{sec:calculus-poly}
-
-We now consider how the calculus might be extended to richer typing features such as parametric
-polymorphism. Specifically, we consider System F-style explicit polymorphism.\todo{write me}
-
 \subsection{Agda Mechanization}
@@
 The standard approach of modeling judgments as inductive datatypes and rules as constructors for
-those datatypes is taken. For marked expressions, we opt for implicitly typed terms [citation],
+those datatypes is taken. For marked expressions, we opt for implicitly typed terms,
 which gives us part of \cref{thm:calculus-marking-well-formedness} for free. For the convenience of
@@
 
-\section{The Marked Lambda in Structured Editing}
+\subsection{Polymorphism and Judgemental Structure}
+\label{sec:calculus-poly}
+
+The case studies above demonstrate that the judgemental structure of the marked lambda calculus is suitable to support a variety of extensions.
+We hope that this spurs language designers to designed mark variants of other calculi, following the recipe discussed above.
+
+In calculi with richer judgemental structure---for example, System F, which introduces type validity and type variables \cite{tapl}---the general approach is again to define marks for any new kind of error that might arise, e.g.
+the appearance of a type with free type variables, and marking judgements corresponding to each judgement in the original language, e.g.
+here defining marking for types.
+One small subtlety in doing so is that a marked free variable might end up on the right side of the typing arrows.
+However, it is straightforward to treat a marked free variable simply as another form of unknown type for the purposes of type consistency and other operations.
+Type functions and type application operate much like standard functions and function application after this additional judgemental structure is accounted for.
+We leave as future work the task of defining a marked version of even more elaborate bidirectional type systems, e.g.
+\citet{dunfield13}'s rather substantial formulation of implicit type application in a bidirectional setting (which would have to handle, for example, the situation where no implicit argument can be resolved).
+
+\section{Integrating Marking with Editing}
 \label{sec:calculus-structured-editing}
+\subsection{Hazel Implementation}
+We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of the Hazel language server.
+In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching \cite{oopsla23}, strings, and explicit polymorphism (all following the same recipe we took above).
+Hazel is implemented in OCaml and compiles to Javascript via \li{js_of_ocaml} \cite{jsoo} for use in the browser.
+The supplemental material contains a pre-built implementation.
 
-[Something about structure editors]
+The result of this effort is the first full-scale typed functional language equipped with a language server that solves the gap problem: all of Hazel's semantic services, which include type hints \cite{potter1}, semantic navigation, semantic code completion \cite{potter1,blinn}, contextualized documentation \cite{potter2}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{HazelnutLive}, even semantic services that require run-time information, like live testing, are available as long as the program sketch is syntactically well-formed.
+%
+
+Hazel offers both a textual syntax, with explicit empty holes, and more interestingly,
+a structure editor that is able to additionally maintain a form
 

### 2023-03-02 related.tex
Updates from Overleaf

--- 2023-02-25 related.tex
@@
 
-Jeremy Siek Gradual Typing with Inference
-- Siek's type inference halts when it can't solve, ours doesn't
+\paragraph{Error Localization}
+
+\paragraph{Type Inference and Unification}
+\citet{GradualInfer} explore a version of unification for gradually typed languages based on Huet's unification algorithm \cite{Huet}.
+Other common approaches to unification are generally based on Hindley-Milner type inference \cite{MilnerInfer}.
+These algorithms are efficient, but can fail to solve constraints in the face of inconsistencies.
+
+% Complex techniques, e.g. based on machine learning or manual weighting, have been proposed to heuristically localize these errors to expressions.
+
+\paragraph{Correcting Sets and Assigning Blame}
+The generation of correcting sets is a common approach to error localization.
+The correcting set is assigned blame for failure in type inference \cite{sherrloc, typeinferDif, Pavlinovic2015}.
+Other approaches to assigning blame include statistically derived approaches that leverage machine learning to predict the best candidates for blame \cite{SeidelBlame}.
+
+\paragraph{Dynamic Witnesses}
+Explanations for type errors can be well illustrated by providing sample inputs (dynamic witnesses) that elicit runtime errors.
+With this approach, one can generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with around 70\% accuracy \cite{Seidel2016}.

### 2023-03-02 intro.tex

@@
 
-Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as is featured in Agda \cite{agda}, GHC Haskell \cite{haskell-holes}, Idris \cite{idris-holes}, and other languages), syntax error recovery \cite{error-reovery}, and structure editing \cite{HazelnutPOPL}.
+Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as is featured in Agda \cite{norell:thesis}, GHC Haskell \cite{haskell-holes}, Idris \cite{DBLP:journals/jfp/Brady13}, and other languages), syntax error recovery \cite{error-reovery}, and structure editing \cite{HazelnutPOPL}.
 Hazel has both a textual syntax and a structure editor, which are briefly described in Sec.~\ref{sec:calculus-hazel}.\todo{sec} 
@@
 However, as a secondary contribution, 
-we briefly discuss in Sec.~\ref{sec:calculus-structured-editing} how (1)  total marking can resolve the problem of undefined editing behavior in the Hazelnut structure editor calculus, 
+we briefly discuss in Sec.~\ref{sec:calculus-hazelnut} how (1)  total marking can resolve the problem of undefined editing behavior in the Hazelnut structure editor calculus, 
 and (2) integrating marking with typed structure editing allows us to re-mark only where necessary based on the edit location and the type and binding structure.
@@
 
-In Sec.~\ref{sec:thi}\todo{sec}, we describe how we blend local and constraint-based type inference in Hazel. In particular,  
+In Sec.~\ref{sec:thi}, we describe how we blend local and constraint-based type inference in Hazel. In particular,  
 we generate constraints for each type hole that arises (either explicitly in the program or during type error localization and recovery as defined in  Sec.~\ref{sec:calculus}). 

### 2023-03-02 calculus.tex
Updates from Overleaf

@@
 
-The marked lambda calculus is based on the gradually typed lambda calculus \cite{Siek06a}\todo{snapl} extended
-with numbers and booleans and empty expression holes. The syntax is given in
+The marked lambda calculus is based on the gradually typed lambda calculus \cite{Siek06a} extended
+with numbers, booleans, and empty expression holes \cite{HazelnutSNAPL}. The syntax is given in
 \cref{fig:calculus-syntax}. There are two expression languages:
@@
 
-The marking rule mirrors \textsc{USPlus} closely. Since an expected type for each operand is known,
+The marking rule parallels \textsc{USPlus} closely. Since an expected type for each operand is known,
 we shift the responsibility for any type errors to them. Hence, we recursively mark each operand in
@@
 
-\subsection{Polymorphism and Judgemental Structure}
+\subsection{Polymorphism and Richer Judgemental Structures}
 \label{sec:calculus-poly}
@@
 \subsection{Hazel Implementation}
+\label{sec:calculus-hazel}
+
 We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of the Hazel language server.
-In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching \cite{oopsla23}, strings, and explicit polymorphism (all following the same recipe we took above).
+In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching \cite{oopsla23}, strings, and explicit polymorphism (all following the same recipe we developed above).
 Hazel is implemented in OCaml and compiles to Javascript via \li{js_of_ocaml} \cite{jsoo} for use in the browser.
@@
 
-The result of this effort is the first full-scale typed functional language equipped with a language server that solves the gap problem: all of Hazel's semantic services, which include type hints \cite{potter1}, semantic navigation, semantic code completion \cite{potter1,blinn}, contextualized documentation \cite{potter2}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{HazelnutLive}, even semantic services that require run-time information, like live testing, are available as long as the program sketch is syntactically well-formed.
+The result of this effort is the first full-scale typed functional language equipped with a language server that solves the semantic gap problem: all of Hazel's semantic services are available as long as the program sketch is syntactically well-formed, which include type hints \cite{potter1}, semantic navigation, semantic code completion \cite{potter1,blinn}, contextualized documentation \cite{potter2}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{HazelnutLive}, even semantic services that require run-time information, like live testing.
 %
@@
 Hazel offers both a textual syntax, with explicit empty holes, and more interestingly,
-a structure editor that is able to additionally maintain a form
+a structure editor that is able to additionally offer partial syntactic error recovery.
+In particular, the \li{tylr} editor uses a form of structured syntax error recovery that automaticallys insert empty holes to maintain syntactic well-formedness as long as matching delimiters are placed by the user \cite{tylr}.
+Efforts outside of the scope of this paper are underway to define syntactic error recovery mechanisms that can handle unmatched delimiters as well, which would achieve total syntax error recovery.
+In combination with our contributions, this would achieve the Hazel project's stated goal of ensuring that \emph{there are no meaningless editor states}.
 
-The previous section presented a formal calculus for error localization and recovery. We now discuss
-its usage in a structured editing setting. In particular, the calculus has been integrated into the
-Hazel programming environment, a live programming environment with support for typed holes.
-
-\subsection{Implementation in Hazel}
-\label{sec:calculus-hazel}
-
-\subsection{Reified Hazelnut Action Semantics}
+\subsection{Fixing Holes in Hazelnut}
 \label{sec:calculus-hazelnut}
 
-The Hazelnut action calculus introduced by \citet{HazelnutPOPL} gives a structure editing scheme
-that employs \emph{expression holes}, denoted $\EEHole$, which represent syntactically incomplete
-part of the program. Not only are terms always syntactically well-formed, they are always
-\emph{well-typed}, ensuring that semantic services are always available. To accomplish this, the
-bidirectional action calculus---given by the synthetic action judgment, written
+Earlier versions of Hazel achieved total syntax error recovery by using a term-based structure editor, which maintained delimiter matching by construction, albeit at the cost of some syntactic rigidity.
+The Hazelnut action calculus introduced by \citet{HazelnutPOPL} specifies such a structure editor formally.
+Notably, this calculus also supports a form of type error localization and recovery: edit actions can automatically insert non-empty holes around type inconsistencies that appear at the cursor after an edit action.
+This requires specifying a bidirectionally typed action calculus that propagates type and binding information to the cursor, given by the synthetic action judgment, written
 $\ASEAction{\ctx}{\EMV}{\TMV}{\EMV'}{\TMV'}{\AMV}$, and the analytic action judgment, written
-$\AAEAction{\ctx}{\EMV}{\EMV'}{\TMV}{\AMV}$---inserts on-demand \emph{error holes}, denoted
-$\hhole{\EMV}{}$, which indicate type errors. \todo{example?}
+$\AAEAction{\ctx}{\EMV}{\EMV'}{\TMV}{\AMV}$.
 
-Hazelnut, however, is unable to cope with actions that require \emph{non-local} hole fixes. For
+A significant issue with Hazelnut is that it does not allow (i.e.
+it leaves undefined) actions that require \emph{non-local} mark (i.e.
+non-empty hole, in Hazelnut's parlance) insertion or removal.
+For
 example, though the system can wrap an arbitrary expression in an arithmetic operation, it cannot
@@
 $\ELam{x}{\TUnknown}{\EMV}$ directly from $\EMV$. The binding of $x$ with the unknown type may
-shadow a previous declaration and require the removal of error holes anywhere in body, but Hazelnut
-may only perform such fixes for immediately local holes. Such actions remain undefined, i.e.
-unperformable.
+shadow a previous declaration and require the removal of error holes in body. The calculus includes manual non-empty hole insertion and removal actions as a workaround.
 
-By applying the marked lambda calculus to Hazelnut, the action calculus may be \emph{wholly decoupled}
-from the typing semantics. Under this new system, the action calculus, given by the singular action
+The fundamental issue is that Hazelnut did not have a total marking system that it could deploy to remark expressions where needed.
+The marked lambda calculus allows us to solve this problem, in two different ways.
+
+One approach is to define an untyped action calculus, \emph{wholly decoupling} edit actions from typing. Under this simplified design, the action calculus, given by the singular action
 judgment, written $\AUEAction{\EMV}{\EMV'}{\AMV}$, is concerned only with the manipulation of
-syntax, and the marking procedure yields statically meaningful terms with error marks, now error
-holes, inserted at the appropriate positions as a kind of editor decoration. This solves the problem
+syntax, and the total marking procedure yields statically meaningful terms with error marks inserted at the appropriate positions as a kind of editor decoration. This solves the problem
 outlined above; since marking operates on the entire program after each action, arbitrary
@@
 
-Alternatively, instead of a wholly untyped action calculus, one may re-adapt the existing Hazelnut
-calculus with scope-based marking. Action operate over \emph{marked expressions}, and error marks
-operate as error holes. The system is once again bidirectional, writing (note $\ECMV$ instead of
-$\EMV$) the judgments $\ASEAction{\ctx}{\ECMV}{\TMV}{\ECMV'}{\TMV'}{\AMV}$ and
-$\AAEAction{\ctx}{\ECMV}{\ECMV'}{\TMV}{\AMV}$. Then, when wrapping $\ECMV$ into
-$\ECLam{x}{\TUnknown}{\ECMV}$, we may simply re-mark the body under the context with $x$ binding
+Alternatively, instead of a wholly untyped action calculus, one may directly integrate re-marking logic into the typed Hazelnut
+calculus, making use of the type and scoping information being propagated to re-mark only in scopes where it might be necessary. For example, we can wrap $\ECMV$ into
+$\ECLam{x}{\TUnknown}{\ECMV}$ by re-mark the body under the context with $x$ binding
 the unknown type:
@@
 \]%
-This system optimizes over the untyped version by performing marking only when necessary. In
-particular, whenever binders are introduced around an expression, marking redraws error holes. The
+The
 supplemental material contains a complete formal description of both variants, and the untyped
 action semantics and related metatheory are mechanized in Agda.
+Because of the simplicity of the untyped version, and because marking is defined separately, this Agda mechanization can serve universally as a baseline for correctness of approaches, like our initial design that aim to make use of type and scope information, and perhaps other analyses, to minimize re-marking.
+We leave a complete assessment of this and other remarking optimization approaches to future work.

### 2023-03-02 let.tex
Updates from Overleaf

@@
 expressions and simple patterns, writing $\PWild$ for the wildcard pattern, $\PVar{x}$ for a
-variable pattern, $\PPair{\PMV_1}{\PMV_2}$ for a pair. 
+variable pattern, and $\PPair{\PMV_1}{\PMV_2}$ for a pair. 
 %
@@
 
-Though the distinction is benign in this example, there is a subtle semantic distinction: synthetic
+Though it is benign in this example, there is a subtle semantic distinction: synthetic
 mode imposes \emph{no} type constraints, whereas analytic mode imposes a \emph{trivial} type

### 2023-03-02 calculus.tex
Re-marking instead of remarking
Updates from Overleaf

@@
 destructing let expressions with granular type annotations, in \cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} briefly explores how
-the system might be extended to more complex typing features, such as polymorphism.
+the system might be extended to more complex judgemental structure, e.g. polymorphism.
 
@@
 supplemental material contains a complete formal description of both variants, and the untyped
-action semantics and related metatheory are mechanized in Agda.
-Because of the simplicity of the untyped version, and because marking is defined separately, this Agda mechanization can serve universally as a baseline for correctness of approaches, like our initial design that aim to make use of type and scope information, and perhaps other analyses, to minimize re-marking.
-We leave a complete assessment of this and other remarking optimization approaches to future work.
+action semantics and related metatheory are mechanized in Agda. Because of the simplicity of the
+untyped version, and because marking is defined separately, this Agda mechanization can serve
+universally as a baseline for correctness of approaches, like our initial design that aim to make
+use of type and scope information, and perhaps other analyses, to minimize re-marking. We leave a
+complete assessment of this and other re-marking optimization approaches to future work.

### 2023-03-02 calculus.tex
Changes from Overleaf

@@
 
-As our starting point, types classify unmarked expressions by a completely standard \emph{bidirectional type system} \cite{BidirTyping,LocalInf}, which employs
+As our starting point, types classify unmarked expressions by a completely standard \emph{bidirectional type system} \cite{BidirTyping,Localinf}, which employs
 two mutually defined judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$,
@@
 synthesize the unknown type. Similarly to the inconsistent type mark, this allows type checking to
-proceed, with usage of the unbound variable permitted in any expression. \todo{intution for
-correctness} \todo{relate to gradual recovery}
+proceed, with usage of the unbound variable permitted in any expression.
 
@@
 considered an arrow type. Defined in \cref{fig:calculus-matched-arrow}, this notion is purely a
-technical mechanism to avoid duplication of rules related to arrow types.
+technical mechanism to avoid duplication of rules related to arrow types \cite{siek2015refined}.
 
@@
 $\TMV_2$, which we define inductively in \cref{fig:calculus-type-join}. We choose the ``more
-specific'' type of the two\todo{intuition for why this is correct}. In analytic position, since
+specific'' type of the two. In analytic position, since
 there is an expected type for both branches, we shift the blame for any errors to them.
@@
 \label{sec:calculus-agda}
-% TODO clarify exactly what is/isn't mechanized
 
 The semantics and metatheory presented above, including the extension with binary products, have
-been fully mechanized in the Agda proof assistant. Though the mechanization's documentation contains
+been fully mechanized in the Agda proof assistant.
+The mechanization additionally includes the decoupled Hazelnut action semantics described in \cref{sec:calculus-hazelnut}.
+Though the mechanization's documentation contains
 more detailed discussion regarding technical decisions made therein, we highlight some important
@@
 
-In calculi with richer judgemental structure---for example, System F, which introduces type validity and type variables \cite{tapl}---the general approach is again to define marks for any new kind of error that might arise, e.g.
+In calculi with richer judgemental structure---for example, System F, which introduces type validity and type variables \cite{reynolds1988introduction}---the general approach is again to define marks for any new kind of error that might arise, e.g.
 the appearance of a type with free type variables, and marking judgements corresponding to each judgement in the original language, e.g.
@@
 We leave as future work the task of defining a marked version of even more elaborate bidirectional type systems, e.g.
-\citet{dunfield13}'s rather substantial formulation of implicit type application in a bidirectional setting (which would have to handle, for example, the situation where no implicit argument can be resolved).
+\citet{dunfield2013complete}'s rather substantial formulation of implicit type application in a bidirectional setting (which would have to handle, for example, the situation where no implicit argument can be resolved).
 
@@
 We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of the Hazel language server.
-In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching \cite{oopsla23}, strings, and explicit polymorphism (all following the same recipe we developed above).
+In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching, strings, and explicit polymorphism (all following the same recipe we developed above).
 Hazel is implemented in OCaml and compiles to Javascript via \li{js_of_ocaml} \cite{jsoo} for use in the browser.
@@
 
-The result of this effort is the first full-scale typed functional language equipped with a language server that solves the semantic gap problem: all of Hazel's semantic services are available as long as the program sketch is syntactically well-formed, which include type hints \cite{potter1}, semantic navigation, semantic code completion \cite{potter1,blinn}, contextualized documentation \cite{potter2}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{HazelnutLive}, even semantic services that require run-time information, like live testing.
+The result of this effort is the first full-scale typed functional language equipped with a language server that solves the semantic gap problem: all of Hazel's semantic services are available as long as the program sketch is syntactically well-formed, which include type hints \cite{potter2020hazel}, semantic navigation, semantic code completion \cite{potter2020hazel,blinn2022integrative}, contextualized documentation \cite{potter2022contextualized}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{HazelLive}, even semantic services that require run-time information, like testing.
 %
@@
 a structure editor that is able to additionally offer partial syntactic error recovery.
-In particular, the \li{tylr} editor uses a form of structured syntax error recovery that automaticallys insert empty holes to maintain syntactic well-formedness as long as matching delimiters are placed by the user \cite{tylr}.
+In particular, the \li{tylr} editor uses a form of structured syntax error recovery that automatically insert empty holes to maintain syntactic well-formedness as long as matching delimiters are placed by the user \cite{moon2022tylr}.
 Efforts outside of the scope of this paper are underway to define syntactic error recovery mechanisms that can handle unmatched delimiters as well, which would achieve total syntax error recovery.

### 2023-03-02 conclusion.tex
Changes from Overleaf

@@
 \label{sec:conclusion}
-Taking a formal approach has several benefits: it allows us to mechanically \emph{prove} that the system exhibits total error recovery, i.e. that recovery can never fail, 
-it gives us the ability to precisely specify the type and binding structure of
+
+Programming is increasingly a live collaboration between human programmers and sophisticated semantic services. 
+These services need to be able to reason throughout the programming process, not just when the program is already complete.
+This paper lays down rigorous type-theoretic foundations for doing just that.
+Local type inference helps make the localization decisions we make systematic and predictable, and type hole inference shows how local and constraint-based type inference might operate hand-in-hand, rather than as alternatives.
+We hope that language designers will use the techniques introduced in this paper to consider more rigorously, perhaps even formally, the problems of type error localization and error recovery when designing future languages.

### 2023-03-02 intro.tex
Intro screenshot and fix compilation issues
Changes from Overleaf

--- 2023-03-02 intro.tex
@@
 reporting error locations in compiler error messages 
-or directly in the editor via markings decorating the localized errors. Developers attend to these error locations when debugging type errors.\todo{cite}
-Many systems also attempt recovery of certain services in certain situations, discussed below.
+or directly in the editor via markings decorating the localized errors. Developers attend primarily to these error locations when debugging type errors \cite{DBLP:journals/jfp/JoostenBH93}.
+Many systems also attempt recovery in certain situations, discussed below.
 
 However, type error localization and recovery mechanisms have developed idiosyncratically, 
-often as folklore amongst language and tool implementors. 
-Different type checkers or language servers \cite{langauge-servers,merlin}\todo{cite}, even for the same language, localize and recover from type errors in different ways, 
+in part as folklore amongst language and tool implementors. 
+Different type checkers or language servers \cite{barros2022editing,bour2018merlin}, even for the same language, localize and recover from type errors in different ways, 
 with little in the way of unifying theory of the sort that grounds the design of modern type systems themselves.
@@
 this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{elm}. Hazel  
-supports local type inference, specified in the well-established bidirectional style \cite{pierce,hazelnut,BidirTyping}:
+supports local type inference, specified in the well-established bidirectional style \cite{Localinf,HazelnutPOPL,BidirTyping}:
 \vspace{-3px}
@@
 
-A language server with support for type error recovery would 
+An implementation with support for type error recovery, like Elm's compiler or OCaml's \li{merlin} \cite{bour2018merlin}, would 
  be tasked with continuing past this first error.
@@
  Intuitively, however, it seems that this type is actually \emph{unnecessary} to make an error localization decision about \li{y}: 
-it is unbound, so a second error can be simultaneously localized despite the missing information.
+it is unbound, so a second error can be simultaneously localized despite the missing information. Elm stops here.
 
@@
 There are several ways to localize this error. 
-One approach would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
+One approach, taken e.g. by \li{merlin}, would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
 Heuristics have been developed to help make this choice less arbitrary, e.g. 
-by observing recent editing history \todo{cite}, 
-or training a machine learning model \todo{cite}. 
-The other approach, which Hazel takes, is to localize the inconsistency to the conditional expression itself, reporting that the problem is that the branch types differ, as indicated in
-the screenshot in \autoref{fig:hazel-intro}.
-When the cursor is on the conditional expression, the Hazel type inspector---a status bar that reports typing information, including type error messages, about the term at the cursor \cite{hannah1}---shows the following error:
+by observing recent editing history \cite{steady-typing}, 
+or training a machine learning model \cite{SeidelBlame}. 
+The other approach, which Hazel takes, is to localize the inconsistency to the conditional expression itself, reporting that the problem is that the branch types differ, as indicated above.
+When the cursor is on the conditional expression, the Hazel type inspector---a status bar that reports typing information, including type error messages, about the term at the cursor \cite{potter2020hazel}---shows:
 
-TODO\todo{tpye inspector}
+\begin{center}
+  \includegraphics[width=0.9\textwidth]{images/haz3l-inconsistent-branches-cursor-inspector.png}
+\end{center}
 
@@
 
-This paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus} (Sec.~\ref{sec:actions}).\todo{sec}
+This paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus} (\cref{sec:calculus}).
 We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
-type annotations to expressions \cite{pierce,BidirTyping}, (2) gradual typing, which offers a principled approach for recovering from missing type information  \cite{GradualTyping,sieksnapl}\todo{siek05?}, and
-(3) non-empty holes, which function as syntactic membranes marking erroneous terms, which the editor can display directly on-screen \cite{HazelnutPOPL}.
+type annotations to expressions \cite{Localinf,BidirTyping}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{Siek06a,siek2015refined}, and 
+(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{HazelnutPOPL}, which the editor can display directly on-screen.
 
-The Hazelnut calculus \cite{HazelnutPOPL} previously brought these ideas together, but its account of type error localization and recovery was partial and tied integrally into its structure editor (see Sec.~\ref{sec:calculus-hazel}).
-We resolve these problems to formally establish, with accompanying metatheory mechanized in Agda, that the marked lambda calculus features \emph{total error recovery}: {every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{sketching}) can be marked, i.e.
+The Hazelnut calculus \cite{HazelnutPOPL} previously brought these ideas together, but its account of type error localization and recovery was partial and tied integrally into its structure editor (see \cref{sec:calculus-hazel}).
+We resolve these problems to formally establish, with accompanying metatheory mechanized in Agda, that the marked lambda calculus features \emph{total error recovery}: {every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solar2013program}) can be marked, i.e.
 its errors can be localized, such that the resulting program has a well-defined type and binding structure.
@@
 
-This system, scaled up and implemented in Hazel (Sec.~\ref{sec:calculus-hazel}), solves the semantic gap problem: Hazel's semantic services are available as long as the program sketch is syntactically well-formed.
+This system, scaled up and implemented in Hazel (\cref{sec:calculus-hazel}), solves the semantic gap problem: Hazel's semantic services are available as long as the program sketch is syntactically well-formed.
 % which include type hints \cite{potter1}, semantic navigation, semantic code completion \cite{potter1,blinn}, contextualized documentation \cite{potter2}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{HazelnutLive}, even semantic services that require run-time information, like live testing,
@@
 
-Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as is featured in Agda \cite{agda}, GHC Haskell \cite{haskell-holes}, Idris \cite{idris-holes}, and other languages), syntax error recovery \cite{error-reovery}, and structure editing \cite{HazelnutPOPL}.
-Hazel has both a textual syntax and a structure editor, which are briefly described in Sec.~\ref{sec:calculus-hazel}.\todo{sec} 
+Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as is featured in Agda \cite{norell:thesis}, GHC Haskell \cite{haskell-holes}, Idris \cite{DBLP:journals/jfp/Brady13}, and other languages), syntax error recovery \cite{medeiros20,sorkin11}, and structure editing \cite{HazelnutPOPL}.
+Hazel has both a textual syntax and a structure editor, which are briefly described in \cref{sec:calculus-hazel}.
 Marking is not integrated with structure editing, as it was in Hazelnut. 
@@
 However, as a secondary contribution, 
-we briefly discuss in Sec.~\ref{sec:calculus-structured-editing} how (1)  total marking can resolve the problem of undefined editing behavior in the Hazelnut structure editor calculus, 
+we briefly discuss in \cref{sec:calculus-hazelnut} how (1)  total marking can resolve the problem of undefined editing behavior in the Hazelnut structure editor calculus, 
 and (2) integrating marking with typed structure editing allows us to re-mark only where necessary based on the edit location and the type and binding structure.
 
-Local type inference pairs well with type error localization and recovery because information flows systematically through the tree, leading to predictable and systematic localization decisions (as informally observed also by \cite{pierce}).
+Local type inference pairs well with type error localization and recovery because information flows systematically through the tree, leading to predictable and systematic localization decisions (as informally observed also by \cite{LocalInf}).
 Languages with constraint-based type inference gather constraints from all relevant locations in a program.
@@
 
-In Sec.~\ref{sec:thi}\todo{sec}, we describe how we blend local and constraint-based type inference in Hazel. In particular,  
-we generate constraints for each type hole that arises (either explicitly in the program or during type error localization and recovery as defined in  Sec.~\ref{sec:calculus}). 
+In \cref{sec:thi}, we describe how we blend local and constraint-based type inference in Hazel. In particular,  
+we generate constraints for each type hole that arises (either explicitly in the program or during type error localization and recovery as defined in \cref{sec:calculus}). 
 We develop a unification algorithm that can recover from inconsistent constraints by maintaining possible type hole solutions. For holes that can be uniquely solved, we suggest the solution. 

### 2023-03-02 related.tex
Changes from Overleaf

@@
 
-\paragraph{Error Localization}
+The contributions of this paper build directly on the Hazelnut type system \cite{HazelnutPOPL}, which is discussed extensively throughout.
+Non-empty holes in Hazelnut generalize to marks in this work.
+In brief, we contribute a total marking procedure and type hole inference procedure for a system based closely on Hazelnut, and use it to fix some expressiveness issues in Hazelnut's edit action calculus.
 
-\paragraph{Type Inference and Unification}
+Our focus is on localization and recovery but not type error repair, as has been considered by other work \cite{lerner07}. We hope that our work will drive future work on rigorous repairs.
+
 \citet{GradualInfer} explore a version of unification for gradually typed languages based on Huet's unification algorithm \cite{Huet}.
@@
 
-\paragraph{Correcting Sets and Assigning Blame}
 The generation of correcting sets is a common approach to error localization.
@@
 Other approaches to assigning blame include statistically derived approaches that leverage machine learning to predict the best candidates for blame \cite{SeidelBlame}.
+Our focus in this paper was to avoid heuristics and data-driven approaches and instead focus on systematic semantic approaches.
+Data-driven approaches could perhaps be layered to improve suggestions.
 
-\paragraph{Dynamic Witnesses}
 Explanations for type errors can be well illustrated by providing sample inputs (dynamic witnesses) that elicit runtime errors.

### 2023-03-23 intro.tex
Updates from Overleaf

@@
 
-Local type inference pairs well with type error localization and recovery because information flows systematically through the tree, leading to predictable and systematic localization decisions (as informally observed also by \cite{LocalInf}).
+Local type inference pairs well with type error localization and recovery because information flows systematically through the tree, leading to predictable and systematic localization decisions (as informally observed also by \cite{Localinf}).
 Languages with constraint-based type inference gather constraints from all relevant locations in a program.

### 2023-07-11 abstract.tex

--- 2023-02-27 abstract.tex
@@
 \begin{abstract}
-Type systems typically only define the conditions under which an expression is well-typed, leaving ill-typed expressions entirely meaningless. 
-This approach is insufficient as the basis for language servers for modern programming environments, which are expected to sensibly localize any type errors and recover, i.e.
-continue to provide a variety of downstream semantic services.
-This paper addresses this problem, contributing the first comprehensive formal account of type error localization and recovery: the marked lambda calculus.
-In particular, we define a gradual type system for expressions with marked errors and holes, together with a total procedure for marking unmarked expressions.
-We mechanize the metatheory of the marked lambda calculus in Agda and implement it, scaled up, as the new basis for Hazel, a full-scale live functional programming environment.
-As a secondary contribution, we show how total marking resolves 
-the problem of undefined behavior in the Hazelnut structure editor calculus
-(though marking itself does not require a structure editor).
+Type systems typically only define the conditions under which an expression is well-typed, leaving ill-typed expressions formally meaningless. 
+This approach is insufficient as the basis for language servers driving modern programming environments, which are expected to recover from simultaneously localized errors and continue to provide a variety of downstream semantic services.
+This paper addresses this problem, contributing the first comprehensive formal account of total type error localization and recovery: the marked lambda calculus.
+In particular, we define a gradual type system for expressions with marked errors, which operate as non-empty holes, together with a total procedure for marking arbitrary unmarked expressions.
+We mechanize the metatheory of the marked lambda calculus in Agda and implement it, scaled up, as the new basis for Hazel, a full-scale live functional programming environment with, uniquely, no meaningless editor states.
+% As a secondary contribution, we show how total marking resolves 
+% the problem of undefined behavior in the Hazelnut structure editor calculus
+% (though marking itself does not require a structure editor).
 
-The marked lambda calculus supports local type inference, i.e.
-it is bidirectionally typed.
-Constraint-based type inference notoriously complicates error localization and recovery.
-We approach this problem by deploying constraint solving as a type-hole-filling layer atop our gradual bidirectionally typed core.
-The unification algorithm we contribute localizes errors arising from inconsistent constraints to type holes with traced  provenances, rather than attempting to localize them in an \emph{ad hoc} manner to particular expressions.
-The user can choose how to shift these errors to expressions by selecting from suggested partial solutions, or allow the unresolved type holes to recover gradually.
+The marked lambda calculus is bidirectionally typed, so localization decisions are systematically predictable based on a local flow of typing information. 
+Constraint-based type inference can bring more distant information to bear in discovering inconsistencies but this notoriously complicates error localization.
+We approach this problem by deploying constraint solving as a type-hole-filling layer atop this gradual bidirectionally typed core.
+Errors arising from inconsistent unification constraints are localized exclusively to type and expression holes, i.e.
+the system identifies unfillable holes using a sytem of traced  provenances, rather than localized in an \emph{ad hoc} manner to particular expressions.
+The user can then interactively shift these errors to particular downstream expressions by selecting from suggested partially consistent type hole fillings, which returns control back to the bidirectional system.
 We implement this type hole inference system in Hazel.

### 2023-07-11 intro.tex

@@
 These service gaps are fundamentally rooted in a definitional gap: a type system defined in the conventional way, 
-e.g. in the tradition of the typed lambda calculus and its derivatives \cite{TaplBook,pfpl},
+e.g. in the tradition of the typed lambda calculus and its derivatives \cite{TaplBook},
 assigns meaning only to well-typed programs. 
@@
 reporting error locations in compiler error messages 
-or directly in the editor via markings decorating the localized errors. Developers attend primarily to these error locations when debugging type errors \cite{DBLP:journals/jfp/JoostenBH93}.
+or directly in the editor via markings decorating the localized errors. Developers are known to attend to these error locations when debugging type errors \cite{DBLP:journals/jfp/JoostenBH93}.
 Many systems also attempt recovery in certain situations, discussed below. 
-
 However, type error localization and recovery mechanisms have developed idiosyncratically, 
@@
 
-Consider, the example program below, which is shown as presented to the user in
+Consider, the example ill-typed program below, which is shown as presented to the user in
 this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{elm}. Hazel  
@@
 A more practical approach, and one common even in production type checkers, 
-is to localize the first error that causes local type inference to fail and emit an explanatory error message.
-In this example, the system might report that the variable \li{f} located on Line 2 is unbound, then stop.
+is to localize the first error that causes local type inference to fail and emit an explanatory error message before terminating.
+In this example, the system might report that the variable \li{f} located on Line 2 is unbound, then terminate.
 
@@
  would appear to demand in order to proceed, here in order to determine which type the argument, \li{y}, is to be analyzed against.
- Intuitively, however, it seems that this type is actually \emph{unnecessary} to make an error localization decision about \li{y}: 
-it is unbound, so a second error can be simultaneously localized despite the missing information. Elm stops here.
+ Intuitively, however, it is clear that this knowledge is \emph{unnecessary} to make an error localization decision about \li{y}: 
+it is unbound, so a second error can be simultaneously localized despite the missing information about \li{f}. Elm's compiler stops here.
 
-To recover further, we might choose to also ignore the bidirectional type system's demand that the guard be confirmed to be a boolean expression 
+To recover further, we might choose to also ignore the bidirectional type system's demand that the conditional guard be confirmed to be a boolean expression 
 (because the type of \li{f(y)} is also not well-defined) 
@@
 There are several ways to localize this error. 
-One approach, taken e.g. by \li{merlin}, would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
+One approach, taken e.g. by Elm and \li{merlin}, would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
 Heuristics have been developed to help make this choice less arbitrary, e.g. 
@@
 or training a machine learning model \cite{SeidelBlame}. 
-The other approach, which Hazel takes, is to localize the inconsistency to the conditional expression itself, reporting that the problem is that the branch types differ, as indicated above.
+The other approach, which Hazel takes above, is to localize the inconsistency to the conditional as a whole, reporting that the problem is that the branch types differ, as indicated above.
 When the cursor is on the conditional expression, the Hazel type inspector---a status bar that reports typing information, including type error messages, about the term at the cursor \cite{potter2020hazel}---shows:
@@
 \begin{center}
-  \includegraphics[width=0.9\textwidth]{images/haz3l-inconsistent-branches-cursor-inspector.png}
+  \includegraphics[width=\textwidth]{images/haz3l-inconsistent-branches-cursor-inspector.png}
 \end{center}
@@
 This localization decision affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
-If localization had assumed that the \li{then} branch correct, as for example in the Elm type checker, then \li{x : Bool} and an error should be reported on the left operand (though this error might mislead the programmer if this earlier localization guess was incorrect).
+If localization had assumed that the \li{then} branch correct, as for example in Elm and \li{merlin}, then \li{x : Bool} and an error should be reported on the left operand (though this error might mislead the programmer if this earlier localization guess was incorrect).
 If the \li{else} branch was chosen, then \li{x : Int} and an error would not be reported. 
@@
 This informal exercise demonstrates that (1) localization choices can vary, particularly with regard to whether they make \emph{ad hoc} guesses about intent;  
-(2) when combined with error recovery, localization decisions can influence downstream localization decisions; and 
+(2) when combined with error recovery, one localization decision can influence downstream localization decisions; and 
 (3) error recovery mechanisms must be able to reason without complete knowledge about types and binding. We argue that such semantic subtleties call for a rigorous theoretical treatment of the topic.

### 2023-07-11 supplement/inferenceFormalism.tex

\newpage
\section{Formalism for Type Hole Inference}

\subsection{Syntax}
\subsubsection{Potential Types and their Graphs}
\begin{mathpar}
    \arraycolsep=4pt\begin{array}{lll}
    PotentialTypeSet~~ s & ::=
    lead(u) ~\vert~
    rep(\tau)
    \\
    PotentialTypes~~ u & ::= 
    single(t) ~\vert~ 
    cons(t, u)
    \\
    PotentialType~~ t & ::= 
      \tnum ~\vert~
      \tbool ~\vert~
      \TUnknown^p ~\vert~
      \tarr{s}{s}
      \\
    PotentialTypeGraph~~ G & ::= \tau_1: s_1, ..., \tau_n : s_n \\
    \end{array}
\end{mathpar}

\subsubsection{Potential Type Snapshots and Solutions}

\begin{mathpar}
    \arraycolsep=4pt\begin{array}{lll}
    PotentialTypesList~~ l & ::= 
    Nil ~\vert~ 
    cons(t, l)
    \\
    PotentialTypesSnapshot~~ j & ::=
    Nil ~\vert~
    cons(h, j)
    \\
    PotentialTypeSnapshot~~ h & ::=
    \tnum ~\vert~
    \tbool ~\vert~
    \TUnknown^p ~\vert~
    \tarr{j}{j}
    \\
    \end{array}
\end{mathpar}

\subsection{Inference Rules} Let $P(\tau)$ be a total function mapping a type $\tau$ to its corresponding PotentialType $t$.
Let $T(t)$ be a partial function mapping a PotentialType $t$ to its corresponding type $\tau$
\vspace{7px}\\
\judgbox{\ensuremath{\incomplete{\tau}}}
\begin{mathpar}
    \judgment{ }{
        \incomplete{\TUnknown^{p}}
    }{I-Unk}

    \judgment{
        \incomplete{\tau_1}
    }{
        \incomplete{\TArrow{\tau_1}{\tau_2}}
    }{I-Arr-L}

    \judgment{
        \incomplete{\tau_2}
    }{
        \incomplete{\TArrow{\tau_1}{\tau_2}}
    }{I-Arr-R}
\end{mathpar}

\judgbox{\ensuremath{\leader{G}{\tau}{\tau'}{u}}}
\begin{mathpar}
    \judgment{
        \isIn{\ptypGraphNode{\tau}{\ptsLead{u}}}{G}
    }{
        \leader{G}{\tau}{\tau}{u}
    }{L-self}

    \judgment{
        \isIn{
            \ptypGraphNode{\tau_1}{\ptsRep{\tau_2}}
        }{G}\\
        \leader{G}{\tau_2}{\tau_3}{u}
    }{
        \leader{G}{\tau_1}{\tau_3}{u}
    }{L-rep}
\end{mathpar}

% \judgbox{\ensuremath{\constrain{\tau_1}{\tau_2}}}
% \begin{mathpar}
%     \judgment{ }{\constrain{\tau}{\tau}}{Constrain-Refl}

%     \judgment{
%         \constrain{\tau_1}{\tau_2}
%     }{
%         \constrain{\tau_2}{\tau_1}
%     }{Constrain-Sym}

%     \judgment{
%         \incomplete{\tau_1}\\
%         \incomplete{\tau_2}\\
%         \constrain{\tau_1}{\tau_2}\\
%         \constrain{\tau_2}{\tau_3}
%     }{
%         \constrain{\tau_1}{\tau_3}
%     }{Constrain-Trans}

%     \judgment{
%         \constrain{\TArrow{\tau_1}{\tau_2}}{\TArrow{\tau_3}{\tau_4}}
%     }{
%         \constrain{\tau_1}{\tau_3}
%     }{Constrain-Arr-L}

%     \judgment{
%         \constrain{\TArrow{\tau_1}{\tau_2}}{\TArrow{\tau_3}{\tau_4}}
%     }{
%         \constrain{\tau_2}{\tau_4}
%     }{Constrain-Arr-R}
% \end{mathpar}

\judgbox{\ensuremath{\ptypsUnion{u_1}{u_2}} \equiv u_{3}}
\[\begin{array}{rcl}
    \ptypsUnion{\ptypSingle{\tarr{s_1}{s_2}}}{\ptypSingle{\tarr{s_3}{s_4}}} & = & \ptypSingle{{\tarr{(\ptypsUnion{s1}{s3})}{(\ptypsUnion{s2}{s4})}}} \\
    \ptypsUnion{\ptypSingle{t}}{\ptypSingle{t'}} & = & \ptypCons{t}{\ptypSingle{t'}} \\
    \ptypsUnion{\ptypCons{t}{u}}{\ptypSingle{t}} & = & \ptypCons{t}{u} \\
    \ptypsUnion{\ptypCons{\tarr{s_1}{s_2}}{u}}{\ptypSingle{\tarr{s_3}{s_4}}} & = & \ptypCons{\tarr{(\ptypsUnion{s1}{s3})}{(\ptypsUnion{s2}{s4})}}{u} \\
    \ptypsUnion{\ptypCons{t}{u}}{\ptypSingle{t'}} & = & \ptypCons{t}{(\ptypsUnion{u}{\ptypSingle{t'}})} \\
    \ptypsUnion{\ptypCons{t}{u}}{\ptypCons{t'}{u'}} & = & \ptypsUnion{\ptypsUnion{\ptypCons{t}{u}}{\ptypSingle{t'}}}{u'} \\
\end{array}\]

\judgbox{\ensuremath{\ptypSetSubset{G}{s_1}{s_2}}}

\begin{mathpar}
\judgment{
    s_1 \equiv rep(\tau)\\
    \leader{G}{\tau}{\tau_l}{s_{leader}}\\
    \ptypSetSubset{G}{s_{leader}}{s_2}
}{
    \ptypSetSubset{G}{s_1}{s_2}
}{S-Subset-Rep-1}

\judgment{
    s_2 \equiv rep(\tau)\\
    \leader{G}{\tau}{\tau_l}{s_{leader}}\\
    \ptypSetSubset{G}{s_1}{s_{leader}}
}{
    \ptypSetSubset{G}{s_1}{s_2}
}{S-Subset-Rep-2}

\judgment{
    s_1 \equiv lead(u_1)\\
    s_2 \equiv lead(u_2)\\
    \ptypesSubset{s_1}{s_2}
}{
    \ptypSetSubset{G}{s_1}{s_2}
}{S-Subset-Lead}
\end{mathpar}

\judgbox{\ensuremath{\ptypesSubset{u_1}{u_2}}}
\begin{mathpar}

\judgment{
    u_1 \equiv single(\tau)\\
    \tau \in u_2
}{
    \ptypesSubset{u_1}{u_2}
}{U-Subset-Single}

\judgment{
    u_1 \equiv cons(\tau, u_{tl})\\
    \tau \in u_2\\
    \ptypesSubset{u_{tl}}{u_2}
}{
    \ptypesSubset{u_1}{u_2}
}{U-Subset-Cons}

\end{mathpar}

\judgbox{\ensuremath{\ptypGraphUpdate{G}{\tau}{u}{G'}}}
\begin{mathpar}
    \judgment{
        \isNotIn{\tau}{\dom(G)}
    }{
        \ptypGraphUpdate{G}{\tau}{u}{G,~\ptypGraphNode{\tau}{\ptsLead{u}}}
    }{Update-New-Key}

    \judgment{
        \leader{G}{\tau}{\tau_l}{u_l}\\
        \ptypsUnion{u_l}{u} \equiv u_l'
    }{
        \ptypGraphUpdate{G}{\tau}{u}{G,~\ptypGraphNode{\tau_l}{\ptsLead{u_l'}}}
    }{Update-Old-Key}
\end{mathpar}

\judgbox{\ensuremath{\ptypGraphAdd{G}{\tau}{G'}}}
\begin{mathpar}
    \judgment{
        \isIn{\tau}{\dom(G)}
    }{
        \ptypGraphAdd{G}{\tau}{G}
    }{Add-Present}

    \judgment{
        \incomplete{\tau_1}\\
        \incomplete{\tau_2}\\
        \ptypGraphAdd{G_1}{\tau_1}{G_2}\\
        \ptypGraphAdd{G_2}{\tau_2}{G_3}\\
        \ptypGraphUpdate{G_3}{\TArrow{\tau_1}{\tau_2}}{\ptypSingle{\TArrow{\ptsRep{\tau_1}}{\ptsRep{\tau_2}}}}{G_4}
    }{
        \ptypGraphAdd{G_1}{\TArrow{\tau_1}{\tau_2}}{G_4}
    }{Add-Arr-I-I}

    \judgment{
        \incomplete{\tau_1}\\
        \ptypGraphAdd{G_1}{\tau_1}{G_2}\\
        \ptypGraphUpdate{G_2}{\TArrow{\tau_1}{\tau_2}}{\ptypSingle{\TArrow{\ptsRep{\tau_1}}{\ptsLead{\ptypSingle{P(\tau_2)}}}}}{G_3}
    }{
        \ptypGraphAdd{G_1}{\TArrow{\tau_1}{\tau_2}}{G_3}
    }{Add-Arr-I-C}

    \judgment{
        \incomplete{\tau_2}\\
        \ptypGraphAdd{G_1}{\tau_2}{G_2}\\
        \ptypGraphUpdate{G_2}{\TArrow{\tau_1}{\tau_2}}{\ptypSingle{\TArrow{\ptsLead{\ptypSingle{P(\tau_1)}}}{\ptsRep{\tau_2}}}}{G_3}
    }{
        \ptypGraphAdd{G_1}{\TArrow{\tau_1}{\tau_2}}{G_3}
    }{Add-Arr-C-I}

    \judgment{
        \incomplete{\tau}\\
        \ptypGraphUpdate{G}{\tau}{\ptypSingle{P(\tau)}}{G'}
    }{
        \ptypGraphAdd{G}{\tau}{G'}
    }{Add-I}
\end{mathpar}

\judgbox{\ensuremath{\unify{G}{C}{G'}}}
\begin{mathpar}
    \judgment{
        \unify{G}{
            \constraintCons{
                \constrain{\tau_{L1}}{\tau_{L2}}
            }{
                \constraintCons{\constrain{\tau_{R1}}{\tau_{R2}}}{C}
            }
        }{G'}
    }{
        \unify{G}{
            \constraintCons{
                \constrain{\TArrow{\tau_{L1}}{\tau_{R1}}}{\TArrow{\tau_{L2}}{\tau_{R2}}}
            }{C}
        }{G'}
    }{U-Arr}

    \judgment{
        \unify{G_1}{C}{G_2}\\
        \incomplete{\tau_1}\\
        \incomplete{\tau_2}\\
        \ptypGraphAdd{G_2}{\tau_1}{G_3}\\
        \ptypGraphAdd{G_3}{\tau_2}{G_4}\\
        \leader{G_4}{\tau_1}{\tau_{1l}}{u_{1l}}\\
        \leader{G_4}{\tau_2}{\tau_{2l}}{u_{2l}}\\
        \ptypsUnion{u_{1l}}{u_{2l}} \equiv u_{12}
    }{
        \unify{G_1}{
            \constraintCons{\constrain{\tau_1}{\tau_2}}{C}
        }{G_4,~\ptypGraphNode{\tau_{1l}}{\ptsLead{u_{12}}},~ \ptypGraphNode{\tau_{2l}}{\ptsRep{\tau_{1l}}}}
    }{U-I-I}

    \judgment{
        \unify{G_1}{C}{G_2}\\
        \incomplete{\tau_1}\\
        \ptypGraphAdd{G_2}{\tau_1}{G_3}\\
        \ptypGraphUpdate{G_3}{\tau_1}{\ptypSingle{P(\tau_2)}}{G_4}
    }{
        \unify{G_1}{
            \constraintCons{\constrain{\tau_1}{\tau_2}}{C}
        }{G_4}
    }{U-I-C}

    \judgment{
        \unify{G_1}{C}{G_2}\\
        \incomplete{\tau_2}\\
        \ptypGraphAdd{G_2}{\tau_2}{G_3}\\
        \ptypGraphUpdate{G_3}{\tau_2}{\ptypSingle{P(\tau_1)}}{G_4}
    }{
        \unify{G_1}{
            \constraintCons{\constrain{\tau_1}{\tau_2}}{C}
        }{G_4}
    }{U-C-I}

    \judgment{
        \unify{G}{C}{G'}
    }{
        \unify{G}{\constraintCons{c}{C}}{G'}
    }{U-C-C}

    \judgment{ }{\unify{G}{\listNil}{G}}{U-Nil}

\end{mathpar}

\judgbox{\ensuremath{\ptsSnapshot{G}{s}{j}}}
\begin{mathpar}
    \judgment{
        \leader{G}{\tau}{\tau_l}{u_l}\\
        list(u_l) \equiv l_l \\
        \listSnapshot{G}{l_l}{j}
    }{
        \ptsSnapshot{G}{\ptsRep{\tau}}{j}
    }{Snap-Rep}

    \judgment{
        list(u_l) \equiv l_l \\
        \listSnapshot{G}{l_l}{j}
    }{
        \ptsSnapshot{G}{\ptsLead{u}}{j}
    }{Snap-Lead}
\end{mathpar}

\judgbox{\ensuremath{\listSnapshot{G}{l}{j}}}
\begin{mathpar}
    \judgment{ }{
        \listSnapshot{G}{\listNil}{\listNil}
    }{Snap-Nil}

    \judgment{
        \ptsSnapshot{G}{s_1}{j_1}\\
        \ptsSnapshot{G}{s_2}{j_2}\\
        \listSnapshot{G}{tl}{j_{tl}}
    }{
        \listSnapshot{G}{ 
            \listCons{\TArrow{s_1}{s_2}}{tl}
        }{
            \listCons{\TArrow{j_1}{j_2}}{j_{tl}}
        }
    }{Snap-Arrow}

    \judgment{
        \listSnapshot{G}{tl}{j_{tl}}
    }{
        \listSnapshot{G}{\listCons{\TUnknown^p}{tl}}{j_{tl}}
    }{Snap-Unk}

    \judgment{
        \listSnapshot{G}{tl}{j_{tl}}
    }{
        \listSnapshot{G}{\listCons{t}{tl}}{\listCons{t}{j_{tl}}}
    }{Snap-Lit}
\end{mathpar}

\judgbox{\ensuremath{\solution{j}{\tau}}}
\begin{mathpar}
    \judgment{
        \fresh{n}
    }{
        \solution{\listNil}{\TUnknown^n}
    }{S-Solvable-Unk}

    \judgment{
        \solution{j_1}{\tau_1}\\
        \solution{j_2}{\tau_2}
    }{
        \solution{\listCons{\TArrow{j_1}{j_2}}{\listNil}}{\TArrow{\tau_1}{\tau_2}}
    }{S-Solvable-Arr}

    \judgment{ }{
        \solution{\listCons{k}{\listNil}}{T(k)}
    }{S-Solvable-Lit}

    \judgment{
        \fresh{n}
    }{
        \solution{\listCons{k}{tl}}{\TUnknown^n}
    }{S-Unsolvable}
\end{mathpar}

\judgbox{\ensuremath{\snapSubset{j_1}{j_2}}}
\begin{mathpar}
\judgment{
    \snapElement{h}{j}\\
    \snapSubset{tl}{j}
}{
    \snapSubset{\listCons{h}{tl}}{j}
}{J-Cons-Subset}

\end{mathpar}

\judgbox{\ensuremath{\snapElement{h}{j}}}
\begin{mathpar}
\judgment{
    h ~=~ hd
}{
    \snapElement{h}{\listCons{hd}{tl}}
}{H-In-Equal}

\judgment{
    \snapSubset{j_1}{j_3}\\
    \snapSubset{j_2}{j_4}
}{
    \snapElement{\TArrow{j_1}{j_2}}{\listCons{\TArrow{j_3}{j_4}}{tl}}
}{H-In-Equal}

\judgment{
    \snapElement{h}{tl}
}{
    \snapElement{h}{\listCons{hd}{tl}}
}{H-In-Equal}
\end{mathpar}

\subsection{Theorems}
\begin{lemma}[name=Potential Types Union Totality] For any $u_1$ and $u_2$, it must be that there exists some $u_3$ such that $\ptypsUnion{u_1}{u_2} \equiv u_3$
\begin{proof}
By induction on the structure of $\ptypsUnion{u_1}{u_2}$
\end{proof}
\end{lemma}

\noindent\rule{\textwidth}{1pt}

\todo{Prove this}
% TODO: needs to be snapshot(u_1) is subset of snapshot(u_1 UNION u_2)
% TODO: and snapshot(u_2) is subset of snapshot(u_1 UNION u_2)
\begin{lemma}[name=Potential Types Union Correctness]\
\begin{enumerate}
    \item $\ptypSetSubset{G}{snapshot_u u_1}{snapshot_u (\ptypsUnion{u_1}{u_2})}$
    \item $\ptypSetSubset{G}{snapshot_u u_2}{snapshot_u (\ptypsUnion{u_1}{u_2})}$
\end{enumerate}
\begin{proof}
\begin{enumerate}
\item By totality (previous lemma) there must be some $u_3$ such that $u_3 \equiv \ptypsUnion{u_1}{u_2}$
\item Let $j_1$ be the snapshot of $u_1$ and $j_2$ be the snapshot of $u_2$.
\item Let $l_1$ be $list(u_1)$ and $l_2$ be $list(u_2)$.
\item fdsa
\end{enumerate}
\end{proof}
\end{lemma}

\noindent\rule{\textwidth}{1pt}

\begin{lemma}[name=Leader Correctness]
Let any "well formed" graph be one for which all potential type sets $s$ associated with some type $\tau$ are either $\ptsLead{u}$ for some $u$ or eventually refer to a $\ptsLead{u}$ through a chain of $\ptsRep{\tau'}$ within $G$ where for all $\tau'$ referred to within some $\ptsRep{\tau'}$, it must be that $\isIn{\tau'}{\dom(G)}$.

For any well formed $G$ and $\tau$, if $\isIn{\tau}{\dom(G)}$ then there must exist some $\tau_l$ and $u_l$ such that $\leader{G}{\tau}{\tau_l}{u_l}$.
\end{lemma}

\begin{proof}
    Proceed by induction.
    Assume $G$ well formed and that $\isIn{\ptypGraphNode{\tau}{s}}{G}$.

    Base Case: $s = \ptsLead{u}$ for some $u$. By L-Self, it must be the case that $\leader{G}{\tau}{\tau_l}{u}$

    Suppose $s = \ptsRep{\tau'}$

    Inductive Hypothesis: There exists some $\tau_l'$ and $u_l'$ such that $\leader{G}{\tau}'{\tau_l'}{u_l'}$.

    Inductive Step: By the inductive hypothesis, we can determine that $\leader{G}{\tau'}{\tau_l'}{u_l'}$. Consequently, by L-Rep, it must be the case that $\leader{G}{\tau}{\tau_l'}{u_l'}$.
\end{proof}

\noindent\rule{\textwidth}{1pt}
% Define graph well formatted-ness in terms of leader correctness

\begin{lemma}[name=Update Correctness]
For any $\tau$ and $u$ and well formed $G$, it must be the case that:
\begin{enumerate}
    \item If $\isNotIn{\tau}{\dom(G)}$, then $\ptypGraphUpdate{G}{\tau}{u}{(G, \ptypGraphNode{\tau}{\ptsLead{u}})}$
    \item If $\isIn{\tau}{\dom(G)}$, then $\ptypGraphUpdate{G}{\tau}{u}{(G, \ptypGraphNode{\tau_l}{\ptsLead{\ptypsUnion{u_l}{u})}}}$
\end{enumerate}
Where the final graph outputted is well formed.
\end{lemma}

\begin{proof}
    Let $G$ be an arbitrary well formed graph, $\tau$ be an arbitrary type, and $u$ be an arbitrary PotentialTypes. Proceed by cases.
    \begin{itemize}
        \item Suppose $\isNotIn{\tau}{\dom(G)}$. Then by Update-New-Key, it must be that $\ptypGraphUpdate{G}{\tau}{u}{(G, \ptypGraphNode{\tau}{\ptsLead{u}})}$. Since $G$ was well formed and $\tau$ was added with a direct link to some leader, the outputted graph is well formed.
        \item Suppose that $\isIn{\tau}{\dom(G)}$. By Leader Correctness, since $G$ is well formatted, it must be the case that $\leader{G}{\tau}{\tau_l}{u_l}$. Furthermore, by Potential Types Union Totality, there exists some $u_l'$ such that $\ptypsUnion{u_l}{u} \equiv u_l'$. By Update-Old-Key, we can then conclude that $\ptypGraphUpdate{G}{\tau}{u}{(G, \ptypGraphNode{\tau_l}{\ptsLead{\ptypsUnion{u_l}{u}}})}$. Since $G$ was well formed and $\tau_l$ was previous a leader of some nodes and remains as such, it must be that the outputted graph is still well formed.
    \end{itemize}
\end{proof}

\noindent\rule{\textwidth}{1pt}

\begin{lemma}[name=Addition Correctness]
Consider any well formed $G$ and $\tau$ where $\incomplete{\tau}$.
There must exist some $G_{new}$ such that $\ptypGraphAdd{G}{\tau}{G \cup G_{new}}$ where  $G_{new}$ consists only of incomplete types not already in $G$ and where if it was not the case that $\isIn{\tau}{\dom(G)}$, then $\ptypGraphNode{\tau}{\ptsLead{\ptypSingle{P(\tau_1)}}} \in G_{new}$. Additionally, any graphs outputted by the judgement must be well formed.
\end{lemma}

\begin{proof}
     Proceed by induction.
     Assume $G$ is a well formed graph and $\incomplete{\tau}$. Let the predicate $I(G, \tau)$ be true if and only if there exists some $G_{new}$ such that $\ptypGraphAdd{G}{\tau}{G \cup G_{new}}$ where  $G_{new} = \{ \ptypGraphNode{\tau_{new}}{\ptsLead{\ptypSingle{\tau_{new}}}} ~|~ \isIn{\tau_{new}}{decompose(\tau)} \wedge \isNotIn{\tau}{dom(G)} \wedge \incomplete{\tau_{new}} \}$\\

     Base Case: $\tau$ does not equal $\TArrow{\tau_1}{\tau_2}$ for any $\tau_1$ and $\tau_2$. Proceed by cases
     \begin{itemize}
         \item $\isNotIn{\tau}{\dom(G)}$: By Update correctness, it must be the case that $\ptypGraphUpdate{G}{\tau}{u}{(G, \ptypGraphNode{\tau}{\ptsLead{u}})}$. By Add-I, it must be the case that $\ptypGraphAdd{G}{\tau}{(G, \ptypGraphNode{\tau}{\ptsLead{u}}}$.
         Since $\tau$ has no children and is incomplete, $decompose(\tau) = \{ \tau \}$. Since $\isNotIn{\tau}{\dom(G)}$, it must be that $G_{new} = \{ \tau \}$. Therefore, $I(G, \tau)$ must be true.
         \item $\isIn{\tau}{\dom(G)}$: Since $\tau$ has no children and is incomplete, $decompose(\tau) = \{ \tau \}$. However, since $\isIn{\tau}{\dom(G)}$, it must be that $G_{new} = \{ \}$. Therefore, $I(G, \tau)$ must be trivially true.
     \end{itemize}

    Let $\tau = \TArrow{\tau_1}{\tau_2}$ for some $\tau_1$ and $\tau_2$. Proceed by cases

    Inductive Hypothesis: $I(G_1, \tau_1)$ and $I(G_2, \tau_2)$

    Inductive Step: $\tau$ must be $\TArrow{\tau_1}{\tau_2}$ for any $\tau_1$ and $\tau_2$. $decompose(\tau) = \{ \tau, decompose(\tau_1), decompose(\tau_2) \}$. Proceed by cases.

    \begin{itemize}
        \item Only $\incomplete{\tau_1}$: 
        \item Only $\incomplete{\tau_2}$: 
        \item $\incomplete{\tau_1}$ and $\incomplete{\tau_2}$: 
    \end{itemize}

\end{proof}

\noindent\rule{\textwidth}{1pt}
% \begin{lemma}[name=Addition of New Arrows]
% If for some well formed $G$ where $\isNotIn{\TArrow{\tau_1}{\tau_2}}{G}$, $\ptypGraphAdd{G}{\TArrow{\tau_1}{\tau_2}}{G'}$, then it must be the case that $\isIn{\tau_1}{G'}$, $\isIn{\tau_2}{G'}$, and that $\isIn{\ptypGraphNode{\TArrow{\tau_1}{\tau_2}}{\ptsLead{\ptypSingle{\TArrow{\ptsRep{\tau_1}}{\ptsRep{\tau_2}}}}}}{G}$
% \end{lemma}

% \begin{proof}
%     Since $\isNotIn{\TArrow{\tau_1}{\tau_2}}{G}$, we cannot utilize Add-Present and must use Add-Arr-I-I.
% \end{proof}

% \noindent\rule{\textwidth}{1pt}

% \todo{merge this with the theorem below}
% \begin{lemma}[name=Unified Graph Comprehensiveness] 
% If $\unify{\cdot}{C}{G}$, then for all types $\tau$ in $C$ for which $\incomplete{\tau}$, there must exist some $\isIn{\ptypGraphNode{\tau}{S}}{G}$ where $\ptypSetSubset{G}{\ptypSingle{P(\tau)}}{s}$.
% \end{lemma}

% \noindent\rule{\textwidth}{1pt}

\begin{theorem}[name=Unified Graph Correctness]
For any constraint set $C$ and well formed graph $G$, must be the case that $\unify{G}{C}{G'}$ for some well formed $G'$ such that for all constraints $\constrain{\tau_1}{\tau_2}$ in $C$:
    \begin{enumerate}
        % \item If $\incomplete{\tau_1}$, then it must be that $\isIn{\ptypGraphNode{\tau_1}{s_1}}{G'}$ where if $\ptypSetSubset{G}{\ptypsUnion{\ptypSingle{P(\tau_1)}}{\ptypSingle{P(\tau_2)}}}{s_1}$. 
        \item If $\incomplete{\tau_1}$, then it must be that $\isIn{\ptypGraphNode{\tau_1}{s_1}}{G'}$ where if $\ptsSnapshot{G}{s_1}{j_1}$, then $\snapElement{H(\tau_1)}{j_1}$. 
        % \item If $\incomplete{\tau_2}$, then it must be that $\isIn{\ptypGraphNode{\tau_2}{s_2}}{G'}$ where $\ptypSetSubset{G}{\ptypsUnion{\ptypSingle{P(\tau_1)}}{\ptypSingle{P(\tau_2)}}}{s_2}$
        \item If $\incomplete{\tau_2}$, then it must be that $\isIn{\ptypGraphNode{\tau_2}{s_2}}{G'}$ where if $\ptsSnapshot{G}{s_2}{j_2}$, then $\snapElement{H(\tau_2)}{j_2}$
        \item If both $\incomplete{\tau_1}$ and $\incomplete{\tau_2}$, then in addition to the previous two conditions, it must be that $\leader{G'}{\tau_1}{\tau_l}{u}$ and $\leader{G'}{\tau_2}{\tau_l}{u}$.
    \end{enumerate}
where the outputted graph must always be well formed.
\end{theorem}

\begin{proof}
\todo{come back to this after anand finalizes proof of union correctness and adjust usage accordingly and then polish it more}
We proceed by induction. Let us define the predicate $I(C, G)$ where $I(C, G)$ is true if conditions 1-3 above hold for all constraints $C$ with graph $G$ and is false otherwise.\\
Base Case:
By U-Nil, it must be that $\unify{G}{\listNil}{G}$. Trivially, the premise holds for all constraints as there are none.\\
For all other cases, let $C = \constraintCons{\constrain{\tau_1}{\tau_2}}{C_{tl}}$.\\
Inductive Hypothesis: 
For $C_{tl}$ and graph $G$, it must be the case that $\unify{G_1}{C_{tl}}{G_2}$ where $I(C_{tl}, G_2)$ is true where $G_2$ is well formed\\
Inductive Step:
We proceed by cases\\
Before commencing unification, repeatedly decompose pairs of arrows constrained to each other into two new constraints linking the left and right hand sides (in accordance with the properties of the constraint relation). Consequently, it must be the case that for all constraints $\constrain{\tau_1}{\tau_2}$, at least one of $\tau_1$ and $\tau_2$ must not be an arrow type.
For all cases below, we begin by applying the inductive hypothesis to determine that there exists some $G_2$ such that $\unify{G_1}{C_{tl}}{G_2}$ and $I(C_{tl}, G_2)$ holds. 
\begin{itemize}
    \item Suppose it is the case that only $\incomplete{\tau_1}$.
    Proceed by cases:
    \begin{itemize}
        \item Suppose $\isIn{\tau_1}{G_2}$. By Add Correctness, it must be the case that $\ptypGraphAdd{G_2}{\tau_1}{G_2}$. Let $G_f \equiv (G, \ptypGraphNode{\tau_l}{\ptsLead{\ptypsUnion{u_l}{\ptypSingle{P(\tau_2)}}}})$. By Update Correctness, it must be the case that $\ptypGraphUpdate{G_2}{\tau_1}{\ptypSingle{P(\tau_2)}}{G_f}$. Therefore, by U-I-C, it must be that $\unify{G_1}{C}{G_f}$. By snapshot totality, there must exist some $j$ such that $\ptsSnapshot{G_f}{\ptsLead{\ptypsUnion{u_l}{\ptypSingle{P(\tau_2)}}}}{j}$. By Potential Types Union Correctness, it must be the case that condition 1 holds. Trivially, conditions 2-3 hold for the first constraint. Furthermore, since $G$ was retained in our final graph and no additions were made to elements of it other than $\tau_l$, for which conditions 1-3 are shown to hold, it must be the case that $I(C, G_f)$.

        \item Suppose $\isNotIn{\tau_1}{G_2}$. By Add Correctness, it must be the case that $\ptypGraphAdd{G_2}{\tau}{G_2 \cup G_{new}}$ where $G_{new}$ consists only of elements not already in $G_2$, including $\ptypGraphNode{\tau_1}{\ptsLead{\ptypSingle{P(\tau_1)}}}$.

        Let $G_f \equiv G_2 \cup G_{new}, \ptypGraphNode{\tau_1}{\ptsLead{\ptypsUnion{\ptypSingle{P(\tau_1)}}{\ptypSingle{P(\tau_2)}}}}$.
        By Update Correctness, it must be the case that $\ptypGraphUpdate{G_2 \cup G_{new}}{\tau_1}{\ptypSingle{P(\tau_2)}}{G_f}$. Therefore, by U-I-C, it must be that $\unify{G_1}{C}{G_f}$. By snapshot totality, there must exist some $j$ such that $\ptsSnapshot{G_f}{\ptsLead{\ptypsUnion{\ptypSingle{P(\tau_1)}}{\ptypSingle{P(\tau_2)}}}}{j}$. By Potential Types Union Correctness, it must be the case that condition 1 holds. Trivially, conditions 2-3 hold for the first constraint. Furthermore, since $G$ was retained in our final graph and no additions were made to elements of it other than $\tau_l$, for which conditions 1-3 are shown to hold, it must be the case that $I(C, G_f)$.

        % By Update Correctness, it must be the case that $\ptypGraphUpdate{G_2 \cup G_{new}}{\tau_1}{\ptypSingle{P(\tau_2)}}{(G_2, \ptypGraphNode{\tau_l}{(\ptypsUnion{u_l}{\ptypSingle{P(\tau_2)}})}, \ptypGraphNode{\tau_1}{\ptsRep{\tau_l}})}$.
    \end{itemize}
    \item Suppose it is the case that only $\incomplete{\tau_2}$. $I(C, G_f)$ can be determined using the same logic above, where all invocations of $\tau_1$ and $\tau_2$ are reversed and U-C-I is used to establish unification success instead of U-I-C.

    \item Suppose it is the case that both $\incomplete{\tau_1}$ and $\incomplete{\tau_2}$. By the inductive hypothesis, $\unify{G_1}{C_{tl}}{G_2}$.

    We begin by noting that for any arbitrary well formed $G_l$ where $G_l$ contains $\tau_1$ and $\tau_2$, we can establish the following derivation, which we will refer to below as Leader Success:

    By leader correctness, it must be that $\leader{G_l}{\tau_1}{\tau_{1l}}{u_{1l}}$ and $\leader{G_l}{\tau_2}{\tau_{2l}}{u_{2l}}$ for some $\tau_{1l}, \tau_{2l}, u_{1l},$ and $ u_{2l}$. Furthermore, by Potential Types Union Totality, there must be some $u_{12}$ such that $\ptypsUnion{u_{1l}}{u_{2l}} \equiv u_{12}$. Consequently, we may apply U-I-I in order to derive that $\unify{G_l}{\listCons{\constrain{\tau_1}{\tau_2}}{C_{tl}}}{G_l, \ptypGraphNode{\tau_1l}{\ptsLead{u_{1l}}}, \ptypGraphNode{\tau_2l}{\ptsRep{\tau_1l}}}$. By Potential Types Union Correctness, the the contents of both of the original potential type sets must be within the union. \todo{need to add part to theorem requiring this; transitivity proof assumes its there, it should be as simple as moving statements around here}

    Since $\tau_{1l}$ and $\tau_{2l}$ were the leaders of $\tau_1$ and $\tau_2$ respectively, and since $\tau_{2l}$ now is represented by $\tau_{1l}$, it must be the case that $\leader{G_l, \ptypGraphNode{\tau_1l}{\ptsLead{u_{1l}}}, \ptypGraphNode{\tau_2l}{\ptsRep{\tau_1l}}}{\tau_1}{\tau_{1l}}{u_{1l}}$ via L-Rep while $\leader{G_l, \ptypGraphNode{\tau_1l}{\ptsLead{u_{1l}}}, \ptypGraphNode{\tau_2l}{\ptsRep{\tau_1l}}}{\tau_2}{\tau_{1l}}{u_{1l}}$ due to no changes in its leader's potential type set in $G_l$.

    With this in mind, proceed by cases
    \begin{itemize}
        \item Suppose $\isIn{\tau_1}{G_2}$ and $\isIn{\tau_2}{G_3}$. By Add Correctness, it must be the case that $\ptypGraphAdd{G_2}{\tau_1}{G_2}$. By Add Correctness, it must be the case that $\ptypGraphAdd{G_2}{\tau_2}{G_2}$.

        Since both $\tau_1$ and $\tau_2$ were both in the graph, they must have been in some constraint within $C_{tl}$. Consequently, it must be that conditions 1 and 2 both hold for each. Furthermore, we may invoke Leader Success we may invoke Leader Success with $G_l = G_2$. Consequently, condition 3 also holds and $I(C, G_l, \ptypGraphNode{\tau_1l}{\ptsLead{u_{1l}}})$ must be true.

        \item Suppose $\isIn{\tau_1}{G_2}$ and $\isNotIn{\tau_2}{G_2}$. By Add Correctness, it must be the case that $\ptypGraphAdd{G_2}{\tau_1}{G_2}$. By Add Correctness, it must be the case that $\ptypGraphAdd{G_2}{\tau_2}{G_2 \cup G_{new}}$ where $G_{new}$ consists only of elements not already in $G_2$, including $\ptypGraphNode{\tau_2}{\ptsLead{\ptypSingle{P(\tau_2)}}}$.

        Since $\tau_1$ was already in the graph $G_2$, it must be that condition 1 holds. Since $\tau_2$ added, where by Addition correctness, the graph is only changed by the addition of new nodes, including $\ptypGraphNode{\tau_2}{\ptsLead{\ptypSingle{P(\tau_2)}}}$, It must be the case that condition 2 also holds.

        Since both $\tau_1$ and $\tau_2$ are both in the graph, we may invoke Leader Success we may invoke Leader Success with $G_l = G_2 \cup G_{new}$. Consequently, condition 3 also holds and $I(C, G_l, \ptypGraphNode{\tau_1l}{\ptsLead{u_{1l}}})$ must be true.

        \item Suppose $\isNotIn{\tau_1}{G_2}$. By Add Correctness, it must be the case that $\ptypGraphAdd{G_2}{\tau_1}{G_2 \cup G_{new}}$ where $G_{new}$ consists only of elements not already in $G_2$, including $\ptypGraphNode{\tau_1}{\ptsLead{\ptypSingle{P(\tau_1)}}}$. Suppose $\isIn{\tau_2}{G_2 \cup G_{new}}$. By Add Correctness, it must be the case that $\ptypGraphAdd{G_2 \cup G_{new}}{\tau_2}{G_2 \cup G_{new}}$.

        \todo{create subset lemma that lets you easily derive trivial subsets like tau is a subset of itself converted to a ptypset}
        Since $\tau_1$ was added, where by Addition correctness, the graph is only changed by the addition of new nodes, including $\ptypGraphNode{\tau_1}{\ptsLead{\ptypSingle{P(\tau_1)}}}$, It must be the case that condition 1 holds by Potential Types Subset Correctness. Since $\tau_2$ is already in $G_2 \cup G_{new}$, it must be the case that either $\tau_2$ was already added after unification on $C_{tl}$, or that it was added at the same time as $\tau_1$. If the former, by the inductive hypothesis, condition 1 must hold. If the latter, Addition correctness guarantees the newly added node must have the form $\ptypGraphNode{\tau_2}{\ptsLead{\ptypSingle{P(\tau_2)}}}$. In both cases, it must be the case that condition 1 holds by Potential Types subset correctness.

        Since both $\tau_1$ and $\tau_2$ are both in the graph, we may invoke Leader Success with $G_l = G_2 \cup G_{new}$. Consequently, condition 3 also holds and $I(C, G_l, \ptypGraphNode{\tau_1l}{\ptsLead{u_{1l}}})$ must be true.

        \item Suppose $\isNotIn{\tau_1}{G_2}$. By Add Correctness, it must be the case that $\ptypGraphAdd{G_2}{\tau_1}{G_2 \cup G_{new}}$ where $G_{new}$ consists only of elements not already in $G_2$, including $\ptypGraphNode{\tau_1}{\ptsLead{\ptypSingle{P(\tau_1)}}}$.

        Suppose $\isNotIn{\tau_2}{G_2 \cup G_{new}}$. By Add Correctness, it must be the case that $\ptypGraphAdd{G_2 \cup G_{new}}{\tau_2}{G_2 \cup G_{new} \cup G_{new'}}$ where $G_{new'}$ consists only of elements not already in $G_2 \cup G_{new}$, including $\ptypGraphNode{\tau_2}{\ptsLead{\ptypSingle{P(\tau_2)}}}$.

        Since $\tau_1$ was added, where by Addition correctness, the graph is only changed by the addition of new nodes, including $\ptypGraphNode{\tau_1}{\ptsLead{\ptypSingle{P(\tau_1)}}}$, It must be the case that condition 1 holds by Potential Types Subset Correctness. Since $\tau_2$ was added, where by Addition correctness, the graph is only changed by the addition of new nodes, including $\ptypGraphNode{\tau_2}{\ptsLead{\ptypSingle{P(\tau_2)}}}$, It must be the case that condition 2 holds by Potential Types Subset Correctness.

        Since both $\tau_1$ and $\tau_2$ are both in the graph, we may invoke Leader Success where $G_l = G_2 \cup G_{new} \cup G_{new'}$. Consequently, condition 3 also holds and $I(C, G_l, \ptypGraphNode{\tau_1l}{\ptsLead{u_{1l}}})$ must be true.
    \end{itemize}
\end{itemize}
\end{proof}

\noindent\rule{\textwidth}{1pt}

\begin{theorem}[name=Unified Graph Transitivity]
Assume $\unify{\cdot}{C}{G}$. Let $C' \equiv \{ \isIn{\constrain{\tau_1}{\tau_2}}{C} ~|~ \constrain{\tau_2}{\tau_1} \}$. Let $C_{trans}$ be an arbitrary set of constraints related via transitivity in $C \cup C'$.
There must exist some $\tau_l$ and $u_l$ such that for all $\isIn{\tau}{C_{trans}}$ where $\incomplete{\tau}$, it is the case that $\leader{G}{\tau}{\tau_l}{u_l}$
\end{theorem}

\begin{proof}
    Proceed by Induction.
    By Unified Graph Correctness, all incomplete types referenced in constraints must have potential type sets in $G$ that contain themselves and their constrained types.

    Base Cases: By Unified Graph Correctness, it must be the case that a constraint set of form $\listCons{\constrain{\tau_1}{\tau_2}}{\listNil}$ has this property. Trivially, it must be the case that a constraint set of form $\listNil$ must also have this property, as it must contain no incomplete types.

    $T = \listCons{\constrain{\tau_1}{\tau_2}}{T_{tl}}$

    Inductive Hypothesis: The theorem holds for all types referenced in $C_{tl}$

    Inductive Step: Since $T$ is a transitively constrained set, it must be the case that $T_{tl} = \listCons{\constrain{\tau_2}{\tau_3}}{T_{end}}$. By the inductive hypothesis, it must be the case that all the theorem holds for all such constraints, and they consequently all share a leader representing their solution. By Unified Graphg correctness, $\constrain{\tau_1}{\tau_2}$ must also share a leader as they are both incomplete. Since $\tau_2$ shares a potential type set leader with both $\tau_1$ and all members of $T_{tl}$, where $\tau_2$ only has one potential type set leader by as $G$ is well formed, it must be the case that the leader of $\tau_1$ is also that of all incomplete types in $T_{tl}$  
\end{proof}

\noindent\rule{\textwidth}{1pt}

\todo{This is wrong. Fix this and the constraint relation definition}
\begin{theorem}[name=Constraint Satisfaction and Synthesizability]
    % Constraints relating to a given type hole represent all required consistencies needed for synthesis to succeed. If all constraints for are satisfied, the program must synthesize a type.
    Assume $\synConstraint{\ctx}{\ECMV}{\TMV}{C}$.\\
    Let $C'$ be the set of all unique constraints that can generated from $C$ using the constraint relation inference rules.\\
    Let $V = \{ \tau \in C \}$, and let $G = (V, C')$.\\
    Consider the transitive closure of $G$, $G^* = (V, C'^*)$. Let $s$ be the neighborhood of $\tau$ in $G^*$ and let $\TMV'$ be any type consistent with all members of $s$. \\
    Suppose we were to replace all annotations in $\ECMV$ containing $\TMV$ with $\TMV'$, thus yielding $\ECMV_{repl}$. There must exist some $\ctx_{sugg}$, $\TMV_{sugg}$, and $C_{sugg}$ such that  $\synConstraint{\ctx_{sugg}}{\ECMV_{repl}}{\TMV_{sugg}}{C_{sugg}}$ (without further marking of $\ECMV_{repl}$).
\end{theorem}

\noindent\rule{\textwidth}{1pt}
\todo{If we have time, it'd be nice to also prove that the gamma', tau', and C' are all the same as the original values, only now the unknown type is replaced everywhere with the solved type}

\begin{theorem}[name=Solution Correctness] Suppose that $\synConstraint{\ctx}{\ECMV}{\TMV}{C}$ where the type annotation $\TUnknown^p$ is in $\ECMV$. Furthermore, suppose that $\unify{\cdot}{C}{G}$,~ $\isIn{\ptypGraphNode{\TUnknown^p}{s}}{G}$,~ $\ptsSnapshot{G}{s}{j}$, and ~$\solution{j}{\tau}$. \\
Let $\ECMV'$ be the result of replacing the annotation $\TUnknown^p$ in $\ECMV$ with $\tau$. \\
There must exist some $\ctx'$, $\TMV'$, and $C'$ that $\synConstraint{\ctx'}{\ECMV'}{\TMV'}{C'}$.
\end{theorem}

\begin{proof}
    By Constraint Satisfaction and Synthesizability, it must be the case that $C$ represents all consistencies necessary for synthesis of $\ECMV$ to succeed. By Unified Graph Correctness, it must be the case that there is some $\isIn{\ptypGraphNode{?^p}{s}}{G}$. By Incomplete Type Transitivity, it must be that this potential type set contains all types ever constrained to $?^p$. By snapshot totality, it must be the case that there exists some $j$ such that $\ptsSnapshot{G}{s}{j}$. By Solution Correctness, it must be the case that there exists some $\tau$ such that $\solution{j}{\tau}$ where $\tau$ is consistent with all types added in the snapshot representing all types constrained to $?^p$. Consequently, it must be that replacing $?^p$ in $\ECMV$ does not violate any constraints, as consistency succeeds for them all. Since the constraints represented all required checks for consistency, it must be that there exists some $\ctx'$, $\TMV'$, and $C'$ that $\synConstraint{\ctx'}{\ECMV'}{\TMV'}{C'}$.
\end{proof}

\noindent\rule{\textwidth}{1pt}

\textit{\textbf{Related Work}}

The contributions of this paper build directly on the Hazelnut type system \cite{HazelnutPOPL}, which is discussed extensively throughout. Non-empty holes in Hazelnut generalize to marks in this work. In brief, we contribute a total marking procedure and type hole inference procedure for a system based closely on Hazelnut, and use it to fix some expressiveness issues in Hazelnut's edit action calculus.

Our focus is on localization and recovery but not type error repair, as has been considered by other work \cite{lerner07}. We hope that our work will drive future work on rigorous repairs.

\citet{GradualInfer} introduces a variation of unification specifically for gradually typed languages based on Huet's unification algorithm \cite{Huet}. Other common approaches to unification are generally based on Hindley-Milner type inference \cite{MilnerInfer}. While \citet{GradualInfer} provides an algorithm which infers the types for a correct program, it does not provide any method for error localization and recovery if the constraint set cannot be solved.

\citet{garcia:2015} presents a static implicitly typed language, where users opt into dynamism by annotating an expression with the gradual type "?", and an associated type inference algorithm. By contrast, the Hazelnut type system assigns gradual types to programs that would ordinarily not type-check in a non-gradual system by wrapping them in expression holes. The type inference algorithm presented in \citet{garcia:2015} also does not specify what to do if the constraint set cannot be solved. If a single static type cannot be determined for an expression, its type is simply "undefined", whereas our Type Hole Inference algorithm provides a list of suggestions derived from any conflicting constraints if a single substitution cannot be determined.

% Complex techniques, e.g. based on machine learning or manual weighting, have been proposed to heuristically localize these errors to expressions.

The generation of correcting sets is a common approach to error localization. The correcting set is assigned blame for failure in type inference \cite{sherrloc, typeinferDif, Pavlinovic2015}. Other approaches to assigning blame include statistically derived approaches that leverage machine learning to predict the best candidates for blame \cite{SeidelBlame}. Our focus in this paper was to avoid heuristics and data-driven approaches and instead focus on systematic semantic approaches. Data-driven approaches could perhaps be layered to improve suggestions.

Explanations for type errors can be well illustrated by providing sample inputs (dynamic witnesses) that elicit runtime errors. With this approach, one can generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with around 70\% accuracy \cite{Seidel2016}.

### 2023-07-11 calculus.tex
Updates from Overleaf

@@
 The marked lambda calculus is based on the gradually typed lambda calculus \cite{Siek06a} extended
-with numbers, booleans, and empty expression holes \cite{HazelnutSNAPL}. The syntax is given in
-\cref{fig:calculus-syntax}. There are two expression languages:
+with numbers, booleans, pairs, and empty expression holes \cite{HazelnutSNAPL}. Given in
+\cref{fig:calculus-syntax}, the syntax consists of two expression languages:
 \begin{itemize}
-  \item The \textbf{unmarked language}, which is our original language. Expressions of this language
+  \item The \textbf{unmarked language}, which is the original language. Expressions of this language
     are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.
@@
 $\EFalse$ are the boolean values and
-$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ is the boolean conditional. Arrow types classify lambdas in the usual way.
+$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ is the boolean conditional. Arrow and product types classify lambda abstractions and pairs, respectively, in the usual way.
 
@@
                      \mid           \EPair{\EMV}{\EMV}
+                     \mid           \EProjL{\EMV} \mid \EProjR{\EMV}
                      \mid           \EEHole \\
@@
            &       & \mid         & \ECTrue \mid \ECFalse \mid \ECIf{\ECMV}{\ECMV}{\ECMV}
+                     \mid           \ECPair{\ECMV}{\ECMV} \mid \ECProjL{\ECMV} \mid \ECProjR{\ECMV}
                      \mid           \ECEHole \\
-             &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \mid \ECInconBr{\ECMV}{\ECMV}{\ECMV} \mid \ECInconAsc{\ECMV} \\
-             &       & \mid         & \ECSynNonMatchedArrow{\ECMV} \mid \ECAnaNonMatchedArrow{\ECMV}
+           &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \\
+           &       & \mid         & \ECLamInconAsc{x}{\TMV}{\ECMV} \mid \ECLamAnaNonMatchedArrow{x}{\TMV}{\ECMV} \mid \ECApSynNonMatchedArrow{\ECMV}{\ECMV} \\
+           &       & \mid         & \ECInconBr{\ECMV}{\ECMV}{\ECMV} \\
+           &       & \mid         & \ECPairAnaNonMatchedProd{\ECMV}{\ECMV} \mid \ECProjLSynNonMatchedProd{\ECMV} \mid \ECProjRSynNonMatchedProd{\ECMV}
   \end{array}\]
@@
 
-The key operation is marking, which transforms an unmarked expression into a marked expression,
+The key operation is \emph{marking}, which transforms an unmarked expression into a marked expression,
 inserting error marks where appropriate.
@@
 Intuitively, each of the mark forms corresponds to a different kind of error message that might be shown by an editor or emitted by a compiler.
-(We do not specify what the error message should say in this paper.)
-
+Note that we do not specify what the error message should say in this paper.
 Throughout the remainder of this section, as we formulate marking for the GTLC, we will motivate and
-give precise semantics for each error mark. When the core is extended, new marks may be needed, as we discuss below.
+give precise semantics for each error mark. Furthermore, when the core is extended, new marks may be needed, as we discuss below.
 
@@
 
-Finally, the key judgement, \emph{marking}, is also specified bidirectionally. The synthetic marking judgment
+Finally, the key judgement, \emph{marking}, is also of a bidirectional nature. The synthetic marking judgment
 $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
@@
 Furthermore, since error marks are effectively annotations on top of the program, marking should
-preserve syntactic structure modulo those marks. \cref{fig:calculus-mark-erasure} defines \emph{mark erasure}, which converts marked expressions back into unmarked expression by
+preserve syntactic structure modulo those marks.
+\cref{fig:calculus-mark-erasure} gives selected clauses of the definition of \emph{mark erasure}, which converts marked expressions back into unmarked ones by
 removing error marks. 
@@
 
-Then, \cref{thm:calculus-marking-well-formedness} provides a
+Then, \cref{thm:calculus-marking-well-formedness} provides the necessary
 \emph{well-formedness} criterion for marking.
@@
 
-These metatheorems imply that to go from a standard bidirectional system for the unmarked language to a marking system, we need to handle all possible failure modes with appropriate marks and marking logic (otherwise totality would be violated) and without otherwise changing the program (otherwise well-formedness would be violated).
-Let us consider each form in turn.
+Together, these metatheorems imply that to go from a standard bidirectional system for the unmarked language to a marking system, we need to handle all possible failure modes with appropriate marks and marking logic (otherwise totality would be violated) and without otherwise changing the program (otherwise well-formedness would be violated).
+Now, let us consider each form in turn.
 
@@
 
-To start, let us consider the simple case of numbers. Because of \emph{subsumption}, which is
+To start, consider the simple case of numbers. Because of \emph{subsumption}, which is
 discussed in more detail next, we need only define a synthesis rule for unmarked numbers, in which
@@
 
-How should numbers be marked? Quite straightforwardly, we simply give the same number as a marked
-expression, synthesizing again $\TNum$. No type errors may occur in a single number, so we only need
+How should numbers be marked? Straightforwardly, we simply give the same number as a marked
+expression, synthesizing again $\TNum$. No type errors may occur in a just single number, so we only need
 the following rules for marking numbers and typing the results: 
@@
 
-Only synthetic rules were necessary for typing numbers and addition because of \emph{subsumption},
+Only synthetic rules are necessary for typing numbers and addition because of \emph{subsumption},
 given below, which states that if an expression synthesizes a type, it may also be analyzed against
-that type or any \emph{consistent} type.
+that type or any consistent type.
 \[%
@@
 We also restrict the usage of subsumption to ``subsumable'' syntactic forms, written
-$\subsumable{\EMV}$. This judgment is defined on all syntactic forms except lambda abstractions and
-conditionals, the only ones with both synthesis and analysis rules (see below). In other words, we restrict subsumption to
+$\subsumable{\EMV}$. This judgment is defined for all syntactic forms except lambda abstractions,
+conditionals, and pairs, the only ones with both synthesis and analysis rules (see below). In other words, we restrict subsumption to
 be the rule of ``last resort'', which is necessary to establish that marking and typing are
-deterministic (discussed later in \cref{thm:calculus-marking-unicity}).
+deterministic (see \cref{thm:calculus-marking-unicity}).
 
@@
 \cref{fig:calculus-examples-inconsistent-types}, subsumption would be used to analyze $\ETrue$
-against $\TNum$ in $\EPlus{\ETrue}{1}$, which fails since $\inconsistent{\TNum}{\TBool}$, i.e.
-\textsc{UASubsume} does not apply, and traditional typing semantics would simply fail the type
-checking process at this point. \textsc{MKASubsume} would also not be applicable, leaving marking
+against $\TNum$ when checking $\EPlus{\ETrue}{1}$, which fails since $\inconsistent{\TNum}{\TBool}$, i.e.
+\textsc{UASubsume} does not apply. At this point, traditional typing semantics would simply fail the type
+checking process. \textsc{MKASubsume} would also not be applicable, leaving marking
 undefined in those cases.
@@
 Observe that the premises of \textsc{MKAInconsistentTypes} are identical to those of
-\textsc{MKASubsume}, except that $\inconsistent{\TMV}{\TMV'}$. By marking an error, type-checking
-may continue---instead of simply failing.
+\textsc{MKASubsume}, except that $\inconsistent{\TMV}{\TMV'}$. By marking an error, the type checking process may carry on and provide semantic feedback for the rest of the program.
 
@@
 However, consider the case that a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
-bound. Again, \textsc{USVar} would not apply. A total marking should, however, report the error and
+bound. Similar to above, \textsc{USVar} would not apply. A total marking procedure should, however, report the error and
 continue, motivating an \emph{unbound variable mark} $\ECUnbound{x}$ and the accompanying rules:
@@
 An unbound variable is marked as such, and since nothing may be said about their types, we
-synthesize the unknown type. Similarly to the inconsistent type mark, this allows type checking to
+synthesize the unknown type. As with the inconsistent type mark, this allows type checking to
 proceed, with usage of the unbound variable permitted in any expression.
@@
 
-The synthesis rule for marking intuitively follows \textsc{USLam} closely. We should recursively
-mark the body with an extended context binding $x$ to $\TMV$ and construct a new marked lambda
+The synthesis rule for marking intuitively follows \textsc{USLam} closely. We recursively
+mark the body with an extended context and construct a new marked lambda
 abstraction:
@@
 \end{mathpar}
-%
+
 However, recalling totality, we look again to the premises of \textsc{UALam} to see what type errors
@@
 
-In this example language, the distinction from the inconsistent type mark is of reduced
+Note that in this example language, the distinction from the inconsistent type mark is of reduced
 significance---all expressions synthesize a type. However, the addition of unannotated lambda
@@
 $\inconsistent{\TMV}{\TMV_1}$, i.e. the actual type annotation for $x$ is inconsistent with the
-expected input type. The \emph{inconsistent ascription mark} $\ECInconAsc{x}$ indicates exactly this
+expected input type. The \emph{inconsistent ascription mark} $\ECLamInconAsc{x}{\TMV}{\ECMV}$ indicates exactly this
 error, and we add a final pair of analytic rules:
@@
 \]%
-Following the same methodology, we have the following marking and typing
+Following the same methodology up to this point, we have the following marking and typing
 rules:
@@
 not quite right to mark $\EMV_1$ with the inconsistent type mark; rather than any single type, it is
-any type within the family of arrow types that is expected. For such a ``constrained'' synthetic
+any member of the family of arrow types that is expected. For such a ``constrained'' synthetic
 mode, we use a specialized mark $\ECSynNonMatchedArrow{\ECMV}$, which indicates that $\ECMV$ was
@@
 It is natural to extend this approach to other elimination forms that require the handling of
-unmatched types, such as products. Indeed, the same approach is taken for projections in
-\cref{sec:calculus-products}. 
-% Another approach might indicate the same kind of error but mark the
-% entire application.
+unmatched types, such as products. Indeed, the same approach is taken for projections below.
+Another similar approach might indicate the same kind of error but mark the entire application.
 
@@
 
-Conditionals, however, are an interesting case. In the unmarked language, we have both explicit
+Conditionals, however, present a more interesting case. In the unmarked language, we have both explicit
 synthetic and analytic rules:
@@
 
-The framework also allows for other localization approaches. For example, as discussed at the
-beginning of the section, we may regard the first branch as ``correct'' and localize any errors to
-the second. The following rule formalizes such a design:
+As previously mentioned, we do not prescribe any single localization design as ``correct'', and the framework freely allows for other approaches.
+For example, as discussed in the introduction, we may choose to regard the first branch as ``correct'' and localize any 
+errors to the second. The following rule formalizes such a design:
 \[%
@@
 % heuristics may be devised for different kinds of branching structures.
+
+\subsubsection{Pairs}
+At this point, the introduction of pairs expressions and the necessary projection operators poses no great challenge.
+Explicit synthesis and analysis rules govern the former.
+%
+\begin{mathpar}
+  \inferrule[USPair]{
+    \ctxSynTypeU{\ctx}{\EMV_1}{\TMV_1} \\
+    \ctxSynTypeU{\ctx}{\EMV_2}{\TMV_2}
+  }{
+    \ctxSynTypeU{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TProd{\TMV_1}{\TMV_2}}
+  }
+
+  \inferrule[UAPair]{
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
+    \ctxAnaTypeU{\ctx}{\EMV_1}{\TMV_1} \\
+    \ctxAnaTypeU{\ctx}{\EMV_2}{\TMV_2}
+  }{
+    \ctxAnaTypeU{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\TMV}
+  }
+\end{mathpar}
+%
+In similar fashion to above, we may derive marking rules from these in an intuitive manner:
+%
+\begin{mathpar}
+  \inferrule[MKSPair]{
+    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\
+    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
+  }{
+    \ctxSynFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TProd{\TMV_1}{\TMV_2}}
+  }
+
+  \inferrule[MSPair]{
+    \ctxSynTypeM{\ctx}{\ECMV_1}{\TMV_1} \\
+    \ctxSynTypeM{\ctx}{\ECMV_2}{\TMV_2}
+  }{
+    \ctxSynTypeM{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TProd{\TMV_1}{\TMV_2}}
+  } \\
+
+  \inferrule[MKAPair1]{
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\\\
+    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\
+    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
+  }{
+    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
+  }
+
+  \inferrule[MAPair1]{
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\\\
+    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TMV_1} \\
+    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TMV_2}
+  }{
+    \ctxAnaTypeM{\ctx}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
+  }
+\end{mathpar}
+
+However, as when analyzing a lambda abstraction against a non-matched arrow type, the type against which a pair is analyzed may not match any product type.
+We solve this by adding an error mark $\ECPairAnaNonMatchedProd{\ECMV_1}{\ECMV_2}$ and the corresponding rules:
+%
+\begin{mathpar}
+  \inferrule[MKAPair2]{
+    \notMatchedProd{\TMV} \\
+    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TUnknown} \\\\
+    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
+  }{
+    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPairAnaNonMatchedProd{\ECMV_1}{\ECMV_2}}{\TMV}
+  }
+
+  \inferrule[MAPair2]{
+    \notMatchedProd{\TMV} \\
+    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TUnknown} \\\\
+    \ctxAnaTypeM{\ctx}{\ECMV_2}{\TUnknown}
+  }{
+    \ctxAnaTypeM{\ctx}{\ECPairAnaNonMatchedProd{\ECMV_1}{\ECMV_2}}{\TMV}
+  }
+\end{mathpar}
+
+As the elimination form for products, projections are handled in a similar manner as applications.
+In the interest of space, we give only the rules for the left projection operator; right projections are governed analogously.
+%
+\begin{mathpar}
+  \inferrule[USProjL]{
+    \ctxSynTypeU{\ctx}{\EMV}{\TMV} \\\\
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynTypeU{\ctx}{\EProjL{\EMV}}{\TMV_1}
+  }
+
+  \inferrule[MKSProjL1]{
+    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV_1}
+  }
+
+  \inferrule[MSProjL1]{
+    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\\\
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynTypeM{\ctx}{\ECProjL{\ECMV}}{\TMV_1}
+  }
+\end{mathpar}
+
+In the case that the subject of the projection does not synthesize a matched product type, we mark with an error, written $\ECProjLSynNonMatchedProd{\ECMV}$:
+%
+\begin{mathpar}
+  \inferrule[MKSProjL2]{
+    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
+    \notMatchedProd{\TMV}
+  }{
+    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjLSynNonMatchedProd{\ECMV}}{\TMV_1}
+  }
+
+  \inferrule[MSProjL2]{
+    \ctxSynTypeM{\ctx}{\ECMV}{\TMV} \\
+    \notMatchedProd{\TMV}
+  }{
+    \ctxSynTypeM{\ctx}{\ECProjLSynNonMatchedProd{\ECMV}}{\TUnknown}
+  }
+\end{mathpar}
 
@@
 
-\subsection{Adding Binary Products}
-\label{sec:calculus-products}
-
-The marked lambda calculus is intended as a minimal core.
-The methodology described above can be used to extend the marked lambda calculus with new typing features.
-For example, let us add binary product types, written $\TProd{\TMV_1}{\TMV_2}$.
-In the interest of space, we focus on
-aspects related to marking and omit standard rules; a comprehensive set of rules is found in the supplemental material.
-
-% First, we extend the syntax of types and the accompanying notions of consistency and join, which
-% closely mirror the definitions for arrow types. Additionally, analogous to the notion of matched
-% arrow types, we need a matched product type relation $\matchedProd{\TMV}{\TMV_1}{\TMV_2}$.
-
-The unmarked language is extended with a pair constructor $\EPair{\EMV}{\EMV}$ and projection
-operators $\EProjL{\EMV}$ and $\EProjR{\EMV}$, governed by standard bidirectional rules. In addition
-to the synthesis rules, an explicit analysis rules is defined for pairs: if the expected type has a matched product type (cf.
-matched arrow types above), the responsibility for type errors is shifted to the components.
-
-We can devise the basic marking rules from the unmarked typing rules.
-%
-\begin{mathpar}
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
-    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
-  }{
-    \ctxSynFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TProd{\TMV_1}{\TMV_2}}
-  }{MKSPair}
-
-  \judgment{
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
-    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TMV_1} \\\\
-    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_2}
-  }{
-    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPair{\ECMV_1}{\ECMV_2}}{\TMV}
-  }{MKAPair1}
-
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
-  }{
-    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjL{\ECMV}}{\TMV_1}
-  }{MKSProjL1}
-
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\\\
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2}
-  }{
-    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjR{\ECMV}}{\TMV_2}
-  }{MKSProjR1}
-\end{mathpar}
-However, as previously seen with lambda abstractions and applications, two sets of rules are further
-necessary to satisfy totality. Firstly, the premise in \textsc{UAPair} that $\TMV$ is a matched
-binary product may fail, requiring the introduction of a new error mark,
-$\ECAnaNonMatchedProd{\ECMV}$, which is analogous to $\ECAnaNonMatchedArrow{\ECMV}$ for lambda
-abstractions. Another kind of error may arise when the subject of a projection does not synthesize a
-matched product type. As in our development of marking for applications, we utilize a new mark,
-written $\ECSynNonMatchedProd{\ECMV}$. These cases give the final marking rules for products:
-%
-\begin{mathpar}
-  \judgment{
-    \notMatchedProd{\TMV} \\
-    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TUnknown} \\
-    \ctxAnaFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TUnknown}
-  }{
-    \ctxAnaFixedInto{\ctx}{\EPair{\EMV_1}{\EMV_2}}{\ECPairAnaNonMatchedProd{\ECMV_1}{\ECMV_2}}{\TMV}
-  }{MKAPair2} \\
-
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
-    \notMatchedProd{\TMV}
-  }{
-    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjLSynNonMatchedProd{\ECMV}}{\TMV_1}
-  }{MKSProjL2}
-
-  \judgment{
-    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
-    \notMatchedProd{\TMV}
-  }{
-    \ctxSynFixedInto{\ctx}{\EProjR{\EMV}}{\ECProjRSynNonMatchedProd{\ECMV}}{\TMV_2}
-  }{MKSProjR2}
-\end{mathpar}
-
-\input{let}
-
 \subsection{Agda Mechanization}
@@
 
-The semantics and metatheory presented above, including the extension with binary products, have
+The semantics and metatheory presented above, have
 been fully mechanized in the Agda proof assistant.
@@
 The standard approach of modeling judgments as inductive datatypes and rules as constructors for
-those datatypes is taken. For marked expressions, we opt for implicitly typed terms,
-which gives us part of \cref{thm:calculus-marking-well-formedness} for free. For the convenience of
-readers interested in browsing the mechanization, all rule names match those presented here.
+those datatypes is taken. By representing marked expressions with implicitly typed terms,
+we get part of \cref{thm:calculus-marking-well-formedness} for free from the definition of marking. For the convenience of
+readers interested in browsing the mechanization, as possible, all rule names match those presented here.
+
+\input{let}
+\input{polymorphism}
 

### 2023-07-11 intro.tex
Updates from Overleaf

@@
 \begin{center}
-    \includegraphics[scale=0.5]{images/hazel-intro-screenshot.png}
+    \includegraphics[scale=0.5]{images/hazel-intro-screenshot-v3.png}
 \end{center}
@@
  Intuitively, however, it is clear that this knowledge is \emph{unnecessary} to make an error localization decision about \li{y}: 
-it is unbound, so a second error can be simultaneously localized despite the missing information about \li{f}. Elm's compiler stops here.
+it is unbound, so a second error can be simultaneously localized despite the missing information.
 
@@
 (because the type of \li{f(y)} is also not well-defined) 
-and continue into its branches, observing that they have inconsistent types, \li{Int} and \li{Bool}. 
+and continue into its branches, observing that they have inconsistent types, \li{Bool} and \li{Int}. 
 There are several ways to localize this error. 
@@
 Heuristics have been developed to help make this choice less arbitrary, e.g. 
-by observing recent editing history \cite{steady-typing}, 
+by observing recent editing history \cite{steady-typing}\todo{fix cite}, 
 or training a machine learning model \cite{SeidelBlame}. 
 The other approach, which Hazel takes above, is to localize the inconsistency to the conditional as a whole, reporting that the problem is that the branch types differ, as indicated above.
-When the cursor is on the conditional expression, the Hazel type inspector---a status bar that reports typing information, including type error messages, about the term at the cursor \cite{potter2020hazel}---shows:
+When the cursor is on the conditional expression, the Hazel type inspector---a status bar that reports typing information, including error messages, about the term at the cursor \cite{potter2020hazel}---shows:
 
@@
 This localization decision affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
-If localization had assumed that the \li{then} branch correct, as for example in Elm and \li{merlin}, then \li{x : Bool} and an error should be reported on the left operand (though this error might mislead the programmer if this earlier localization guess was incorrect).
-If the \li{else} branch was chosen, then \li{x : Int} and an error would not be reported. 
-If the inconsistency is localized to the conditional expression as a whole, as in our language server, then we again confront the problem of missing type information: 
+If localization had assumed that the \li{then} branch correct, as for example in \li{merlin}, then 
+\li{x : Bool} and an error should be reported on only the second use of \li{x}. 
+If the \li{else} branch was chosen, then \li{x : Int} and an error would be reported on only the first use of \li{x}. 
+ In either case, this error might mislead the programmer if the earlier localization guess was incorrect.
+If the inconsistency is localized to the conditional expression as a whole, as in Hazel, then we again confront the problem of missing type information: 
 \li{x} does not have a known type,
 though it \emph{is} known to be bound. 
-There is no definitive type or binding error here, 
-so we choose not to report an error in Hazel.
-No matter how localization handles the left operand, we would like to be able to recover and 
-localize the type inconsistency on the right operand (the \li{+} operator is integer addition in Hazel).
+There is no definitive type or binding error at either use of \li{x}, 
+so we choose not to report an error in Hazel (and in fact treat \li{x}'s type as the unknown a.k.a. dynamic type from gradual typing).
+In any of these cases, we would like to be able to recover and 
+localize the type inconsistency on the right addend because the \li{+} operator is integer addition in Hazel.
+For the same reason, there is no definitive inconsistency with the branch types in the second conditional, so no 
+error is marked there.
 
@@
 (2) when combined with error recovery, one localization decision can influence downstream localization decisions; and 
-(3) error recovery mechanisms must be able to reason without complete knowledge about types and binding. We argue that such semantic subtleties call for a rigorous theoretical treatment of the topic.
+(3) error recovery mechanisms must be able to reason without complete knowledge about types and binding.
+We argue that such semantic subtleties call for a rigorous theoretical treatment of these topics.
 

### 2023-07-11 let.tex
Changes from Overleaf
Updates from Overleaf

@@
 
-To ease the use of products, many languages feature destructuring bindings.
+To ease the use of products and other datatypes, many languages feature destructuring bindings.
 In a typed setting, we may want to granularly add type annotations in patterns as well.
@@
 The marking rule is directly analogous and utilizes the marking of patterns, which we omit---they
-may be derived in the same way as those for expressions. In the analytic cases, we introduce error
-marks that parallel to $\ECInconType{\ECMV}$ and $\ECAnaNonMatchedProd{\ECMV}$.
+may be derived in the same way as those for expressions and are governed by similar metatheorems. 
+In the analytic cases, we introduce error marks that parallel to $\ECInconType{\ECMV}$ and 
+$\ECAnaNonMatchedProd{\ECMV}$.
 \[%

### 2023-07-11 polymorphism.tex
Updates from Overleaf

\subsection{Parametric Polymorphism}
\label{sec:calculus-polymorphism}

To demonstrate that the judgmental structure of the marked lambda calculus may be applied to richer typing features, we now explore an extension of the core language developed above to System F-style parametric polymorphism.

\[\begin{array}{rrcl}
  \TMName  & \TMV  & \Coloneqq & \cdots \mid \TForall{\TVarMV}{\TMV} \mid \TVarMV \\
  \MTMName & \MTMV & \Coloneqq & \cdots
                     % \MTUnknown \mid \MTNum \mid \MTBool
                     % \mid \MTArrow{\MTMV}{\MTMV} \mid \MTProd{\MTMV}{\MTMV}
                     \mid \MTForall{\MTVarMV}{\MTMV} \mid \MTVarMV \mid \MTUnbound{\MTVarMV} \\
  \EMName  & \EMV  & \Coloneqq & \cdots \mid \ETypeLam{\TVarMV}{\EMV} \mid \ETypeAp{\EMV}{\TMV} \\
  \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECTypeLam{\MTVarMV}{\ECMV} \mid \ECTypeAp{\ECMV}{\MTMV} \\
           &       & \mid      & \ECSynNonMatchedForall{\ECMV} \mid \ECAnaNonMatchedForall{\ECMV}
\end{array}\]

### 2023-07-12 calculus.tex
Updates from Overleaf

@@
 environment. 
-%For example, a simple error that may occur is the use of an unbound variable (\cref{fig:calculus-examples-unbound}).
+%For example, a simple error that may occur is the use of an Free variable (\cref{fig:calculus-examples-Free}).
 % As in
-% \cref{fig:calculus-examples-unbound}, the editor indicates that the usage of an unbound $y$ is
+% \cref{fig:calculus-examples-Free}, the editor indicates that the usage of an Free $y$ is
 % invalid by highlighting it in red.
@@
                      \mid           \ECEHole \\
-           &       & \mid         & \ECUnbound{x} \mid \ECInconType{\ECMV} \\
+           &       & \mid         & \ECFree{x} \mid \ECInconType{\ECMV} \\
            &       & \mid         & \ECLamInconAsc{x}{\TMV}{\ECMV} \mid \ECLamAnaNonMatchedArrow{x}{\TMV}{\ECMV} \mid \ECApSynNonMatchedArrow{\ECMV}{\ECMV} \\
@@
 
-However, consider the case that a variable, such as $y$ in \cref{fig:calculus-examples-unbound}, is \emph{not}
+However, consider the case that a variable, such as $y$ in \cref{fig:calculus-examples-Free}, is \emph{not}
 bound. Similar to above, \textsc{USVar} would not apply. A total marking procedure should, however, report the error and
-continue, motivating an \emph{unbound variable mark} $\ECUnbound{x}$ and the accompanying rules:
+continue, motivating an \emph{Free variable mark} $\ECFree{x}$ and the accompanying rules:
 %
@@
   }{
-    \ctxSynFixedInto{\ctx}{x}{\ECUnbound{x}}{\TUnknown}
-  }{MKSUnbound}
+    \ctxSynFixedInto{\ctx}{x}{\ECFree{x}}{\TUnknown}
+  }{MKSFree}
 
@@
   }{
-    \ctxSynTypeM{\ctx}{\ECUnbound{x}}{\TUnknown}
-  }{MSUnbound}
-\end{mathpar}
-%
-An unbound variable is marked as such, and since nothing may be said about their types, we
+    \ctxSynTypeM{\ctx}{\ECFree{x}}{\TUnknown}
+  }{MSFree}
+\end{mathpar}
+%
+An Free variable is marked as such, and since nothing may be said about their types, we
 synthesize the unknown type. As with the inconsistent type mark, this allows type checking to
-proceed, with usage of the unbound variable permitted in any expression.
+proceed, with usage of the Free variable permitted in any expression.
 
@@
 For completeness, we finish the development of the marked lambda calculus on the extended GTLC with
-the marking of empty holes, which are never marked.
+the marking of empty holes, which are never marked (see Sec.~\ref{sec:thi} for marking unfillable holes using constraint solving).
 %
@@
 
-\subsection{Polymorphism and Richer Judgemental Structures}
-\label{sec:calculus-poly}
-
-The case studies above demonstrate that the judgemental structure of the marked lambda calculus is suitable to support a variety of extensions.
-We hope that this spurs language designers to designed mark variants of other calculi, following the recipe discussed above.
-
-In calculi with richer judgemental structure---for example, System F, which introduces type validity and type variables \cite{reynolds1988introduction}---the general approach is again to define marks for any new kind of error that might arise, e.g.
-the appearance of a type with free type variables, and marking judgements corresponding to each judgement in the original language, e.g.
-here defining marking for types.
-One small subtlety in doing so is that a marked free variable might end up on the right side of the typing arrows.
-However, it is straightforward to treat a marked free variable simply as another form of unknown type for the purposes of type consistency and other operations.
-Type functions and type application operate much like standard functions and function application after this additional judgemental structure is accounted for.
-We leave as future work the task of defining a marked version of even more elaborate bidirectional type systems, e.g.
-\citet{dunfield2013complete}'s rather substantial formulation of implicit type application in a bidirectional setting (which would have to handle, for example, the situation where no implicit argument can be resolved).
-
 \section{Integrating Marking with Editing}
@@
 a structure editor that is able to additionally offer partial syntactic error recovery.
-In particular, the \li{tylr} editor uses a form of structured syntax error recovery that automatically insert empty holes to maintain syntactic well-formedness as long as matching delimiters are placed by the user \cite{moon2022tylr}.
+In particular, the \li{tylr} editor uses a form of structured syntax error recovery that automatically insert empty holes to maintain syntactic well-formedness as long as matching delimiters are placed by the user \cite{moon2022tylr,DBLP:conf/vl/Moon023}.
 Efforts outside of the scope of this paper are underway to define syntactic error recovery mechanisms that can handle unmatched delimiters as well, which would achieve total syntax error recovery.
@@
 calculus, making use of the type and scoping information being propagated to re-mark only in scopes where it might be necessary. For example, we can wrap $\ECMV$ into
-$\ECLam{x}{\TUnknown}{\ECMV}$ by re-mark the body under the context with $x$ binding
+$\ECLam{x}{\TUnknown}{\ECMV}$ by re-marking the body under the context with $x$ binding
 the unknown type:
@@
   }{
-    \ASEConLam{\ctx}{\ECMV}{\TMV}{\ECMV'}{\TMV'}{x}
+    \ASCon{\ctx}{\ZCursor{\ECMV}}{\TMV}{\ZLamT{x}{\ZTCursor{\TUnknown}}{\ECMV'}}{\TArrow{\TUnknown}{\TMV'}}{\SLam{x}}
   }{ASEConLam}
@@
 use of type and scope information, and perhaps other analyses, to minimize re-marking. We leave a
-complete assessment of this and other re-marking optimization approaches to future work.
+complete assessment of this and other re-marking optimization approaches to future work, particularly
+because Hazel is now moving toward the decoupled approach for future development.

### 2023-07-12 intro.tex
Updates from Overleaf

@@
 (1)~\textbf{type error localization}: mechanisms for identifying the location(s) in a program that explain a type error, and 
-(2) \textbf{type error recovery}: mechanisms that allow the language server to optimistically recover from a localized type error 
+(2) \textbf{type error recovery}: mechanisms that allow the system to optimistically recover from a localized type error 
 and continue on to locate other errors and provide downstream semantic services, 
@@
 Essentially all widely-used programming systems have some support for type error localization, 
-reporting error locations in compiler error messages 
-or directly in the editor via markings decorating the localized errors. Developers are known to attend to these error locations when debugging type errors \cite{DBLP:journals/jfp/JoostenBH93}.
+e.g. in compiler error messages 
+or directly in the editor via markings decorating localized errors. Developers are known to attend to reported error locations when debugging type errors \cite{DBLP:journals/jfp/JoostenBH93}.
 Many systems also attempt recovery in certain situations, discussed below. 
@@
 
-Consider, the example ill-typed program below, which is shown as presented to the user in
+Consider, for example, the ill-typed program below, which is shown as presented to the user in
 this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{elm}. Hazel  
@@
  Intuitively, however, it is clear that this knowledge is \emph{unnecessary} to make an error localization decision about \li{y}: 
-it is unbound, so a second error can be simultaneously localized despite the missing information.
+it is unbound, so a second error can be simultaneously localized despite the missing information about \li{f}.
 
@@
 Heuristics have been developed to help make this choice less arbitrary, e.g. 
-by observing recent editing history \cite{steady-typing}\todo{fix cite}, 
+by observing recent editing history \cite{steady-typing}, 
 or training a machine learning model \cite{SeidelBlame}. 
-The other approach, which Hazel takes above, is to localize the inconsistency to the conditional as a whole, reporting that the problem is that the branch types differ, as indicated above.
-When the cursor is on the conditional expression, the Hazel type inspector---a status bar that reports typing information, including error messages, about the term at the cursor \cite{potter2020hazel}---shows:
+A less \emph{ad hoc} approach, which Hazel takes above, is to localize the inconsistency to the conditional as a whole, reporting that the problem is that the branch types differ.
+When the cursor is on this conditional expression, the Hazel type inspector---a status bar that reports typing information, including error messages, about the term at the cursor \cite{potter2020hazel}---displays:
 
 \begin{center}
-  \includegraphics[width=\textwidth]{images/haz3l-inconsistent-branches-cursor-inspector.png}
+  \includegraphics[width=0.8\textwidth]{images/haz3l-inconsistent-branches-cursor-inspector.png}
 \end{center}
 
-This localization decision affects how the system recovers as it proceeds into the let body, \li{x + 1}. 
-If localization had assumed that the \li{then} branch correct, as for example in \li{merlin}, then 
+This localization decision affects how the system recovers as it proceeds into the let body. 
+If localization had assumed that the \li{then} branch is correct, as for example in \li{merlin}, then 
 \li{x : Bool} and an error should be reported on only the second use of \li{x}. 
@@
 There is no definitive type or binding error at either use of \li{x}, 
-so we choose not to report an error in Hazel (and in fact treat \li{x}'s type as the unknown a.k.a. dynamic type from gradual typing).
+so we do not report an error in Hazel. More specifically, we treat \li{x}'s type as the unknown a.k.a. dynamic type from gradual typing.
 In any of these cases, we would like to be able to recover and 
-localize the type inconsistency on the right addend because the \li{+} operator is integer addition in Hazel.
-For the same reason, there is no definitive inconsistency with the branch types in the second conditional, so no 
-error is marked there.
+localize the type inconsistency on the string addend because the \li{+} operator is integer addition in Hazel.
+Recovering from this error, we can assume that both branch types in the second conditional will have \li{Int} type, so no 
+error is marked on the conditional as a whole.
 
-This informal exercise demonstrates that (1) localization choices can vary, particularly with regard to whether they make \emph{ad hoc} guesses about intent;  
+This informal exercise demonstrates that (1) localization choices can vary, particularly with regard to the extent to which they make \emph{ad hoc} guesses about intent;  
 (2) when combined with error recovery, one localization decision can influence downstream localization decisions; and 
-(3) error recovery mechanisms must be able to reason without complete knowledge about types and binding.
-We argue that such semantic subtleties call for a rigorous theoretical treatment of these topics.
+(3) error recovery necessitates reasoning without complete knowledge about types and binding. We argue that such semantic subtleties call for a rigorous theoretical treatment of these topics.
 
-This paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus} (\cref{sec:calculus}).
+Section~\ref{sec:calculus} of this paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus}.
 We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
 type annotations to expressions \cite{Localinf,BidirTyping}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{Siek06a,siek2015refined}, and 
-(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{HazelnutPOPL}, which the editor can display directly on-screen.
+(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{HazelnutPOPL}, operationalizing the ``red lines'' that an editor displays to the user.
+The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solar2013program}) can be marked, i.e.
+its errors can be simultaneously localized, such that the resulting program has a well-defined type and binding structure.
+We establish this and other properties as metatheorems that we mechanize in the Agda proof assistant. 
+As we define the calculus, we consider a number of situations where error localization decisions are subtle and conclude by defining some extensions of the intentionally minimal core calculus, e.g.
+with System-F-style polymorphism and pattern matching,
+that serve as case studies of the more general approach that we hope will be adopted by language designers defining practical type systems.
 
-The Hazelnut calculus \cite{HazelnutPOPL} previously brought these ideas together, but its account of type error localization and recovery was partial and tied integrally into its structure editor (see \cref{sec:calculus-hazel}).
-We resolve these problems to formally establish, with accompanying metatheory mechanized in Agda, that the marked lambda calculus features \emph{total error recovery}: {every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solar2013program}) can be marked, i.e.
-its errors can be localized, such that the resulting program has a well-defined type and binding structure.
-As we define the calculus, we consider a number of situations where error localization decisions are subtle, including conditionals and destructuring let with granular type annotations.
+\cref{sec:calculus-hazel} describes our own effort in this direction, which is to scale up the marked lambda calculus as the new basis for Hazel, a typed functional programming environment that fully solves the semantic gap problem: Hazel's semantic services are available as long as the program sketch is syntactically well-formed.
+Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as in Agda \cite{norell:thesis}, GHC Haskell \cite{haskell-holes}, Idris \cite{DBLP:journals/jfp/Brady13}, and others), syntax error recovery \cite{medeiros20,sorkin11}, and structure editing \cite{HazelnutPOPL,teitelbaum_cornell_1981}.
+Hazel has both a textual syntax and a structure editor \cite{HazelnutPOPL,DBLP:conf/vl/Moon023}. 
+%offering (1) a textual syntax, which allows the programmer to manually insert empty holes as necessary, (2) a text-like structure editor that offers partial syntax error recovery (inserting holes to maintain well-formedness as long as all matching delimiters have been placed) \cite{tylr}, and (3) a term-based structure editor with total syntax error recovery, i.e.
+a guarantee that every edit maintains syntactic well-formedness \cite{HazelnutPOPL}\todo{popl17}.
+As a secondary contribution, 
+\cref{sec:calculus-hazelnut} describes how (1)  total marking can resolve the problem of undefined behavior in the Hazelnut structure editor calculus, 
+and (2) integrating marking with typed structure editing allows us to incrementally re-mark only where necessary based on the edit location and the type and binding structure.
 
-This system, scaled up and implemented in Hazel (\cref{sec:calculus-hazel}), solves the semantic gap problem: Hazel's semantic services are available as long as the program sketch is syntactically well-formed.
 % which include type hints \cite{potter1}, semantic navigation, semantic code completion \cite{potter1,blinn}, contextualized documentation \cite{potter2}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{HazelnutLive}, even semantic services that require run-time information, like live testing,
@@
 
-Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as is featured in Agda \cite{norell:thesis}, GHC Haskell \cite{haskell-holes}, Idris \cite{DBLP:journals/jfp/Brady13}, and other languages), syntax error recovery \cite{medeiros20,sorkin11}, and structure editing \cite{HazelnutPOPL}.
-Hazel has both a textual syntax and a structure editor, which are briefly described in \cref{sec:calculus-hazel}.
-Marking is not integrated with structure editing, as it was in Hazelnut. 
-%offering (1) a textual syntax, which allows the programmer to manually insert empty holes as necessary, (2) a text-like structure editor that offers partial syntax error recovery (inserting holes to maintain well-formedness as long as all matching delimiters have been placed) \cite{tylr}, and (3) a term-based structure editor with total syntax error recovery, i.e.
-a guarantee that every edit maintains syntactic well-formedness \cite{HazelnutPOPL}\todo{popl17}.
-However, as a secondary contribution, 
-we briefly discuss in \cref{sec:calculus-hazelnut} how (1)  total marking can resolve the problem of undefined editing behavior in the Hazelnut structure editor calculus, 
-and (2) integrating marking with typed structure editing allows us to re-mark only where necessary based on the edit location and the type and binding structure.
+The developments in Section~\ref{sec:calculus} support the intuition (remarked upon by \citet{Localinf} and \citet{BidirTyping}) that local type inference pairs well with type error localization and recovery because information flows systematically and locally through the tree.
+However, many functional languages including Elm, OCaml and Haskell feature constraint-based type inference, with constraints gathered globally.
+For type systems where inference is decidable, this is powerful, but it is also notorious for complicating error localization, because inconsistencies can arise through a confluence of constraints originating from any number of locations.
 
-Local type inference pairs well with type error localization and recovery because information flows systematically through the tree, leading to predictable and systematic localization decisions (as informally observed also by \cite{Localinf}).
-Languages with constraint-based type inference gather constraints from all relevant locations in a program.
-This approach is powerful, but it is also notorious for complicating type error localization and recovery, because inconsistencies can arise through a confluence of constraints originating from any number of locations in the program.
-
-In \cref{sec:thi}, we describe how we blend local and constraint-based type inference in Hazel. In particular,  
-we generate constraints for each type hole that arises (either explicitly in the program or during type error localization and recovery as defined in \cref{sec:calculus}). 
-We develop a unification algorithm that can recover from inconsistent constraints by maintaining possible type hole solutions. For holes that can be uniquely solved, we suggest the solution. 
-If conflicting constraints arise, we localize the errors to the relevant type holes, rather 
-than to expressions.
-
-For example, in the following Hazel expression, bidirectional error localization 
-does not mark any expressions as erroneous because a type for \li{x} cannot locally be inferred (the bound expression is a hole). The type hole at the cursor therefore behaves gradually. 
-However, by gathering constraints from the three uses of \li{x} and attempting to unify, we see that this type hole is unfillable -- it is being used at three different types.
-This error is localized, with a \li{!}, to the type hole itself.
+For example, in the following Hazel expression (ignoring the error on the type hole for the moment), bidirectional error localization 
+does not mark any uses of \li{x} as erroneous because a type for \li{x} cannot locally be inferred (the let-bound expression is an empty hole, which has unknown type).
+However, by gathering constraints from the three uses of \li{x} and attempting to unify, we see that the type hole is unfillable.
+Rather than arbitrarily privileging one of these uses, as is the case in languages like OCaml (and which has led to the development of complex heuristics, e.g.
+using machine learning \cite{SeidelBlame}, to avoid misleading programmers), this error is localized, with a \li{!}, to the type hole itself, with the error message providing partially consistent solutions as shown.
 \begin{center}
 \vspace{-3px}
-    \includegraphics[scale=0.55]{images/intro-thi.png}
-    \includegraphics[scale=0.4]{images/example_CI.png}
+    % \includegraphics[scale=0.55]{images/intro-thi.png}
+    % \includegraphics[scale=0.4]{images/example_CI.png}
+    \includegraphics[scale=0.4]{images/figSugg.png}
 \end{center}
-The user can investigate by choosing a partial solution, which (temporarily) fills the type hole (not shown) and thus allows the local inference system to systematically localize the error.
-Here, each choice causes different set of two uses of \li{x} to be marked, e.g.
-the first and third use for \li{Int -> Int} (not shown).
-Until the choice is finalized, the system continues to recover gradually.
+
+The user can interactively explore possible error localizations by hovering over a partially consistent suggestion, which (temporarily) fills the type hole and thus returns the localization decision to the local inference system.
+Here, each choice causes a different set of two uses of \li{x} to be marked, e.g.
+the first and third use for \li{Int -> Int} (shown above).
+Until the choice is finalized, the type hole continues to operate gradually.
+\cref{sec:thi} describes this distinctively neutral approach to blending local and constraint-based type inference in Hazel, where local inference is used to systematically mark erroneous expressions in the program and constraint-based inference (using entirely standard algorithms, which we do not repeat in this paper) is used exclusively to locate unfillable holes.

### 2023-07-12 let.tex
Updates from Overleaf

@@
 In a typed setting, we may want to granularly add type annotations in patterns as well.
-As it turns out this combination of features is surprisingly tricky to get right!
-Let us add let
-expressions and simple patterns, writing $\PWild$ for the wildcard pattern, $\PVar{x}$ for a
-variable pattern, and $\PPair{\PMV_1}{\PMV_2}$ for a pair. 
+As it turns out, this combination of features is surprisingly tricky to get right!
+Let us add let expressions and simple patterns, writing $\PWild$ for the wildcard pattern, $\PVar{x}$ for a variable pattern, and $\PPair{\PMV_1}{\PMV_2}$ for a pair:
 %
@@
 If the pattern $\PVar{a}$ suggests that the conditional is in synthetic position, it will be marked
-with an inconsistent branches error following our development of marking above. If instead it is in
+with an inconsistent branches error mark following our development above. If instead it is in
 analytic position against the unknown type, no mark will be produced.
@@
   \judgment{
-    \ctxSynTypeU{\ctx}{\EMV}{\ctx'}
+    \ctxSynTypeU{\ctx}{\EMV}{\TMV}
   }{
@@
 The marking rule is directly analogous and utilizes the marking of patterns, which we omit---they
-may be derived in the same way as those for expressions and are governed by similar metatheorems. 
+may be derived in the same way as those for expressions and are governed by similar metatheorems (again, totality guides the derivation).
 In the analytic cases, we introduce error marks that parallel to $\ECInconType{\ECMV}$ and 
@@
 \]%
-
-% TODO conclude briefly?

### 2023-07-12 polymorphism.tex
Updates from Overleaf

@@
 
-To demonstrate that the judgmental structure of the marked lambda calculus may be applied to richer typing features, we now explore an extension of the core language developed above to System F-style parametric polymorphism.
+% TODO less explicit development?
 
+To further demonstrate that the judgmental structure of the marked lambda calculus may be applied to richer typing features, we now explore an extension of the core language developed above to System F-style parametric polymorphism.
+Toward this end, we supplement the language in \cref{fig:calculus-polymorphism-syntax} with type abstractions $\ETypeLam{\TVarMV}{\EMV}$ and applications $\ETypeAp{\EMV}{\TMV}$.
+These operate on forall types $\TForall{\TVarMV}{\TMV}$ and type variables $\TVarMV$, which are governed by a well-formedness judgment, written $\tvarCtxWFU{\tvarCtx}{\TMV}$, in the standard way.
+%
+\begin{figure}[htbp]
     \[\begin{array}{rrcl}
@@
                          % \mid \MTArrow{\MTMV}{\MTMV} \mid \MTProd{\MTMV}{\MTMV}
-                     \mid \MTForall{\MTVarMV}{\MTMV} \mid \MTVarMV \mid \MTUnbound{\MTVarMV} \\
+                         \mid \MTForall{\MTVarMV}{\MTMV} \mid \MTVarMV \mid \MTFree{\MTVarMV} \\
         \EMName  & \EMV  & \Coloneqq & \cdots \mid \ETypeLam{\TVarMV}{\EMV} \mid \ETypeAp{\EMV}{\TMV} \\
         \ECMName & \ECMV & \Coloneqq & \cdots \mid \ECTypeLam{\MTVarMV}{\ECMV} \mid \ECTypeAp{\ECMV}{\MTMV} \\
-           &       & \mid      & \ECSynNonMatchedForall{\ECMV} \mid \ECAnaNonMatchedForall{\ECMV}
+               &       & \mid         & \ECTypeLamAnaNonMatchedForall{\MTVarMV}{\ECMV} \mid \ECTypeApSynNonMatchedForall{\ECMV}{\MTMV}
     \end{array}\]
+    \caption{Extension of the marked lambda calculus for parametric polymorphism}
+    \label{fig:calculus-polymorphism-syntax}
+\end{figure}
+
+Crucially, just as expression variables may be free, arbitrary programs may contain free type variables; similar to the introduction of both unmarked and marked patterns in the previous section, we now need separate notions of unmarked and marked types.
+The latter we denote by the metavariable $\MTMV$.
+
+A marking judgment, written $\tvarCtxTypeMarkedInto{\tvarCtx}{\TMV}{\MTMV}$, then, relates the two sorts.
+It is parameterized over a context of type variables $\tvarCtx$, and most of the rules recurse straightforwardly.
+When a free type variable $\MTVarMV$ is encountered, however, we mark it as such, writing $\MTFree{\MTVarMV}$.
+This gives the following marking and well-formedness rules:
+%
+\begin{mathpar}
+  \inferrule[MKTFree]{
+    \notInTvarCtx{\tvarCtx}{\TVarMV}
+  }{
+    \tvarCtxTypeMarkedInto{\tvarCtx}{\TVarMV}{\MTFree{\MTVarMV}}
+  }
+
+  \inferrule[MTWFFree]{
+    \notInTvarCtx{\tvarCtx}{\MTVarMV}
+  }{
+    \tvarCtxWFM{\tvarCtx}{\MTFree{\MTVarMV}}
+  }
+\end{mathpar}
+%
+Just as expression error marks synthesize the unknown type to allow type-checking to continue, these marked types operate identically to the unknown type with respect to consistency and other auxiliary notions.
+Furthermore, we define analogous notions of mark erasure on types, and, as with pattern marking above, metatheoretic statements ensure their correctness.
+
+% TODO make reference to other polymorphic GTLC systems in relation to how auxiliary judgment rules are formulated
+
+The development of marking for type abstractions is similar to that for ordinary lambda abstractions.
+The synthetic case is straightforward; in analytic position, if the type being analyzed against is not a matched forall type, an error is marked:
+%
+\begin{mathpar}
+  \inferrule[MKATypeLam2]{
+    \notMatchedForall{\MTMV} \\
+    \bothCtxAnaFixedInto{\extendTvarCtx{\tvarCtx}{\MTVarMV}}{\ctx}{\EMV}{\ECMV}{\MTUnknown}
+  }{
+    \bothCtxAnaFixedInto{\tvarCtx}{\ctx}{\ETypeLam{\TVarMV}{\EMV}}{\ECTypeLamAnaNonMatchedForall{\MTVarMV}{\ECMV}}{\MTMV}
+  }
+
+  \inferrule[MATypeLam2]{
+    \notMatchedForall{\MTMV} \\
+    \bothCtxAnaTypeM{\extendTvarCtx{\tvarCtx}{\MTVarMV}}{\ctx}{\ECMV}{\MTUnknown}
+  }{
+    \bothCtxAnaTypeM{\tvarCtx}{\ctx}{\ECTypeLamAnaNonMatchedForall{\MTVarMV}{\ECMV}}{\MTMV}
+  }
+\end{mathpar}
+
+Likewise, type application mirrors ordinary application: in the case that the expression being applied does not synthesize a matched forall type, an error is marked.
+With this, totality of marking is satisfied.
+%
+\begin{mathpar}
+  \inferrule[MKSTypeAp2]{
+    \bothCtxSynFixedInto{\tvarCtx}{\ctx}{\EMV}{\ECMV}{\MTMV} \\
+    \tvarCtxTypeMarkedInto{\tvarCtx}{\TMV_2}{\MTMV_2} \\
+    \notMatchedForall{\MTMV}
+  }{
+    \bothCtxSynFixedInto{\tvarCtx}{\ctx}{\ETypeAp{\EMV}{\TMV_2}}{\ECTypeAp{\ECMV}{\MTMV_2}}{\MTUnknown}
+  }
+
+  \inferrule[MSTypeAp2]{
+    \bothCtxSynTypeM{\tvarCtx}{\ctx}{\ECMV}{\MTMV} \\
+    \tvarCtxWFM{\tvarCtx}{\MTMV_2} \\
+    \notMatchedForall{\MTMV}
+  }{
+    \bothCtxSynTypeM{\tvarCtx}{\ctx}{\ECTypeApSynNonMatchedForall{\ECMV}{\MTMV_2}}{\MTUnknown}
+  }
+\end{mathpar}
+
+The case studies above demonstrate that the framework of the marked lambda calculus is suitable to support a variety of extensions to richer judgmental structures.
+We hope that this spurs language designers to design marked variants of other calculi, following the recipe we have demonstrated.
+% TODO explicitly give recipe.
+
+We leave as future work the task of defining a marked version of even more elaborate bidirectional type systems, e.g.
+\citet{dunfield2013complete}'s rather substantial formulation of implicit type application in a bidirectional setting (which would have to handle, for example, the situation where no implicit argument can be resolved).
+
+% TODO cite gradual dependent types papers, etc.

### 2023-07-12 related.tex
Updates from Overleaf

@@
 Our focus is on localization and recovery but not type error repair, as has been considered by other work \cite{lerner07}. We hope that our work will drive future work on rigorous repairs.
+
+\citet{garcia:2015} presents a static implicitly typed language, where users opt into dynamism by annotating an expression with the gradual type "?", and an associated type inference algorithm.
+By contrast, the Hazelnut type system assigns gradual types to programs that would ordinarily not type-check in a non-gradual system by wrapping them in expression holes.
+The type inference algorithm presented in \citet{garcia:2015} also does not specify what to do if the constraint set cannot be solved.
+If a single static type cannot be determined for an expression, its type is simply "undefined", whereas our Type Hole Inference algorithm provides a list of suggestions derived from any conflicting constraints if a single substitution cannot be determined.
 

### 2023-07-12 calculus.tex
Copy over updated symbols

@@
   }{
-    \ctxSynTypeU{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
+    \ctxSynTypeU{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TMeet{\TMV_1}{\TMV_2}}
   }{USIf}
@@
 %
-In synthetic position, conditionals synthesize the \emph{join} of the branch types $\TMV_1$ and
-$\TMV_2$, which we define inductively in \cref{fig:calculus-type-join}. We choose the ``more
+In synthetic position, conditionals synthesize the \emph{meet} of the branch types $\TMV_1$ and
+$\TMV_2$, which we define inductively in \cref{fig:calculus-type-meet}. We choose the ``more
 specific'' type of the two. In analytic position, since
@@
 
-\newcommand{\joinsTo}[3]{\ensuremath{\TJoin{#1}{#2} & = & #3}}
+\newcommand{\meetsTo}[3]{\ensuremath{\TMeet{#1}{#2} & = & #3}}
 \begin{figure}[htbp]
   \raggedright
-  \judgbox{\ensuremath{\TJoin{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
+  \judgbox{\ensuremath{\TMeet{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
   %
   \[\begin{array}{rcl}
-    \joinsTo{\TUnknown}{\TMV}{\TMV} \\
-    \joinsTo{\TMV}{\TUnknown}{\TMV} \\
-    \joinsTo{\TNum}{\TNum}{\TNum} \\
-    \joinsTo{\TBool}{\TBool}{\TBool} \\
-    \joinsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TJoin{\TMV_1}{\TMV_1'})}{(\TJoin{\TMV_2}{\TMV_2'})}}
+    \meetsTo{\TUnknown}{\TMV}{\TMV} \\
+    \meetsTo{\TMV}{\TUnknown}{\TMV} \\
+    \meetsTo{\TNum}{\TNum}{\TNum} \\
+    \meetsTo{\TBool}{\TBool}{\TBool} \\
+    \meetsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TMeet{\TMV_1}{\TMV_1'})}{(\TMeet{\TMV_2}{\TMV_2'})}}
   \end{array}\]
   \vspace{-10px}
-  \caption{Type join.}
-  \label{fig:calculus-type-join}
+  \caption{Type meet.}
+  \label{fig:calculus-type-meet}
 \end{figure}
@@
   }{
-    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
+    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMeet{\TMV_1}{\TMV_2}}
   }{MKSIf}
@@
   }{
-    \ctxSynTypeM{\ctx}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TJoin{\TMV_1}{\TMV_2}}
+    \ctxSynTypeM{\ctx}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMeet{\TMV_1}{\TMV_2}}
   }{MSIf} \\
@@
 
-However, in synthetic position, it may be the case that the two branch types have no join. This
+However, in synthetic position, it may be the case that the two branch types have no meet. This
 occurs, in fact, when they are inconsistent, motivating the \emph{the inconsistent branches mark}

### 2023-07-12 intro.tex
Rename unbound stuff to free

@@
 is to localize the first error that causes local type inference to fail and emit an explanatory error message before terminating.
-In this example, the system might report that the variable \li{f} located on Line 2 is unbound, then terminate.
+In this example, the system might report that the variable \li{f} located on Line 2 is free, then terminate.
 
@@
  Intuitively, however, it is clear that this knowledge is \emph{unnecessary} to make an error localization decision about \li{y}: 
-it is unbound, so a second error can be simultaneously localized despite the missing information about \li{f}.
+it is free, so a second error can be simultaneously localized despite the missing information about \li{f}.
 

### 2023-07-12 supplement/preface.tex
Compile updated supplement

\documentclass[supplement.tex]{subfiles}

\begin{document}
\section{Preface}
\label{sec:preface}
This is a formalism demonstrating the \emph{marked lambda calculus}, a judgmental framework for
precise bidirectional error localization and recovery that employs gradual typing.

\subsection{Organization}
Though more is said in each individual section, the overall structure of the document is as follows:
%
\begin{itemize}
  \item \cref{sec:marked} employs the framework on a gradually typed lambda calculus.

  \item \cref{sec:patterned} extends the demonstration with patterned let expressions.

  \item \cref{sec:polymorphism} extends the demonstration with System F-style parametric
    polymorphism.

  \item \cref{sec:untyped} gives a version of the Hazelnut structure editor calculus that uses the
    marked lambda calculus to solve Hazelnut's deficiency with regards to non-local hole fixing.

  \item \cref{sec:typed} is similar, except that it employs the marking procedure in a roughly
    incremental fashion.
\end{itemize}
%
Note that each of the following sections after \cref{sec:marked} build upon the same core language.

\subsection{Mechanization}
Not all parts of the formalism are mechanized in Agda. In particular, the core of the marked lambda
calculus (\cref{sec:marked}) and the untyped Hazelnut calculus that employs marking
(\cref{sec:untyped}) \emph{are} mechanized, while all others \emph{are not}. Crucially, the
metatheorems of the aforementioned sections are verified, whereas those of all other sections are
not.

As possible, the names of judgments and rules that appear in the mechanization have been made to
follow those in this formalism. Refer also to the mechanization's README for more details.

\end{document}

### 2023-07-12 calculus.tex
Updates from Overleaf

@@
 
-We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of the Hazel language server.
+We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of the Hazel.
+% TODO citation and explain what in the world Hazel is (reviewer A stated that the lack of this made this section pretty much impossible to understand)
 In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching, strings, and explicit polymorphism (all following the same recipe we developed above).

### 2023-07-12 intro.tex
Updates from Overleaf

@@
 \begin{center}
-  \includegraphics[width=0.8\textwidth]{images/haz3l-inconsistent-branches-cursor-inspector.png}
+  \includegraphics[width=0.9\textwidth]{images/haz3l-inconsistent-branches-cursor-inspector.png}
 \end{center}
@@
 This localization decision affects how the system recovers as it proceeds into the let body. 
-If localization had assumed that the \li{then} branch is correct, as for example in \li{merlin}, then 
+If localization had assumed that the \li{then} branch were correct, as for example in \li{merlin}, then 
 \li{x : Bool} and an error should be reported on only the second use of \li{x}. 
-If the \li{else} branch was chosen, then \li{x : Int} and an error would be reported on only the first use of \li{x}. 
+If the \li{else} branch were chosen, then \li{x : Int} and an error would be reported on only the first use of \li{x}. 
  In either case, this error might mislead the programmer if the earlier localization guess was incorrect.
-If the inconsistency is localized to the conditional expression as a whole, as in Hazel, then we again confront the problem of missing type information: 
+If the inconsistency were localized to the conditional expression as a whole, as in Hazel, then we again confront the problem of missing type information: 
 \li{x} does not have a known type,
@@
 As we define the calculus, we consider a number of situations where error localization decisions are subtle and conclude by defining some extensions of the intentionally minimal core calculus, e.g.
-with System-F-style polymorphism and pattern matching,
+with destructuring patterns and System F-style polymorphism,
 that serve as case studies of the more general approach that we hope will be adopted by language designers defining practical type systems.

### 2023-07-12 calculus.tex
Fix some typos

@@
       $\erasesTo{\ECMV}{\EMV}$.
-    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then  and
+    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then
       $\erasesTo{\ECMV}{\EMV}$.
@@
 Now, to define analytic marking on forms without explicit analytic typing rules, we also need
-subsumption rules for marking and marked expression typing. Note that we define a analogous notion
+subsumption rules for marking and marked expression typing. Note that we define an analogous notion
 of ``subsumability'' for marked expressions, written $\subsumable{\ECMV}$.
@@
 significance---all expressions synthesize a type. However, the addition of unannotated lambda
-abstractions, for example, would necessitate such a disctinction.
+abstractions, for example, would necessitate such a distinction.
 

### 2023-07-12 calculus.tex
Clean up some writing in section 2

@@
 environment. 
-%For example, a simple error that may occur is the use of an Free variable (\cref{fig:calculus-examples-Free}).
-% As in
-% \cref{fig:calculus-examples-Free}, the editor indicates that the usage of an Free $y$ is
+% For example, a simple error that may occur is the use of an free variable (\cref{fig:calculus-examples-free}).
+% As in \cref{fig:calculus-examples-unbound}, the editor indicates that the usage of an free $y$ is
 % invalid by highlighting it in red.
@@
 
-The key contribution of this section is the \emph{marked lambda calculus}, a calculus based on the gradually typed lambda calculus (GTLC) that formalizes the mechanism by which errors like these can be localized, and how to  recover, in all cases, from such errors.
-% In brief, the marked lambda calculus formalizes 
-% bidirectional type error localization with gradual recovery. 
-\Cref{sec:calculus-calculus} introduces the syntax, judgemental structure, guiding metatheory, and then goes through the rules, organized by syntactic form rather than judgement, to build intuition about how the various judgements relate to one another \cite{Siek06a}.
-All rules and theorems may be found organized by judgement form in the supplementary appendix for reference,
-alongside a complete mechanization in the Agda proof assistant \cite{norell:thesis} (discussed in
-\cref{sec:calculus-agda}).
-We intentionally keep the marked lambda calculus minimal, because it is intended to capture the essential idea and introduce a general pattern that language designers can employ to created mark variants of their own type systems.
-As an initial example, we consider product types (in \cref{sec:calculus-products}).
-We then consider a surprisingly subtle combination of features as a more substantial case study: the combination of
-destructing let expressions with granular type annotations, in \cref{sec:calculus-let}. Finally, \cref{sec:calculus-poly} briefly explores how
-the system might be extended to more complex judgemental structure, e.g. polymorphism.
+The key contribution of this section is the \emph{marked lambda calculus}, a calculus based on the
+gradually typed lambda calculus (GTLC) that formalizes the mechanism by which errors like these can
+be localized, and how to  recover, in all cases, from such errors. 
+% In brief, the marked lambda calculus formalizes bidirectional type error localization with gradual
+% recovery. 
+\Cref{sec:calculus-calculus} introduces the syntax, judgemental structure, guiding metatheory, and
+then goes through the rules, organized by syntactic form rather than judgement, to build intuition
+about how the various judgements relate to one another \cite{Siek06a}.
+All rules and theorems may be found organized by judgement form in the supplementary appendix for
+reference, alongside a complete mechanization in the Agda proof assistant \cite{norell:thesis}
+(discussed in \cref{sec:calculus-agda}).
+We intentionally keep the marked lambda calculus minimal, because it is intended to capture the
+essential idea and introduce a general pattern that language designers can employ to created mark
+variants of their own type systems.
+As an initial example, we consider a surprisingly subtle combination of features as a more
+substantial case study: the combination of destructing let expressions with granular type
+annotations, in \cref{sec:calculus-let}.
+Finally, \cref{sec:calculus-polymorphism} briefly explores how the system might be extended to more
+complex judgemental structures, such as parametric polymorphism.
 
@@
 The marked lambda calculus is based on the gradually typed lambda calculus \cite{Siek06a} extended
-with numbers, booleans, pairs, and empty expression holes \cite{HazelnutSNAPL}. Given in
-\cref{fig:calculus-syntax}, the syntax consists of two expression languages:
+with numbers, booleans, pairs, and empty expression holes \cite{HazelnutSNAPL}.
+Given in \cref{fig:calculus-syntax}, the syntax consists of two expression languages:
+%
 \begin{itemize}
@@
 
-However, consider the case that a variable, such as $y$ in \cref{fig:calculus-examples-Free}, is \emph{not}
+However, consider the case that a variable, such as $y$ in \cref{fig:calculus-examples-free}, is \emph{not}
 bound. Similar to above, \textsc{USVar} would not apply. A total marking procedure should, however, report the error and
-continue, motivating an \emph{Free variable mark} $\ECFree{x}$ and the accompanying rules:
+continue, motivating an \emph{free variable mark} $\ECFree{x}$ and the accompanying rules:
 %
@@
 %
-An Free variable is marked as such, and since nothing may be said about their types, we
-synthesize the unknown type. As with the inconsistent type mark, this allows type checking to
-proceed, with usage of the Free variable permitted in any expression.
+A free variable is marked as such, and since nothing may be said about their types, we synthesize
+the unknown type. As with the inconsistent type mark, this allows type checking to proceed, with
+usage of the free variable permitted in any expression.
 
@@
 %
-Together, totality and unicity give that marking may be implemented as a total
-function. Indeed, given the algorithmic nature of bidirectional typing, it is fairly direct to implement these rules.
+Together, totality and unicity give that marking may be implemented as a total function.
+Indeed, given the algorithmic nature of bidirectional typing, it is fairly direct to implement these
+rules.
 
@@
 
-The semantics and metatheory presented above, have
-been fully mechanized in the Agda proof assistant.
-The mechanization additionally includes the decoupled Hazelnut action semantics described in \cref{sec:calculus-hazelnut}.
-Though the mechanization's documentation contains
-more detailed discussion regarding technical decisions made therein, we highlight some important
-aspects here.
+The semantics and metatheory presented above have been fully mechanized in the Agda proof assistant.
+The mechanization additionally includes the decoupled Hazelnut action semantics described in
+\cref{sec:calculus-hazelnut}.
+Though the mechanization's documentation contains more detailed discussion regarding technical
+decisions made therein, we highlight some important aspects here.
 
 The standard approach of modeling judgments as inductive datatypes and rules as constructors for
-those datatypes is taken. By representing marked expressions with implicitly typed terms,
-we get part of \cref{thm:calculus-marking-well-formedness} for free from the definition of marking. For the convenience of
-readers interested in browsing the mechanization, as possible, all rule names match those presented here.
+those datatypes is taken.
+By representing marked expressions with implicitly typed terms, we get part of
+\cref{thm:calculus-marking-well-formedness} for free from the definition of marking.
+For the convenience of readers interested in browsing the mechanization, as possible, rule names
+match those presented here.
 

### 2023-07-12 intro.tex

@@
 There is no definitive type or binding error at either use of \li{x}, 
-so we do not report an error in Hazel. More specifically, we treat \li{x}'s type as the unknown a.k.a. dynamic type from gradual typing.
+so we do not report an error in Hazel. More specifically, we treat \li{x}'s type as the unknown a.k.a. dynamic type from gradual typing \cite{Siek06a}.
 In any of these cases, we would like to be able to recover and 
@@
 
-Section~\ref{sec:calculus} of this paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus}.
+\Cref{sec:calculus} of this paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus}.
 We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
 type annotations to expressions \cite{Localinf,BidirTyping}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{Siek06a,siek2015refined}, and 
-(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{HazelnutPOPL}, operationalizing the ``red lines'' that an editor displays to the user.
+(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{HazelnutPOPL}, operationalizing the ``red lines'' that an editor displays.
 The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solar2013program}) can be marked, i.e.
@@
 with destructuring patterns and System F-style polymorphism,
-that serve as case studies of the more general approach that we hope will be adopted by language designers defining practical type systems.
+that serve as case studies of the general approach that we hope this paper will be adopted by language designers defining practical type systems.
 
-\cref{sec:calculus-hazel} describes our own effort in this direction, which is to scale up the marked lambda calculus as the new basis for Hazel, a typed functional programming environment that fully solves the semantic gap problem: Hazel's semantic services are available as long as the program sketch is syntactically well-formed.
+\Cref{sec:calculus-hazel} describes our own effort in this direction, which is to scale up the marked lambda calculus as the new basis for Hazel, a typed functional programming environment that fully solves the semantic gap problem: Hazel's semantic services are available as long as the program sketch is syntactically well-formed.
 Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as in Agda \cite{norell:thesis}, GHC Haskell \cite{haskell-holes}, Idris \cite{DBLP:journals/jfp/Brady13}, and others), syntax error recovery \cite{medeiros20,sorkin11}, and structure editing \cite{HazelnutPOPL,teitelbaum_cornell_1981}.

### 2023-07-12 related.tex
Updates from Overleaf

@@
 Non-empty holes in Hazelnut generalize to marks in this work.
-In brief, we contribute a total marking procedure and type hole inference procedure for a system based closely on Hazelnut, and use it to fix some expressiveness issues in Hazelnut's edit action calculus.
+In brief, we contribute a total marking procedure (Section~\ref{sec:calculus}) and type hole inference scheme (Section~\ref{sec:thi}) for a system based closely on Hazelnut, and use it to fix some expressiveness issues in Hazelnut's edit action calculus (Section~\ref{sec:calculus-structured-editing}).
+
+Hazelnut is in turn rooted in gradual type theory \cite{Siek06a, siek2015refined}, and we make extensive use of (only) the static aspects of gradual typing, namely the universal consistency of the unknown type, to enable recovery from marked errors, which can leave missing type information.
+
+Our focus was exclusively on static typing in this paper, and the results are relevant to the design of language servers for any statically typed language, but it is worth noting that the results in this paper, taken together with Hazel's support for maintaining syntactic well-formedness using structure editing \cite{DBLP:conf/vl/Moon023,moon2022tylr} and for running programs with holes and marked errors \cite{HazelLive}, allow our implementation of Hazel to achieve \emph{total liveness}: every editor state is syntactically, statically, and dynamically meaningful, without gaps.
+
+Type error localization is a well-studied problem in practice.
+This paper is the first to formally support the intuition that, in the words of \cite{BidirTyping}, ``bidirectional typing improves error locality''.
+Although there has been considerable folklore around error localization
+for systems with local type inference, the problem has received little formal attention. We hope that this paper, with its rigorous formulation
+of type error localization and recovery for bidirectionally typed languages, will provide more rigorous grounding to language server development,
+much as bidirectional typing has done for type checker development.
+
+For systems rooted in constraint solving, there has been considerable work in improving error localization because it is notoriously well-understood that such systems make error localization difficult, and programmers are often confused by localization decisions \cite{DBLP:conf/popl/Wand86} because they are rooted in \emph{ad hoc} traversal orders \cite{mcadam1998unification,DBLP:journals/toplas/LeeY98}.
+More recently, there has been a series of papers on finding the most likely location for an error based on either manual weights \cite{DBLP:conf/popl/ZhangM14,DBLP:conf/oopsla/PavlinovicKW14} or learned weights \cite{SeidelBlame}.
+While improving the situation somewhat, these remain fundamentally \emph{ad hoc} in their need to guess intent.
+However, data-driven approaches could perhaps be layered atop type hole inference to improve the ranking or filtering of suggestions.
+
+A more neutral alternative is to derive a set of terms that contribute to an error, an approach known as type error slicing \cite{DBLP:conf/esop/HaackW03,DBLP:journals/tosem/TipD01,DBLP:conf/sfp/Schilling11}.
+This creates a large amount of information for the programmer to consume.
+Our approach is to instead simply report the constraint inconsistencies on a hole in the program and allow for the programmer to interactively refine their intent, so only the local inference system is responsible for identifying particular erroneous expressions.
+We do not make particular usability claims about the interactive affordances related to type hole inference in this paper, but rather simply claim a novel neutral point in the overall design space that uniquely combines local and global approaches.
+
+Recent work on gradual liquid type inference described an exploratory interface for filling holes in refinement types by selecting from partial solutions to conflicting refinement type constraints \cite{DBLP:journals/pacmpl/VazouTH18}.
+This is similar in spirit to type hole inference as described in this paper, albeit targeting program verification predicates.
 
 Our focus is on localization and recovery but not type error repair, as has been considered by other work \cite{lerner07}. We hope that our work will drive future work on rigorous repairs.
+
+The underlying unification algorithm is essentially standard. In particular, we base our approach on the system described by \cite{GradualInfer}
 
@@
 
-The generation of correcting sets is a common approach to error localization.
-The correcting set is assigned blame for failure in type inference \cite{sherrloc, typeinferDif, Pavlinovic2015}.
-Other approaches to assigning blame include statistically derived approaches that leverage machine learning to predict the best candidates for blame \cite{SeidelBlame}.
-Our focus in this paper was to avoid heuristics and data-driven approaches and instead focus on systematic semantic approaches.
-Data-driven approaches could perhaps be layered to improve suggestions.
-
 Explanations for type errors can be well illustrated by providing sample inputs (dynamic witnesses) that elicit runtime errors.
 With this approach, one can generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with around 70\% accuracy \cite{Seidel2016}.
+
+- Local constraint solving in the style of HM(X) ("Type Inference with Constrained Types", Odersky et al), interleaves local reasoning and constraint solving.
+
+- Inferno ("Hindley-milner elaboration in applicative style: functional pearl", by Pottier) showcases how to elaborate and solve constraints at the same time, which would allow you to emit marked terms and build up constraints in one fell swoop.
+
+- Pottier & Remy's "The essence of ML type inference" (for instance, OutsideIn(X) for GHC Haskell
+
+- Garcia and Cimini's POPL 2015 paper on type inference for gradually-typed languages, which improves substantially on the prior work of Siek and Vachharajani
+
+- Lennon-Bertrand et al.'s TOPLAS 2022 paper on Gradual CIC includes an integration of bidirectional typing and gradual typing

### 2023-07-12 intro.tex

@@
 However, many functional languages including Elm, OCaml and Haskell feature constraint-based type inference, with constraints gathered globally.
-For type systems where inference is decidable, this is powerful, but it is also notorious for complicating error localization, because inconsistencies can arise through a confluence of constraints originating from any number of locations.
+For type systems where inference is decidable, this is powerful, but it is also notorious for complicating error localization, because inconsistencies can arise through a confluence of constraints originating from any number of locations \cite{DBLP:conf/popl/Wand86}.
 

### 2023-07-12 calculus.tex
Writing, cleanup

@@
 
-Programmers often work with program states that are syntactically well-formed but not well-typed.
-\Cref{fig:calculus-examples} shows common type errors as they appear in the Hazel programming
-environment. 
+To begin to motivate the development of this section, consider \cref{fig:calculus-examples}, which
+shows common type errors as they appear in the Hazel programming environment. These programs are
+syntactically well-formed but ill-typed.
 % For example, a simple error that may occur is the use of an free variable (\cref{fig:calculus-examples-free}).

### 2023-07-12 polymorphism.tex
Clean up some writing in section 2
Writing, cleanup

@@
-\subsection{Parametric Polymorphism}
+\subsection{Parametric Polymorphism and Richer Judgmental Structures}
 \label{sec:calculus-polymorphism}
@@
 The case studies above demonstrate that the framework of the marked lambda calculus is suitable to support a variety of extensions to richer judgmental structures.
-We hope that this spurs language designers to design marked variants of other calculi, following the recipe we have demonstrated.
-% TODO explicitly give recipe.
+We hope that this spurs language designers to design marked variants of other calculi, following the
+recipe we have demonstrated: starting with a gradual, bidirectional type system, by considering each
+possible failure case, one may systematically derive the necessary error marks and marking rules.
 
-We leave as future work the task of defining a marked version of even more elaborate bidirectional type systems, e.g.
-\citet{dunfield2013complete}'s rather substantial formulation of implicit type application in a bidirectional setting (which would have to handle, for example, the situation where no implicit argument can be resolved).
+We leave as future work the task of defining marked versions of even more elaborate bidirectional
+type systems, e.g. \citet{dunfield2013complete}'s rather substantial formulation of implicit type
+application in a bidirectional setting (which would have to handle, for example, the situation where
+no implicit argument can be resolved).
 

### 2023-07-12 related.tex

@@
 
-Our focus is on localization and recovery but not type error repair, as has been considered by other work \cite{lerner07}. We hope that our work will drive future work on rigorous repairs.
-
-The underlying unification algorithm is essentially standard. In particular, we base our approach on the system described by \cite{GradualInfer}
-
-\citet{garcia:2015} presents a static implicitly typed language, where users opt into dynamism by annotating an expression with the gradual type "?", and an associated type inference algorithm.
+The underlying unification algorithm is essentially standard--the novelty is in how the unification results are used and how failures are handled, rather than in the inference itself.
+In particular, we base our approach on the system described by \cite{GradualInfer}, because it also identifies type inference variables with the unknown type from gradual typing and the union find data structure is useful for computing possible type sets.
+\citet{garcia:2015} similarly present a static implicitly typed language, where users opt into dynamism by annotating an expression with the gradual type "?", and an associated type inference algorithm and accompanying metatheory.
 By contrast, the Hazelnut type system assigns gradual types to programs that would ordinarily not type-check in a non-gradual system by wrapping them in expression holes.
 The type inference algorithm presented in \citet{garcia:2015} also does not specify what to do if the constraint set cannot be solved.
-If a single static type cannot be determined for an expression, its type is simply "undefined", whereas our Type Hole Inference algorithm provides a list of suggestions derived from any conflicting constraints if a single substitution cannot be determined.
+If a single static type cannot be determined for an expression, its type is simply undefined, whereas our approach provides a list of suggestions derived from any conflicting constraints if a single substitution cannot be determined.
+As our focus is on failure cases and partially consistent suggestions, the metatheory in this prior work is less relevant in guiding the design of the type hole inference system.
 
-\citet{GradualInfer} explore a version of unification for gradually typed languages based on Huet's unification algorithm \cite{Huet}.
-Other common approaches to unification are generally based on Hindley-Milner type inference \cite{MilnerInfer}.
-These algorithms are efficient, but can fail to solve constraints in the face of inconsistencies.
+Of note, however, is that type inference approaches that eagerly solve and substitute for type inference variables \cite{DBLP:journals/tapos/OderskySW99,DBLP:conf/icfp/Pottier14,mcadam1998unification} are not well-suited to the type hole inference approach, as they lose information necessary for computing partially consistent suggestions.
+
+% In addition to type error localization and recovery, type hole inference touches on the problem of type error repair, as has been considered by other work \cite{lerner07}.
+We hope that our work will drive future work on rigorous repairs.
 
@@
 
-Explanations for type errors can be well illustrated by providing sample inputs (dynamic witnesses) that elicit runtime errors.
+Our focus in this paper has not been on the error messages displayed on screen, about which we make no specific claims.
+There has been considerable work on improving explanations for type errors, e.g.
+by providing sample inputs (dynamic witnesses) that elicit runtime errors.
 With this approach, one can generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with around 70\% accuracy \cite{Seidel2016}.
-
-- Local constraint solving in the style of HM(X) ("Type Inference with Constrained Types", Odersky et al), interleaves local reasoning and constraint solving.
-
-- Inferno ("Hindley-milner elaboration in applicative style: functional pearl", by Pottier) showcases how to elaborate and solve constraints at the same time, which would allow you to emit marked terms and build up constraints in one fell swoop.
-
-- Pottier & Remy's "The essence of ML type inference" (for instance, OutsideIn(X) for GHC Haskell
-
-- Garcia and Cimini's POPL 2015 paper on type inference for gradually-typed languages, which improves substantially on the prior work of Siek and Vachharajani
-
-- Lennon-Bertrand et al.'s TOPLAS 2022 paper on Gradual CIC includes an integration of bidirectional typing and gradual typing
+These and a variety of other approaches to explaining errors are a fruitful avenue for future integration into the system.

### 2023-07-12 conclusion.tex
Updates from Overleaf

@@
 
+\begin{quote}
+    \emph{Nothing will ever be attempted if all possible objections must first be overcome.} 
+    \begin{flushright}- Samuel Johnson\end{flushright}
+\end{quote}
+
 Programming is increasingly a live collaboration between human programmers and sophisticated semantic services. 
-These services need to be able to reason throughout the programming process, not just when the program is already complete.
+These services need to be able to reason throughout the programming process, not just when the program is formally complete.
 This paper lays down rigorous type-theoretic foundations for doing just that.

### 2023-07-12 polymorphism.tex
Updates from Overleaf

@@
 possible failure case, one may systematically derive the necessary error marks and marking rules.
+The metatheorems, particularly totality, ensure that no rules or premises have been missed.
 
@@
 application in a bidirectional setting (which would have to handle, for example, the situation where
-no implicit argument can be resolved).
+no implicit argument can be resolved) or \citet{DBLP:phd/hal/LennonBertrand22}'s gradual bidirectional
+variant of the dependently typed calculus of constructions.
 
 % TODO cite gradual dependent types papers, etc.
+% DBLP:phd/hal/LennonBertrand22

### 2023-09-16 review/icfp23/response.md
Move ICFP 2023 reviews and response to subdirectory

Thank you for your thoughtful and detailed comments! We believe that the issues raised can be
addressed with only small revisions.

# Extensions (RA, RB, RD)

The reviews suggested extensions of the marked lambda calculus to handle polymorphism (RA, RD),
linear types (RB), dependent types (RB), and subtyping (RD). Our intention was to present a minimal
framework that, like the STLC, is intended to establish guiding metatheory–most importantly,
totality–in support of conservative extensions in many subsequent directions. We dedicate Sec. 2.2
and 2.3 to two small examples of such extensions, emphasizing how totality guides us. We do
similarly consider System-F-style polymorphism in Sec. 2.5, albeit less formally for space reasons;
we would be happy to use an extra page in the final version to include the straightforward rules for
“marked System F” if requested.

We conclude Sec. 2.5 with a perhaps too brief discussion of more sophisticated extensions, e.g.
implicit type application. We would be happy to more explicitly lay out the general approach, which
is still to first define a bidirectional + gradual version of the feature, then use totality as a
guide. We will cite relevant work on gradual implicit type application [1], gradual dependent types
[2], and gradual subtyping [3] in the final version of this section; we aren’t aware of a gradual
linear type system, though there are gradual session type systems [4]. However, it is beyond the
scope of this initial paper to define marked dialects of these systems, which each comprise dozens
of typing rules.

Our implementation does scale beyond the calculus to those features found in ML-family languages
like Elm, e.g. full pattern matching.

We agree that marked dialects of advanced type systems might require a designer to experiment with
different localization heuristics when errors are due to a confluence of factors (e.g. as in Sec.
2.1.6 with conditionals). Our contribution is to provide a framework for clearly specifying these
design decisions, rather than leaving them implementation-defined.

# Mechanizing constraint solving / Sec. 4 contributions (RB, RC, RD)

While there are several mechanizations of unification-based type inference (e.g. [5]), the
metatheory of these systems (e.g. soundness, principality) is relevant only when the constraint set
is consistent. Our focus in Sec. 4 is on inconsistent constraints: suggested fillings can be
inconsistent with the constraints. There is no critical metatheorem, like totality, that must guide
this development; the approach in Sec. 4 is instead fundamentally rooted in the heuristic that even
inconsistent constraints likely contain information about user intent. The key technical challenge
is therefore in efficiently computing partially consistent suggestions, so we intentionally favored
an algorithmic presentation in Sec. 4.4.

We also intentionally used simple, well-studied techniques, like union find, because the real
contribution of Sec. 4 is the system architecture: we demonstrate that one can dramatically clean up
type error localization by using constraint solving only for localizing erroneous type holes,
layering it atop a bidirectional + gradual core where localizing erroneous expressions is more
systematic. Previously, these approaches were seen as alternatives. We agree that Sec. 4 could use a
clearer “punchline” along these lines (RC, RD) and general clarity improvements (many suggested by
the reviewers).

# Related work: interleaving (RA)

We agree that the approach taken in contemporary constraint-based type inference implementations,
cited by RA [6, 7], of interleaving constraint generation and solving could inform efforts to
interleave marking (Sec. 2) and constraint solving (Sec. 4). However, because Sec. 2 can be adopted
on its own, we believe that introducing these topics separately, based on more elementary and
widely-understood foundations (e.g. STLC, HM) is expositionally preferable. We will discuss.

# Empirical evaluation (RB, RC)

The primary contribution of Sec. 2 is a metatheoretic framework, so our primary evaluation of this
contribution is the mechanized metatheory. While it would be interesting for future work to, for
example, empirically evaluate different approaches to localizing inconsistencies between conditional
branches, we make no specific usability claims and formalize each possible approach.

The primary contribution of Sec. 4 is a novel architecture, described above, that simplifies and
systematizes a previously complex and ad hoc task: constraint inconsistency error localization.
Certainly it would be interesting to empirically evaluate the “code completion” affordance we
present that allows quick toggling between partially consistent type hole fillings, but we again
make no specific usability claims about this particular interface. It is optional and can be toggled
off without any loss of liveness elsewhere in the system, which relies only on the localization
mechanisms detailed in Sec. 2.

We have deployed Hazel in the classroom recently, so we hope to observe how students interact with
all aspects of the relevant affordances, including localization choices, error messages, and code
completions, over the coming years, but we believe that the metatheoretic and architectural
contributions of this paper open up an interesting and novel region in the design space on their
own. We will be careful to make clear that there are no specific HCI claims related to the UI
screenshots or specific error message text in this paper.

# What is “gradual recovery”? (RC, RD)

We will clarify in the introduction that by “gradual recovery” we refer to our use of gradual typing
to both recover from missing type information (pg. 3, Sec. 2) and to provide a site for
interactively recovering from conflicting constraints (pg. 4, Sec. 4).

# Motivation for Sec 3 (RA, RD)

Section 3 is a short section pointing out that marking allows us to resolve a significant problem
related to undefined behavior in the Hazelnut edit action calculus. We introduce the judgemental
forms but indeed, a deep understanding of this section would require familiarity with Hazelnut. We
can note that this section can be safely skipped for readers uninterested in formal models of
structure editing. (It can also be omitted from the final version if requested.)

# Additional Responses

We respond to some more specific questions brought up by each reviewer below.

## RA

    A good part of the early presentation is a bit repetitive in nature, and could be summarized by
    stating that valid unmarked typing derivations are left as-is in the marked word. Is that
    correct?

Yes, this is formalized with Theorem 2.3, part 1. We can mention this sooner.

    Why aren't destructuring bindings implemented like this at all? The proposed way seems much more
    complicated, more expansive, and it's unclear to me why it would provide better error messages.

Sec. 2 is intended as a standalone system that does not incorporate constraint solving, so this was
not an option. We included this in the paper because a strictly bidirectional approach is indeed
rather subtle and we hadn’t previously seen this issue brought up, e.g. in Dunfield’s 2018 review
paper. The approach we specify ensures that information from both the expression and pattern is
unified, while taking pattern type annotations as “ground truth” for localizing errors in the
expression. Sec. 4’s approach would generate the constraints as RA describes.

    The use of ?u for unification variables confuses me a lot. It seems to mix up unification
    variables with the gradual unknown, which seems completely wrong to me. A gradual unknown is any
    type (possibly several different ones during the execution), a unification variable is a single
    type that isn't determined yet. Am I missing something motivating your usage? (RA)

We consider missing type annotations type holes for the purposes of inference, i.e. we presume the
user intends to fill type holes with a known type. To expose gradual typing for intentional use, we
could introduce a “dyn” type that would fill a type hole and behave identically except that it would
not be subject to type hole filling. (This is related to the approach explored by Garcia and Cimini,
which was mentioned by RD.)

    If I remember correctly, Elm also uses this graph to diagnose type errors, by recognizing
    specific patterns in the graph of types and/or constraints. Is that something you do as well,
    and can you develop a bit on that? (RA)

We take the opposite approach, as we described above. Briefly, we do not attempt to try to localize
constraint solving failures to expressions, leaving that to the marking system. This avoids the need
to define this style of ad hoc system – type inference variables that cannot be solved can be
systematically localized to type holes, and the user can interactively choose from partial fillings
to see where the problem would be localized in a second step.

# RB

    can I essentially think of this as type checking, but that it will always produce a type where
    there may nevertheless be marks to indicate the location of type errors? And an unmarked
    expression is not necessarily well typed but is well formed?

Yes, precisely!

# RC

    Why are your rules algorithmic? (RC)

The rules are incidentally algorithmic from our perspective – bidirectional typing is useful because
it specifies a static “direction” that information flows, which thereby systematizes error
localization. In addition, we want to ensure unicity, i.e. that there are not multiple possible
markings, because our goal is to formalize what would normally be an implementation decision.

# RD

    Could you clarify what would happen if a conditional would be subsumable? (RD)

This would violate unicity (i.e. make marking non-deterministic) in the case where a conditional
expression with inconsistent branches appeared in an analytic position, e.g.

```
let x : Num = tt ? ff : 2
```

The analysis rule would mark the error on the “then” branch, while the synthesis rule would mark the
error on the conditional as a whole (or arbitrarily on one of the branches if the alternative rules
are used).

## Related Work (RD)

Thank you for bringing up some more distantly related work on type inference, which we respond to
below. We will include this in the final related work section.

    Garcia and Cimini's POPL 2015 paper on type inference for gradually-typed languages, which
    improves substantially on the prior work of Siek and Vachharajani.

Indeed, this paper is a cleaner presentation of the approach explored initially by Siek and
Vachharajani. It distinguishes explicit unknowns from types that must be inferred, whereas we
consider type holes as types that need to be inferred (and also allow other internally generated
unknowns to be inferred, with full provenance). We could add a “dyn” type to allow users to mark
that they intend the gradual unknown semantics and do not want a solution to be inferred (see
response to RA above) which would be consistent with Garcia and Cimini. This paper does not consider
the key problem we focused on, which is generating partial solutions, but we agree that this paper
should be discussed in more detail alongside Siek and Vachharajani in the final version.

    The second is Vazoy et al's OOPSLA 2018 paper on gradual liquid type inference, which features a
    similar treatment of holes and interactive exploration of possible solutions (in that case, for
    unknown refinements, but the underlying concepts are very close).

We agree that the liquid type exploration outlined in this paper is similar in spirit to our
approach in that it allows for interactive exploration of conflicting refinement type constraints,
localized to holes in a refinement type specification. Marking is not detailed in this paper, and
the technical approach to synthesizing refinements is quite different, but we will certainly mention
that the exploratory interface we propose has connections to this prior work!

    Also, Lennon-Bertrand et al.'s TOPLAS 2022 paper on Gradual CIC includes an integration of
    bidirectional typing and gradual typing.

We mention this at the beginning of the author response – we agree that this system would be a
suitable basis for creating a marked CIC and look forward to future explorations in this direction!

[1] Xie et al, 2019 (based on Dunfield and Krishanswami, 2013)

[2] Lennon-Bertrand et al, 2022

[3] Siek and Taha, 2007

[4] Igarashi et al, 2017

[5] Paulson, 1993 (and several other subsequent efforts, which we will cite as well)

[6] Odersky et al, 1999

[7] Pottier, 2014

### 2023-09-16 review/popl24/response.md
Add POPL 2024 reviews and author response

Thank you for your thoughtful and detailed comments. We address questions brought up by the reviewers below. The issues raised can be addressed with revisions that will correct the typos and small expositional clarity issues identified by the reviewers and expand on the discussion of the design space, as suggested by the reviewers.

# Review A

> In the rules you have chosen τ₁ from the environment. I would argue that τ is the more natural
> choice from the programmer's point of view (that's what they wrote!) and ? is the more principled
> one. At the very least, you need to acknowledge that there is a choice to make here.

This is actually a typo—indeed, we agree that it is more natural to choose τ (and in fact this is
the choice found in the Agda mechanization)! We will fix this in the final version and, in general,
more clearly highlight these spots where language designers must make such decisions.

> It's not clear at all why the first two premises [of USLetPat] (involving τp) are needed at all,
> since no information flows out of them.

Yes, those two premises are unnecessary. They were included as, following the marking logic from
ISLetPat, we must ultimately analyze the definition against the pattern type to properly attribute
any inconsistency error to the definition, given that we want any annotations in the pattern to be
canonical. However, for USLetPat, the analysis of the pattern against the definition type suffices
to ensure consistency, so the second and hence the first premise are redundant, and will be removed
from the final version, where we will clarify this point.

> Is the correspondence between this approach to polymorphism (using the gradual unknown type for
> unification variables) and standard Hindley-Milner known? Are there any systems implementing this
> currently?

Garcia and Cimini (2015) have previously developed the connection between polymorphism, gradual
typing, and constraint-based (Hindley-Milner) type inference. We discussed this paper in the Related
Work section, but will also mention it alongside Siek and Vachharajani (2008) in Section 4.4 in the
final version (and expand our discussion of it in the Related Work section).

> In section 3.1 you write that you have implemented marking for algebraic datatypes and explicit
> polymorphism, but there are no examples for this in the supplied Hazel build, and I can't figure out
> what the syntax is.

These features were in different branches of Hazel at submission time, so they did not end up in the
build provided in the submitted supplement. They will be merged in the final artifact. We will also
include an Examples tab in the UI which will demonstrate the relevant syntax.

# Review B

> Lines 644-645 say that "[i]f the pattern a suggests that the conditional is in synthetic position,
> it will be marked with an inconsistent branches error mark following our development above. If
> instead it is in analytic position against the unknown type, no mark will be produced." Is this
> desirable? I would expect to get a type error pointing out inconsistent branch types regardless of
> how the expression is type checked.

The gradually typed interpretation of type holes as the unknown (i.e. dynamic) type suggests
allowing each branch to be checked independently (as only one of the branches will be taken at
run-time). It is essentially as if each branch was independently annotated with a ? type ascription,
which would cause both branches to become consistently typed. The inconsistency between branches
would still be caught by the type hole inference mechanism described in Section 4, localizing the
problem to the ? itself. We will clarify this in the final version.

> Have you thought about how this (e.g., Section 2.1.1) would work with operator overloading (e.g.
> if you add a float type and overload + to operate on both ints and floats?)

Overloading would require changing one or both operand modes to synthetic (depending on how operator
overloading is defined—the left operand is often arbitrarily privileged in languages with this
feature) so that the overloading can be resolved during typing/marking. If an operator is not
defined given the synthesized type of the operand(s), the error might be marked on the operator
itself. Type classes (a.k.a. traits / implicits) present a symmetric alternative to operator
overloading where the operands can remain in analytic position. We will briefly discuss this
situation in the final version.

> I accept that producing error messages was not the focus of the paper, but are there more general
> lessons you can draw on error localization (e.g., for other languages/paradigms, or if one wants to
> do better error localization but without buying fully into the mechanisms of typed holes)?

The marked lambda calculus relies in some instances on having an unknown type à la gradual typing
(which has been studied and applied across many language paradigms, including imperative / OO
languages). However, we will note more clearly that empty expression holes are not integral to the
core calculus and recipe—they are included to illustrate how the marked lambda calculus fits into
the larger problem of modeling incomplete program states and to discuss polymorphic generalization
in Section 4.4, but could be removed entirely from the marked lambda calculus without issue. As
such, the techniques described in the paper should be relevant to a wide variety of language
designs, including those where hole-driven development is not part of the user workflow.

# Review C

> line 288: I was expecting some kind of "Localisation" principle about pushing errors to
> a maximally local position. Is that not possible? It seems related to "the neutral approach."

We did not consider the problem of measuring locality but we note that the typing rules for marks
ensures that the marks are placed precisely at locations where inconsistencies exist, and therefore
less local marks (e.g. on parent expressions) would violate the metatheory. Bidirectional typing
specifies a flow of typing information that ultimately justifies the localization decisions in the
marked lambda calculus; the justification for the type flow of bidirectional typing is discussed by
other authors, notably Dunfield and Krishnaswami (2019).

> line 442: what would break if you synthesized the body in the premise of the rule rather than
checking against? Anything? Is there a guiding principle around this?

It is indeed possible to instead synthesize the body here, but we chose analysis because the overall
lambda is in analytic position so, were the expected type an arrow type, the body would be in
analytic position and therefore we retain this mode using the unknown type. The design principle
here is perhaps some notion of “conservation of mode”, which we can discuss in the final version,
but it is not critical here.

> line 452: In the premise of MKALAM3, why is x typed \tau_1, as opposed to \tau or "?"? What is the
> guiding principle at play here?

As mentioned above in response to a similar question in Review A, this is actually a typo, and
indeed we believe it more natural to choose τ (and this choice is the one found in the Agda
mechanization). This will be corrected in the final version.

> line 458: In the USAp rule, is it necessary for e2 to be checked against \tau_1? Is it plausible
> to instead synthesize the argument type and then "half-check/half-synthesize" the function? It would
> be nice to understand what the boundaries of this design are.

We are not entirely sure what is meant by “half-check/half-synthesize”, but our development of the
unmarked typing rules follow the typical formulation of function application in bidirectional
typing. Dunfield and Krishnaswami (2019) discuss other “modes” for function application where
information could flow in other directions, and if these alternatives were chosen, error
localization would need to be updated accordingly.

> line 531: Why must localization be deterministic? Given the number of design decisions involved,
> is it plausible to defer them or choose in some delayed manner, or by using more program context
> to decide?

The algorithmic and deterministic nature of the localization calculus makes it straightforward to
implement as part of a real type checker. Ultimately, how errors are localized depends heavily on
the formulation of the original type system. However, it may be possible to involve the user or
other heuristic mechanisms to resolve ambiguity rather than specifying a decision
deterministically—we explore this in concept in Section 4 for type inference errors in particular,
and will mention it more broadly as a potential direction for future research.

> Section 2.3 seems to reuse the ? type to model phenomena that should make sense in a language that
> is bidirectionally-typed, has pattern matching, and has no notion of gradual typing whatsoever.
> I think that this would be a much better contribution if it started with explicating a purely static
> version of bidirectional pattern matching and then demonstrated that the marking recipe transfers
> over to this purely-static version. I expect that such a system would need some way of accounting
> for partial analysis and synthesis, but without using gradual typing (albeit something related but
> static).

The type ?⇒ for switching to synthesis does indeed not require gradual typing and is merely
a marker for switching to synthesis because we are using analysis to propagate information from the
pattern. It never needs to participate in a type consistency relation (via subsumption) because the
UASynSwitch rule would always apply there. It need not look like an unknown type, i.e. using ? may
be misleading notation, so we will clarify this as suggested in the final version.

> line 678: the USLETPAT rule seems to depend on all expressions being able to synthesize, which is
> not the case in all bidirectional systems (e.g. the common treatment of unannotated lambdas). Can
> this be generalized?

Yes, we do assume that successful analysis implies synthesis, but a more general treatment here
would choose to use the pattern’s synthesized type to analyze the bound expression in situations
where no type can be synthesized. We will describe this design dimension in the final version.

> line 730: "these marked types operate identically..." why is that the "right" thing to do? What is
> the guiding principle (aside from it seeming to work)?

In the same way that unbound variables and other error-marked expressions in our core development
synthesized the unknown type to allow type-checking to continue, so too do unbound type variables
behave like the unknown type.

> line 816: what is the box superscript?

This is mark erasure, as defined in Figure 3.

> line 828: I find it strange to call bidirectional typing "local type inference" since local type
> inference (at least in Pierce and Turner) was a combination of bidirectional typing and constraint
> solving.

We will make this distinction more clear in the final version.

> line 884: What is the red angle bracket? I didn't see it explained.

The editor cursor in Hazel is drawn as an angle bracket at expression boundaries to convey the
“hexagonal block structure” of Hazel expressions (see Moon et al, VL/HCC 2023 for details – we will
clarify in the final version).

> line 963: "without needing to introduce explicit constraints" I did not understand this, since the
> rules seem to be adding explicit constraints.

We will clarify that the purpose of the matched arrow provenances here is to avoid generating fresh
type inference variables each time a matched arrow type is invoked, which would lead to a blow up in
the number of inference variables that must necessarily be equal to one another.

> line 969: "A marked expression has already been deemed erroneous" but why is it the expression,
> rather than the boundary between outside and inside?

The bidirectional flow in Section 2 determines expression error localization, and indeed it is
typically due to inconsistencies between the type information “outside” and “inside” the marked
expression that causes issues.

> line 982: "a standard union-find based unification algorithm" but this algorithm must account for
> consistency, right?

Yes, the algorithm favors more precise types as previously described e.g. by Siek and Vahharajani
(2008) and Garcia and Cimini (2015). We will make this point more clear.

### 2023-10-24 let.tex
Updates from Overleaf
USLetPat and related adjustments

@@
 In a typed setting, we may want to granularly add type annotations in patterns as well.
-As it turns out, this combination of features is surprisingly tricky to get right!
+In a bidirectional setting, as it turns out, this combination of features is surprisingly tricky to get right!
 Let us add let expressions and simple patterns, writing $\PWild$ for the wildcard pattern, $\PVar{x}$ for a variable pattern, and $\PPair{\PMV_1}{\PMV_2}$ for a pair:
@@
 
-To type this, the most obvious approach is to synthesize a type from the pattern and analyze the
+To type this, one approach is to synthesize a type from the pattern and analyze the
 definition against that type. However, this may run afoul of user expectation, which might
@@
 
-To ensure propagation of type information between pattern, definition, and body, the destructuring
-let requires some extra machinery.
+Let synthesis is straightforward: we synthesize the definition's type, then analyze the pattern against it:
+
 \[%
   \judgment{
-    \ctxSynPatU{\ctx}{\PMV}{\TMV_p} \\
-    \ctxAnaTypeU{\ctx}{\EMV_{1}}{\TMV_p} \\
-    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\\\
+    %\ctxSynPatU{\ctx}{\PMV}{\TMV_p} \\
+    %\ctxAnaTypeU{\ctx}{\EMV_{1}}{\TMV_p} \\
+    \ctxSynTypeU{\ctx}{\EMV_{1}}{\TMV_{1}} \\
     \ctxAnaPatU{\ctx}{\PMV}{\TMV_{1}}{\ctx'} \\
@@
 \]%
-First, to ensure that the pattern and definition types are consistent, the pattern synthesizes a
-type, against which the definition is analyzed. Then, the definition synthesizes a type, against
-which the pattern is analyzed. This analysis is guaranteed to succeed; we need the context it
-produces for use in the synthesis of the body's type. The analytic rule is similar, with the final
+
+The analytic rule is similar, with the final
 premise and conclusion changed to analysis.
+\\
+Attempting an analogous marking rule runs into trouble.
+We want any type annotations in the pattern to be canonical, which means analyzing the definition to attribute to it any inconsistencies.
+But we still must analyze the pattern, incorporating the definition's type to produce the context needed by the body.
+So we do a 'round trip': Analyze the definition against the pattern's type, and then the pattern against the definition's.
+Since the first analysis establishes consistency, this second analysis is guaranteed to succeed; we only care about the context produced:
 
-The marking rule is directly analogous and utilizes the marking of patterns, which we omit---they
-may be derived in the same way as those for expressions and are governed by similar metatheorems (again, totality guides the derivation).
-In the analytic cases, we introduce error marks that parallel to $\ECInconType{\ECMV}$ and 
-$\ECAnaNonMatchedProd{\ECMV}$.
 \[%
@@
 \]%
+
+We omit the marking of patterns---they
+may be derived in the same way as those for expressions and are governed by similar metatheorems (again, totality guides the derivation).
+In the analytic cases, we introduce error marks that parallel to $\ECInconType{\ECMV}$ and 
+$\ECAnaNonMatchedProd{\ECMV}$.

### 2023-10-25 abstract.tex
Updates from Overleaf
Fix a bunch of typos

@@
 Errors arising from inconsistent unification constraints are localized exclusively to type and expression holes, i.e.
-the system identifies unfillable holes using a sytem of traced  provenances, rather than localized in an \emph{ad hoc} manner to particular expressions.
+the system identifies unfillable holes using a system of traced  provenances, rather than localized in an \emph{ad hoc} manner to particular expressions.
 The user can then interactively shift these errors to particular downstream expressions by selecting from suggested partially consistent type hole fillings, which returns control back to the bidirectional system.

### 2023-10-25 calculus.tex
Fix a bunch of typos

@@
 We intentionally keep the marked lambda calculus minimal, because it is intended to capture the
-essential idea and introduce a general pattern that language designers can employ to created mark
+essential idea and introduce a general pattern that language designers can employ to created marked
 variants of their own type systems.
@@
     }{TCArr}
+
+    \inferrule[TCProd]{
+      \consistent{\TMV_1}{\TMV_1'} \\
+      \consistent{\TMV_2}{\TMV_2'} \\
+    }{
+      \consistent{\TProd{\TMV_1}{\TMV_2}}{\TProd{\TMV_1'}{\TMV_2'}}
+    }
   \end{mathpar}
@@
   \judgment{
-    \ctxSynTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
+    \ctxSynTypeU{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\TMV_2}
   }{
@@
     \judgment{ }{
-      \matchedArrow{\TArrow{\TMV}{\TMV}}{\TMV}{\TMV}
+      \matchedArrow{\TArrow{\TMV_1}{\TMV_2}}{\TMV_1}{\TMV_2}
     }{TMAArr}
@@
   \judgment{
-    \ctxSynTypeM{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
+    \ctxSynTypeM{\extendCtx{\ctx}{x}{\TMV_1}}{\ECMV}{\TMV_2}
   }{
@@
     \inconsistent{\TMV}{\TMV_1} \\\\
-    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
+    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
   }{
@@
     \meetsTo{\TBool}{\TBool}{\TBool} \\
-    \meetsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TMeet{\TMV_1}{\TMV_1'})}{(\TMeet{\TMV_2}{\TMV_2'})}}
+    \meetsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TMeet{\TMV_1}{\TMV_1'})}{(\TMeet{\TMV_2}{\TMV_2'})}} \\
+    \meetsTo{(\TProd{\TMV_1}{\TMV_2})}{(\TProd{\TMV_1'}{\TMV_2'})}{\TProd{(\TMeet{\TMV_1}{\TMV_1'})}{(\TMeet{\TMV_2}{\TMV_2'})}}
   \end{array}\]
@@
   }{
-    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjLSynNonMatchedProd{\ECMV}}{\TMV_1}
+    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjLSynNonMatchedProd{\ECMV}}{\TUnknown}
   }
@@
 
-We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of the Hazel.
+We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of Hazel.
 % TODO citation and explain what in the world Hazel is (reviewer A stated that the lack of this made this section pretty much impossible to understand)

### 2023-10-25 intro.tex
Writing, cleanup
Fix a bunch of typos

@@
 with destructuring patterns and System F-style polymorphism,
-that serve as case studies of the general approach that we hope this paper will be adopted by language designers defining practical type systems.
+that serve as case studies of the general approach that we hope will be adopted by language designers defining practical type systems.
 
@@
 using machine learning \cite{SeidelBlame}, to avoid misleading programmers), this error is localized, with a \li{!}, to the type hole itself, with the error message providing partially consistent solutions as shown.
+\\
 \begin{center}
-\vspace{-3px}
     % \includegraphics[scale=0.55]{images/intro-thi.png}

### 2023-10-25 let.tex
Fix a bunch of typos

@@
 
+  \inferrule[USPPair]{
+    \ctxSynPatU{\ctx}{\PMV_1}{\TMV_1} \\
+    \ctxSynPatU{\ctx}{\PMV_2}{\TMV_2}
+  }{
+    \ctxSynPatU{\ctx}{\PPair{\PMV_1}{\PMV_2}}{\TProd{\TMV_1}{\TMV_2}}
+  }
+
   \judgment{
@@
 
+  \inferrule[UAPPair]{
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
+    \ctxAnaPatU{\ctx}{\PMV_1}{\TMV_1}{\ctx_1} \\
+    \ctxAnaPatU{\ctx_1}{\PMV_2}{\TMV_2}{\ctx_2}
+  }{
+    \ctxAnaPatU{\ctx}{\PPair{\PMV_1}{\PMV_2}}{\TMV}{\ctx_2}
+  }
+
   \judgment{
@@
 
-Let synthesis is straightforward: we synthesize the definition's type, then analyze the pattern against it:
+Synthesis for let expressions is straightforward: we synthesize the definition's type, then analyze the pattern against it:
 
@@
 premise and conclusion changed to analysis.
-\\
-Attempting an analogous marking rule runs into trouble.
-We want any type annotations in the pattern to be canonical, which means analyzing the definition to attribute to it any inconsistencies.
-But we still must analyze the pattern, incorporating the definition's type to produce the context needed by the body.
-So we do a 'round trip': Analyze the definition against the pattern's type, and then the pattern against the definition's.
-Since the first analysis establishes consistency, this second analysis is guaranteed to succeed; we only care about the context produced:
+
+Attempting an analogous marking rule runs into trouble. We want any type annotations in the pattern
+to be canonical, which means analyzing the definition to attribute to it any inconsistencies. But we
+still must analyze the pattern, incorporating the definition's type to produce the context needed by
+the body. So we do a ``round trip'': analyze the definition against the pattern's type, and then
+the pattern against the definition's. Since the first analysis establishes consistency, this second
+analysis is guaranteed to succeed; we only care about the context produced:
 

### 2023-10-25 calculus.tex
Fix non-THI typos from Review C

@@
 bound. Similar to above, \textsc{USVar} would not apply. A total marking procedure should, however, report the error and
-continue, motivating an \emph{free variable mark} $\ECFree{x}$ and the accompanying rules:
+continue, motivating a \emph{free variable mark} $\ECFree{x}$ and the accompanying rules:
 %
@@
 
-One approach is to define an untyped action calculus, \emph{wholly decoupling} edit actions from typing. Under this simplified design, the action calculus, given by the singular action
+One approach is to define an untyped action calculus, \emph{wholly decoupling} edit actions from typing. Under this simplified design, the action calculus, given by a singular action
 judgment, written $\AUEAction{\EMV}{\EMV'}{\AMV}$, is concerned only with the manipulation of
 syntax, and the total marking procedure yields statically meaningful terms with error marks inserted at the appropriate positions as a kind of editor decoration. This solves the problem
-outlined above; since marking operates on the entire program after each action, arbitrary
+outlined above: since marking operates on the entire program after each action, arbitrary
 constructions are permitted.
@@
 untyped version, and because marking is defined separately, this Agda mechanization can serve
-universally as a baseline for correctness of approaches, like our initial design that aim to make
+universally as a baseline for correctness of approaches, like our initial design that aims to make
 use of type and scope information, and perhaps other analyses, to minimize re-marking. We leave a

### 2023-10-25 calculus.tex
Address some of the detailed comments by review c

@@
 then goes through the rules, organized by syntactic form rather than judgement, to build intuition
-about how the various judgements relate to one another \cite{Siek06a}.
+about how the various judgements relate to one another.
 All rules and theorems may be found organized by judgement form in the supplementary appendix for

### 2023-10-25 intro.tex
Address some of the detailed comments by review c

@@
 
-The developments in Section~\ref{sec:calculus} support the intuition (remarked upon by \citet{Localinf} and \citet{BidirTyping}) that local type inference pairs well with type error localization and recovery because information flows systematically and locally through the tree.
-However, many functional languages including Elm, OCaml and Haskell feature constraint-based type inference, with constraints gathered globally.
+The developments in Section~\ref{sec:calculus} support the intuition (remarked upon by \citet{Localinf} and \citet{BidirTyping}) that bidirectional typing pairs well with type error localization and recovery because information flows systematically and locally through the tree.
+However, many functional languages including Elm, OCaml and Haskell feature constraint-based type inference, where constraints are gathered globally.
 For type systems where inference is decidable, this is powerful, but it is also notorious for complicating error localization, because inconsistencies can arise through a confluence of constraints originating from any number of locations \cite{DBLP:conf/popl/Wand86}.

### 2023-10-25 calculus.tex
Fix more typos and address some of the feedback

@@
 \begin{itemize}
-  \item The \textbf{unmarked language}, which is the original language. Expressions of this language
+  \item The \emph{unmarked language}, which is the original language. Expressions of this language
     are called \emph{unmarked expressions}, denoted by the metavariable $\EMV$.
 
-  \item The \textbf{marked language}, which mirrors the structure of the unmarked language but is
-    extended with explicit \emph{error marks}. We call expressions of this language
-    \emph{marked expressions}, denoted $\ECMV$.
+  \item The \emph{marked language}, which mirrors the structure of the unmarked language but is
+    extended with explicit \emph{error marks} which, intuitively, may be viewed as exactly the
+    errors marked in an editor. We call expressions of this language \emph{marked expressions},
+    denoted $\ECMV$.
 \end{itemize}
@@
 used to represent syntactically incomplete portions of the program à la Hazelnut
-\cite{HazelnutPOPL}. Note, however, that empty expression holes are not semantically critical to the calculus---they are included to emphasize that the system can handle program sketches.
+\cite{HazelnutPOPL}. Note, however, that these empty expression holes are not at all semantically
+critical to the calculus---they are included illustrate how the calculus fits into the larger
+problem of modelling incomplete program states and to discuss polymorphic generalization in
+\cref{sec:polymorphicGlobal}.
 
@@
 
-Finally, the key judgement, \emph{marking}, is also of a bidirectional nature. The synthetic marking judgment
+Finally, the \emph{marking} judgment is also of bidirectional nature. The synthetic marking judgment
 $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$ establishes that, under the context $\Gamma$, the
@@
 Though no expected output type is known, we still need to check and mark the body; it is thus
-analyzed against the unknown type in \textsc{MKALam2}.
+analyzed against the unknown type in \textsc{MKALam2}. It is possible to instead synthesize the
+body, but we choose analysis so that the body is always in analytic mode. The guiding design
+principle in this decision is is a notion of ``conservation of mode,'' but it is not critical to the
+calculus.
 

### 2023-10-25 conclusion.tex
Updates from Overleaf
Fix more typos and address some of the feedback

@@
     \emph{Nothing will ever be attempted if all possible objections must first be overcome.} 
-    \begin{flushright}- Samuel Johnson\end{flushright}
+    \begin{flushright}-- Samuel Johnson\end{flushright}
 \end{quote}
@@
 This paper lays down rigorous type-theoretic foundations for doing just that.
-Local type inference helps make the localization decisions we make systematic and predictable, and type hole inference shows how local and constraint-based type inference might operate hand-in-hand, rather than as alternatives.
+Bidirectional type checking helps make the localization decisions we make systematic and predictable, and type hole inference shows how local and constraint-based type inference might operate hand-in-hand, rather than as alternatives.
 We hope that language designers will use the techniques introduced in this paper to consider more rigorously, perhaps even formally, the problems of type error localization and error recovery when designing future languages.
+
+% \section*{Acknowledgements}
+% (omitted for review)
+
+% Thank you for reading this paper!

### 2023-10-25 intro.tex
Fix more typos and address some of the feedback

@@
 supports local type inference, specified in the well-established bidirectional style \cite{Localinf,HazelnutPOPL,BidirTyping}:
-\vspace{-3px}
+\\\vspace{-5px}
 \begin{center}
@@
 \end{center}
-\vspace{-3px}
+
 A type checker with no support for error localization or recovery---%
@@
 A more practical approach, and one common even in production type checkers, 
-is to localize the first error that causes local type inference to fail and emit an explanatory error message before terminating.
+is to localize the first error that causes type-checking to fail and emit an explanatory error message before terminating.
 In this example, the system might report that the variable \li{f} located on Line 2 is free, then terminate.

### 2023-10-25 polymorphism.tex
Fix non-THI typos from Review C
Fix more typos and address some of the feedback

@@
   }{
-    \bothCtxSynFixedInto{\tvarCtx}{\ctx}{\ETypeAp{\EMV}{\TMV_2}}{\ECTypeAp{\ECMV}{\MTMV_2}}{\MTUnknown}
+    \bothCtxSynFixedInto{\tvarCtx}{\ctx}{\ETypeAp{\EMV}{\TMV_2}}{\ECTypeApSynNonMatchedForall{\ECMV}{\MTMV_2}}{\MTUnknown}
   }
@@
 no implicit argument can be resolved) or \citet{DBLP:phd/hal/LennonBertrand22}'s gradual bidirectional
-variant of the dependently typed calculus of constructions.
-
-% TODO cite gradual dependent types papers, etc.
-% DBLP:phd/hal/LennonBertrand22
+variant of the dependently typed calculus of constructions. Languages with operator overloading may
+provide interesting localization decisions. So that overloading can be resolved during typing and
+marking, the modes of one or both operand must be changed to synthesis. If the operator is not
+defined given the synthesized operand type(s), an error might be marked on the operator itself.

### 2023-10-25 related.tex
Fix more typos and address some of the feedback

@@
 
-Hazelnut is in turn rooted in gradual type theory \cite{Siek06a, siek2015refined}, and we make extensive use of (only) the static aspects of gradual typing, namely the universal consistency of the unknown type, to enable recovery from marked errors, which can leave missing type information.
+Hazelnut is in-turn rooted in gradual type theory \cite{Siek06a, siek2015refined}, and we make extensive use of (only) the static aspects of gradual typing, namely the universal consistency of the unknown type, to enable recovery from marked errors, which can leave missing type information.
 
@@
 Type error localization is a well-studied problem in practice.
-This paper is the first to formally support the intuition that, in the words of \cite{BidirTyping}, ``bidirectional typing improves error locality''.
-Although there has been considerable folklore around error localization
+This paper is the first to formally support the intuition that, in the words of \citet{BidirTyping}, ``bidirectional typing improves error locality.'' Although there has been considerable folklore around error localization
 for systems with local type inference, the problem has received little formal attention. We hope that this paper, with its rigorous formulation

### 2023-10-25 intro.tex
Remove most usages of "local type inference"

@@
 
-The user can interactively explore possible error localizations by hovering over a partially consistent suggestion, which (temporarily) fills the type hole and thus returns the localization decision to the local inference system.
+The user can interactively explore possible error localizations by hovering over a partially consistent suggestion, which (temporarily) fills the type hole and thus returns the localization decision to the bidirectional type system.
 Here, each choice causes a different set of two uses of \li{x} to be marked, e.g.
@@
 Until the choice is finalized, the type hole continues to operate gradually.
-\cref{sec:thi} describes this distinctively neutral approach to blending local and constraint-based type inference in Hazel, where local inference is used to systematically mark erroneous expressions in the program and constraint-based inference (using entirely standard algorithms, which we do not repeat in this paper) is used exclusively to locate unfillable holes.
+\cref{sec:thi} describes this distinctively neutral approach to blending local and constraint-based
+type inference in Hazel, where bidirectional typing is used to systematically mark erroneous expressions in the program and constraint-based inference (using entirely standard algorithms, which we do not repeat in this paper) is used exclusively to locate unfillable holes.

### 2023-10-25 related.tex
Remove most usages of "local type inference"

@@
 This creates a large amount of information for the programmer to consume.
-Our approach is to instead simply report the constraint inconsistencies on a hole in the program and allow for the programmer to interactively refine their intent, so only the local inference system is responsible for identifying particular erroneous expressions.
+Our approach is to instead simply report the constraint inconsistencies on a hole in the program and allow for the programmer to interactively refine their intent, so only the bidirectional type system is responsible for identifying particular erroneous expressions.
 We do not make particular usability claims about the interactive affordances related to type hole inference in this paper, but rather simply claim a novel neutral point in the overall design space that uniquely combines local and global approaches.
@@
 
-The underlying unification algorithm is essentially standard--the novelty is in how the unification results are used and how failures are handled, rather than in the inference itself.
-In particular, we base our approach on the system described by \cite{GradualInfer}, because it also identifies type inference variables with the unknown type from gradual typing and the union find data structure is useful for computing possible type sets.
+The underlying unification algorithm is essentially standard---the novelty is in how the unification results are used and how failures are handled, rather than in the inference itself.
+In particular, we base our approach on the system described by \citet{GradualInfer}, because it also identifies type inference variables with the unknown type from gradual typing and the union find data structure is useful for computing possible type sets.
 \citet{garcia:2015} similarly present a static implicitly typed language, where users opt into dynamism by annotating an expression with the gradual type "?", and an associated type inference algorithm and accompanying metatheory.

### 2023-10-26 calculus.tex
Move meet to premise in if rules
Updates from Overleaf

@@
 Furthermore, since error marks are effectively annotations on top of the program, marking should
-preserve syntactic structure modulo those marks.
-\cref{fig:calculus-mark-erasure} gives selected clauses of the definition of \emph{mark erasure}, which converts marked expressions back into unmarked ones by
+preserve syntactic structure modulo those marks. \cref{fig:calculus-mark-erasure} gives the definition of \emph{mark erasure}, which converts marked expressions back into unmarked ones by
 removing error marks. 
@@
   \begin{enumerate}
-    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then 
-      $\erasesTo{\ECMV}{\EMV}$.
-    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$, then
-      $\erasesTo{\ECMV}{\EMV}$.
+    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
+      then $\ctxSynTypeM{\ctx}{\ECMV}{\TMV}$
+        and $\erasesTo{\ECMV}{\EMV}$.
+    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
+      then $\ctxAnaTypeM{\ctx}{\ECMV}{\TMV}$
+        and $\erasesTo{\ECMV}{\EMV}$.
   \end{enumerate}
@@
   \judgment{
-    \ctxAnaTypeU{\ctx}{\EMV_1}{\TBool} \\\\
-    \ctxSynTypeU{\ctx}{\EMV_2}{\TMV_1} \\
-    \ctxSynTypeU{\ctx}{\EMV_3}{\TMV_2}
-  }{
-    \ctxSynTypeU{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TMeet{\TMV_1}{\TMV_2}}
+    \ctxAnaTypeU{\ctx}{\EMV_1}{\TBool} \\
+    \ctxSynTypeU{\ctx}{\EMV_2}{\TMV_1} \\\\
+    \ctxSynTypeU{\ctx}{\EMV_3}{\TMV_2} \\
+    \TMV_3 = \TMeet{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynTypeU{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\TMV_3}
   }{USIf}
@@
   \judgment{
-    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\\\
-    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\
-    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2}
-  }{
-    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMeet{\TMV_1}{\TMV_2}}
+    \ctxAnaFixedInto{\ctx}{\EMV_1}{\ECMV_1}{\TBool} \\
+    \ctxSynFixedInto{\ctx}{\EMV_2}{\ECMV_2}{\TMV_1} \\\\
+    \ctxSynFixedInto{\ctx}{\EMV_3}{\ECMV_3}{\TMV_2} \\
+    \TMV_3 = \TMeet{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynFixedInto{\ctx}{\EIf{\EMV_1}{\EMV_2}{\EMV_3}}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV_3}
   }{MKSIf}
@@
   \judgment{
-    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TBool} \\\\
-    \ctxSynTypeM{\ctx}{\ECMV_2}{\TMV_1} \\
-    \ctxSynTypeM{\ctx}{\ECMV_3}{\TMV_2}
-  }{
-    \ctxSynTypeM{\ctx}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMeet{\TMV_1}{\TMV_2}}
+    \ctxAnaTypeM{\ctx}{\ECMV_1}{\TBool} \\
+    \ctxSynTypeM{\ctx}{\ECMV_2}{\TMV_1} \\\\
+    \ctxSynTypeM{\ctx}{\ECMV_3}{\TMV_2} \\
+    \TMV_3 = \TMeet{\TMV_1}{\TMV_2}
+  }{
+    \ctxSynTypeM{\ctx}{\ECIf{\ECMV_1}{\ECMV_2}{\ECMV_3}}{\TMV_3}
   }{MSIf} \\
@@
 To conclude this section, we present two more metatheorems that help ensure correctness of the
-system. Though \cref{thm:calculus-marking-well-formedness} guarantees that marking does not change
+system. Although \cref{thm:calculus-marking-well-formedness} guarantees that marking does not change
 the syntactic structure of a program, it makes no statement about the presence of error marks in the
 resulting marked program. 
-\cref{thm:calculus-marking-well-ill-typed} established that well-typed expressions are left unmarked and ill-typed expressions have at least one mark.
+\cref{thm:calculus-marking-well-ill-typed} establishes that well-typed expressions are left unmarked and ill-typed expressions have at least one mark.
 
@@
     \item \begin{enumerate}
-        \item If $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$ and $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
+        \item If $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$
+            and $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
           then $\markless{\ECMV}$.
-        \item If $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$ and $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
+        \item If $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$
+            and $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV}$,
           then $\markless{\ECMV}$.
@@
     \item \begin{enumerate}
-        \item If there does not exist $\TMV$ such that $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$, then for
-          all $\ECMV$ and $\TMV'$ such that $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, it is not
-          the case that $\markless{\ECMV}$.
-        \item If there does not exist $\TMV$ such that $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$, then for
-          all $\ECMV$ and $\TMV'$ such that $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$, it is not
-          the case that $\markless{\ECMV}$.
+        \item If there does not exist $\TMV$
+            such that $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$,
+          then for all $\ECMV$ and $\TMV'$
+            such that $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$,
+            it is not the case that $\markless{\ECMV}$.
+        \item If there does not exist $\TMV$
+            such that $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$,
+          then for all $\ECMV$ and $\TMV'$
+            such that $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV}{\TMV'}$,
+            it is not the case that $\markless{\ECMV}$.
       \end{enumerate}
@@
   \begin{enumerate}
-    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV_1}$ and
-      $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV_2}$, then $\ECMV_1 = \ECMV_2$ and $\TMV_1 =
-      \TMV_2$.
-    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV}$ and
-      $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV}$, then $\ECMV_1 = \ECMV_2$.
+    \item If $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV_1}$
+        and $\ctxSynFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV_2}$,
+      then $\ECMV_1 = \ECMV_2$
+        and $\TMV_1 = \TMV_2$.
+    \item If $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_1}{\TMV}$
+        and $\ctxAnaFixedInto{\ctx}{\EMV}{\ECMV_2}{\TMV}$,
+      then $\ECMV_1 = \ECMV_2$.
   \end{enumerate}
@@
 
-We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of Hazel.
-% TODO citation and explain what in the world Hazel is (reviewer A stated that the lack of this made this section pretty much impossible to understand)
+We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of Hazel \cite{hazel},
+a live functional programming environment in which incomplete programs, i.e., programs that contain holes, may be type-checked, manipulated, and even executed.
 In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching, strings, and explicit polymorphism (all following the same recipe we developed above).
@@
 
-Earlier versions of Hazel achieved total syntax error recovery by using a term-based structure editor, which maintained delimiter matching by construction, albeit at the cost of some syntactic rigidity.
-The Hazelnut action calculus introduced by \citet{HazelnutPOPL} specifies such a structure editor formally.
-Notably, this calculus also supports a form of type error localization and recovery: edit actions can automatically insert non-empty holes around type inconsistencies that appear at the cursor after an edit action.
-This requires specifying a bidirectionally typed action calculus that propagates type and binding information to the cursor, given by the synthetic action judgment, written
-$\ASEAction{\ctx}{\EMV}{\TMV}{\EMV'}{\TMV'}{\AMV}$, and the analytic action judgment, written
-$\AAEAction{\ctx}{\EMV}{\EMV'}{\TMV}{\AMV}$.
-
-A significant issue with Hazelnut is that it does not allow (i.e.
-it leaves undefined) actions that require \emph{non-local} mark (i.e.
-non-empty hole, in Hazelnut's parlance) insertion or removal.
-For
-example, though the system can wrap an arbitrary expression in an arithmetic operation, it cannot
-wrap an arbitrary expression $\EMV$ in a lambda abstraction, i.e. construct
-$\ELam{x}{\TUnknown}{\EMV}$ directly from $\EMV$. The binding of $x$ with the unknown type may
-shadow a previous declaration and require the removal of error holes in body. The calculus includes manual non-empty hole insertion and removal actions as a workaround.
+Earlier versions of Hazel achieved total type error recovery by using a term-based structure editor, which maintained delimiter matching by construction, albeit at the cost of some syntactic rigidity.
+To specify such a structure editor formally, \citet{HazelnutPOPL} introduced the Hazelnut action calculus, which operates over a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{cursor}.
+The synthetic action judgment, written $\ASEAction{\ctx}{\ZMV}{\TMV}{\ZMV'}{\TMV'}{\AMV}$, performs some edit action $\AMV$ on some \emph{zippered expression} $\ZMV$ (which is an expression with a superimposed cursor, denoted $\hat{\EMV}$ in the original work) that synthesizes type $\TMV$, producing some new $\ZMV'$ that synthesizes types $\TMV'$.
+Analogously, the analytic action judgment is written $\AAEAction{\ctx}{\ZMV}{\ZMV'}{\TMV}{\AMV}$.
+Type and binding information is propagated to the location of the cursor during these operations, allowing them to insert the necessary error marks (i.e. non-empty holes, in Hazelnut's parlance).
+When, for example, wrapping an expression whose type is inconsistent with $\TNum$ in an arithmetic operation, the calculus inserts an error mark around that operand:
+%
+\begin{mathpar}
+    \inferrule{
+        \inconsistent{\TMV}{\TNum}
+    }{
+        \ASCon{\ctx}{\ZCursor{\EMV}}{\TMV}{\ZPlusL{\ECInconType{\ECMV}}{\ZCursor{\EEHole}}}{\TNum}{\SPlusL}
+    }
+\end{mathpar}
+
+However, a significant issue with Hazelnut is that it does not allow---and simply leaves undefined---actions that require \emph{non-local} mark insertions or removals. 
+For example, although the system can wrap an arbitrary expression $\EMV$ in an arithmetic operation, it cannot wrap it in a lambda abstraction, i.e.
+construct $\ELam{x}{\TUnknown}{\EMV}$ directly from $\EMV$.
+The binding of $x$ with the unknown type may shadow a previous binding and require the removal of error holes in the body.
+As a workaround, the calculus includes manual mark insertion and removal actions.
 
@@
 
-One approach is to define an untyped action calculus, \emph{wholly decoupling} edit actions from typing. Under this simplified design, the action calculus, given by a singular action
-judgment, written $\AUEAction{\EMV}{\EMV'}{\AMV}$, is concerned only with the manipulation of
-syntax, and the total marking procedure yields statically meaningful terms with error marks inserted at the appropriate positions as a kind of editor decoration. This solves the problem
-outlined above: since marking operates on the entire program after each action, arbitrary
-constructions are permitted.
+One approach is to define an untyped action calculus, wholly decoupling edit actions from typing.
+Under this simplified design, the action calculus, given by a singular action judgment, written $\AUEAction{\ZMV}{\ZMV'}{\AMV}$, is concerned only with the manipulation of syntax,
+and the total marking procedure developed above yields statically meaningful terms with error marks inserted at the appropriate positions as a kind of editor decoration.
+This solves the problem outlined above: since marking operates on the entire program after each action, arbitrary constructions are permitted.
 
 Alternatively, instead of a wholly untyped action calculus, one may directly integrate re-marking logic into the typed Hazelnut
-calculus, making use of the type and scoping information being propagated to re-mark only in scopes where it might be necessary. For example, we can wrap $\ECMV$ into
-$\ECLam{x}{\TUnknown}{\ECMV}$ by re-marking the body under the context with $x$ binding
-the unknown type:
-\[%
-  \judgment{
+calculus, making use of the type and scoping information being propagated to re-mark only in positions where it might be necessary, in a roughly incremental fashion.
+For example, we can wrap $\ECMV$ into $\ECLam{x}{\TUnknown}{\ECMV}$ by re-marking the body under the context with $x$ binding the unknown type:
+%
+\begin{mathpar}
+  \inferrule[ASEConLam]{
     \ctxSynFixedInto{\extendCtx{\ctx}{x}{\TUnknown}}{\erase{\ECMV}}{\ECMV'}{\TMV'}
   }{
-    \ASCon{\ctx}{\ZCursor{\ECMV}}{\TMV}{\ZLamT{x}{\ZTCursor{\TUnknown}}{\ECMV'}}{\TArrow{\TUnknown}{\TMV'}}{\SLam{x}}
-  }{ASEConLam}
-\]%
-The
-supplemental material contains a complete formal description of both variants, and the untyped
-action semantics and related metatheory are mechanized in Agda. Because of the simplicity of the
-untyped version, and because marking is defined separately, this Agda mechanization can serve
-universally as a baseline for correctness of approaches, like our initial design that aims to make
-use of type and scope information, and perhaps other analyses, to minimize re-marking. We leave a
-complete assessment of this and other re-marking optimization approaches to future work, particularly
-because Hazel is now moving toward the decoupled approach for future development.
+    \ASCon{\ctx}{\ZCCursor{\ECMV}}{\TMV}{\ZCLamT{x}{\ZTCursor{\TUnknown}}{\ECMV'}}{\TArrow{\TUnknown}{\TMV'}}{\SLam{x}}
+  }
+\end{mathpar}
+%
+In this formulation, the action calculus is bidirectional and operates on \emph{zippered marked expressions}, written $\ZCMV$, where the cursor is superimposed atop marked expressions.
+With an analogous definition of mark erasure on these expressions, the correctness of this typed action calculus may be defined in relation to the behavior of the untyped variant.
+
+The supplemental material contains a complete formal sketch of both variants, and the untyped calculus and its related metatheory are mechanized in Agda.
+Because of the untyped variant's simplicity, and because marking is defined separately, this mechanization can serve universally as a baseline for the correctness of approaches like the typed action calculus and perhaps other analyses that minimize re-marking.
+We leave a complete assessment of this and other re-marking optimization approaches to future work, particular because Hazel is now moving toward the decoupled approach for future development.

### 2023-10-26 let.tex
Fix more typos and address some of the feedback
Updates from Overleaf

@@
 with an inconsistent branches error mark following our development above. If instead it is in
-analytic position against the unknown type, no mark will be produced.
+analytic position against the unknown type, each branch is checked independently against this
+trivial constraint, and no mark will be produced.
 
@@
 
-This system is achieved by augmenting the type system with a variant of the unknown type, written
-$\TUnknownSwitch$, that triggers a ``switch'' to synthesis and otherwise behaves identically to
-$\TUnknown$. In the previous example, the unannotated pattern variable $\PVar{a}$ synthesizes
+This is achieved by adding a ``switch type'', denoted
+$\TUnknownSwitch$, existing entirely to trigger a ``switch'' to synthesis.
+$\TUnknownSwitch$ represents a trivial constraint, instantiated as a type for the sole purpose of using the existing machinery to propagate the appropriate mode switch to the relevant sub-expression of the definition.
+This switch type carries over identically to a non-gradual setting, but here we visually represent as a variant of $\TUnknown$ as it behaves identically with respect to type consistency and matched arrow/product judgements.
+In the previous example, the unannotated pattern variable $\PVar{a}$ synthesizes
 $\TUnknownSwitch$, and the annotated pattern $\PAsc{\PVar{b}}{\TUnknown}$ synthesizes $\TUnknown$.
@@
 in which case they produce an output context $\ctx'$, which extends $\ctx$ with bindings introduced
-by the pattern. Note that $\TUnknownSwitch$ exists entirely to ensure that sub-expressions of the
+by the pattern. Note again that $\TUnknownSwitch$ exists entirely to ensure that sub-expressions of the
 definition are assigned an appropriate typing mode---it is never added to the context, i.e. it
@@
     \ctxSynFixedInto{\ctx}{\ELet{\PMV}{\EMV_{1}}{\EMV_{2}}}{\ELet{\PCMV}{\ECMV_{1}}{\ECMV_{2}}}{\TMV_2}
-  }{ISLetPat}
+  }{MKSLetPat}
 \]%

### 2023-10-27 let.tex
Updates from Overleaf

@@
 $\TUnknownSwitch$ represents a trivial constraint, instantiated as a type for the sole purpose of using the existing machinery to propagate the appropriate mode switch to the relevant sub-expression of the definition.
-This switch type carries over identically to a non-gradual setting, but here we visually represent as a variant of $\TUnknown$ as it behaves identically with respect to type consistency and matched arrow/product judgements.
+This switch type carries over identically to a non-gradual setting, but here we represent it as a variant of $\TUnknown$ as it behaves identically with respect to consistency and matched arrow/product judgements.
+%Note that in a non-gradual setting, the matched product rule (or equivalent) would still be required to support the cases such as $\ELet{\PVar{a}}{\EPair{1}{2}}{a}$.
 In the previous example, the unannotated pattern variable $\PVar{a}$ synthesizes
@@
 $\ECAnaNonMatchedProd{\ECMV}$.
+\\\\
+Note that both let rules assume the let definition synthesizes.
+In our system this is always the case, but if we added (e.g.) unannotated lambdas, we might wish to relax this restriction.
+We can freely add variations of the rules where if the definition fails to synthesize, it is instead analyzed against the synthesized type of the pattern.
+To avoid an additional redundant analysis, we would further want to modify the pattern synthesis judgement to produce a context as well.

### 2023-10-27 related.tex
Updates from Overleaf

@@
 
-% Complex techniques, e.g. based on machine learning or manual weighting, have been proposed to heuristically localize these errors to expressions.
+Regarding error messages, there has been recent work on using subtyping constraints to model the flow of type information throughout the program to produce more detailed error messages when unification fails \cite{Bhanuka_Parreaux_Binder_Brachthäuser_2023}.
+Our focus in this paper has not been on the error messages displayed on screen, about which we make no specific claims.
+However, it may be beneficial in the future to indicate to the user where in the program type suggestions originated from by using type provenances.
+Documenting the flow of type information would require using asymmetric constraints rather than typical bidirectional equality constraints.
 
-Our focus in this paper has not been on the error messages displayed on screen, about which we make no specific claims.
-There has been considerable work on improving explanations for type errors, e.g.
+Work on the Helium Haskell compiler \cite{DBLP:conf/haskell/HeerenLI03} explores an approach to constraint-based type inference where different solvers can be selected based on the properties desired of provided feedback.
+Heeren et al.
+present two constraint solvers: a global, type graph based constraint solver, which provides detailed error messages and feedback, and a lightweight greedy constraint solver with high performance.
+When using the global constraint solver, various heuristics can be employed to tune the likelihood of different parts of programs being reported as the source of errors.
+Additionally, hints can be emitted if corrections are identified that fully resolve type inconsistencies, or if any common sources of errors are identified.
+
+In addition, there has been considerable work on improving explanations for type errors, e.g.
 by providing sample inputs (dynamic witnesses) that elicit runtime errors.

### 2023-10-27 calculus.tex
Clean up references.bib

@@
 All rules and theorems may be found organized by judgement form in the supplementary appendix for
-reference, alongside a complete mechanization in the Agda proof assistant \cite{norell:thesis}
+reference, alongside a complete mechanization in the Agda proof assistant \cite{norell2007}
 (discussed in \cref{sec:calculus-agda}).
@@
 
-The marked lambda calculus is based on the gradually typed lambda calculus \cite{Siek06a} extended
-with numbers, booleans, pairs, and empty expression holes \cite{HazelnutSNAPL}.
+The marked lambda calculus is based on the gradually typed lambda calculus \cite{siek2006} extended
+with numbers, booleans, pairs, and empty expression holes \cite{omar2017a}.
 Given in \cref{fig:calculus-syntax}, the syntax consists of two expression languages:
@@
 
-$\TUnknown$ is a type hole, which we identify with the unknown type from gradual type theory \cite{Siek06a}.
+$\TUnknown$ is a type hole, which we identify with the unknown type from gradual type theory \cite{siek2006}.
 Finally, to model the edit state of a program in development, $\EEHole$ denotes an \emph{empty expression hole},
 used to represent syntactically incomplete portions of the program à la Hazelnut
-\cite{HazelnutPOPL}. Note, however, that these empty expression holes are not at all semantically
+\cite{omar2017b}. Note, however, that these empty expression holes are not at all semantically
 critical to the calculus---they are included illustrate how the calculus fits into the larger
@@
 
-As our starting point, types classify unmarked expressions by a completely standard \emph{bidirectional type system} \cite{BidirTyping,Localinf}, which employs
+As our starting point, types classify unmarked expressions by a completely standard \emph{bidirectional type system} \cite{dunfield2019,pierce2000}, which employs
 two mutually defined judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$,
@@
 The synthesis rule is standard. Analysis employs the judgment $\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$,
-which establishes that $\TMV$ is a \emph{matched arrow type} \cite{cimini:2016}, i.e. it may be
+which establishes that $\TMV$ is a \emph{matched arrow type} \cite{cimini2016}, i.e. it may be
 considered an arrow type. Defined in \cref{fig:calculus-matched-arrow}, this notion is purely a
-technical mechanism to avoid duplication of rules related to arrow types \cite{siek2015refined}.
+technical mechanism to avoid duplication of rules related to arrow types \cite{siek2015}.
 
@@
 In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching, strings, and explicit polymorphism (all following the same recipe we developed above).
-Hazel is implemented in OCaml and compiles to Javascript via \li{js_of_ocaml} \cite{jsoo} for use in the browser.
+Hazel is implemented in OCaml and compiles to Javascript via \li{js_of_ocaml} \cite{vouillon2014} for use in the browser.
 The supplemental material contains a pre-built implementation.
 
-The result of this effort is the first full-scale typed functional language equipped with a language server that solves the semantic gap problem: all of Hazel's semantic services are available as long as the program sketch is syntactically well-formed, which include type hints \cite{potter2020hazel}, semantic navigation, semantic code completion \cite{potter2020hazel,blinn2022integrative}, contextualized documentation \cite{potter2022contextualized}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{HazelLive}, even semantic services that require run-time information, like testing.
+The result of this effort is the first full-scale typed functional language equipped with a language server that solves the semantic gap problem: all of Hazel's semantic services are available as long as the program sketch is syntactically well-formed, which include type hints \cite{potter2020}, semantic navigation, semantic code completion \cite{potter2020,blinn2022}, contextualized documentation \cite{potter2022}, and, because Hazel is able to evaluate programs with empty and non-empty holes due to prior work \cite{omar2019}, even semantic services that require run-time information, like testing.
 %
@@
 a structure editor that is able to additionally offer partial syntactic error recovery.
-In particular, the \li{tylr} editor uses a form of structured syntax error recovery that automatically insert empty holes to maintain syntactic well-formedness as long as matching delimiters are placed by the user \cite{moon2022tylr,DBLP:conf/vl/Moon023}.
+In particular, the \li{tylr} editor uses a form of structured syntax error recovery that automatically insert empty holes to maintain syntactic well-formedness as long as matching delimiters are placed by the user \cite{moon2022,moon2023}.
 Efforts outside of the scope of this paper are underway to define syntactic error recovery mechanisms that can handle unmatched delimiters as well, which would achieve total syntax error recovery.
@@
 Earlier versions of Hazel achieved total type error recovery by using a term-based structure editor, which maintained delimiter matching by construction, albeit at the cost of some syntactic rigidity.
-To specify such a structure editor formally, \citet{HazelnutPOPL} introduced the Hazelnut action calculus, which operates over a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{cursor}.
+To specify such a structure editor formally, \citet{omar2017b} introduced the Hazelnut action calculus, which operates over a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{cursor}.
 The synthetic action judgment, written $\ASEAction{\ctx}{\ZMV}{\TMV}{\ZMV'}{\TMV'}{\AMV}$, performs some edit action $\AMV$ on some \emph{zippered expression} $\ZMV$ (which is an expression with a superimposed cursor, denoted $\hat{\EMV}$ in the original work) that synthesizes type $\TMV$, producing some new $\ZMV'$ that synthesizes types $\TMV'$.

### 2023-10-27 intro.tex
Clean up references.bib

@@
 The problem is that when the program being edited is ill-typed, 
-these semantic services can become degraded or unavailable \cite{HazelnutSNAPL}. 
+these semantic services can become degraded or unavailable \cite{omar2017a}. 
 These gaps in service are not always transient. 
@@
 These service gaps are fundamentally rooted in a definitional gap: a type system defined in the conventional way, 
-e.g. in the tradition of the typed lambda calculus and its derivatives \cite{TaplBook},
+e.g. in the tradition of the typed lambda calculus and its derivatives \cite{pierce2002},
 assigns meaning only to well-typed programs. 
@@
 e.g. in compiler error messages 
-or directly in the editor via markings decorating localized errors. Developers are known to attend to reported error locations when debugging type errors \cite{DBLP:journals/jfp/JoostenBH93}.
+or directly in the editor via markings decorating localized errors. Developers are known to attend to reported error locations when debugging type errors \cite{joosten1993}.
 Many systems also attempt recovery in certain situations, discussed below. 
@@
 in part as folklore amongst language and tool implementors. 
-Different type checkers or language servers \cite{barros2022editing,bour2018merlin}, even for the same language, localize and recover from type errors in different ways, 
+Different type checkers or language servers \cite{barros2022,bour2018}, even for the same language, localize and recover from type errors in different ways, 
 with little in the way of unifying theory of the sort that grounds the design of modern type systems themselves.
@@
 Consider, for example, the ill-typed program below, which is shown as presented to the user in
-this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{elm}. Hazel  
-supports local type inference, specified in the well-established bidirectional style \cite{Localinf,HazelnutPOPL,BidirTyping}:
+this paper's version of Hazel \cite{hazel}, a typed functional dialect of Elm \cite{czaplicki2013}. Hazel  
+supports local type inference, specified in the well-established bidirectional style \cite{pierce2000,omar2017b,dunfield2019}:
 \\\vspace{-5px}
@@
 
-An implementation with support for type error recovery, like Elm's compiler or OCaml's \li{merlin} \cite{bour2018merlin}, would 
+An implementation with support for type error recovery, like Elm's compiler or OCaml's \li{merlin} \cite{bour2018}, would 
  be tasked with continuing past this first error.
@@
 Heuristics have been developed to help make this choice less arbitrary, e.g. 
-by observing recent editing history \cite{steady-typing}, 
-or training a machine learning model \cite{SeidelBlame}. 
+by observing recent editing history \cite{chuchem2019}, 
+or training a machine learning model \cite{seidel2017}. 
 A less \emph{ad hoc} approach, which Hazel takes above, is to localize the inconsistency to the conditional as a whole, reporting that the problem is that the branch types differ.
-When the cursor is on this conditional expression, the Hazel type inspector---a status bar that reports typing information, including error messages, about the term at the cursor \cite{potter2020hazel}---displays:
+When the cursor is on this conditional expression, the Hazel type inspector---a status bar that reports typing information, including error messages, about the term at the cursor \cite{potter2020}---displays:
 
@@
 There is no definitive type or binding error at either use of \li{x}, 
-so we do not report an error in Hazel. More specifically, we treat \li{x}'s type as the unknown a.k.a. dynamic type from gradual typing \cite{Siek06a}.
+so we do not report an error in Hazel. More specifically, we treat \li{x}'s type as the unknown a.k.a. dynamic type from gradual typing \cite{siek2006}.
 In any of these cases, we would like to be able to recover and 
@@
 We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
-type annotations to expressions \cite{Localinf,BidirTyping}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{Siek06a,siek2015refined}, and 
-(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{HazelnutPOPL}, operationalizing the ``red lines'' that an editor displays.
-The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solar2013program}) can be marked, i.e.
+type annotations to expressions \cite{pierce2000,dunfield2019}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{siek2006,siek2015}, and 
+(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{omar2017b}, operationalizing the ``red lines'' that an editor displays.
+The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solarlezama2013}) can be marked, i.e.
 its errors can be simultaneously localized, such that the resulting program has a well-defined type and binding structure.
@@
 \Cref{sec:calculus-hazel} describes our own effort in this direction, which is to scale up the marked lambda calculus as the new basis for Hazel, a typed functional programming environment that fully solves the semantic gap problem: Hazel's semantic services are available as long as the program sketch is syntactically well-formed.
-Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as in Agda \cite{norell:thesis}, GHC Haskell \cite{haskell-holes}, Idris \cite{DBLP:journals/jfp/Brady13}, and others), syntax error recovery \cite{medeiros20,sorkin11}, and structure editing \cite{HazelnutPOPL,teitelbaum_cornell_1981}.
-Hazel has both a textual syntax and a structure editor \cite{HazelnutPOPL,DBLP:conf/vl/Moon023}. 
+Prior work has separately considered mechanisms for maintaining syntactically well-formed program sketches during development: manual hole insertion (as in Agda \cite{norell2007}, GHC Haskell \cite{haskell-holes}, Idris \cite{brady2013}, and others), syntax error recovery \cite{medeiros2020,sorkin2011}, and structure editing \cite{omar2017b,teitelbaum1981}.
+Hazel has both a textual syntax and a structure editor \cite{omar2017b,moon2023}. 
 %offering (1) a textual syntax, which allows the programmer to manually insert empty holes as necessary, (2) a text-like structure editor that offers partial syntax error recovery (inserting holes to maintain well-formedness as long as all matching delimiters have been placed) \cite{tylr}, and (3) a term-based structure editor with total syntax error recovery, i.e.
@@
 
-The developments in Section~\ref{sec:calculus} support the intuition (remarked upon by \citet{Localinf} and \citet{BidirTyping}) that bidirectional typing pairs well with type error localization and recovery because information flows systematically and locally through the tree.
+The developments in Section~\ref{sec:calculus} support the intuition (remarked upon by \citet{pierce2000} and \citet{dunfield2019}) that bidirectional typing pairs well with type error localization and recovery because information flows systematically and locally through the tree.
 However, many functional languages including Elm, OCaml and Haskell feature constraint-based type inference, where constraints are gathered globally.
-For type systems where inference is decidable, this is powerful, but it is also notorious for complicating error localization, because inconsistencies can arise through a confluence of constraints originating from any number of locations \cite{DBLP:conf/popl/Wand86}.
+For type systems where inference is decidable, this is powerful, but it is also notorious for complicating error localization, because inconsistencies can arise through a confluence of constraints originating from any number of locations \cite{wand1986}.
 
@@
 Rather than arbitrarily privileging one of these uses, as is the case in languages like OCaml (and which has led to the development of complex heuristics, e.g.
-using machine learning \cite{SeidelBlame}, to avoid misleading programmers), this error is localized, with a \li{!}, to the type hole itself, with the error message providing partially consistent solutions as shown.
+using machine learning \cite{seidel2017}, to avoid misleading programmers), this error is localized, with a \li{!}, to the type hole itself, with the error message providing partially consistent solutions as shown.
 \\

### 2023-10-27 polymorphism.tex
Updates from Overleaf
CCS, type classes
Clean up references.bib

@@
 We leave as future work the task of defining marked versions of even more elaborate bidirectional
-type systems, e.g. \citet{dunfield2013complete}'s rather substantial formulation of implicit type
+type systems, e.g. \citet{dunfield2013}'s rather substantial formulation of implicit type
 application in a bidirectional setting (which would have to handle, for example, the situation where
-no implicit argument can be resolved) or \citet{DBLP:phd/hal/LennonBertrand22}'s gradual bidirectional
-variant of the dependently typed calculus of constructions. Languages with operator overloading may
-provide interesting localization decisions. So that overloading can be resolved during typing and
-marking, the modes of one or both operand must be changed to synthesis. If the operator is not
-defined given the synthesized operand type(s), an error might be marked on the operator itself.
+no implicit argument can be resolved) or \citet{lennonbertrand2022}'s gradual bidirectional
+variant of the dependently typed calculus of constructions.
+Languages with operator overloading may provide interesting localization decisions.
+So that overloading can be resolved during typing and marking, the modes of one or both operand must be changed to synthesis.
+If the operator is not defined given the synthesized operand type(s), an error might be marked on the operator itself.
+Type classes present a symmetric alternative where the operands may stay in analytic mode.
+With either approach, the necessary error marks and localization decisions may be arrived at by considering the possible failure cases.

### 2023-10-27 related.tex
Clean up references.bib

@@
 
-The contributions of this paper build directly on the Hazelnut type system \cite{HazelnutPOPL}, which is discussed extensively throughout.
+The contributions of this paper build directly on the Hazelnut type system \cite{omar2017b}, which is discussed extensively throughout.
 Non-empty holes in Hazelnut generalize to marks in this work.
@@
 
-Hazelnut is in-turn rooted in gradual type theory \cite{Siek06a, siek2015refined}, and we make extensive use of (only) the static aspects of gradual typing, namely the universal consistency of the unknown type, to enable recovery from marked errors, which can leave missing type information.
+Hazelnut is in-turn rooted in gradual type theory \cite{siek2006, siek2015}, and we make extensive use of (only) the static aspects of gradual typing, namely the universal consistency of the unknown type, to enable recovery from marked errors, which can leave missing type information.
 
-Our focus was exclusively on static typing in this paper, and the results are relevant to the design of language servers for any statically typed language, but it is worth noting that the results in this paper, taken together with Hazel's support for maintaining syntactic well-formedness using structure editing \cite{DBLP:conf/vl/Moon023,moon2022tylr} and for running programs with holes and marked errors \cite{HazelLive}, allow our implementation of Hazel to achieve \emph{total liveness}: every editor state is syntactically, statically, and dynamically meaningful, without gaps.
+Our focus was exclusively on static typing in this paper, and the results are relevant to the design of language servers for any statically typed language, but it is worth noting that the results in this paper, taken together with Hazel's support for maintaining syntactic well-formedness using structure editing \cite{moon2023,moon2022} and for running programs with holes and marked errors \cite{omar2019}, allow our implementation of Hazel to achieve \emph{total liveness}: every editor state is syntactically, statically, and dynamically meaningful, without gaps.
 
 Type error localization is a well-studied problem in practice.
-This paper is the first to formally support the intuition that, in the words of \citet{BidirTyping}, ``bidirectional typing improves error locality.'' Although there has been considerable folklore around error localization
+This paper is the first to formally support the intuition that, in the words of \citet{dunfield2019}, ``bidirectional typing improves error locality.'' Although there has been considerable folklore around error localization
 for systems with local type inference, the problem has received little formal attention. We hope that this paper, with its rigorous formulation
@@
 
-For systems rooted in constraint solving, there has been considerable work in improving error localization because it is notoriously well-understood that such systems make error localization difficult, and programmers are often confused by localization decisions \cite{DBLP:conf/popl/Wand86} because they are rooted in \emph{ad hoc} traversal orders \cite{mcadam1998unification,DBLP:journals/toplas/LeeY98}.
-More recently, there has been a series of papers on finding the most likely location for an error based on either manual weights \cite{DBLP:conf/popl/ZhangM14,DBLP:conf/oopsla/PavlinovicKW14} or learned weights \cite{SeidelBlame}.
+For systems rooted in constraint solving, there has been considerable work in improving error localization because it is notoriously well-understood that such systems make error localization difficult, and programmers are often confused by localization decisions \cite{wand1986} because they are rooted in \emph{ad hoc} traversal orders \cite{mcadam1999,lee1998}.
+More recently, there has been a series of papers on finding the most likely location for an error based on either manual weights \cite{zhang2014,pavlinovic2014} or learned weights \cite{seidel2017}.
 While improving the situation somewhat, these remain fundamentally \emph{ad hoc} in their need to guess intent.
@@
 
-A more neutral alternative is to derive a set of terms that contribute to an error, an approach known as type error slicing \cite{DBLP:conf/esop/HaackW03,DBLP:journals/tosem/TipD01,DBLP:conf/sfp/Schilling11}.
+A more neutral alternative is to derive a set of terms that contribute to an error, an approach known as type error slicing \cite{haack2003,tip2001,schilling2011}.
 This creates a large amount of information for the programmer to consume.
@@
 
-Recent work on gradual liquid type inference described an exploratory interface for filling holes in refinement types by selecting from partial solutions to conflicting refinement type constraints \cite{DBLP:journals/pacmpl/VazouTH18}.
+Recent work on gradual liquid type inference described an exploratory interface for filling holes in refinement types by selecting from partial solutions to conflicting refinement type constraints \cite{vazou2018}.
 This is similar in spirit to type hole inference as described in this paper, albeit targeting program verification predicates.
@@
 The underlying unification algorithm is essentially standard---the novelty is in how the unification results are used and how failures are handled, rather than in the inference itself.
-In particular, we base our approach on the system described by \citet{GradualInfer}, because it also identifies type inference variables with the unknown type from gradual typing and the union find data structure is useful for computing possible type sets.
-\citet{garcia:2015} similarly present a static implicitly typed language, where users opt into dynamism by annotating an expression with the gradual type "?", and an associated type inference algorithm and accompanying metatheory.
+In particular, we base our approach on the system described by \citet{siek2008}, because it also identifies type inference variables with the unknown type from gradual typing and the union find data structure is useful for computing possible type sets.
+\citet{garcia2015} similarly present a static implicitly typed language, where users opt into dynamism by annotating an expression with the gradual type "?", and an associated type inference algorithm and accompanying metatheory.
 By contrast, the Hazelnut type system assigns gradual types to programs that would ordinarily not type-check in a non-gradual system by wrapping them in expression holes.
-The type inference algorithm presented in \citet{garcia:2015} also does not specify what to do if the constraint set cannot be solved.
+The type inference algorithm presented in \citet{garcia2015} also does not specify what to do if the constraint set cannot be solved.
 If a single static type cannot be determined for an expression, its type is simply undefined, whereas our approach provides a list of suggestions derived from any conflicting constraints if a single substitution cannot be determined.
@@
 
-Of note, however, is that type inference approaches that eagerly solve and substitute for type inference variables \cite{DBLP:journals/tapos/OderskySW99,DBLP:conf/icfp/Pottier14,mcadam1998unification} are not well-suited to the type hole inference approach, as they lose information necessary for computing partially consistent suggestions.
+Of note, however, is that type inference approaches that eagerly solve and substitute for type inference variables \cite{odersky1999,pottier2014,mcadam1999} are not well-suited to the type hole inference approach, as they lose information necessary for computing partially consistent suggestions.
 
@@
 
-Regarding error messages, there has been recent work on using subtyping constraints to model the flow of type information throughout the program to produce more detailed error messages when unification fails \cite{Bhanuka_Parreaux_Binder_Brachthäuser_2023}.
+Regarding error messages, there has been recent work on using subtyping constraints to model the flow of type information throughout the program to produce more detailed error messages when unification fails \cite{bhanuka2023}.
 Our focus in this paper has not been on the error messages displayed on screen, about which we make no specific claims.
@@
 
-Work on the Helium Haskell compiler \cite{DBLP:conf/haskell/HeerenLI03} explores an approach to constraint-based type inference where different solvers can be selected based on the properties desired of provided feedback.
+Work on the Helium Haskell compiler \cite{heeren2003} explores an approach to constraint-based type inference where different solvers can be selected based on the properties desired of provided feedback.
 Heeren et al.
@@
 by providing sample inputs (dynamic witnesses) that elicit runtime errors.
-With this approach, one can generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with around 70\% accuracy \cite{Seidel2016}.
+With this approach, one can generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with around 70\% accuracy \cite{seidel2016}.
 These and a variety of other approaches to explaining errors are a fruitful avenue for future integration into the system.

### 2023-10-27 calculus.tex
Updates from Overleaf

@@
   \item The \emph{marked language}, which mirrors the structure of the unmarked language but is
-    extended with explicit \emph{error marks} which, intuitively, may be viewed as exactly the
-    errors marked in an editor. We call expressions of this language \emph{marked expressions},
+    extended with \emph{error marks}. We call expressions of this language \emph{marked expressions},
     denoted $\ECMV$.
 \end{itemize}
+%
 In this simple setting, we only need one sort for types, $\tau$. 
@@
 
-Before giving any of the rules, let us summarize the overall judgemental structure of the calculus.
-Note that colors in the judgement forms below are entirely redundant reading aids---color has no semantic significance.
-
-As our starting point, types classify unmarked expressions by a completely standard \emph{bidirectional type system} \cite{dunfield2019,pierce2000}, which employs
-two mutually defined judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$,
-establishes that, under the typing context $\ctx$, the expression $e$ synthesizes or locally infers the type $\TMV$. \emph{Type analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$,
-states that the expression $\EMV$ may appear where an expression of type $\TMV$ is expected.
-
 The key operation is \emph{marking}, which transforms an unmarked expression into a marked expression,
 inserting error marks where appropriate.
-%
 This corresponds to a type checking
@@
 Note that we do not specify what the error message should say in this paper.
+
 Throughout the remainder of this section, as we formulate marking for the GTLC, we will motivate and
 give precise semantics for each error mark. Furthermore, when the core is extended, new marks may be needed, as we discuss below.
+As we shall see, the rules for marking may be systematically derived by considering the error cases, yielding a recipe for developing error localization and recovery semantics for gradual, bidirectional type systems in general.
+
+Before giving any of the rules, let us summarize the overall judgemental structure of the calculus.
+Note that colors in the judgement forms below are entirely redundant reading aids---color has no semantic significance.
+
+As our starting point, types classify unmarked expressions by a completely standard \emph{bidirectional type system} \cite{dunfield2019,pierce2000}, which employs
+two mutually defined judgments. \emph{Type synthesis}, written $\ctxSynTypeU{\ctx}{\EMV}{\TMV}$,
+establishes that, under the typing context $\ctx$, the expression $e$ synthesizes or locally infers the type $\TMV$. \emph{Type analysis}, written $\ctxAnaTypeU{\ctx}{\EMV}{\TMV}$,
+states that the expression $\EMV$ may appear where an expression of type $\TMV$ is expected.
 
@@
 Furthermore, since error marks are effectively annotations on top of the program, marking should
-preserve syntactic structure modulo those marks. \cref{fig:calculus-mark-erasure} gives the definition of \emph{mark erasure}, which converts marked expressions back into unmarked ones by
+preserve syntactic structure modulo those marks. \cref{fig:calculus-mark-erasure} gives part of the definition of \emph{mark erasure}, which converts marked expressions back into unmarked ones by
 removing error marks. 
@@
 The synthetic action judgment, written $\ASEAction{\ctx}{\ZMV}{\TMV}{\ZMV'}{\TMV'}{\AMV}$, performs some edit action $\AMV$ on some \emph{zippered expression} $\ZMV$ (which is an expression with a superimposed cursor, denoted $\hat{\EMV}$ in the original work) that synthesizes type $\TMV$, producing some new $\ZMV'$ that synthesizes types $\TMV'$.
-Analogously, the analytic action judgment is written $\AAEAction{\ctx}{\ZMV}{\ZMV'}{\TMV}{\AMV}$.
 Type and binding information is propagated to the location of the cursor during these operations, allowing them to insert the necessary error marks (i.e. non-empty holes, in Hazelnut's parlance).

### 2023-10-27 let.tex
Updates from Overleaf

--- 2023-10-26 let.tex
@@
 impose typing constraints---the trivial constraint, if they wish---on the definition.
-
 As illustration, consider the following program:
@@
 
-This is achieved by adding a ``switch type'', denoted
-$\TUnknownSwitch$, existing entirely to trigger a ``switch'' to synthesis.
+This is achieved by adding a ``switch type'', denoted $\TUnknownSwitch$, which exists entirely to trigger a ``switch'' to synthesis.
 $\TUnknownSwitch$ represents a trivial constraint, instantiated as a type for the sole purpose of using the existing machinery to propagate the appropriate mode switch to the relevant sub-expression of the definition.
-This switch type carries over identically to a non-gradual setting, but here we visually represent as a variant of $\TUnknown$ as it behaves identically with respect to type consistency and matched arrow/product judgements.
-In the previous example, the unannotated pattern variable $\PVar{a}$ synthesizes
-$\TUnknownSwitch$, and the annotated pattern $\PAsc{\PVar{b}}{\TUnknown}$ synthesizes $\TUnknown$.
-We write $\ctxSynPatU{\ctx}{\PMV}{\TMV}$ to say that the pattern $\PMV$ synthesizes type $\TMV$.
+This is expressed by the following rule:
 %
@@
   }{UASynSwitch}
+\end{mathpar}
+%
+This switch type carries over identically to a non-gradual setting, but here we represent it as a variant of $\TUnknown$ as it behaves identically with respect to consistency and our notions of matched arrow/product.
+% Note that in a non-gradual setting, the matched product rule (or equivalent) would still be required to support the cases such as $\ELet{\PVar{a}}{\EPair{1}{2}}{a}$.
 
+In the previous example, the unannotated pattern variable $\PVar{a}$ synthesizes
+$\TUnknownSwitch$, and the annotated pattern $\PAsc{\PVar{b}}{\TUnknown}$ synthesizes $\TUnknown$.
+We write $\ctxSynPatU{\ctx}{\PMV}{\TMV}$ to say that the pattern $\PMV$ synthesizes type $\TMV$.
+%
+\begin{mathpar}
   \judgment{ }{
@@
 Given by the judgment $\ctxAnaPatU{\ctx}{\PMV}{\TMV}{\ctx'}$, patterns are also typed analytically,
-in which case they produce an output context $\ctx'$, which extends $\ctx$ with bindings introduced
+in which case they produce an output context $\ctx'$ that extends $\ctx$ with bindings introduced
 by the pattern. Note again that $\TUnknownSwitch$ exists entirely to ensure that sub-expressions of the
-definition are assigned an appropriate typing mode---it is never added to the context, i.e. it
+definition are assigned an appropriate typing mode---it is never added to the context, i.e., it
 cannot escape into the rest of the program and cause body expressions to synthesize accidentally.
@@
   \inferrule[UAPPair]{
-    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
+    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\\\
     \ctxAnaPatU{\ctx}{\PMV_1}{\TMV_1}{\ctx_1} \\
@@
   \judgment{
-    \ctxAnaPatU{\ctx}{\PMV}{\TMV'}{\ctx'} \\
+    \ctxAnaPatU{\ctx}{\PMV}{\TMV'}{\ctx'} \\\\
     \consistent{\TMV}{\TMV'}
@@
 
-Synthesis for let expressions is straightforward: we synthesize the definition's type, then analyze the pattern against it:
-
+Synthesis for let expressions is straightforward: we synthesize the type of the definition and analyze the pattern against it. We then synthesize the body under the new context:
 \[%
@@
 \]%
+The analytic rule is similar, with the final premise and conclusion changed to analysis.
 
-The analytic rule is similar, with the final
-premise and conclusion changed to analysis.
-
-Attempting an analogous marking rule runs into trouble. We want any type annotations in the pattern
-to be canonical, which means analyzing the definition to attribute to it any inconsistencies. But we
-still must analyze the pattern, incorporating the definition's type to produce the context needed by
-the body. So we do a ``round trip'': analyze the definition against the pattern's type, and then
-the pattern against the definition's. Since the first analysis establishes consistency, this second
-analysis is guaranteed to succeed; we only care about the context produced:
-
+Unfortunately, attempting an analogous marking rule runs into some trouble.
+We want any type annotations in the pattern to be canonical, which means analyzing the definition to attribute to it any inconsistencies.
+But we still must analyze the pattern, incorporating the definition's type to produce the context needed by the body.
+So we do a ``round trip'': analyze the definition against the pattern's type, and then the pattern against the definition's.
+Since the first analysis establishes consistency, this second analysis is guaranteed to succeed; we only care about the context produced:
 \[%
@@
 $\ECAnaNonMatchedProd{\ECMV}$.
+
+Note that both \textsc{USLetPat} and \textsc{MKSLetPat} assume that the definition synthesizes.
+In our system this is always the case, but if we added, for example, unannotated lambda abstractions, we might wish to relax this restriction.
+We can freely add variations of the rules where if the definition fails to synthesize, it is instead analyzed against the synthesized type of the pattern.
+To avoid an additional redundant analysis, we would further want to modify the pattern synthesis judgement to produce a context as well.

### 2023-10-27 polymorphism.tex
Updates from Overleaf

@@
 If the operator is not defined given the synthesized operand type(s), an error might be marked on the operator itself.
-Type classes present a symmetric alternative where the operands may stay in analytic mode.
-With either approach, the necessary error marks and localization decisions may be arrived at by considering the possible failure cases.
+Type classes \cite{wadler1989} present a symmetric alternative where the operands may stay in analytic mode.
+In either approach, the necessary error marks and localization decisions may be arrived at via the recipe we have described.

### 2023-10-27 related.tex
Updates from Overleaf

@@
 
-Hazelnut is in-turn rooted in gradual type theory \cite{siek2006, siek2015}, and we make extensive use of (only) the static aspects of gradual typing, namely the universal consistency of the unknown type, to enable recovery from marked errors, which can leave missing type information.
+Hazelnut is in-turn rooted in gradual type theory \cite{siek2006, siek2015}.
+We make extensive use of (only) the static aspects of gradual typing---namely, the universal consistency of the unknown type---to enable recovery from marked errors, which can leave missing type information.
 
@@
 
-Of note, however, is that type inference approaches that eagerly solve and substitute for type inference variables \cite{odersky1999,pottier2014,mcadam1999} are not well-suited to the type hole inference approach, as they lose information necessary for computing partially consistent suggestions.
+Of note, however, is that approaches that eagerly solve and substitute for type inference variables \cite{odersky1999,pottier2014,mcadam1999} are not well-suited to the type hole inference approach, as they lose information necessary for computing partially consistent suggestions.
 
@@
 
-Regarding error messages, there has been recent work on using subtyping constraints to model the flow of type information throughout the program to produce more detailed error messages when unification fails \cite{bhanuka2023}.
-Our focus in this paper has not been on the error messages displayed on screen, about which we make no specific claims.
-However, it may be beneficial in the future to indicate to the user where in the program type suggestions originated from by using type provenances.
-Documenting the flow of type information would require using asymmetric constraints rather than typical bidirectional equality constraints.
-
-Work on the Helium Haskell compiler \cite{heeren2003} explores an approach to constraint-based type inference where different solvers can be selected based on the properties desired of provided feedback.
-Heeren et al.
-present two constraint solvers: a global, type graph based constraint solver, which provides detailed error messages and feedback, and a lightweight greedy constraint solver with high performance.
-When using the global constraint solver, various heuristics can be employed to tune the likelihood of different parts of programs being reported as the source of errors.
-Additionally, hints can be emitted if corrections are identified that fully resolve type inconsistencies, or if any common sources of errors are identified.
+In the realm of error messages for constraint-based inference, work on the Helium Haskell compiler \cite{heeren2003} offers two constraint solvers based on the desired feedback: a global, type graph-based constraint solver which provides detailed error messages, and a lightweight and high-performance greedy solver.
+When using the global constraint solver, various heuristics can tune the likelihood of different parts of programs reported as error sources.
+Recently, \citet{bhanuka2023} model the flow of type information throughout the program using subtyping constraints to produce detailed error messages when unification fails.
+Our focus in this paper has not be on the error messages displayed on-screen, about which we make no specific claims.
+It may, however, be beneficial to indicate to the user from where in the program type suggestions originated via type provenances.
+Documenting the flow of type information would require using assymetric constraints rather than typical bidirectional equality constraints.
 
 In addition, there has been considerable work on improving explanations for type errors, e.g.
-by providing sample inputs (dynamic witnesses) that elicit runtime errors.
-With this approach, one can generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with around 70\% accuracy \cite{seidel2016}.
+by providing sample inputs (dynamic witnesses) that elicit runtime errors \cite{seidel2016}.
+With this approach, one can generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with around 70\% accuracy.
 These and a variety of other approaches to explaining errors are a fruitful avenue for future integration into the system.

### 2023-10-27 abstract.tex
Some final tweaks and changelist

--- 2023-07-11 abstract.tex
@@
 The marked lambda calculus is bidirectionally typed, so localization decisions are systematically predictable based on a local flow of typing information. 
-Constraint-based type inference can bring more distant information to bear in discovering inconsistencies but this notoriously complicates error localization.
-We approach this problem by deploying constraint solving as a type-hole-filling layer atop this gradual bidirectionally typed core.
-Errors arising from inconsistent unification constraints are localized exclusively to type and expression holes, i.e.
-the system identifies unfillable holes using a sytem of traced  provenances, rather than localized in an \emph{ad hoc} manner to particular expressions.
+Constraint-based type inference can bring more distant information to bear in discovering
+inconsistencies but this notoriously complicates error localization. We approach this problem by
+deploying constraint solving as a type-hole-filling layer atop this gradual bidirectionally typed
+core. Errors arising from inconsistent unification constraints are localized exclusively to type and
+expression holes, i.e., the system identifies unfillable holes using a system of traced  provenances, rather than localized in an \emph{ad hoc} manner to particular expressions.
 The user can then interactively shift these errors to particular downstream expressions by selecting from suggested partially consistent type hole fillings, which returns control back to the bidirectional system.

### 2023-10-27 calculus.tex
Some final tweaks and changelist

@@
 But what happens when the synthesized type of an expression is \emph{not} consistent with the
-expected type, i.e. that the premise $\consistent{\TMV}{\TMV'}$ fails? Recalling the example of
+expected type, i.e., that the premise $\consistent{\TMV}{\TMV'}$ fails? Recalling the example of
 \cref{fig:calculus-examples-inconsistent-types}, subsumption would be used to analyze $\ETrue$
-against $\TNum$ when checking $\EPlus{\ETrue}{1}$, which fails since $\inconsistent{\TNum}{\TBool}$, i.e.
-\textsc{UASubsume} does not apply. At this point, traditional typing semantics would simply fail the type
+against $\TNum$ when checking $\EPlus{\ETrue}{1}$, which fails since
+$\inconsistent{\TNum}{\TBool}$---\textsc{UASubsume} does not apply. At this point, traditional typing semantics would simply fail the type
 checking process. \textsc{MKASubsume} would also not be applicable, leaving marking
@@
 The synthesis rule is standard. Analysis employs the judgment $\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$,
-which establishes that $\TMV$ is a \emph{matched arrow type} \cite{cimini2016}, i.e. it may be
+which establishes that $\TMV$ is a \emph{matched arrow type} \cite{cimini2016}, i.e., it may be
 considered an arrow type. Defined in \cref{fig:calculus-matched-arrow}, this notion is purely a
@@
 Another error arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
-$\inconsistent{\TMV}{\TMV_1}$, i.e. the actual type annotation for $x$ is inconsistent with the
+$\inconsistent{\TMV}{\TMV_1}$, i.e., the actual type annotation for $x$ is inconsistent with the
 expected input type. The \emph{inconsistent ascription mark} $\ECLamInconAsc{x}{\TMV}{\ECMV}$ indicates exactly this
@@
 To specify such a structure editor formally, \citet{omar2017b} introduced the Hazelnut action calculus, which operates over a small bidirectionally typed lambda calculus extended with \emph{holes} and a \emph{cursor}.
-The synthetic action judgment, written $\ASEAction{\ctx}{\ZMV}{\TMV}{\ZMV'}{\TMV'}{\AMV}$, performs some edit action $\AMV$ on some \emph{zippered expression} $\ZMV$ (which is an expression with a superimposed cursor, denoted $\hat{\EMV}$ in the original work) that synthesizes type $\TMV$, producing some new $\ZMV'$ that synthesizes types $\TMV'$.
+The synthetic action judgment, written $\ASEAction{\ctx}{\ZMV}{\TMV}{\ZMV'}{\TMV'}{\AMV}$, performs an edit action $\AMV$ on some \emph{zippered expression} $\ZMV$ (which is an expression with a superimposed cursor, denoted $\hat{\EMV}$ in the original work) that synthesizes type $\TMV$, producing some new $\ZMV'$ that synthesizes types $\TMV'$.
 Type and binding information is propagated to the location of the cursor during these operations, allowing them to insert the necessary error marks (i.e. non-empty holes, in Hazelnut's parlance).

### 2023-10-27 intro.tex
Some final tweaks and changelist

@@
 These service gaps are fundamentally rooted in a definitional gap: a type system defined in the conventional way, 
-e.g. in the tradition of the typed lambda calculus and its derivatives \cite{pierce2002},
+e.g., in the tradition of the typed lambda calculus and its derivatives \cite{pierce2002},
 assigns meaning only to well-typed programs. 
@@
 Essentially all widely-used programming systems have some support for type error localization, 
-e.g. in compiler error messages 
+e.g., in compiler error messages 
 or directly in the editor via markings decorating localized errors. Developers are known to attend to reported error locations when debugging type errors \cite{joosten1993}.
@@
 There are several ways to localize this error. 
-One approach, taken e.g. by Elm and \li{merlin}, would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
-Heuristics have been developed to help make this choice less arbitrary, e.g. 
+One approach, taken e.g., by Elm and \li{merlin}, would be to assume, arbitrarily, that one branch is correct, localizing the error to the other branch. 
+Heuristics have been developed to help make this choice less arbitrary, e.g., 
 by observing recent editing history \cite{chuchem2019}, 
@@
 (3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{omar2017b}, operationalizing the ``red lines'' that an editor displays.
-The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solarlezama2013}) can be marked, i.e.
-its errors can be simultaneously localized, such that the resulting program has a well-defined type and binding structure.
+The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solarlezama2013}) can be marked, i.e., its errors can be simultaneously localized, such that the resulting program has a well-defined type and binding structure.
 We establish this and other properties as metatheorems that we mechanize in the Agda proof assistant. 
-As we define the calculus, we consider a number of situations where error localization decisions are subtle and conclude by defining some extensions of the intentionally minimal core calculus, e.g.
-with destructuring patterns and System F-style polymorphism,
+As we define the calculus, we consider a number of situations where error localization decisions are
+subtle and conclude by defining some extensions of the intentionally minimal core calculus, e.g., with destructuring patterns and System F-style polymorphism, 
 that serve as case studies of the general approach that we hope will be adopted by language designers defining practical type systems.
@@
 For example, in the following Hazel expression (ignoring the error on the type hole for the moment), bidirectional error localization 
-does not mark any uses of \li{x} as erroneous because a type for \li{x} cannot locally be inferred (the let-bound expression is an empty hole, which has unknown type).
-However, by gathering constraints from the three uses of \li{x} and attempting to unify, we see that the type hole is unfillable.
-Rather than arbitrarily privileging one of these uses, as is the case in languages like OCaml (and which has led to the development of complex heuristics, e.g.
-using machine learning \cite{seidel2017}, to avoid misleading programmers), this error is localized, with a \li{!}, to the type hole itself, with the error message providing partially consistent solutions as shown.
+does not mark any uses of \li{x} as erroneous because a type for \li{x} cannot locally be inferred
+(the let-bound expression is an empty hole, which has unknown type). However, by gathering
+constraints from the three uses of \li{x} and attempting to unify, we see that the type hole is
+unfillable. Rather than arbitrarily privileging one of these uses, as is the case in languages like
+OCaml (and which has led to the development of complex heuristics, e.g., using machine learning \cite{seidel2017}, to avoid misleading programmers), this error is localized, with a \li{!}, to the type hole itself, with the error message providing partially consistent solutions as shown.
 \\
@@
 The user can interactively explore possible error localizations by hovering over a partially consistent suggestion, which (temporarily) fills the type hole and thus returns the localization decision to the bidirectional type system.
-Here, each choice causes a different set of two uses of \li{x} to be marked, e.g.
-the first and third use for \li{Int -> Int} (shown above).
+Here, each choice causes a different set of two uses of \li{x} to be marked, e.g., the first and third use for \li{Int -> Int} (shown above).
 Until the choice is finalized, the type hole continues to operate gradually.

### 2023-10-27 polymorphism.tex
Some final tweaks and changelist

@@
 We leave as future work the task of defining marked versions of even more elaborate bidirectional
-type systems, e.g. \citet{dunfield2013}'s rather substantial formulation of implicit type
+type systems, e.g., \citet{dunfield2013}'s rather substantial formulation of implicit type
 application in a bidirectional setting (which would have to handle, for example, the situation where

### 2023-10-27 related.tex
Some final tweaks and changelist

@@
 
-In addition, there has been considerable work on improving explanations for type errors, e.g.
-by providing sample inputs (dynamic witnesses) that elicit runtime errors \cite{seidel2016}.
+In addition, there has been considerable work on improving explanations for type errors, e.g., by providing sample inputs (dynamic witnesses) that elicit runtime errors \cite{seidel2016}.
 With this approach, one can generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with around 70\% accuracy.

### 2023-11-15 related.tex
Updates from Overleaf
Fix typos

@@
 Recently, \citet{bhanuka2023} model the flow of type information throughout the program using subtyping constraints to produce detailed error messages when unification fails.
-Our focus in this paper has not be on the error messages displayed on-screen, about which we make no specific claims.
+Our focus in this paper has not been on the error messages displayed on-screen, about which we make no specific claims.
 It may, however, be beneficial to indicate to the user from where in the program type suggestions originated via type provenances.
-Documenting the flow of type information would require using assymetric constraints rather than typical bidirectional equality constraints.
+Documenting the flow of type information would require using asymmetric constraints rather than typical bidirectional equality constraints.
 

### 2023-11-15 calculus.tex
Fix typos
Data availability statement

@@
 body, but we choose analysis so that the body is always in analytic mode. The guiding design
-principle in this decision is is a notion of ``conservation of mode,'' but it is not critical to the
+principle in this decision is a notion of ``conservation of mode,'' but it is not critical to the
 calculus.
@@
 The semantics and metatheory presented above have been fully mechanized in the Agda proof assistant.
-The mechanization additionally includes the decoupled Hazelnut action semantics described in
+The mechanization \cite{zhao2023} additionally includes the decoupled Hazelnut action semantics described in
 \cref{sec:calculus-hazelnut}.

### 2023-11-15 conclusion.tex
Updates from Overleaf
CCS, type classes
Data availability statement

@@
 
-% \section*{Acknowledgements}
-% (omitted for review)
+\section*{Acknowledgements}
+The authors would like to thank the anonymous referees at POPL 2024 and ICFP 2023 for helpful feedback on earlier drafts of this paper.
+This work was partially funded through the NSF grant \#CCF-2238744.
+
+\section*{Data Availability Statement}
+An artifact \cite{zhao2023} containing the Agda mechanization and the implementation of Hazel
+including type hole inference is available.
 

### 2023-11-15 related.tex
Updates from Overleaf

@@
 
-For systems rooted in constraint solving, there has been considerable work in improving error localization because it is notoriously well-understood that such systems make error localization difficult, and programmers are often confused by localization decisions \cite{wand1986} because they are rooted in \emph{ad hoc} traversal orders \cite{mcadam1999,lee1998}.
-More recently, there has been a series of papers on finding the most likely location for an error based on either manual weights \cite{zhang2014,pavlinovic2014} or learned weights \cite{seidel2017}.
+For systems rooted in constraint solving, there has been considerable work in improving error localization because such systems are notorious for making error localization difficult, and programmers are often confused by localization decisions \cite{wand1986} because they are rooted in \emph{ad hoc} traversal orders \cite{mcadam1999,lee1998}.
+More recently, there has been a series of papers that discuss finding the most likely location for an error based on maximum likelihood estimation \cite{zhang2014} or manual/learned weights \cite{pavlinovic2014,seidel2017}.
 While improving the situation somewhat, these remain fundamentally \emph{ad hoc} in their need to guess intent.

### 2023-11-15 let.tex
Add reference to pattern matching with holes paper

@@
 To avoid an additional redundant analysis, we would further want to modify the pattern synthesis judgement to produce a context as well.
+
+Recent work by \citet{yuan2023} addresses the problem of general pattern matching with typed holes,
+providing mechanisms to reason statically about redundancy and exhaustiveness in the presence of
+pattern holes. A comprehensive account incorporating the marked lambda calculus would employ error
+marks indicating inexhaustive matches and redundant patterns.

### 2023-11-15 related.tex
Updates from Overleaf

@@
 For systems rooted in constraint solving, there has been considerable work in improving error localization because such systems are notorious for making error localization difficult, and programmers are often confused by localization decisions \cite{wand1986} because they are rooted in \emph{ad hoc} traversal orders \cite{mcadam1999,lee1998}.
-More recently, there has been a series of papers that discuss finding the most likely location for an error based on maximum likelihood estimation \cite{zhang2014} or manual/learned weights \cite{pavlinovic2014,seidel2017}.
+More recently, there has been a series of papers that discuss finding the most likely location for an error based on a maximum likelihood criterion applied to type flows \cite{zhang2014} or manual/learned weights \cite{pavlinovic2014,seidel2017}.
 While improving the situation somewhat, these remain fundamentally \emph{ad hoc} in their need to guess intent.
@@
 Recently, \citet{bhanuka2023} model the flow of type information throughout the program using subtyping constraints to produce detailed error messages when unification fails.
+\citet{seidel2016} provide sample inputs (dynamic witnesses) that elicit runtime errors.
+With this approach, one can generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with around 70\% accuracy.
 Our focus in this paper has not been on the error messages displayed on-screen, about which we make no specific claims.
-It may, however, be beneficial to indicate to the user from where in the program type suggestions originated via type provenances.
+It may, however, be beneficial in future work to indicate to the user from where in the program type suggestions originated via type provenances and by incorporating the techniques in this prior work, e.g.
+to explain why a particular suggestion arises.
 Documenting the flow of type information would require using asymmetric constraints rather than typical bidirectional equality constraints.
-
-In addition, there has been considerable work on improving explanations for type errors, e.g., by providing sample inputs (dynamic witnesses) that elicit runtime errors \cite{seidel2016}.
-With this approach, one can generate graphs for visualizing the execution of witnesses and heuristically identify the source of errors with around 70\% accuracy.
-These and a variety of other approaches to explaining errors are a fruitful avenue for future integration into the system.

### 2023-11-15 intro.tex

@@
 \cref{sec:thi} describes this distinctively neutral approach to blending local and constraint-based
-type inference in Hazel, where bidirectional typing is used to systematically mark erroneous expressions in the program and constraint-based inference (using entirely standard algorithms, which we do not repeat in this paper) is used exclusively to locate unfillable holes.
+type inference in Hazel, where bidirectional typing is used to systematically mark erroneous expressions in the program and constraint-based inference (using entirely standard algorithms, which we do not repeat in this paper) is used exclusively to mark unfillable holes.

### 2023-11-15 related.tex
Updates from Overleaf

@@
 More recently, there has been a series of papers that discuss finding the most likely location for an error based on a maximum likelihood criterion applied to type flows \cite{zhang2014} or manual/learned weights \cite{pavlinovic2014,seidel2017}.
-While improving the situation somewhat, these remain fundamentally \emph{ad hoc} in their need to guess intent.
-However, data-driven approaches could perhaps be layered atop type hole inference to improve the ranking or filtering of suggestions.
+While improving the situation somewhat, these remain fundamentally \emph{ad hoc} in their need to guess the user's likely intent.
+However, these approaches could perhaps be layered atop type hole inference to improve the ranking or filtering of suggestions.
 
@@
 Our focus in this paper has not been on the error messages displayed on-screen, about which we make no specific claims.
-It may, however, be beneficial in future work to indicate to the user from where in the program type suggestions originated via type provenances and by incorporating the techniques in this prior work, e.g.
+It may, however, be beneficial in future work to indicate to the user from where in the program type suggestions originated via our type provenances and by incorporating the techniques in this prior work, e.g.
 to explain why a particular suggestion arises.
-Documenting the flow of type information would require using asymmetric constraints rather than typical bidirectional equality constraints.
+% Note that documenting the flow of type information would require using asymmetric constraints rather than typical bidirectional equality constraints.

### 2023-11-15 conclusion.tex
Updates from Overleaf
Update artifact ref and data availability statement

@@
 
+\section*{Data Availability Statement}
+An artifact \cite{zhao2023} containing the complete formalization of the marked lambda calculus and the extensions described, the Agda mechanization, and the implementation of Hazel including type hole inference is available.
+Up-to-date versions of the formalism and mechanization may be found at \url{https://github.com/hazelgrove/error-localization-agda}.
+Hazel is being actively developed---more information is available at \url{https://hazel.org}, and the Hazel source code is maintained at \url{https://github.com/hazelgrove/hazel}.
+
 \section*{Acknowledgements}
@@
 
-\section*{Data Availability Statement}
-An artifact \cite{zhao2023} containing the Agda mechanization and the implementation of Hazel
-including type hole inference is available.
-
 % Thank you for reading this paper!

### 2023-11-15 calculus.tex
tweak to conclusion

@@
 Because of the untyped variant's simplicity, and because marking is defined separately, this mechanization can serve universally as a baseline for the correctness of approaches like the typed action calculus and perhaps other analyses that minimize re-marking.
-We leave a complete assessment of this and other re-marking optimization approaches to future work, particular because Hazel is now moving toward the decoupled approach for future development.
+We leave a complete assessment of this and other re-marking optimization approaches to future work, particularly because Hazel is now moving toward the decoupled approach for future development.

### 2023-11-15 conclusion.tex
Updates from Overleaf
tweak to conclusion

@@
 Bidirectional type checking helps make the localization decisions we make systematic and predictable, and type hole inference shows how local and constraint-based type inference might operate hand-in-hand, rather than as alternatives.
+Throughout, we focused on maintaining neutrality about user intent whenever possible.
 We hope that language designers will use the techniques introduced in this paper to consider more rigorously, perhaps even formally, the problems of type error localization and error recovery when designing future languages.
@@
 \section*{Data Availability Statement}
-An artifact \cite{zhao2023} containing the complete formalization of the marked lambda calculus and the extensions described, the Agda mechanization, and the implementation of Hazel including type hole inference is available.
+An artifact \cite{zhao2023} containing the complete formalization of the marked lambda calculus and the extensions described above, the Agda mechanization, and the implementation of Hazel including type hole inference is available.
 Up-to-date versions of the formalism and mechanization may be found at \url{https://github.com/hazelgrove/error-localization-agda}.
